### 2019-03-22 Code/cluster.Rmd
genesis

---
title: "Clustering"
output:
  pdf_document:
    dev: png
---

<!--
cd phdproj/platformMS/
rmarkdown::render('./analyses/cluster.Rmd', 'pdf_document')
-->

```{r codeOpts, echo = F}
# Code chunk option
knitr::opts_chunk$set(echo = F,
                      eval = F)
```

# Cluster analysis

- Cluster analysis
- Dissimilarity metric
- Selecting number of clusters $k$
- Validating clusters (cross-validation)
- Intra-cluster variability
- Inter-cluster variability

We identified regions with similar cumulative exposure patterns, referred to as
threat complexes after [@bowler2018], using a partional *k-medoids* clustering
algorithm, CLARA [CLustering for Large Applications; @kaufman1990].
This algorithm extends the PAM (Partition Around Medoids) algorithm to allow the
analysis of large datasets. The PAM algorithm seeks to identify $k$ objects in the
dataset that are representative of all other objects, *i.e.* medoids (also
referred to as examplars) and that are central to the cluster they represent.
The goal of the algorithm is to iteratively minimize dissimilarity between
objects in the cluster and the representative medoid. Sets of $k$ medoids are
thus selected and replaced in order to iteratively minimize the intra-cluster
dissimilarity.

Because the algorithm identifies clusters by minimizing dissimilarity around a
set of *k* medoids, the algorithm requires a user defined number of clusters
composing the dataset under investigation. When the number of cluster is
unknown, there are strategies to choose the optimal number of clusters. We
compared the

The CLARA algorithm was designed to use with large datasets. It uses the
PAM algorithm on a sample from the original dataset to identify a set of optimal
medoids for the sample, and subsequently assign all objects in the dataset to
the identified clusters. This process is repeated and iterations are compared
on the basis of their average dissimilarity between objects and cluster medoids
to select the optimal set of $k$ medoids that minimizes average dissimilarity.

The number of iterations and the sample size will influence the
quality of the results, so the user must find a balance between the number
of sampling runs, the sample size and analysis time.
<!-- There has to be a reference for this. Look up in the article I already read -->

Analyses were performed using the `cluster` package in R [@machler2018].

## Figures

![](./validation.png)

\newpage

![](./clusterComp.png)

\newpage

![](./indCluster.png)

\newpage

![](./totDiss.png)

- Group according to similarity -> unsupervised machine learning
- k-medoids clustering:
  - minimize variation between points in a cluster
    - Object in the cluster serves as its center
    - Object: representative of the cluster
    - *"The k-medoids approach attempts to identify $k$ objects from the dataset that best represent all objects. Clusters are created by assigning each object in the dataset to its nearest representative (*i.e.* medoid)."*
    - Partition around medoids: PAM
    - Clustering for large applications: CLARA
- Hierarchical clustering
  - *"Organizes states into nested sequences of clusters forming a growing aggregation tres"*
- Partitional clustering:
  - *"Iterativelu performs the classification from randomly predefined initial centroids according to a pregiven number $k$ of clusters"*

# Lexicon

- PAM
- CLARA
- Medoids
- Clustering
- Silhouette
- Cluster
- Similarity / dissimilarity
- Manhattan distance
- Average silhouette width (ASW) (to maximize)
- Davies-Bouldin (DB) index (to minimize)
- Partitional vs hierarchical clustering

# To figure out

- Sample
- Data standardization
- Examplars = medoids?
- Testing with other clustering methods
- Cross-validation
- Bootstraping
- Sum total variation / k -> elbow plot

# Relevant papers

- Long et al. 2010: landscape pattern, CLARA analysis
- Halkidi et al. 2001: Clustering review, number of $k$, silhouette (average silhouette width)
- Kaufman and Rousseeuw 1990: CLARA algorithm developers, data normalization
- Davies-Bouldin 1979: DB index for evaluation of number of $k$
- Bowler et al. 2018: Threat complexes
- Machler et al 2018: Cluster package
- Hamilton et al. 2011: CLARA, multi-beam surveys
- Anderberg 1973: Partitional vs hierarchical clustering
- Cassou et al. 2004: Winter regimes (not very interesting, but discusses clustering)
- Shertzer and Williams 2007: Bootstrap validation, figure with ASW and confidence interval from bootstraping validation

# Relevant links

- http://www.sthda.com/english/wiki/print.php?id=234#partitioning-clustering
- https://github.com/bowlerbear/geographyDrivers/blob/master/analysis.R
- http://cc.oulu.fi/~jarioksa/opetus/metodi/sessio3.pdf
- https://www.rdocumentation.org/packages/cluster/versions/2.0.7-1/topics/clara
- https://uc-r.github.io/kmeans_clustering#silo
-

# Analysis

The time required to perform the analysis grows exponentially with the sample
size, and linearly with the number of iterations. With a dataset composed of
250000 rows and 19 columns, and using a sample size of 10000, an iteration
takes a little over 1.5 minutes to run. This means that the total analysis
would take approximately 2.5 hours for 100 iterations and a sample size of
10000 data points.

Testing: Sample of 10000 data points results in an analysis that takes approximately 1.5 minutes to run.

nS <- number of samples
sS <- sample size

# Code

## Parameters setup

```{r setup}
# Packages
library(magrittr)
library(sf)
library(cluster)
library(vegan)
library(slmeta)
library(graphicsutils)
```

## Data preparation

```{r data}
# Load data
library(slmeta)
data(drivers, package = 'slmeta')
dr <- drivers[, !colnames(drivers) %in% c('UVpositive','UVnegative','ID')]

# Scale data
# Z_{if} = \frac{x_{if} - m_f}{s_f} -> Long et al. 2010
dr <- scale(dr)

# Load other data
load('./analyses/clMed.RData')
load('./analyses/simpDr.RData')
```

## k-mediods clustering

### Clustering

```{r kmedoids}
# Cluster analysis
nSample <- 50
sampleSize <- 10000
k <- 2:10
out <- list('list', length(k))
for(i in k) {
  cat("   k: ", i, "\r")
  out[[i]] <- clara(dr,
                    k = i,
                    sampsize = sampleSize,
                    samples = nSample,
                    metric="manhattan",
                    stand = TRUE,
                    keep.data = FALSE)
}
out[[1]] <- NULL
save(out, file = './analyses/kmedoid.RData')
```

### Validating the number of clusters

#### Average silhouette width

```{r silhouettes1}
# load('./analyses/kmedoid.RData')
# for(i in 1:9) plot(silhouette(out[[i]]))
aswMedoid <- unlist(lapply(out, function(x) x$silinfo$avg.width))
```

#### Elbow method
```{r elbow1}
# Verify number of clusters by using the elbow method,
# i.e. the inflexion point in the sum of squared error (SSE) scree plot
# Extract SSE from each clustering run
# wss function
wss <- function(x) (nrow(x)-1) * sum(apply(x,2,var))

# Measure for each k and each cluster
wssKMedoid <- wss(dr)
for(i in 1:length(out)) { # For each k value tested
  wssTemp <- numeric()
  for(j in 1:k[i]) { # For each cluster
    id <- out[[i]]$clustering == j # Identify observations in cluster j
    wssTemp[j] <- wss(dr[id, ])
  }
  wssKMedoid[k[i]] <- sum(wssTemp)
}
```

## k-means clustering

### Clustering

```{r kmeans}
# K-means cluster analysis
# Stored in a list
out2 <- list('list', length(k))
for(i in k) {
  cat("   k: ", i, "\r")
  out2[[i]] <- kmeans(x = dr,
                      centers = i,
                      iter.max = 1000,
                      nstart = 25,
                      algorithm = 'Lloyd')
}
out2[[1]] <- NULL
save(out2, file = './analyses/kmeans.RData')
```

### Validating the number of clusters

#### Average silhouette width

```{r silhouettes2}
# load('./analyses/.RData')
# Silhouette needs to be computed manually here, because the general function crashes
# This will make a very inefficient function, but one that will work
sil <- function(x, cl) {
  # Euclidian distance function between a single point and another point or a string of points
  euc <- function(x, y) {
    if(is.null(dim(y))) {
      sqrt(sum((x - y) ^ 2))
    } else {
      sqrt(colSums((x - t(y)) ^ 2))
    }
  }

  # Number of rows in dataset x
  nr <- length(cl)

  # Clusters in dataset
  k <- unique(cl)

  # Empty data.frame to store silhouette information
  df <- data.frame(cluster = numeric(nr),
                   neighbor = numeric(nr),
                   sil_width = numeric(nr),
                   stringsAsFactors = F)
  # Calculate silhouette for all data points in dataset x
  for(i in 1:nr) {
    # Cluster of data point i
    df$cluster[i] <- cl[i]

    # ID of points in cluster of point i
    id <- cl == df$cluster[i]

    # a = mean distance of point i to all points within its cluster
      a <- euc(x[i, ], x[id, ]) %>%
           mean()

    # b =  bi is the smallest average distance of i to all points in any other
    #      cluster of which i is not a member.
      # Clusters of which i is not a member
      kb <- k[k != df$cluster[i]]

      # Empty names string to store distances of i to objects in other clusters
      b <- numeric(length(kb))
      names(b) <- kb

      # Distance between i and objects in other clusters
      for(j in 1:length(kb)) {
        # ID of points in first kb cluster
        id <- cl == kb[j]
        b[j] <- euc(x[i, ], x[id, ]) %>%
                mean()
      }

      # Identify neighbour of i, which is the cluster of which i is not a member
      # and with minimum distance to i
      df$neighbor[i] <- as.numeric(names(which.min(b)))

      # b is the average distance of i to its neighbor cluster
      b <- min(b)

      # Evaluate silhouette width: si = (bi - ai) / (max(ai, bi))
      df$sil_width[i] <- (b - a) / max(c(a,b))
  }
  return(df)
}

# Now calculate the silhouette width for all observations
silKM <- vector('list', length(k))
for(i in 1:length(k)) {
  cat("   i: ", i, ' of ', length(k), "\r")
  silKM[[i]] <- sil(dr, out2[[i]]$cluster)
  save(silKM, file = './analyses/silKM.RData')
}

# Next get mean silhouette width for each k value
load('./analyses/silKM.RData')
aswKM <- unlist(lapply(silKM, function(x) mean(x$sil_width)))
```

#### Elbow method
```{r elbow2}
# Verify number of clusters by using the elbow method,
# i.e. the inflexion point in the sum of squared error (SSE) scree plot
# Extract SSE from each clustering run
# The kmeans object already includes the total within-cluster SSE in $tot.withinss
wssKMeans <- wssKMedoid[1]
for(i in k) wssKMeans[i] <- out2[[i-1]]$tot.withinss

```

### Validation

```{r validation}
png('./analyses/validation.png', width = 1280, height = 920, res = 200, pointsize = 7)
layout(matrix(1:4, ncol = 2))
par(mar = c(5,5,2,1))
# k-medoid
  # Silhouette
  plot(aswMedoid ~ k,
       pch = 20,
       ylab = 'Average Silhouette Width',
       type = 'b',
       frame = F,
       xlab = '',
       xlim = c(0,10),
       cex = 1.5,
       cex.axis = .8)
  mtext(side = 3, 'k-medoids clustering', font = 2, line = .75)
  # SSE as a function of the number of clusters k
  plot(x = c(1,k),
       y = wssKMedoid,
       type="b",
       frame = F,
       xlab="Number of Clusters (k)",
       ylab="Total within-clusters sum of squares",
       pch = 20,
       xlim = c(0,10),
       cex = 1.5,
       cex.axis = .8)
# k-means
  # Silhouette
  plot(aswKM ~ k,
       pch = 20,
       frame = F,
       xlim = c(0,10),
       xlab = '',
       ylab = 'Average Silhouette Width',
       type = 'b',
       cex = 1.5,
       cex.axis = .8)
  mtext(side = 3, 'k-means clustering', font = 2, line = .75)
  # SSE as a function of the number of clusters k
  plot(x = c(1,k),
       y = wssKMeans,
       type="b",
       frame = F,
       xlim = c(0,10),
       xlab="Number of Clusters (k)",
       ylab="Total within-clusters sum of squares",
       pch = 20,
       cex = 1.5,
       cex.axis = .8)
dev.off()
```

## Selected k value and clustering

```{r k}
k <- 7
clMed <- out[[k-1]]
clMean <- out2[[k-1]]
```

## k-medoids & k-means comparison

```{r clusComp}
cl1 <- clMed$clustering
cl2 <- clMean$cluster

# Make clusters similar
id1 <- cl2 == 1
id2 <- cl2 == 2
id3 <- cl2 == 3
id4 <- cl2 == 4
id5 <- cl2 == 5
id6 <- cl2 == 6
id7 <- cl2 == 7
cl2[id1] <- 2
cl2[id2] <- 5
cl2[id3] <- 7
cl2[id4] <- 3
cl2[id5] <- 1
cl2[id6] <- 4
cl2[id7] <- 6

# Differences
cl12 <- as.numeric(cl1 == cl2) + 1
corr <- round((sum(cl1==cl2)/length(cl1)*100))

png('./analyses/clusterComp.png', width = 2560, height = 1840, res = 300, pointsize = 6)
par(mfrow = c(2,2))
pal <- colorRampPalette(slmetaPal('platform'))
# K-medoids clustering
cols <- pal(k)[cl1]
plotEGSL('egslSimple')
plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
plot(st_geometry(egslSimple), col = 'transparent', border = '#25364A', lwd = .75, add = T)
text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 75000, 'k-medoid clustering', cex = 2, font = 2)
# K-means clustering
cols <- pal(k)[cl2]
plotEGSL('egslSimple')
plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
plot(st_geometry(egslSimple), col = 'transparent', border = '#25364A', lwd = .75, add = T)
text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 75000, 'k-means clustering', cex = 2, font = 2)
# Difference
pal2 <- colorRampPalette(c('#00000000', pal(1)))
cols <- pal2(2)[cl12]
plotEGSL('egslSimple')
plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
plot(st_geometry(egslSimple), col = 'transparent', border = '#25364A', lwd = .75, add = T)
text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 75000, 'Correspondance', cex = 2, font = 2)
text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 125000, paste0('Cell correspondance: ', corr, '%'), cex = 1.5)
# Legend
cols <- pal(k)
y <- seq(.85, .15, length.out = 7)
yG <- .035
plot0(xlim = c(0,1), ylim = c(0,1))
for(i in 1:k) polygon(x = c(.1,.2,.2,.1), y = c(y[i]-yG, y[i]-yG, y[i]+yG, y[i]+yG), col = cols[i])
text(x = .21, y = y, labels = paste('Cluster', 1:7), cex = 1.5, adj = c(0, .5))
dev.off()

```

## Individual clusters

```{r indGraph}

data(egslGrid)
egsl <- st_centroid(egslGrid) %>%
        cbind(clMed)

png('./analyses/indCluster.png', width = 3840, height = 2760, res = 300, pointsize = 10)
par(mfrow = c(3,3))
for(i in 1:7) {
  cl = i
  id = egsl$clMed == cl
  plotEGSL('egslSimple', borders = '#74747444')
  plot(st_geometry(egsl[id, 2]), cex = .01, add = T, col = '#187962', border = '#187962', pch = 20)
  text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 75000, i, cex = 2)
}
dev.off()
```

## Inter-cluster variability

Not possible to do it in a single batch, so will have to do it iteratively

```{r simper}
# x <- dr
# iter <- 10
# prop <- .02 # .05: ~500s/iter | .01: ~20s/iter | .02: ~80s/iter
# cl <- cl1
subSimper <- function(x, cl, iter = 100, prop = .1, accr = '') {
  # List to store simper results for post-processing
  simpDr <- vector('list', iter)

  # Determine the number of data points to sample from each cluster
    clSub <- ceiling(table(cl) * prop)

  # Repeat the process for the desired number of iterations
  for(i in 1:iter) {
    cat("   i: ", i, ' of ', iter, "\r")

    # Subsampled dataset
    samp <- vector('list', length(clSub))
    for(j in 1:length(clSub)) samp[[j]] <- sample(x = which(cl == j), size = clSub[j])
    samp <- unlist(samp)

    # Run simper analysis on subset of data
    simpDr[[i]] <- simper(x[samp, ], cl[samp])

    # Save file
    save(simpDr, file = paste0('./analyses/simpDr', accr, '.RData'))
  }
  return(simpDr)
}

# Run iterative simper analysis
iter <- 100
prop <- .05
simpDr <- subSimper(dr, clMed, iter = iter, prop = prop, accr = '1')

#
load('./analyses/simpDr1.RData')
x1 <- simpDr
load('./analyses/simpDr2.RData')
x2 <- simpDr
load('./analyses/simpDr3.RData')
x3 <- simpDr
load('./analyses/simpDr4.RData')
x4 <- simpDr
simpDr <- c(x1,x2,x3,x4)
id <- which(unlist(lapply(simpDr, is.null)))
for(i in rev(id)) simpDr[[i]] <- NULL

# Summarize in array format
iter <- length(simpDr)
nList <- length(simpDr[[1]])
varNames <- c('average')
drNames <- colnames(dr)
simp <- vector('list', nList)
names(simp) <- names(simpDr[[1]])
for(i in 1:nList) simp[[i]] <- array(data = 0, dim = c(ncol(dr), 1, iter), dimnames = list(drNames, varNames))

# Extract species, average and sd
for(i in 1:iter) {
  temp <- simpDr[[i]]
  for(j in 1:nList) {
    # simp[[j]][,'species',i] <- temp[[j]]$species
    simp[[j]][,'average',i] <- temp[[j]]$average
    # simp[[j]][,'sd',i] <- temp[[j]]$sd
  }
}

# Summarize the results over all iterations
simpMean <- simpSd <- vector('list', nList)
names(simpMean) <- names(simpSd) <- names(simp)
for(i in 1:nList) simpMean[[i]] <- apply(X = simp[[i]], MARGIN = c(1,2), FUN = mean, na.rm = T) # average of average
for(i in 1:nList) simpSd[[i]] <- apply(X = simp[[i]], MARGIN = c(1,2), FUN = sd, na.rm = T) # average of average
```

```{r interDiss}
# Let's begin with the total intercluster dissimilarity
totDiss <- matrix(0,k,k)
totDiss[lower.tri(totDiss)] <- unlist(lapply(simpMean, sum))
totDiss[upper.tri(totDiss)] <- t(totDiss)[upper.tri(totDiss)]
totDiss[totDiss == 0] <- NA
xmn <- floor(min(totDiss, na.rm = T))
xmx <- ceiling(max(totDiss, na.rm = T))
xr <- (range(xmn,xmx)+abs(xmn)) * 100
totDiss <- rotate(totDiss)

# Plot
png('./analyses/totDiss.png', width = 1000, height = 980, res = 200, pointsize = 7)
mat <- matrix(ncol = 2, data = 1:2)
layout(mat, widths = c(.85,.15))
par(mar = c(2,2,2,2))
plot0(xlim = c(0,k), ylim = c(0,k))
for(x in 1:k) {
  for(y in 1:k) {
    cols <- pal(xr[2])[totDiss[y,x]*100]
    polygon(x = c(x-1,x,x,x-1), y = c(y-1,y-1,y,y), col = cols)
    text(x = x-.5, y = y-.5, round(totDiss[y,x], 2), col = '#bfbfbf')
  }
}
text(x = (1:k)-.5, y = (k:1)-.5, 1:k, cex = 1.5)
mtext(side = 3, 'Total cluster dissimilarity', font = 2, cex = 1.75)
colorBar(colRamp = pal(100), min = 0, max = 1, align = 'vertical', mar = c(10,0,10,3))
dev.off()

```

```{r simperSummary1}
# Driver names vector
nDr <- ncol(dr)
drNames <- rep(rownames(simp[[1]]), each = iter)

# Data frame to store results
simpSummary <- data.frame(Drivers = drNames, stringsAsFactors = F)

# Average dissimilarity contribution values
for(j in 1:nList) {
  y <- numeric(0)
  for(i in 1:nDr) y <- c(y, simp[[j]][i,'average',])
  simpSummary <- cbind(simpSummary, y)
  colnames(simpSummary)[j+1] <- names(simp)[j]
}

# Function to change names of
changeNames <- function(x) {
  drNames <- data.frame(ori = c("Aragonite","aquacultureInv","coastDev","dirHumImpact",
                                "fisheriesDD","fisheriesDNH","fisheriesDNL",
                                "fisheriesPHB","fisheriesPLB","Hypoxia","inorgPol",
                                "invasives", "marinePol","nutrientInput",
                                "orgPol","seaLevel","shipping","sst", "toxicAlgae"),
                        new = c('ARAG','AQUA','CD','DHI','DD','DNH','DNL','PHB','PLB', "HYP",
                                 'IP','INV','MP','NI','OP','SL','SHIP','SST',"TOX"),
                        stringsAsFactors = F)
  for(i in 1:nrow(drNames)) x$Drivers[x$Drivers == drNames$ori[i]] <- drNames$new[i]
  x
}
simpSummary <- changeNames(simpSummary)
# simpSummary$Drivers <- factor(simpSummary$Drivers, drNames$new)

# Figure
# Set layout view
mat <- matrix(0,k,k)
# codens <- mat[upper.tri(mat)] <- 1:sum(upper.tri(mat))
kern <- mat[lower.tri(mat)] <- (max(mat)+1):(sum(lower.tri(mat)) + max(mat))
diagMat <- diag(mat) <- (max(mat)+1):(max(mat)+k)

# Graph
png('./analyses/simper.png', width = 3700, height = 1840, res = 200, pointsize = 15)
layout(mat)
par(mar = c(2,2,0,0))
for(i in 2:ncol(simpSummary)) boxplot(simpSummary[,i] ~ simpSummary$Drivers, cex.axis = .2, frame = F)
for(i in 1:k) { graphicsutils::plot0(); text(0, 0, i, cex = 2) }
dev.off()
```

```{r contDiss}

# Driver contribution to overall dissimilarity
contDiss <- lapply(simpMean, function(x) x[,1] / sum(x[,1])) %>%
            as.data.frame() %>%
            dplyr::mutate(driver = rownames(.), mean = rowMeans(.), sum = rowSums(.)) %>%
            dplyr::select(driver, mean, sum)
```

```{r cumulativeDiss}
# Accronyms
accr <-  c('ARAG','AQUA','CD','DHI','DD','DNH','DNL','PHB','PLB',
         "HYP",'IP','INV','MP','NI','OP','SL','SHIP','SST',"TOX")

# Set layout view
mat <- matrix(0,k,k)
# codens <- mat[upper.tri(mat)] <- 1:sum(upper.tri(mat))
mat[lower.tri(mat)] <- (max(mat)+1):(sum(lower.tri(mat)) + max(mat))
diag(mat) <- (max(mat)+1):(max(mat)+k)
mat <- t(mat)
mat[lower.tri(mat)] <- (max(mat)+1):length(mat)
mat <- t(mat)

# Compute cumulative dissimilarity
# cumDiss <- lapply(simpMean, function(x) sort((x[,1] / sum(x[,1])), decreasing = T))
# Percent similarity
cumDiss <- lapply(simpMean, function(x) (x[,1] / sum(x[,1])))
p1 <- vector('list', length(cumDiss))
for(i in 1:length(cumDiss)) {
  data <- data.frame(individual = accr,
                     group = c('O','C','C','C','O','O','O','O','O','O','C','C','O','C','C','O','O','O','C'),
                     value = round(cumDiss[[i]]*100))
  rownames(data) <- NULL
  p1[[i]] <- circularPlot(data)
}

# Individual clusters
p2 <- vector('list', k)
for(i in 1:length(p2)) {
  id <- egsl$clMed == i

  temp <- egsl %>%
          filter(clMed == i) %>%
          st_coordinates() %>%
          as.data.frame()

  p2[[i]] <- ggplot() +
             geom_sf(data = egslSimple, fill = NA) +
             geom_point(aes(x = X, y = Y), colour = '#187962', size = .01, data = temp) +
             coord_sf(crs = st_crs(egslSimple), datum = NA) +
             theme(
                panel.ontop = TRUE,   ## Note: this is to make the panel grid visible in this example
                panel.grid = element_blank(),
                line = element_blank(),
                rect = element_blank(),
                text = element_blank(),
                plot.background = element_rect(fill = "transparent"))
}

# Total dissimilarity
totDiss <- round(unlist(lapply(simpMean, sum)),2)
names(totDiss) <- NULL
diss <- totDiss*100
diss <- diss - min(diss)
pal3 <- colorRampPalette(c('#c2c2c2', '#494949'))
cols <- pal3(max(diss)+1)[diss+1]
p3 <- vector('list', length(cols))
for(i in 1:length(p3)) {
  datapoly <- data.frame(id = 1, x = c(0,0,1,1), y = c(0,1,1,0))
  p3[[i]] <- ggplot(datapoly, aes(x = x, y = y)) +
              geom_polygon(colour = '#4e4e4e', fill = cols[i]) +
              theme(
                legend.position = "none",
                axis.text = element_blank(),
                axis.title = element_blank(),
                axis.ticks = element_blank(),
                panel.grid = element_blank(),
                panel.background = element_blank(),
                plot.margin = unit(rep(.5,4), "cm")) +
             annotate('text', x = .5, y = .5, label = totDiss[i], size = 6, colour = '#000000')
}

p <- c(p1,p2,p3)

png('./analyses/simper2.png', width = 2000, height = 2000, res = 200, pointsize = 7)
grid.arrange(grobs = p, layout_matrix = mat)
dev.off()
```

## Map of selected cluster

```{r mapCluster}

png('/users/davidbeauchesne/desktop/cluster.png', width = 1280, height = 920, res = 300, pointsize = 10)
pal <- colorRampPalette(slmetaPal('platform'))
# K-medoids clustering
cols <- pal(k)[clMed$clustering]
plotEGSL('egslSimple')
plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
plot(st_geometry(egslSimple), col = 'transparent', border = '#25364A', lwd = .75, add = T)
dev.off()

```

```{r sim}
# data(egslGrid)
# egsl <- st_centroid(egslGrid) %>%
#         cbind(clMed)
#
k <- 7
sim <- vector('list', k)
# for(i in 1:k) {
for(i in c(2,3,5,6,7,1,4)) {
  cat("   k: ", i, "\r")
  id <- clMed == i
  sim[[i]] <- bootSimper2(dr[id, ], iter = 50, samp = 1200)
  save(sim, file = './analyses/similarity.RData')
}

# Extract species, average and sd
# Summarize in array format
iter = 50
varNames <- c('average')
drNames <- colnames(dr)
simSummary <- vector('list', k)
for(i in 1:k) simSummary[[i]] <- array(data = 0, dim = c(ncol(dr), 1, iter), dimnames = list(drNames, varNames))

for(i in 1:k) {
  # Check if there are multiple iterations or not.
  # If length(sim[[i]]) == 2 it means that there were no iterations and the whole cluster was parsed through the simper2 function
  # If not, then sim[[i]] should be equal to the number of iterations
  if(length(sim[[i]]) == 2) {
    simSummary[[i]] <- matrix(ncol = 1,
                                  data = sim[[i]]$similarityContribution$average,
                                  dimnames = list(drNames, 'average'))
  } else {
    for(j in 1:iter) {
      simSummary[[i]][,'average',j] <- sim[[i]][[j]]$similarityContribution$average
    }
  }
}

# Summarize the results over all iterations
simMean <- vector('list', k)
for(i in 1:k) simMean[[i]] <- apply(X = simSummary[[i]], MARGIN = c(1,2), FUN = mean, na.rm = T) # average of average
# for(i in 1:nList) simpSd[[i]] <- apply(X = simp[[i]], MARGIN = c(1,2), FUN = sd, na.rm = T) # average of average
simTot <- lapply(simMean, sum)

png('./analyses/indCluster.png', width = 3840, height = 2760, res = 300, pointsize = 10)
par(mfrow = c(3,3))
for(i in 1:7) {
  cl = i
  id = egsl$clMed == cl
  plotEGSL('egslSimple', borders = '#74747444')
  plot(st_geometry(egsl[id, 2]), cex = .01, add = T, col = '#187962', border = '#187962', pch = 20)
  text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 75000, i, cex = 2)
}
dev.off()

```

### 2019-03-22 eDriversMS-SI.rmd
genesis

---
title: Supplementary Material
fontsize: 12pt
output:
  pdf_document:
    toc: false
    number_sections: true
  html_document:
    number_sections: true
relativeurls: true
bibliography: eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
cd phdproj/edrivers/edriversms
rmarkdown::render('./eDriversMS-SI.rmd', 'pdf_document')
rmarkdown::render('./eDriversMS-SI.rmd', 'html_document')
-->

# Drivers description

Table S1. List of drivers currently available on *eDrivers* along with their respective accronym used in the figures in the supplementary material.

```{r tableS1, echo=F, fontsize=4}
load('./tables/tableS1.RData')
tableS1
```

## Climate

### Aragonite

***Describe data***

We interpolated aragonite distribution using the exponential kriging model.
We then built an index of aragonite level stress. When $\Omega$ aragonite levels
decrease below 1, the water becomes corrosive to calcifying organisms such as
echinoderms, corals and mollusks larvae. These organisms thus have to
continually recalcify their shells to maintain their structure, which becomes
harder to do as $\Omega$ aragonite levels further decrease towards 0. We therefore
considered stress associated to decreasing $\Omega$ aragonite levels to increase
linearly between values of 1 and 0. The index of stress therefore has a value of
0 and 1 for values of $\Omega$ aragonite equal to 1 and 0, respectively.

### Hypoxia

***Describe data***

We interpolated the dissolved oxygen using cokriging with depth as a covariable,
as done in [@dutil2011]. According to @diaz1995, severe hypoxia can be observed
when oxygen saturation falls below 2 $ml$ $L^{-1}$ or 62.5 $\mu mol$ $L^{-1}$, which is considered
the level necessary to maintain most animal life. This threshold is equal to 1.4
$ml$ $L^{-1}$ (1 $mg$ $L^{-1}$ = 0.7 $ml$ $L^{-1}$).
We used this threshold to create an index of hypoxia using an inverted logistic curve
to transform the dissolved oxygen values as an index of hypoxic stress $H_s$ ranging
from 0 to 1, under the assumption that hypoxic stress will increase following
a logistic curve as it approaches the hypoxic threshold:

$$H_s = \frac{-1}{200 * e^{-2.5 * DO_2}} + 1$$

### Bottom temperature anomalies

The data used to characterize bottom temperature anomalies come from the
Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
[AZMP; @galbraith2018]. We provide a brief summary of data and methods
to characterize surface temperature climatology and anomalies in this document.
For more details, refer to @galbraith2018.

Bottom temperatures are interpolated in the Gulf using
conductivity-temperature-depth (CTD) sampling performed annually through DFO's
multispecies surveys for the northern Gulf in August and for the Magdalen
Shallows in September. Using this sampling survey, temperatures are interpolated
at each 1 m depth layer on a 2 km resolution grid. Bottom temperature are then
extracted from this grid by using a bathymetry layer from the Canadian
Hydrographic Survey [@dutil2012].

We used temperature anomalies, *i.e.* deviations from long-term normal conditions,
to measure an annual index of stress associated with extreme temperatures between
2013 and 2017. Temperature anomalies were calculated using the difference between
grid cell monthly climatologies to the associated long-term averages generated
from the reference period between 1981 and 2010. Grid cells whose monthly value
exceeded ±0.5 standard deviation (SD) from the long-term average were considered
as anomalous [@galbraith2018].

Outliers in the data were considered as those that fell beyond the
interquartile range (IQR) * 3, identified as extreme ouliers by @tukey1977.
Outlier values were capped to correspond to the 5th and 95th percentiles.
Anomalies were divided into positive and negative anomalies and
summed for the available months of September and August.
By adding the monthly anomaly values, we essentially considered the summed
deviation from the mean as an indicator of the annual intensity of surface
temperature anomalies in the St. Lawrence.

### Surface temperature anomalies

The data used to characterize surface temperature anomalies come from the
Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
[AZMP; @galbraith2018]. We provide a brief summary of data and methods
to characterize surface temperature climatology and anomalies in this document.
For more details, refer to @galbraith2018.

The surface layer is characterized using a variety of methods:

1. Temperature-salinity sensors installed on various sihps and forming the
shipboard thermosalinograph network,
2. Moored instruments recording water temperature every 30 minutes and forming
the thermograph network,
3. Sea surface temperature (SST) monthly composite climatologies using Advanced
Very High Resolution Radiometer (AVHRR) satellite images from the National
Oceanic and Atmospheric Administration (NOAA) and European Organisation for the
Exploitation of Meteorological Satellites (EUMETSAT) available from the Maurice
Lamontagne Institute sea surface temperature processing facility at a 1km
resolution from 1985-2013 and from the Bedford Institute of Oceanography (BIO)
Operational Remote Sensing group at a 1.5km resolution since 2014.

Positive and negative surface temperature anomalies were identified following
the same approach used for bottom temperature anomalies. Compared to the
bottom temperature anomalies, however, data is available thoughout the year.
Only the months of April to November were included to avoid biases associated
with anomalies measurements due to the presence of ice cover.

### Sea level rise

The data used to characterize sea level rise risk comes from the global
cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
available on the NCEAS online data repository [@halpern2015b]. We provide a
brief summary of data and methods in this document. For more details, refer to
@halpern2015.

Sea level rise was characterized by @nicholls2010 using NASA's satellite
altimetry data (Topex/Poseidon, Jason-1&2, GFO, ERS-1&2, and Envisat
missions) and available at
http://www.aviso.altimetry.fr/en/data/products/ocean-indicatorsproducts/mean-sea-level/products-images.html

The rate of sea level rise ($mm/year$) was measured between 1992 and 2012 and
transformed as a net change value ($mm$) by multiplying by the number of years
considered. Only positive values were selected under the assumption that
only positive sea level rise is likely to cause environmental stress.

For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
1 $km^2$ grid cell using weighted area average.

## Coastal

### Aquaculture

Aquaculture data comes from a variety of sources in the St. Lawrence because
aquaculture sites are mostly managed at the provincial level. We therefore had
to gather the data on aquaculture sites from the 5 provinces dividing the
St. Lawrence.

Invertebrates aquaculture is especially important in the southern and western
Gulf. Fish and algae aquaculture, on the other hand, remains marginal.
Considering this, we only considered invertebrates aquaculture for the
aquaculture driver layer. However, if fish or algae farming were to become more
important, these driver should be incorporated in future analyses as individual
layers, as impacts vary between types of aquaculture.

Aquaculture activities are highly localized and potential effects do not or
rarely extend beyond the location of the farms. We therefore only considered
the actual location of sites to characterize the distribution of this driver.
We were unable to characterize site production in terms of biomass farmed, which
could provide an indication of the intensity of aquaculture activities. As such,
we considered aquaculture as binary presence-absence data in our analyses.

### Coastal development

We used lights at night as a proxy of coastal infrastructure development, as
terrestrial stable lights at night represent light from human settlements and
industrial sites with electricity.

The data comes from the Nighttime Lights Time Series. Nighttime light products
are compiled by the Earth Observation Group at the National Oceanic and
Atmospheric Administration's (NOAA) National Centers for Environmental
Information (NCEI). They use glogally available nighttime data obtained from
the Visible Infrared Imaging Radiometer Suite (VIIRS) Day/Night Band (DNB) of
the Defense Meteorological Satellite Program (DMSP) to characterize global
average radiance ($nanoWatts$ $cm^{-2}$ $sr^{-1}$) composite images at a
15 arc-second (~200 m) resolution.

We used the annual Version 1 Nighttime VIIRS DNB composites between 2015 and
2016 [@eog2019] to characterize coastal development in coastal areas of the
St. Lawrence. As the effects of coastal development are likely acute in its
direct vicinity, we extracted average radiance values using a 2 km buffer
around grid cells within 2 km of the coast. We used a weighted area average to
extract the radiance values.

### Direct human impact

As in @halpern2008 and @halpern2015, we used the sum of coastal populations as
a proxy of direct human impact. We used Statistics Canada dissemination area
population count from the 2016 census to obtain coastal population size
around the St. Lawrence [@statscan2017]. Dissemination areas are the smallest
standard geographic area in which census data are disseminated and they combine
to cover all of Canada. The census provides population count within the boundary
of each dissemination area, which we used to evaluate total coastal population.

As the effects of direct human impacts are likely acute mostly in coastal areas
we calculated total population in grid cells within 2 km of the coast. Total
population was measured in a 10 km buffer around each coastal cell.
The total population in each buffer was the sum of intersecting dissemination
areas divided by the intersection area between buffers and dissemination areas:

$$DHI_j = \sum_{k=1}^{n_j} P_k * \frac{A_{j,k}}{A{tot, k}}$$

where $j$ is a buffered grid cell, $k$ is a dissemination area intersecting $j$,
$P$ is the population in $k$, $A$ is the area of the $k$ overlapping with $j$
and $A_{tot}$ is the total area of $k$. This approach was favoured to reduce the
effects of very large dissemination areas overlapping with buffers on a very
small percentage of their total area.

### Inorganic pollution

The data used to characterize inorganic pollution comes from the global
cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
available on the NCEAS online data repository [@halpern2015b]. We provide a
brief summary of data and methods in this document. For more details, refer to
@halpern2015.

Inorganic pollution was modelled using impervious surface area (*i.e.*
artificial surfaces such as paved roads) under the assumption that most of this
pollution source comes from urban runoff. Inorganic pollution originating from
point-sources or in areas lacking paved roads is therefore not captured by
this layer. The data obtained was aggregated at the watershed scale and spread
into coastal and marine environments was modelled using a diffusive plume model
from each watershed pourpoints (*e.g.* river mouths).

For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
1 $km^2$ grid cell using weighted area average.

### Nutrient pollution

The data used to characterize nutrient pollution comes from the global
cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
available on the NCEAS online data repository [@halpern2015b]. We provide a
brief summary of data and methods in this document. For more details, refer to
@halpern2015.

Annual fertilizer use in tonnes ($t$) was used as a proxy of nutrient pollution.
The data used came from the Food and Agriculture Organization of the United
Nations (FAO). Gaps in data were modelled using a linear regression between
fertilizer and pesticides or agricultural gross domestic product (GDP).
Dasymetric maps were then used to distribute fertilizer data over the landscape
using 2009 data from the Moderate Resolution Imaging Spectroradiometer (MODIS)
at ~500 m resolution and aggregated to watersheds. Diffusive plume models
from each watershed pourpoint (*e.g.* river mouths) were then used to model the
distribution and intensity of nutrient pollution in coastal and marine
environments.

For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
1 $km^2$ grid cell using weighted area average.

### Organic pollution

The data used to characterize organic pollution comes from the global
cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
available on the NCEAS online data repository [@halpern2015b]. We provide a
brief summary of data and methods in this document. For more details, refer to
@halpern2015.

Annual pesticide use in tonnes ($t$) was used as a proxy of organic pollution.
The data used came from the FAO and gaps in data were modelled using a linear
regression between pesticides and fertilizers or agricultural GDP. The same
methodology as that used to characterized nutrient pollution was then applied
to organic polllution.

For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
1 $km^2$ grid cell using weighted area average.

### Toxic algae

The data we use to describe the risk of toxic algae comes from an expert based
map delineating the areas where coastal areas are at risk from five different
toxins [@bates2019]. The map presents coastal areas at risk from 5
different toxins: 1) paralytic shellfish poisoning (PSP) toxins from the regular
presence of the dinoflagellate *Alexandrium catenella* (previously
*Alexandrium tamarense*) at high concentrations, 2) amnesic shellfish
poisoning (ASP) toxins from domoic acid 3) diarrhetic shellfish poisoning (DSP)
toxins, 4) spirolides and 5) pectenotoxins,  two toxins produced by
dinoflagellates occurring in the St. Lawrence.

The information provided on this expert map on the 5 toxins [@bates2019].
was georeferenced and transformed as vecctorized objects.
We calculated a toxic algae risk ($T$) index for each cell ($x$) in the 1 $km^2$
study grid. For each toxin ($t$), a value of 1 was attributed to all grid cells
overlapping with areas identified at risk on the expert map and a value of 0.5
for grid cells overlapping with areas where ASP and DSP toxins were observed
without exceeding legal thresholds. The value for all 5 toxins was them summed
for all grid cells:

$$TA_{i,x} = \sum_{i = 1}^{5} i_x$$

## Fisheries

The impacts of fisheries activities in the St. Lawrence are evaluated using
DFO's fisheries logbooks program [@dfo2016]. While logbooks are not mandatory
for all fisheries in the St. Lawrence, they still provide a very thorough
overview of the spatial distribution and intensity of fishing activities in the
St. Lawrence. The data we used spans 6 years from 2010 to 2015 and details
218323 fishing events (36387 $\pm$ 3147 fishing events per year). There were 31
targetted species and a total of 53 caught species in the dataset.

Fishing activities are performed using a variety of gear types: trap, trawl,
dredge, driftnet, hand line, longline, scuba diving, purse seine, seine, beach
seine and jig fishing. Intensity of fishing activities was divided among gear
types and based on their respective types of environmental impacts (Table S2).
For example, traps and trawls have very different effects on a system. Gear
classification was done using the classification presented in @halpern2008 and
@halpern2015b and is broken down into 5 distinct classes: demersal destructive
(DD), demersal, non-destructive, low-bycatch (DNL), demersal, non-destructive,
high-bycatch (DNH), pelagic, low-bycatch (PLB) and pelagic, high-bycatch (PHB).
This categorization therefore divides the fisheries data into 5 distinct
driver layers characterizing fishing activities.

Gear types can also be further classified into fixed or mobile engines based
on their mobility. We used these two mobility classes to generate a buffer of
impact around each fishing activity coordinates to consider potential spatial
uncertainty associated with locations and the fact that mobile engines can be
tracted over several kilometers during fishing activities and that we do not
have the beginning and end points of mobile fishing events. Buffer sizes for
fixed and mobile engine was of 200 and 2000 meters, respectively.

Table S2. Classification of gear types in the fisheries dataset based on their
environmental impact and mobility

Gear type (EN)    | Classification    | Mobility
--------------    | --------------    | --------
Trap              | DNH               | Fixed
Trawl             | DD                | Mobile
Dredge            | DD                | Mobile
Driftnet          | PHB               | Fixed
Hand lines        | PLB               | Fixed
Longline          | PHB               | Fixed
Scuba diving      | DNL               | Fixed
Purse seine       | PLB               | Fixed
Seine             | DNH               | Fixed
Beach seine       | DNH               | Fixed
Trap              | DNH               | Fixed
Jig fishing       | PLB               | Fixed

In order to characterize the intensity of fishing activities ($FI$), we used
a biomass yield density index. We multipled the total annual biomass captured
in each grid cell *j*, regardless of species, by the proportion of fishing area
in each grid cell:

$$FI_j = \sum_{k=1}^{n_j} B_{tot, k} * \frac{A_{j,k}}{A_{tot,k}}$$

where $j$ is a grid cell, $k$ is a fishing event, $B_{tot}$ is the total biomass
of a fishing event $k$, $A$ is the area of a fishing event *k* overlapping a
cell $j$ and $A_{tot}$ is the total area of the fishing event $k$. This formula
gives an intensity measurement in biomass units, which is kg in our case. Since
we measure the intensity within a 1 $km^2$ grid cell, the intensity evaluation
is in $kg * km^{-2}$. This metric distributes the biomass captured within each
grid cell as a function of overlapping fishing area and provides an overview of
how impacted each grid cell is in terms of extracted biomass.

## Marine traffic

### Shipping

The data used to characterize shipping comes from the global cumulative impacts
assessment on habitats [@halpern2008; @halpern2015] and available on the NCEAS
online data repository [@halpern2015b]. We provide a brief summary of data and
methods in this document. For more details, refer to @halpern2015.

Two data sources were used to characterize shipping. The first set of data is
gathered as part of the World Meteorological Organization Voluntary Observing
Ships' (VOS) scheme [@jcomm2018]. Ships participating in the program gather
meteorological data along with observation location as part of an open-ocean
climate dataset. The data spans 20 years and annually covers 10-20% of ships
worlwide. Data used spanned 2003 to 2011.

The second set of data comes from the Automatic Identification System (AIS),
an initiative launched in 2002 that sought to improve marine safety by providing
mariners with real-time vessel traffic [@tetreault2002]. Through the
International Maritime Organization SOLAS agreement, all vessels of over 300
gross tonnage on international voyages and those carrying passengers are now
required to be equipped with AIS transceivers. These transceivers use Global
Positioning System technology to locate vessels every 10 minutes. The data used
was from November 2010 to December 2011.

Data used come mostly from vessels that move globally (*i.e.* cargo, tanker and
passenger), as they are required to carry AIS transceivers, but also include
data from fishing, high-speed, pleasure and support classes. Shipping intensity
was evaluated as the number of fishing tracks at a 0.1 decimal degrees resolution.
For more details on data and methods used, consult [@walbridge2013].

For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
1 $km^2$ grid cell using weighted area average.

### Invasive species

The data used to characterize invasive species risk comes from the global
cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
available on the NCEAS online data repository [@halpern2015b]. We provide a
brief summary of data and methods in this document. For more details, refer to
supplementary materials provided in @halpern2008 and @halpern2015.

Cargo volume was used as a proxy of invasion risk under the assumption that
risk of invasion is proportional to tonnes of goods transferred through ports.
Cargo throughput in metric tonnes for the year 2011 was accessed through a
variety of sources [see supplementary material in @halpern2015 for more details]
and cross-matched with entries in the World Port Index database (WPI; available
from the National Geospatial-Intelligence Agency). A gap-filling procedure
using linear regression and sets of predictors related to port volume and
available in the WPI dataset was then applied to the WPI dataset to predict
missing cargo volume entries. Finally, volume data was distributed in marine
environments adjacent to ports using a diffusive plume model with an exponential
decay function that set the maximum spread distance to approximately 1000 km.
The plume model was then clipped to areas less than 60 m deep, as
invasives species are more likely to invade shallow areas.

For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
1 $km^2$ grid cell using weighted area average.

### Marine pollution

The data used to characterize marine pollution risk comes from the global
cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
available on the NCEAS online data repository [@halpern2015b].
Marine pollution was considered to be mainly driver by the shipping industry.
As such, the driver layer was constructed by combining the shipping (*i.e.*
shipping lanes) and invasives species (*i.e.* cargo volume) layers.
invasive. For more details, refer to supplementary materials provided in
@halpern2008 and @halpern2015.

For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
1 $km^2$ grid cell using weighted area average.

# Driver intensity and distribution

We evaluated the frequency distribution of each drivers to verify whether data
should be transformed (Figure S\ref{hist}). In light of this, we log-transformed
the following driver layers

- Coastal development
- Direct human impact
- All fisheries data
- Hypoxia
- Inorganic pollution
- Invasive species
- Nutrient pollution
- Organic pollution
- Sea bottom temperature anomalies
- Shipping

To allow for relative intensity comparison, all driver layers were subsequently
normalized between 0 and 1 using the 99th quantile to further control for
extreme values (Figure S\ref{drivers}).

![Frequency distribution of the untransformed data for all driver layers. \label{hist}](./figures/drHist.png)

![Distribution and intensity of transformed and normalized drivers in the Estuary and Gulf of St. Lawrence available on *eDrivers*. \label{drivers}](./figures/drivers.png)

\newpage

# Cumulative exposure

![Frequency distribution of cumulative exposure (*i.e.* sum of normalized driver intensity in each grid cell) and percent contribution of each driver to the frequency distribution of cumulative exposure in the Estuary and Gulf of St. Lawrence \label{marimekko}](./figures/marimekko.png)

# Threat complexes

## Clustering

We identified regions with similar cumulative exposure regimes, referred to as
threat complexes after [@bowler2019], using a partional *k-medoids* clustering
algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
designed to use with large datasets. The CLARA algorithm uses the PAM
(Partition Around Medoids) algorithm on a sample from the original dataset
to identify a set of $k$ objects that are representative of all other objects,
*i.e.* medoids (also referred to as examplars) and that are central to the
cluster they represent. The goal of the algorithm is to iteratively minimize
dissimilarity between objects in the cluster and the representative medoid.
Sets of $k$ medoids are thus selected and replaced in order to iteratively
minimize the intra-cluster dissimilarity. With the CLARA algorithm, this
process is repeated and iterations are compared on the basis of their average
dissimilarity between objects and cluster medoids to select the optimal set of
$k$ medoids that minimizes average dissimilarity. We performed the analyses for
a total of 100 iterations using samples of 10000 observations, representing ~5%
of the drivers dataset. Analyses were performed using the *cluster* R package
[@maechler2018].

Because the algorithm identifies clusters by minimizing dissimilarity around a
set of *k* medoids, it requires a pre-defined number of clusters. The
appropriate number of clusters $k$ was thus tested using values ranging from
2 to 10 and validated by selecting the number of clusters that maximized the average
silhouette width [@kaufman1990] and minimized the total within-cluster sum of
squares (WSS; Figure S\ref{valid}).

We also validated the clustering by comparing *k-medoids* clustering with
*k-means* clustering with the *Lloyd* algorithm [@lloyd1982]. The *k-means*
approach is similar to the *k-medoids*, but identifies observations belonging
to a cluster iteratively by minimizing the mean intra-cluster squared distance
until it converges to an optimal solution. We used 25 random sets and set a
maximum of 1000 iterations for the analysis. Analyses were performed using
the *stats* R package [@rcoreteam2018]. We used the same validation procedure
to select the optimal number of clusters $k$ than with the *k-medoids*
clustering (Figure S\ref{valid}).

Both the *k-medoids* and *k-means* clustering validation procedure suggest that
there are 7 distinc threat complexes in the St. Lawrence (Figure S\ref{valid}).
While the comparison of of the spatial distribution of clusters results in only
a 66% cell correspondance between the *k-medoids* and *k-means* approaches, the
bulk of the difference is attributable to a single undivided cluster that
combines most of the southern and norther Gulf in a single cluster in favor of
a more localised cluster by the *k-means* algorithm (Figure S\ref{comparison}).
While *k-means* algorithms are much faster than *k-medoids* since it does not
need a pairwise dissimilarity measurement of all observations, it is also more
sensitive to outliers since it focuses on the mean rather than a centroid. We
therefore opted the use of *k-medoids* clustering.

![Validation procedure for the *k-medoids* and *k-means* clustering algorithms based on the number of cluster that maximizes average silhouette width [upper panels; @kaufman1990] and minimizes the total within-cluster sum of squares (WSS; lower panels). \label{valid}](./figures/validation.png)

![Comparison of the distribution of the 7 clusters identified using the *k-medoids* and *k-means* clustering algoriths and the percent grid cell correspondance between the two approaches \label{comparison}](./figures/clusterComp.png)

## Inter-cluster dissimilarity

The difference between clusters was explored by measuring the total inter-cluster
dissimilarity and the contribution of each driver to the total inter-cluster
dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis
dissimilarity [Figure S\ref{inter}; @clarke1993]. As the drivers dataset is too
large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting
5% of each cluster to run the analysis and repeating the process over 300 iterations.
We also compared the mean intensity of each driver within each cluster to
better capture the inter-cluster dissimilarity. Analyses were performed using
the *vegan* R package [@oksanen2018].

![Evaluation of inter-cluster dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis dissimilarity [@clarke1993]. The figure diagonal presents the distribution of the 7 clusters identified using the *k-medoids* clustering algorithm. The lower triangle shows all combinations of inter-cluster dissimilarity with circular barplots showing the percent contribution to total dissimilarity of each driver and with the total inter-cluster dissimilarity in the center of the barplots. The upper triangle shows the average relative intensity of each driver for all driver combinations, with barplots to the left and the right representing the row and columns clusters, respectively. \label{inter}](./figures/interDissimilarity.png)

## Intra-cluster similarity

Intra-cluster similarity was evaluated using the Bray-Curtis similarity index
(Figure S\ref{intra}). As with the inter-cluster dissimilarity, we used a
bootstrap procedure for the intra-cluster similarity, randomly selecting 5% of
each cluster observation to run the analysis and repeating the process over 300
iterations. We however did not use the bootstraping procedure for clusters 2
and 6, as they both contain less than 10000 observations, making the calculation
time more manageable.

![Evaluation of intra-cluster similarity using the Bray-Curtis similarity index. The distribution of all 7 clusters is presented along with circular barplots showing the percent contribution to total similarity of each driver and with the total intra-cluster similarity in the center of the barplots. \label{intra}](./figures/intraSimilarity.png)

# References

### 2019-03-22 eDriversMS.Rmd
genesis

---
title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
author: David Beauchesne^1,2,\*^, Rémi M. Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^5^, Simon Bélanger^4^, Pascal Bernatchez^4^, Hugo Bourdages^7^, Clément Chion^6^, Peter S. Galbraith^7^, Benjamin Halpern^8,9,10^, Camille Lavoie^2^, Christopher W. McKindsey^7^, Alfonso Mucci^11^, Simon Pineault^12^, Michel Starr^7^, Anne-Sophie Ste-Marie^5^, Philippe Archambault^2^
date: 2019-02-06
fontsize: 12pt
output:
  pdf_document:
    toc: false
    number_sections: true
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   # - \usepackage{titling}
   # - \pretitle{\begin{flushleft}}
   # - \posttitle{\end{flushleft}}
   - \usepackage{lscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
relativeurls: true
bibliography: eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
cd phdproj/eDrivers/eDriversMS/
rmarkdown::render('./eDriversMS.rmd', 'pdf_document')
rmarkdown::render('./eDriversMS.rmd', 'word_document')
-->

<!--
<style type="text/css">
/* body{
  font-size: 12px;
}
td {
  font-size: 8px;
} */
h1 {
  font-size: 38px;
  color: DarkRed;
}

Style specs
Title{
  font-size: 16
  font: times new roman
  align: center
}
Headings & subheadings{
  font-size: 12
  font: times new roman
  bold
  numbered
}
Abstract{
  font-size: 12
  font: times new roman
  normal
}
Body{
  font-size: 12
  font: times new roman
  single-spaced
  page numbers
  line numbers
}

</style>
-->

<!-- Setting up parameters for code snippet presented in Box 1 -->
\lstset{language=R,
    basicstyle=\small\ttfamily,
    stringstyle=\color{DarkGreen},
    otherkeywords={fetchDrivers, install_github, getData, kde2d},
    morekeywords={TRUE,FALSE},
    deletekeywords={data,frame,length,as,character},
  keywordstyle=\color{blue},
  commentstyle=\color{DarkGreen},
}

^1^Institut des sciences de la mer, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^2^Québec Océan, Département de biologie, Université Laval, Québec, QC, Canada \newline
^3^Département de biologie, Université de Sherbrooke, Sherbrooke, QC, Canada \newline
^4^Département de Biologie, Chimie et Géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^5^St. Lawrence Global Observatory, Rimouski, QC, Canada \newline
^6^Institut des sciences de la forêt tempérée, Université du Québec en Outaouais, Gatineau, QC, Canada \newline
^7^Fisheries and Oceans Canada, Maurice Lamontagne Institute, Mont-Joli, QC, Canada \newline
^8^National Center for Ecological Analysis and Synthesis, University of California, Santa Barbara, CA, United States \newline
^9^Bren School of Environmental Science and Management, University of California, Santa Barbara, CA, United States \newline
^10^Imperial College London, Ascot, United Kingdom \newline
^11^Department of Earth & Planetary Sciences, McGill University, Montréal, QC, Canada \newline
^12^Ministère Environnement et Lutte contre les changements climatiques, Québec, QC, Canada  \newline

**Correspondence**: \newline
David Beauchesne \newline
david.beauchesne@uqar.com \newline

**Number of figures**: 5 \newline
**Number of tables**: 1 \newline

\newpage

# Abstract

Gathering environmental data for large scale, systematic initiatives such as
ecosystem-based management is a very challenging process. Building such
holistic datasets is made increasingly possible through initiatives like the
Ocean Biogeographic Information System (OBIS) and Bio-ORACLE. Data on
environmental drivers (*e.g.* climate change and fisheries) are however largely
missing even though their characterization is crucial to a better understanding
of the effects of multiple environmental threats. In this paper we discuss how
we are addressing this issue by launching an open data platform called *eDrivers*, focused
on characterizing and sharing data on the distribution and intensity of drivers
and using the Estuary and Gulf of St. Lawrence in eastern Canada as a case study.
The core objective underlying this initiative is to create a community-led
project built on a series of guiding principles aiming at upholding the highest
standards of data management and open science. The platform currently shares data for 22 drivers
including coastal, climate, fisheries and marine traffic related layers. Using
the currently available data, we show that few areas are free of cumulative
exposure in the St. Lawrence. The Estuary, the Anticosti Gyre and coastal areas
are particularly exposed, especially in the vicinity of urban centers. We also
identify 7 distinct areas of similar cumulative exposure regime (*i.e.* threat
complexes) that show that numerous drivers typically co-occur in different
regions of the St. Lawrence and that coastal areas are exposed to all driver
types. These observations are destined to improve as *eDrivers* evolves
through time to address gaps in knowledge, incorporates additional driver
layers and refines the characterization of current driver layers. Ultimately,
we believe that *eDrivers* represents a much needed solution that could
radically influence broad scale research and management practices by increasing
data accessibility and interoperability and by increasing research and
decision-making efficiency. All that is required is a commitment to open data,
adaptive monitoring and an integrated vision of ecosystem management.

**Keywords**: ocean observing systems, St. Lawrence, environmental drivers,
cumulative exposure, threat complex, multiple stressors, global change

# Introduction

With awareness of human impacts on the oceans increasing globally, there has
been a growing demand for systematic and large scale environmental management
schemes, such as ecosystem-based management, systematic conservation planning,
and strategic environmental assessment [@margules2000; @jones2016]. Beyond
substantial political and scientific commitment, such initiatives require the
proper infrastructures and tools allowing for efficient data reporting and
adaptive, continuous monitoring, all of which hinge on the construction of
large, holistic datasets.

Gathering environmental data for large scale, systematic initiatives can however
be a very challenging and time consuming - not to say painful - process. On one
hand, there is an overwhelming and expanding wealth of data available. Such
information overload may reduce ability to take decision based on scientific
information, cause massive effort duplication, disproportionately appropriate
research funds for certain sectors, and obscure knowledge gaps amid a sea of information
[@eppler2004]. On the other hand, crucial data are lacking and remain largely
unavailable or inaccessible for a variety of reasons, including proprietary
rights, lack of organizational time, capacity and training, and unwillingness
to share, curtailing our ability for appropriate decision-making. When coupled
with political intricacies, such as multi-jurisdictional management, the data
gathering process can quickly become intractable.

Moreover, the current digital infrastructure is highly
decentralized and the practices are heterogeneous, preventing us from maximizing
benefits from research investments [@wilkinson2016]. Per @wilkinson2016, *"the
reason that we often need several weeks (or months) of specialist technical
effort to gather the data necessary to answer such research questions is not
the lack of appropriate technology; the reason is, that we do not pay our
valuable digital objects the careful attention they deserve when we create and
preserve them"*.

There now exists multiple initiatives that address this issue by assembling,
organizing and sharing vast arrays of environmental data.
Biotic data can be accessed through web portals such as the Ocean Biogeographic
Information System [OBIS; @obis2018], the Global Biodiversity Information Facility
[GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
and the World Register of Marine Species [WoRMS; @worms2017]. Abiotic data can
also be accessed through WorldClim [@hijmans2005], Bio-ORACLE
[@tyberghein2012], and MARSPEC [@sbrocco2013]. Initiatives focused on sharing
environmental data for specific areas also exist, such as the St. Lawrence
Global Observatory (SLGO; https://ogsl.ca/en), the European Marine Observation
and Data Network (EMODnet; http://www.emodnet.eu/) and the U.S. Integrated Ocean
Observing System (IOOS; https://ioos.noaa.gov/). Essential environmental
parameters are also organized, coordinated and acquired through global
initiatives like the Group on Earth Observations Biodiversity Observation
Network [GEO BON; @scholes2012], the Census of Marine Life [CoML; @coml2010]
and the Global Ocean Observing System (GOOS; http://www.goosocean.org).
However, environmental drivers, *i.e.* any externality that affects environmental processes
and disturbs natural systems and often referred to as stressors or pressures, remain
conspicuously missing from most of these initiatives [but see @halpern2015].
This is in spite of the need for high resolution data on their distribution and
intensity at a scale suitable for environmental management and impact assessments.

Drivers may be natural (e.g. sea surface
temperature anomalies and hypoxia) or anthropogenic (e.g. fisheries and marine
pollution) in origin.
The potential for complex interactions between drivers remains the largest
uncertainty when studying or predicting environmental change
[@darling2008; @cote2016]. The effects of multiple drivers can combine
non-linearly and result in effects that are greater (synergistic effect) or
lower (antagonistic effect) than the sum of individual effects
[@crain2008; @darling2008; @cote2016]. The uncertainty associated with complex
driver interactions must therefore be taken into account [@cote2016], yet most
research on driver effects in marine environments remains overwhelmingly focused
on single driver assessments [@obrien2019]. This is partly explained by the
complexity of studying multiple drivers simultaneously [@cote2016], but also
by the need for standardized data management practices and tools tailored to
the study of the effects of multiple drivers [@dafforn2016; @stock2018].

To address these issues, we are launching an initiative called *eDrivers*, an
open data platform focused on the description of the spatial distribution and
intensity of drivers and using the Estuary and Gulf of St. Lawrence in Canada
as a case study. The overarching goal of this platform is to create an expert
community committed to structuring, sharing and standardizing drivers data in
support to science and management.

The objective of this paper is twofold. The first is to present the guiding
principles underlying the inception and structure of *eDrivers*. We also wish
this paper to serve as a call for collaboration to any person, group or
organization who might be interested in contributing to this initiative. The
second objective focuses using the platform content to describe the intensity and distribution of drivers
in the St. Lawrence system and aims at 1) identifying areas of high cumulative exposure,
2) at characterizing areas with similar cumulative exposure regimes, dubbed threat
complexes [@bowler2019] and 3) at identifying drivers that are likely to interact
in the St. Lawrence.

# The St. Lawrence System

The St. Lawrence system, formed by one of the largest estuaries in the world and a
vast interior sea, is an complex social-ecological system characterized by highly
variable environmental conditions and oceanographic processes, both in space
and time [@el-sabh1990; @white1997; @dufour2007]. It thus offers a unique and
heterogeneous array of habitats suited for the establishment of diverse and
productive ecological communities [@savenkoff2000]. It concomitantly provides
a wealth of ecosystem services that have historically and contemporarily
benefited the Canadian economy. It holds a rich fisheries industry targeting
over 50 species, serves as the gateway to eastern North-America by granting
access to more than 40 ports, has a booming tourism industry and an expanding
aquaculture production, develops emerging activities and has a yet untapped
hydrocarbon potential [@beauchesne2016; @schloss2017]. With major
investments recently made and more forthcoming in economic and infrastructure
development and research [*e.g.* @goq2015; @rqm2018], an intensification of the
human footprint is expected in the St. Lawrence system.

Arising from this expected intensification and the experience of past and
current ecological tragedies [*e.g.* the collapse of cod fisheries and the
decline of the beluga population; @frank2005; @plourde2014; @dempsey2018] is a
call for an integrative and holistic approach to manage the St. Lawrence as a
whole. Yet, jurisdiction over the St. Lawrence is divided between the Canadian
federal government and five provincial governments (Newfoundland and
Labrador, New Brunswick, Nova Scotia, Prince Edward Island, and Québec) with
each governmental entity further divided into multiple departments.
Researchers from dozens of academic institutions and
countless local organizations also share an interest in the St. Lawrence. This
wide range of actors results in highly scattered data management
and sharing processes, which we wish to address through this open data platform
initiative.

# Open Data Platform

## Guiding principles

Five guiding principles motivate the development of the platform and shape its
structure (Figure \ref{diag}).

### Unity and inclusiveness

***Why***: Operating over such large scales in time, space and subject matter
requires a vast and diverse expertise that cannot possibly be possessed by any
one individual or organization. Consequently, we envision an initiative that
seeks to mobilize all individuals and entities with relevant expertise.

***How***: By promoting, consolidating and working with experts involved in
existing and highly valuable environmental initiatives already in place in the
St. Lawrence. Notable examples of environmental initiatives are the annual
review of physical [@galbraith2018], chemical, and biological [@blais2018]
oceanographic conditions in the St. Lawrence, the fisheries monitoring program
[@dfo2016], the annual groundfish and shrimp multidisciplinary survey
[@bourdages2018], the characterisation of benthic [@dutil2011], epipelagic and
coastal [@dutil2012] habitats of the St. Lawrence, and Canada's shoreline
classification [@eccc2018]. There are also nascent efforts to make available
information on several human activities in the St. Lawrence such as the Marine
Spatial Data Infrastructure portal, which provides data on zoning, shipping,
port activities, and other human activities in Canadian waters, including the
St. Lawrence system [@goc2016].

By working with existing data portals whose objective is to share
environmental data. We are thus collaborating actively with the St. Lawrence
Global Observatory (SLGO) to develop the initiative and to host the
platform on their web portal (https://ogsl.ca/en). The mission of SLGO is to
promote and facilitate the accessibility, dissemination and exchange of official
and quality data and information on the St. Lawrence ecosystem through the
networking of organisations and data holders to meet their needs and those of
users, to improve knowledge and to assist decision-making in areas such as
public safety, climate change, transportation, resources and biodiversity
conservation. The SLGO is also one of three regional association spearheading
the Canadian Integrated Ocean Observing System
(CIOOS; http://meopar.ca/research/cioos-call-for-proposals/),
which will focus on integrating oceanographic data from multiple sources to
make them accessible to end-users and to enable the national coordination of
ocean observing efforts by integrating isolated or inaccessible data, and by
identifying gaps or duplications in observations and research efforts.
We are also developing collaborations with the Portal on water knowledge
(http://www.environnement.gouv.qc.ca/eau/portail/), an initiative from the
Québec provincial government. This portal aims at collecting and sharing
accurate, complete and updated ressources on water and aquatic ecosystems to
support the mandate of relevant actors and stakeholders working in
water and aquatic ecosystems management.

By actively inviting, seeking, and developing collaborations as well as
encouraging constructive criticism from the inception and throughout the
lifetime of the platform.

By inviting external community contributions (Figure \ref{diag}).
External researchers or entities wishing to submit marine data will be able to
do so through the SLGO web portal (https://ogsl.ca/en). Submissions through
other data portals will also be accepted either through the development of data
sharing agreements or with the caveat that shared data are under an open source
license and that they adhere to the platform data standards.

### Findability, accessibility, interoperability and reusability

***Why***: Open data has been propelled to the forefront of scientific research
in an era of open, collaborative and reproducible science. By moving towards
large scale, cross-disciplinary research and management projects, there is a
growing need to increase the efficiency of data discovery, access,
interoperability and analysis, for both humans and computationally automated
processes [@reichman2011; @wilkinson2016]. Embracing open-data accelerates research
outcomes, improves transparency and replicability, assists in discovering
erroneous data and corroborating study findings, and decreases overall costs of
research initiatives [@reichman2011; @poisot2013; @lowndes2017]. As such, many
scientific journals (*e.g.* Frontiers, PLoS, Nature, Science), governmental agencies <!-- mettre nom de la revue ou on soumet -->
[*e.g.* @goc2016] and other institutions are adopting open
data-sharing policies and detailed management plans as part of their mandate.
Our goal is to foster efficient and functional open science by creating a
fully open, transparent and replicable open data platform.

***How***:
By building an infrastructure adhering to the FAIR Data Principles,
which states that data and metadata must be Findable, Accessible, Interoperable
and Reusable. These principles focus on the ability of humans and machines to
automatically find and (re)use data and knowledge [@wilkinson2016].

By making data and associated tools accessible through a variety of ways, namely
the SLGO web portal (https://ogsl.ca/en), two R packages called *eDrivers*
(https://github.com/orgs/eDrivers/eDrivers) and *eDriversEx*
(https://github.com/orgs/eDrivers/eDriversEx) to access the data through SLGO's
API and to provide analytical tools to explore data, respectively, and a Shiny
application (https://david-beauchesne.shinyapps.io/eDriversApp/) that allows
users to explore drivers data interactively (Figure \ref{diag}).
Note that the data are currently contained within and accessible through the
*eDrivers* R package only, as we are actively working to allow users to
download selected layers from SLGO's web portal and geoserver. The functions
available in *eDrivers* to access the data have however been developed to
ensure forward compatibility once the data migrate to SLGO's geoserver.

By defining clear data and metadata standards and specifications to support
the regional standardization of protocols and practices and to favour
interoperability with national and international programs like the SLGO web
portal, the upcoming CIOOS, Essential Ocean Variables (EOV) identified by the
Global Ocean Observing System [GOOS; http://www.goosocean.org] and the Ocean
Best Practices repository [OBP; https://www.oceanbestpractices.net/] from the
International Oceanographic Data and Information Exchange (IODE). The standard
used for the platform will thus be the North American Profile of ISO
19115:2003 - Geographic information - Metadata, a schema favoured for
geospatial data in Canada and the United-States.

By providing version control and code access to the workflows set up to generate
driver layers from raw data, the R packages and the Shiny application through
a GitHub organization called *eDrivers* (https://github.com/orgs/eDrivers/).

By committing to provide an increasingly FAIR infrastructure as we continue
developing it.

### Adaptiveness

***Why***: In the face of uncertainty and in an effort to address impending
environmental changes, adaptive management has been identified as the chief
strategy to guide efficient decision-making
[*e.g.* @costanza1998; @margules2000; @keith2011; @jones2016] and has already
been discussed in the context of multi-drivers and cumulative impact assessments
[@halpern2015; @beauchesne2016; @cote2016; @schloss2017]. Adaptive management can only be
truly achieved through a commitment to adaptive monitoring and data reporting
[@margules2000; @halpern2012; @lubchenco2015]. We further contend that adaptive
management requires the development of adaptive monitoring tools and
infrastructures, which we seek to address through a continuously-evolving platform.

***How***:
By setting up mechanisms structuring cyclic reviews of platform content,
for the integration of new material (*e.g.* data and methods) as it becomes
available or accessible, and by striving to provide time-series data that are
crucial to assess temporal trends and potentially early-warning signals of
ecosystem change (Figure \ref{diag}).

### Validation

***Why***: Using high-quality data is crucial to ensure robust analyses and
decision-making [@cai2015; @veiga2017]. Data quality assessment (*i.e.* quality
measurement, validation and classification) and management (*i.e.* quality
control and assurance) are therefore crucial steps in data sharing [@veiga2017].

***How***: By selecting a framework to assess and manage data quality and by
providing a data quality report with for all driver layers shared on the
platform.

### Recognition

***Why***: Like peer-reviewed publications, data must also be given its due
importance in scientific endeavors and thus be considered as legitimate citable
products contributing to the overall scientific output of data providers
[@codata2013; @force112014]. Appropriate citations should
therefore be provided for all data layers used and shared by the platform.

***How***: By including the persistent identifier, when possible, and full
citation of all data layers used and shared on the platform.

By listing all known projects and peer-reviewed publications using or citing
the data shared on the platform.

By adhering to the Data Citation Principles [@force112014], which
focus on citation practices that cover purpose, function and attributes of
citations in an effort to make citations human and machine readable as well as
providing appropriate credit to data products. As such, all data layers shared
on the platform will be identified by a persistent identifier (*e.g.* DOI),
possess a license outlining conditions of use and a clear *how to cite* label
including the authors, the data version and year, an informative title, the
persistent identifier and how to access the data (*e.g.* url).

# St. Lawrence drivers

## Data

The list of drivers for which we sought data was informed by global cumulative
impacts assessments initiatives [@halpern2008; @halpern2015], regional holistic
evaluation of the state of the St. Lawrence [@dufour2007; @benoit2012], and
through communications with regional experts. Drivers are varied in origin,
from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping), and from
anthropogenic (*e.g.* fisheries) to natural (*e.g.* temperature anomalies).
Through the data gathering process, we developed and continue developing
collaborations with regional experts and data holders.
We also use global data from seminal works on cumulative impacts on global
habitats [@halpern2008; @halpern2015] and available on the National Center for
Ecological Analysis and Synthesis (NCEAS) online data repository
[Table 1; @halpern2015b]. We selected global data that
were unavailable at the regional scale and that were available at a resolution
adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).

The platform currently provides access to 22 driver layers (Table 1) and we are
actively working on updating or developing additional driver layers.
We divided drivers into 4 groups: coastal, climate, fisheries and marine traffic
related drivers (Table 1). All data layers and methodologies used to produce
driver layers are described in the supplementary materials of this paper and on
the *eDrivers* GitHub organization (https://github.com/orgs/eDrivers/).
Also note that the specific
methodologies underlying the figures presented in the paper are described in
the figure captions to ease the flow of text.

## Cumulative exposure

Apart from the northeastern Gulf, cumulative exposure (*i.e.* sum of normalized
driver intensity) is ubiquitous in the St. Lawrence (Figure \ref{footprint}).
Cumulative exposure is generally highest along the coast (Figure \ref{footprint}),
with most intense hotposts (*i.e.* areas of overlapping drivers at high relative
intensity) located in the vicinity of coastal cities (Figure \ref{hotspot}).
In general, offshore areas are less exposed to cumulative drivers, with the
Estuary and the Anticosti Gyre being notable exceptions (Figures \ref{footprint}
and \ref{hotspot}). That is however not to say that offshore areas are free of
exposure, as most of the St. Lawrence is exposed to multiple drivers
(Figures \ref{footprint} and \ref{hotspot}).

These results are consistent with observations elsewhere in the world, where
cumulative exposure notably arises from and markedly intensifies close to
coastal cities and at the mouth of rivers draining highly populated areas
[e.g. @halpern2015; @feist2016; @mach2017; @stock2018]. These are areas
where human uses (*e.g.* coastal development and shipping) and footprint
(pollution runoff) are the most intense [@feist2016], and on which is
overlaid a background of natural disturbances [@micheli2016]. They are also the
areas in which the most dramatic increases in exposure are expected, with
populations increasing more rapidly along the coast than inland [@feist2016].

In the St. Lawrence, large coastal cities are mostly located along the Estuary
and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
or home to small coastal towns. Moreover, the Estuary, along with the
St. Lawrence river, provide access to and serve as the primary drainage
outflow of the Great Lakes Basin, which is the most densely populated region
in Canada [@statcan2017].

## Driver interactions

By providing data on driver intensity and distribution, the platform will
greatly facilitate the study of the effects of multiple drivers, their interaction
likelihood and the areas where they may interact.
For instance, drivers like hypoxia (Figure \ref{kernel}A) and demersal destructive
fisheries (Figure \ref{kernel}B) are known to occur mainly in deeper areas of the
St. Lawrence, and hence an interaction between the effects of the two drivers
could be anticipated. Fisheries in the St. Lawrence have historically affected
biodiversity distribution and habitat quality [@moritz2015]. Hypoxia, meanwhile,
decreases overall habitat quality but has variable effects that are
species-dependent, ranging from well-adapted [*e.g.* northern shrimp
*Pandalus borealis* and Greenland halibut *Reinhardtius hippoglossoides*;
@pillet2016] to reduced growth rates [@dupont2013] and avoidance of oxygen
depleted habitats [*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased
mortality [*e.g.* sessile benthic invertebrates; @eby2005; @gilbert2007; @belley2010].
Certain species may thus be negatively affected by fisheries and withstand hypoxia
but still experience a decrease in prey availability, while others may be
negatively affected by the compounded effect of both drivers [@deleo2017].

By combining both drivers (Figure \ref{kernel}C), we can evaluate
which regions in the St. Lawrence are more likely to be affected by their
simultaneous presence and evaluate the relative intensity at which they tend to
co-occur (Figure \ref{kernel}D). Doing this, we observe that hypoxia and bottom
fisheries tend to co-occur in the St. Lawrence and that they do so at relatively
high intensities, with extensively used fishing grounds also characterized by
low levels of oxygen neighboring the limits of hypoxia (*i.e.* ~0.7 mg*L^-1^,
corresponding to ~30% saturation). We demonstrate the ease with which this
can be accomplished using *eDrivers* by providing the code necessary to reproduce
Figure \ref{kernel} in box 1.

## Threat complexes

The number of driver overlapping in the St. Lawrence increases with cumulative
exposure (Figure SX). Areas with high exposure such as the Estuary, the
Anticosti Gyre and the southwestern Gulf coastline (Figure \ref{footprint} and
\ref{hotspot}) are thus areas in the St. Lawrence where driver interactions are
most likely, and where they can arise between a host of different drivers.
While we cannot ascertain that high exposure areas are also the most impacted,
we can safely predict that these are the areas where studying ecosystem state
will be the most complex due to the uncertainty associated with driver
interactions, an uncertainty certain to increase rapidly with the increasing
number of interacting drivers [@cote2016]. It is however important to note
that there are very few areas of the St. Lawrence that do not experience driver
overlap (Figure \ref{hotspot}) and that drivers do not necessarily need to be
co-occurring at high intensities to cause significant adverse effects
[*e.g.* @relyea2009; @liess2016].

In an effort to describe areas where interactions are most likely between
multiple and typical combinations of drivers, we identified areas with similar cumulative
exposure regimes, which we refer to as threat complexes after @bowler2019. Seven
distinct threat complexes were identified in the St. Lawrence, which can be
divided into 4 offshore and 3 coastal complexes (Figure \ref{cluster}). Refer
to the supplementary information for the figures related to the following
description of threat complexes.

Coastal threat complexes (2, 4 and 6; Figure \ref{cluster}) are differentiated
from offshore complexes by the inclusion of
all coastal drivers. They however also include all other types of drivers except
hypoxia. They are also the most exposed threat complexes, both in terms of
driver overlap and intensity. Among coastal complexes, complex 2 is
differentiated by the presence of aquaculture sites. Threat complex 4, which is
widespread through the coastal Estuary and Gulf, includes areas associated with
port activities at higher relative intensity. Finally, complex 6, in the
southeastern and northernmost Gulf, is characterized by negative bottom
temperature anomalies.

Offshore threat complexes (1, 3, 5 and 7; Figure \ref{cluster}) are generally
characterized by climate and marine traffic related drivers. Threat complex 1
mostly incorporates the shallow areas of the northern Gulf and is characterized
by lower marine traffic intensity and with higher intensity surface temperature
anomalies. Threat complex 3 encompasses most of the southern Gulf and is
characterized by medium to high marine traffic intensity and higher aragonite
stress. Threat complex 5, meanwhile, is mostly delineated by the deep channels
of the northern Gulf and is characterized by positive bottom temperature
anomalies. Finally, complex 7 corresponds to the Laurentian channel, which is a
deep channel connecting the Estuary to the Atlantic. This threat complex is
characterized by high intensity hypoxia, marine traffic and pollution, and
demersal destructive fisheries.

Fisheries are generally distributed among all offshore threat complexes and
coastal complex 4. This widespread distribution among clusters causes fisheries
to have limited contribution to differences between complexes. There are however
certain threat complexes characterized by higher proportion of fishing
activities. Threat complex 3 holds higher intensities of pelagic low-bycatch and
demersal high-bycatch fisheries. Threat complex 5, meanwhile, has higher intensities of
demersal destructive fisheries. Finally, threat complex 7 is characterized by higher intensities
of demersal destructure and pelagic high-bycatch fisheries.

Describing these threat complexes allows us to see that each of those areas
are characterized by distinct overlapping suites of drivers. This holds
significant importance for the species, habitats and ecosystems located
within or moving through those threat complexes, as they are likely to be
affected by the joint effects of all drivers typical to those
complexes [@bowler2019].

# Perspectives and next steps

Understanding how ecosystem state will be affected by global change necessitates
a comprehensive understanding of how threats are distributed and interacting in
space and time, which hinges on appropriate data tailored to multi-drivers
studies [@dafforn2016; @stock2018; @bowler2019]. In the St. Lawrence,
we found that few areas are free of cumulative exposure and that the whole
Estuary, the Anticosti Gyre and coastal southwestern Gulf are particularly
exposed to cumulative drivers, especially close to urban areas. We also
identified seven geographically distinct threat complexes that show similar
cumulative exposure regimes. These complexes also show that coastal areas are
particularly exposed to all types of drivers and that multiple drivers are
typically co-occurring in space. The *eDrivers* initiative allows us to
efficiently identify these areas in need of heightened scrutiny. It also ensures
an iterative improvement of cumulative exposure assessments and threat complexes
identification as gaps in knowledge are addressed and approaches to describe
drivers are refined.

Arguably the most meaningful benefit anticipated from *eDrivers* is the
significant gain in data accessibility through the creation of a
cross-jurisdictional collaborative tool. This will effectively curtail the need
to reach dozens of experts across multiple organizations and over extensive
periods of time to assemble the data needed for systematic research. These
benefits could pay quick scientific and management dividends by drawing on the
knowledge and efforts of a wide range of contributors, by expanding avenues of
scientific inquiry, by decreasing overall effort duplication and research costs,
and by increasing research efficiency [@franzoni2014]. Critically - and we
emphasize this point - the platform will allow the scientific and governmental
communities to identify key knowledge gaps that will assist in prioritizing
and optimizing research efforts. For example, the platform could assist in
designing targeted experiments to study the effects of multiple drivers by
identifying sampling sites capturing appropriate exposure gradients
[@stock2018; @bowler2019].

Another key issue with data sharing that will be addressed by the platform is
data interoperability. By using and enforcing (meta)data standards the platform
will strongly encourage the standardization of future data collection protocols
to maintain compatibility. Standardized protocols will ease the regional
aggregation of local datasets that are currently highly fragmented in
availability, quality and experimental approaches [@reichman2011]. It will also
enable interoperability with other open data platforms (*e.g.* OBIS). This
standardization could also be accompanied by an initiative to extend
the Essential Ocean Variables (http://www.goosocean.org) to include a list of
essential driver variables in marine environments. Additional steps towards
standardization could also be taken by emulating protocols elaborated through
international programs like GEO BON [@scholes2012] and CoML [@coml2010].

We also believe that the *eDrivers* initiative will operationalize
evidence-based decision-making by streamlining data management and research,
allowing science output to be available and interpretable on a time scale
relevant to management. The platform will thus greatly facilitate the
application of broad scale, holistic research and management approaches such as
ecosystem-based management, strategic environmental assessments and
social-ecological metanetworks [*e.g.* @halpern2015; @jones2016; @dee2017].

Significant effort is still needed to bring our vision to fruition. Foremost on
that list is to maintain our efforts to foster collaborations, develop platform
content and identify key knowledge gaps. A fair and efficient organizational
structure will also be developed in order to manage the platform as a community.
Appropriate funding must also be secured to continue building this community and
ensure the long-term viability of the initiative, although the partnership with
SLGO partly addresses this issue.

Finally, terrestrial and coastal environments must also be incorporated, as
sources of stress within those habitats extend to the marine environments.
Moreover, despite coastal areas being recognized as the most exposed to
environmental threats we continue delineating
terrestrial and marine realms, using coastlines as we would an impermeable
barrier. While there is a sensible rationale for this division, we must strive
to eliminate this barrier if we are to appropriately study and predict the
impacts of global change [*e.g.* see @bowler2019].

Despite the work still ahead, we are hopeful that this initiative will be a very
successful one. Ultimately, this platform represents a much needed solution to
address important issues in data management that could radically shift broad
scale research and management practices towards efficient, adaptive and holistic
ecosystem-based management in the St. Lawrence and elsewhere in the world. All
it requires to be successful is for the scientific and political communities to
fully commit to open data, adaptive monitoring and, most of all, an integrated
vision of ecosystem management.

# Acknowledgements

We thank the Fond de Recherche Québécois Nature et Technologie (FRQNT) and the
Natural Science and Engineering Council of Canada (CRSNG) for financial support.
This project is also supported by Québec Océan, the Quebec Centre for
Biodiversity Science (QCBS), Takuvik, and the Notre Golfe networks. This
research is also sponsored by the NSERC Canadian Healthy Oceans Network and its
Partners: Department of Fisheries and Oceans Canada and INREST (representing the
Port of Sept-Îles and City of Sept-Îles).

# Author contributions statement

TO WRITE

# Conflict of interest statement

The authors declare that the submitted work was carried out in the absence of any personal, professional or financial relationships that could potentially be construed as a conflict of interest.

\newpage

# Listings

Box 1. Code snippet demonstrating how to use the *eDrivers* to reproduce
figure \ref{kernel} in R.

\begin{lstlisting}[frame=single]
# Install and load eDrivers package
devtools::install_github('eDrivers/eDrivers')
library(eDrivers)

# Load data
drivers <- fetchDrivers(drivers = c('hypoxia','fishDD'))

# Get data from `eDrivers` class object
driverData <- getData(drivers)

# Normalize data
driverData <- driverData / cellStats(driverData, 'max')

# Visualize data and combination
plot(driverData$fishDD)   # Demersal destructive fisheries
plot(driverData$hypoxia)  # Hypoxia
plot(sum(driverData))     # Combination

# Identify values > 0 and not NAs
driverData$fishDD[driverData$fishDD < 0] <- NA
driverData$fishDNH[driverData$hypoxia < 0] <- NA
id0 <- !is.na(values(driverData$fishDD)) &
       !is.na(values(driverData$hypoxia))

# 2D kernel for driver co-intensity
library(MASS)
coInt  <- kde2d(x = values(driverData$fishDD)[id0],
                y = values(driverData$hypoxia)[id0],
                n = 500, lims = c(0, 1, 0, 1))
image(coInt, zlim = c(0,max(coInt$z)))

# Driver density distribution
plot(density(driverData$fishDD[id0]))   # Demersal destructive
plot(density(driverData$hypoxia[id0]))  # Hypoxia
\end{lstlisting}

\newpage

# Figures

![Diagram of the platform structure. Community input in the form of raw data is accessed through the St. Lawrence Global Observatory (SLGO); https://ogsl.ca/en) repository - the platform host - or through open access repositories (*e.g.* NASA data). The raw data are then processed through a workflow hosted on the *eDrivers* GitHub organization (https://github.com/orgs/eDrivers/). Data processing may be as simple as data rescaling (*e.g.* night lights) or make use of more complex methodologies (*e.g.* waste water). All data is then hosted on SLGO's geoserver and accessible through their API. We developed a R package called *eDrivers* to access the driver layers through R and we are actively developing a second R package called *eDriversEx* that includes analytical tools to explore drivers data. Finally, we have developed a Shiny app that allows users to explore drivers data interactively (https://david-beauchesne.shinyapps.io/eDriversApp/). All R components of the project are also hosted on the *eDrivers* GitHub organization. \label{diag}](./figures/structure.png)

\newpage

![Distribution of driver footprint in the Estuary and Gulf of St. Lawrence. The footprint is measured by summing the relative intensity of all drivers for each grid cell: $F_d = \sum_{i=1}^{n} D_i$. Driver layers were also log-transformed when their frequency distribution was non-normal. All driver layers were also normalized between 0 and 1 using the 99th quantile to allow for direct comparison of relative intensities between drivers and control for extreme values. \label{footprint}](./figures/footprint.png)

\newpage

![Distribution of hotspots of driver footprint in the Estuary and Gulf od St. Lawrence. Hotspots were identified using the sum of all driver's 80th intensity quantile. This results in the distribution of areas with overlapping drivers at high intensity. \label{hotspot}](./figures/hotspot.png)

\newpage

![Example of platform content using the spatial distribution and the intensity of hypoxia and demersal destructive fisheries in the St. Lawrence. An index of hypoxia (**A**) was created using bottom oxygen concentrations from a newly instated survey during the months of August and September [@starr2019]. We interpolated the survey data using cokriging with depth as a covariable [@dutil2011]. A hypoxic index was then measured using a hypoxic threshold of 2 $mlL^{-1}$ bottom oxygen concentration [@diaz1995]. An inverted logistic curve was used to create an index of hypoxic stress ranging from close to hypoxic (*i.e.* 0) to anoxic (*i.e.* 1), under the assumption that hypoxic stress increases following a logistic curve as it approaches the hypoxic threshold. Demersal destructive fisheries (**B**) correspond to fishing activities using trawl and dredges. Intensity of fishing activities was measured using fisheries location data collected between 2010 and 2015 [@dfo2016] and a 2km buffer to consider trawling and dredging area. For each study area cell ($1 km^2$), fishing intensity (FI) was measured as the annual area weighted total biomass ($kg$) yield for all species captured: $FI = \sum_{i = 1}^{n} B_i * \frac{Area_{Fish}}{Area_{Cell}}$, where $B_i$ is the total annual biomass captured in each cell. As fishing intensity was highly skewed towards zeroes, we FI was log transformed to obtain a workable distribution. Values were then normalized to allow comparison with hypoxia values. Drivers intensity was summed (**C**) to visualize the combination of hypoxia and demersal destructive fisheries. Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a kernel analysis (**D**).\label{kernel}](./figures/kernel.png)

\newpage

![Distribution of threat complexes in the Estuary and Gulf of St. Lawrence. Threat complexes (a term coined by [@bowler2019]) are areas with similar cumulative driver exposure regimes. Threat complexes were identified using a partional *k-medoids* clustering algorithm, CLARA [CLustering for Large Applications; @kaufman1990]. The appropriate number of clusters $k$ was tested using a range of values and validated by selecting the number of clusters that maximized the average silhouette width [@kaufman1990] and minimized the total within-cluster sum of squares (WSS). We also validated the clustering by comparing *k-medoids* clustering with *k-means* clustering with the *Lloyd* algorithm [@lloyd1982] and found a 66% cell correspondence between both approaches, with most of the differences attributable to a single undivided cluster in the *k-means* approach. Seven distinct threat complexes were identified in the St. Lawrence. The difference between clusters was explored by measuring the total inter-cluster dissimilarity and the contribution of each driver to the total inter-cluster dissimilarity through a similarity percentage analysis (SIMPER) with Bray-Curtis dissimilarity [@clarke1993]. As the drivers data is too large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting 5% of each cluster to run the analysis and repeating the process over 300 iterations. A similar bootstrapping procedure was used to evaluate the intra-cluster similarity with the Bray-Curtis similarity using 300 iterations using 5% of each cluster data when the total number of observations in cluster exceeded 10000 observations. The *k-medoid*, *k-means* and SIMPER analyses were performed using the *cluster*, *stats* and *vegan* R packages, respectively [@maechler2018; @oksanen2018; @rcoreteam2018]. Refer to the Supplementary Materials for more details. \label{cluster}](./figures/cluster.png)

\newpage
\blandscape

# Tables

Table 1. List of drivers currently available on *eDrivers* and used for the analyses presented in this paper.

```{r table1, echo=F, fontsize=4}
load('./tables/table1.RData')
table1
```

\elandscape
\newpage

# References

<!--
Additional notes:

*Elements needed*:
  - Platform host (F-A-I-R): SLGO, (present CIOOS)
X  - Data access (F-A-I-R):
X    - SLGO: Rawdata, Geoserver, API
X    - eDrivers R package
X    - eDriversEx R package
X    - Shiny application
X  - Code access and version control (F-A-I-R):  eDrivers organization
X    - Drivers workflow, R packages, Shiny app
O  - Essential Drivers Variables (I): EOV, GOOS (EBV)
B1  - Rich metadata standard and ontology (F-I-R): SLGO, CIOSS, Ocean Best Practices(?)
B1  - Data and metadata format (I-R): JSON & TXT
B1  - Data citation (F-R): Data citation principles Force11
B1  - Persistent Identifier (F): DOI for all datasets used (if available) and shared
B1  - License agreement (A-R): detailed for each data shared
B1  - How to cite

Reusability:
- Describe the scope of your data: for what purpose was it generated/collected?
- Mention any particularities or limitations about the data that other users should be aware of.
- Specify the date of generation/collection of the data, the lab conditions, who prepared the data, the parameter settings, the name and version of the software used.
- Is it raw or processed data?
- Ensure that all variable names are explained or self-explanatory (i.e., defined in the research field’s controlled vocabulary).
- Clearly specify and document the version of the archived and/or reused data.
- Where the data came from (i.e., clear story of origin/history, see R1),
- who to cite and/or how you wish to be acknowledged.
- Include a description of the workflow that led to your data:
  - Who generated or collected it? How has it been processed?
  - Has it been published before?
  - Does it contain data from someone else that you may have transformed or completed?
  - Ideally, this workflow is described in a machine-readable format.
 -->

### 2019-03-22 eDriversMSv1.rmd
genesis

---
title: Next Generation Planning - Structuring and Sharing Drivers of Ecosystem Change Data for the St. Lawrence
author: David Beauchesne^1^, Rémi Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^4^, Simon Bélanger^1^, Pascal Bernatchez^1^, Clément Chion^5^, Peter S. Galbraith^6^, Benjamin Halpern^7^, Camille Lavoie^2^, Christopher W. McKindsey^6^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
date: 2018-09-19
fontsize: 12pt
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    fig_width: 3
  pdf_document:
    toc: false
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
relativeurls: true
bibliography: platformMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
setwd('/users/davidbeauchesne/dropbox/phd/phd_obj1/stressor_spat_dist/platformMS/')
rmarkdown::render('./platformMS.rmd', 'html_document')
rmarkdown::render('./platformMS.rmd', 'pdf_document')
rmarkdown::render('./platformMS.rmd', 'word_document')
-->

^1^Université du Québec à Rimouski \newline
^2^Université Laval \newline
^3^Université de Sherbrooke \newline
^4^St. Lawrence Global Observatory \newline
^5^Université du Québec en Outaouais \newline
^6^Department of Fisheries and Oceans Canada \newline
^7^University of California \newline

# Abstract

Gathering data for large scale, systematic research initiatives such as regional
multi-driver analyses can be a very challenging - not to say painful - process.
On one hand, there is an overwhelming wealth of data available, while on the
other hand, some important data remain largely unavailable or inaccessible.
Coupled with political intricacies, such as the multi-jurisdictional division
of the St. Lawrence system, the data gathering process can quickly become
intractable. To address this issue in the context of ecosystem based management,
we launched an open data platform focused on the spatial distribution and
intensity of drivers in the St. Lawrence. The objective of this platform is to
initiate a community project based on transparency, inclusivity and adaptiveness,
and to promote existing environmental monitoring programs while identifying key
knowledge gaps. We believe that this initiative is urgently needed for the
benefit of both scientific research and ecosystem based management initiatives
within the St. Lawrence.

# Context

With awareness of human impacts on the oceans increasing globally, there has
been a growing demand for systematic and large scale environmental management
schemes, such as ecosystem-based management, systematic conservation planning,
and strategic environmental assessment [@margules2000; @jones2016]. Beyond
substantial political and scientific commitment, such initiatives require the
proper infrastructures and tools allowing for efficient data reporting and
adaptive, continuous monitoring, all of which hinge on the construction of
large, holistic datasets.

Gathering data for large scale, systematic initiatives can however be a very
challenging - not to say painful - process. On one hand, there is an
overwhelming and expanding wealth of data available at our fingertips. On the
other hand, crucial data are lacking and remain largely unavailable or
inaccessible for a variety of reasons, including proprietary rights and
unwillingness to share. Both are fraught with pitfalls. Information overload
can lead to decreased decision-making ability, financial losses, delayed
decisions, massive effort duplication, disproportionate appropriation of
research funds, and it may obscure knowledge gaps amid a sea of information
[@oreilly1980]. The impacts of a lack of information are self-evident: without
knowledge, appropriate decision-making is beyond our grasp. Coupled with
political intricacies, such as multi-jurisdictional management, the data
gathering process can quickly become intractable.

There are now multiple initiatives that partly address this issue by assembling,
organizing and sharing vast arrays of relevant data. Biotic data can be accessed
through web portals such as the Ocean Biogeographic Information System
[OBIS; @obis2018], the Global Biodiversity Information Facility
[GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
and the World Register of Marine Species [WoRMS; @worms2017]. Abiotic data can
similarly be accessed through WorldClim [@hijmans2005], Bio-ORACLE
[@tyberghein2012], and MARSPEC [@sbrocco2013]. There are also global initiatives
such as the Group on Earth Observations Biodiversity Observation Network
[GEO BON; @scholes2012] and the Census of Marine Life [CoML; @coml2010] which
are focused on organizing, coordinating and acquiring essential ecological
parameters to support decision-making and management.

Drivers of ecosystem change are conspicuously missing from this list
[but see @halpern2015], even though data describing their distribution and
intensity at a resolution informative to decision-making and scientific research
are crucial to large-scale environmental management. For this paper, we define
drivers of ecosystem change (hereafter referred to as drivers) as any process
that may disturb a system beyond natural thresholds. Drivers may be
natural (*e.g.* sea surface temperature anomalies and hypoxia) or anthropogenic
(*e.g.* fisheries and marine pollution) in origin.

To address this issue, we are launching an open data platform focused on the
description of the spatial distribution and intensity of drivers in the Estuary
and Gulf of St. Lawrence in Canada, a unique and complex social-ecological
system. The overarching goal of this platform is to initiate a community-led
project based on transparency, inclusivity and adaptiveness that will increase
accessibility to critical data for scientific inquiries and decision-making.
We also wish to use this platform to promote pre-existing environmental
monitoring initiatives and identify key knowledge gaps. The objective of this
paper is to present the initiative, the guiding principles underlying its
inception, and its potential applications and benefits. Most importantly, we
wish to initiate a call for collaboration to any person, group or organization
who might be interested in contributing to this initiative.

<br/>

# The St. Lawrence System

The St. Lawrence, formed by one of the largest estuaries in North America and a
vast interior sea, is an ecologically complex system characterized by highly
variable environmental conditions and oceanographic processes, both in space
and time [@el-sabh1990; @white1997; @dufour2007]. It thus offers a unique and
heterogeneous array of habitats suited for the establishment of diverse and
productive ecological communities [@savenkoff2000]. It concomitantly provides
a wealth of ecosystem services that have historically and contemporarily
benefited the Canadian economy. It holds a rich fisheries industry targeting
over 50 species, serves as the gateway to eastern North-America by granting
access to more than 40 ports, has a booming tourism industry and an expanding
aquaculture production, develops emerging activities and a yet untapped
hydrocarbon potential [@beauchesne2016]. With major investments recently made
and more forthcoming in economic and infrastructure development and research
[*e.g.* @goq2015; @rqm2018], an intensification of the human footprint is
expected in the St. Lawrence.

Arising from this expected intensification and the experience of past and
current ecological tragedies [*e.g.* the collapse of cod fisheries and the
decline of the beluga population; @frank2005; @plourde2014; @dempsey2018] is a
call for an integrative and holistic approach to managing the St. Lawrence as a
whole. Yet jurisdiction over the St. Lawrence is divided between the Canadian
federal government and five provincial governments, *i.e.* Newfoundland and
Labrador, New Brunswick, Nova Scotia, Prince Edward Island, and Québec, with
each governmental entity further divided into multiple departments, each with
different mandates. Researchers from dozens of academic institutions and
countless local organizations also share an interest in the St. Lawrence.
This results in highly scattered expertises and heterogeneous data management
and sharing processes, which we wish to address through this open data platform
initiative.

# Open Data Platform

## Guiding principles

In accordance to the overarching goals of the platform, which is to establish a
transparent, inclusive and adaptive community-led project, to promote pre-existing
environmental monitoring initiatives and identify knowledge gaps, we propose
four foundational guiding principles:

- Create a unifying and inclusive project
- Foster open, transparent and replicable science
- Embrace adaptiveness as a core principle
- Adhere to the principles of scientific peer-review

### Unity and inclusiveness

**How**: By actively inviting, seeking, and developing collaborations as well as
encouraging constructive criticism from the inception and throughout the
lifetime of the platform.

**Why**: Operating over such large scales in time, space and subject matter
requires a vast and diverse expertise that cannot possibly be possessed by any
one individual or organization. Consequently, we are launching a crowd
science-like project [see @franzoni2014] that seeks to mobilize all entities
with relevant expertise. As such, we are inviting any individual or organization
interested in collaborating in this project to contact us.

### Openness, transparency and replicability

**How**: By striving to provide all data, metadata, methods descriptions, and
code used to generate platform content and structure.

**Why**: Open data has been propelled to the forefront of scientific research
in an era of big data science. By moving towards large scale, cross-disciplinary
research and management projects, there is a growing need to increase the
efficiency of data access, integration and analysis [@reichman2011]. Embracing
open-data accelerates research outcomes, improves transparency and replicability,
assists in discovering incorrect data and corroborating study findings, and
decreases overall costs of research initiatives
[@reichman2011; @poisot2013; @lowndes2017]. As such, many scientific journals
(*e.g.* PLoS, Nature, Science), governmental agencies [*e.g.* @goc2016] and
other institutions are increasingly adopting open data-sharing policies as
part of their mandate. Our goal is to foster open data science by creating a
fully open, transparent and replicable open data platform.

### Adaptiveness

**How**: By setting up mechanisms allowing for cyclic reviews of platform
content and for the integration of new material (*e.g.* data and methods) as it
becomes available or accessible.

**Why**: In the face of uncertainty and in an effort to address impending
environmental changes, adaptive management has been identified as the chief
strategy to guide efficient decision-making
[*e.g.* @costanza1998; @margules2000; @keith2011; @jones2016] and has already
been discussed in the context of multi-drivers and cumulative impact assessments
[@halpern2015; @beauchesne2016; @cote2016]. Adaptive management can only be
truly achieved through a commitment to adaptive monitoring and data reporting
[@margules2000; @halpern2012; @lubchenco2015]. We further contend that adaptive
management requires the development of adaptive monitoring tools, which we seek
to address for the St. Lawrence through a continuously-evolving platform.

### Peer-review

**How**: By ensuring that platform content has been processed through a rigorous
peer-review process.

**Why**: Peer-review is a fundamental tenet of science that ensures robustness
and quality of published research and lends credibility to communicated
scientific outputs [@ware2008; @rennie2016]. In an effort to
uphold the highest standards of scientific research, we wish to provide data
layers that have been thoroughly reviewed, ideally through peer-reviewed
journals. Peer-reviewed material will thus be clearly identified on the
platform, with appropriate citations provided.

## Benefits

Arguably the most meaningful benefit emerging from operating under our guiding
principles is the collective recognition of the depth of our current
understanding of the threats to the St. Lawrence. This community-wide surge in
consciousness could pay quick dividends in scientific and management
collaborations and efficiency, data management, and environmental monitoring.

The platform will also create a significant multi-jurisdictional collaborative
capacity between regional experts and create a community-led movement under
which local and specialized initiatives may be unified under a single umbrella.
This type of crowd science can, by drawing on the knowledge and efforts of a
wide range of contributors, expand avenues of scientific inquiry, decrease
overall research costs, increase research speed, and decrease effort duplication
[@franzoni2014]. This could easily lead to new and unforeseen avenues of
research and funding for new initiatives in the St. Lawrence. Critically,
awareness of the state of our knowledge will enable the identification of key
knowledge gaps, which will assist in prioritizing and optimizing research
efforts.

The platform will also promote and consolidate existing and highly valuable
environmental initiatives already in place in the St. Lawrence. Notable examples
are the annual review of physical [@galbraith2017], chemical and biological
[@devine2017] oceanographic conditions in the St. Lawrence, the fisheries
monitoring program that provides details on a significant proportion of fishing
activities in the St. Lawrence [@dfo2016] and the annual groundfish and shrimp
multidisciplinary survey [@bourdages2017]. There is also a valuable
characterisation of benthic [@dutil2011], epipelagic (*i.e.* 0-30m surface
layer) and coastal [@dutil2012] habitats of the St. Lawrence available as well
as a portal on Canada's shoreline classification [@eccc2018]. There are also
nascent efforts to make available information on several human activities in the
St. Lawrence. For example, the Marine Spatial Data Infrastructure portal
provides data on zoning, shipping, port activities, and other human activities
in Canadian waters, including the St. Lawrence [@goc2018].

Typically, however, many such initiatives originate from local projects
and their resulting datasets are often highly fragmented in availability,
quality, and experimental approaches [@reichman2011]. The platform project
could address this issue by supporting a push towards a regional standardization
of local protocols and practices, the identification of indicators of stress
within the St. Lawrence, and the establishment of a list of Essential Driver
Variables [EDV; see @pereira2013 for a description of the equivalent Essential
Biodiversity Variables]. Additional steps towards standardization could also be
taken by emulating protocols elaborated through international programs like
GEO BON [@scholes2012] and CoML [@coml2010].

Over the long-term, cyclic updates of platform content will also provide
time-series for drivers and provide valuable insights on the temporal trends of
drivers intensity and distribution. This could provide the necessary mechanisms
to allow for adaptive monitoring and help identify early-warning signals for
looming significant ecosystem transitions [*e.g.* @boettiger2013].

Finally, major gains in data availability and accessibility, for everyone, will
constitute major breakthroughs for environmental sciences. It will effectively
end the need to reach dozens of experts across dozens of departments over
extensive periods of time in order to assemble the data needed for integrated
research. Standardized protocols will also enable the production of comparable
and robust metadata and ease the regional aggregation of local datasets
[@reichman2011].

## Content

Building the backbone of the platform allowed us to initiate collaborations,
identify relevant experts in each field, integrate data on drivers that are
already freely and readily available, and begin identifying knowledge gaps.
It also serves as the stepping stone towards the long-term community project
we are envisioning.

We first identified a list of 34 drivers (Table 1) known to potentially impact
the St. Lawrence. This list was informed by global cumulative impacts
assessments initiatives [@halpern2008; @halpern2015], regional holistic
evaluation of the state of the St. Lawrence [@dufour2007; @benoit2012], and
through communications with regional experts.

We initialized the content of the platform using data from NCEAS’s online data
repository that was built through a global cumulative impact assessment
initiative [@halpern2015b]. Available global data for drivers deemed important
in the St. Lawrence thus serve as version 1.0 of driver layers on the platform
(Table 1). While some data layers are rather uninformative at the scale of the
St. Lawrence (*e.g.* acidification), some layers provide valuable information at
a resolution adequate for use at regional scales and currently constitute the
best available knowledge for the St. Lawrence (*e.g.* marine pollution).

We then initiated the production of version 2.0 of driver layers, either for
drivers for which data are unavailable globally (*e.g.* aquaculture) or for which
alternative data constitute an upgrade over the existing global data, typically
from regional or local initiatives (*e.g.* fisheries). We have thus far updated
or characterized 12 drivers and are actively working on the characterization of
12 drivers, with their development at various stages of completion (Table 1).
We are also actively seeking collaborations or data for 10 drivers. All versions
of driver layers will remain available on the platform. The development version
allowing data visualization is currently available at webpage.

# Applications

Assembling all drivers in a single platform will provide a wide array of
visualization and analytical possibilities that can be useful for both decision
making and scientific research. Here we describe two such potential applications.

## Multi-drivers assessment

This platform initiative will enable usual, single driver impact assessments,
but it will also pave the way towards holistic assessments of the state of the
St. Lawrence by enabling the study of combinations of multiple drivers. For
instance, drivers like hypoxia (Figure \ref{fig1}A) and demersal destructive
fisheries (Figure \ref{fig1}B) are known to occur mainly in deeper areas of the
St. Lawrence. By combining them (Figure \ref{fig1}C), we can evaluate which
regions in the St. Lawrence are more likely to be affected by both drivers
simultaneously and evaluate the relative intensity at which they tend to
co-occur (Figure \ref{fig1}D). Doing this, we observe that hypoxia and bottom
fisheries tend to co-occur in the St. Lawrence and that they do so at relatively
high intensities, with extensively used fishing grounds also characterized by
low levels of oxygen neighboring the limits of hypoxia (*i.e.* ~0.7 mg*L^-1^,
corresponding to ~30% saturation).

Fisheries in the St. Lawrence have historically affected fish stocks
and habitat quality [@moritz2015]. Hypoxia, meanwhile, decreases overall
habitat quality but has variable effects that are species-dependent, ranging
from well-adapted (*e.g.* northern shrimp *Pandalus borealis* and Greenland
halibut *Reinhardtius hippoglossoides*; [@pillet2016] to reduced growth rates
[@dupont2013] and avoidance of oxygen depleted habitats
[*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality
[*e.g.* sessile benthic invertebrates; @eby2005; @belley2010]. Certain species
may thus be negatively affected by fisheries and withstand hypoxia but still
experience a decrease in prey availability, while others may be negatively
affected by the compounded effect of both drivers. By providing data on multiple
drivers and their spatial interaction, the platform would yield valuable
insights for the integrated management of the St. Lawrence and the ability to
tackle complex problems not restricted to single drivers.

## Multi-scale management

Management is typically distributed across multiple jurisdictional, spatial and
temporal scales. For example, in the St. Lawrence, commercial fisheries are
managed at the federal level, with 3 regional jurisdictions. Ports and shipping,
meanwhile, are managed by a combination of local, provincial and federal
institutions. Such divisions, while necessary, isolate decision-making into
contextualized boxes even though decisions may have consequences reaching far
beyond its immediate context. With the appropriate time-series data provided by
cyclic updates of platform content and a commitment to adaptive monitoring, the
platform will allow us to visualize and evaluate management outcomes and
efficiency across scales and holistically. We draw on two topical issues in
the St. Lawrence and use simulated data to illustrate this.

The first issue is the case of the North Atlantic Right Whale
(NARW; *Eubalaena glacialis*), an endangered species that received a lot of media,
political and scientific attention in the summer of 2017 when 12 dead NARW and
5 live-entanglements were documented in the St. Lawrence [@daoust2017]. It was
established that the most likely causes of death were vessel collisions and
fishing activities [@daoust2017], and a series of management actions were put
in place to limit impacts to NARW [@dfo2018b; @tc2018]. One such action is a
static fixed engine fisheries closure in the Gulf of St. Lawrence in an area
encompassing 90% of 2017 NARW observations. With the platform, we can visualize
the fisheries intensity before (Figure \ref{fig2}B) and after the management
action (Figure \ref{fig2}D) to evaluate effects of management actions on fixed
engine fisheries.

The second issue is the case of port development initiatives in the
St. Lawrence. Five local port projects scattered through the St. Lawrence have
been submitted for an environmental assessment review at the Canadian
Environmental Assessment Agency [@ssl2018]. Of those project, we focus on the
marine terminal project on the north shore of the Saguenay River, for which a
draft environment assessment report was published [@ceaa2018] as well as a
science response by the Canadian Science Advisory Secretariat [@dfo2018]. It is
reported that the establishment of this port alone would lead to a 2.33-fold
increase in the number of ships in the Saguenay River by 2030, with vessels
transiting through the St. Lawrence to reach international markets
[@ceaa2018; @dfo2018]. With the platform, we can visualize shipping intensity
before (Figure \ref{fig2}C) and after (Figure \ref{fig2}E) to evaluate
the effects of this project on local marine traffic, but along the whole river
and the estuary.

Critically, the platform will allow users to explore the compounded effects of
management outcomes by providing the ability to explore effects on the
integrative distribution and intensity of drivers by considering them all
simultaneously (Figure \ref{fig2}A, F), or using any combination of drivers desired.
Similarly, the effects of multiple decision-making processes could be evaluated
in combination to see if their outcomes overlapped. This could provide an
invaluable tool for management as well as provide the tools necessary to justify
and encourage a strong commitment to adaptive monitoring.

# Ongoing development & perspectives

In this perspective paper we presented an adaptive open data platform focused on
the description and intensity of drivers of ecosystem change specifically built
for the Estuary and Gulf of St. Lawrence, Canada. The core objectives underlying
this initiative are to create a community-led project based on transparency,
inclusivity, and adaptiveness that will increase accessibility to critical data
for scientific inquiries and management schemes and help in identifying key
knowledge gaps.

With the project now officially launched, there are a number of initiatives
which we must continue to build and some that are in need of development.
Foremost on that list is to continue to foster collaborations, develop platform
content with relevant data and methods, and identify knowledge gaps. Concrete
mechanisms to adhere to all guiding principles must also be detailed, including
drafting a data sharing policy, developing standard metadata, building a fair
and efficient organizational structure, and secure funding for the long-term
integration of the platform and its guiding principles to the management of the
St. Lawrence.

Beyond what we already discussed, we also believe that the potential
applications of this platform are vast and far-reaching. It will undoubtedly
serve as critical information to broader research and management initiatives,
such as cumulative impacts assessments
[*e.g.* @halpern2008; @halpern2015; @beauchesne2016], the creation of a Canadian
Ocean Health Index [@halpern2012; @daigle2017] and the use of social-ecological
metanetworks for the management of ecosystem functions and services in the
St. Lawrence [@dee2017].

It could also be coupled with or serve as a means of operationalizing a network
of sampling stations across the St. Lawrence and provide real-time, integrated
data on drivers within the St. Lawrence to guide efficient decision-making and
monitoring [*e.g.* @onc2018]. Decision-making scenarios could also be tested
by simulating management outcomes and evaluating their influence on the
distribution and intensity of drivers in the St. Lawrence. We also expect other
unforeseen applications to emerge as the initiative gains momentum and users
begin using the content of the platform to address their own specific
objectives.

Ultimately, this platform represents a much needed solution to address important
issues in data management that could help in radically shifting environmental
impact assessment practices towards efficient, adaptive and holistic
ecosystem-based management of whole ecosystems, in the St. Lawrence and
elsewhere in the world. All it requires to be successful is for the scientific
and political communities to fully commit to open data, adaptive monitoring and,
most of all, an integrated vision of ecosystem management.

# Acknowledgements

TO WRITE

\newpage

# Tables

Table 1. List of drivers listed for inclusion in the multi-drivers platform. Their development status is detailed as such: data unavailable: no data is known to be available; seeking collaboration: we are actively seeking collaborations and expertise to develop this drivers layer; in development: we are in the process of building the driver layer, at varying levels of progress; complete: driver layer is complete for this version of the platform. Note that drivers identified as complete but identified with an * are were completed using data from [@halpern2015b] and that development of these drivers using regional data may still be pursued. Finally, a relative evaluation of progress made towards completion of each driver is presented. Black and grey boxes are for regional and global data development, respectively.

```{r table1, echo=F, fontsize=4}
readRDS('./table1.RDS')
```

\newpage

# Figures

![Figure 1. Example of platform content using the spatial distribution and the intensity of hypoxia and demersal destructive fisheries in the St. Lawrence. An index of hypoxia (**A**) was measured by normalizing values of bottom oxygen saturation percentage [@dutil2011] between 0 and 1 so that values close to 0 are fully saturated in $O_2$, while values close to 1 are hypoxic. Areas with less than 30% $O_2$ saturation (index $>$ 0.7) are generally considered hypoxic. Demersal destructive fisheries (**B**) correspond to fishing activities using trawl and dredges. Intensity of fishing activities was measured using fisheries location data collected between 2010 and 2015 [@dfo2016] and a 2km buffer to consider trawling and dredging area. For each study area cell ($1 km^2$), fishing intensity (FI) was measured as the annual area weighted total biomass ($kg$) yield for all species captured: $FI = \sum_{i = 1}^{n} B_i * \frac{Area_{Fish}}{Area_{Cell}}$, where $B_i$ is the total annual biomass captured in each cell. As fishing intensity was highly skewed towards zeroes, we FI was log transformed to obtain a workable distribution. Values were then normalized to allow comparison with hypoxia values. drivers intensity was summed (**C**) to visualize the combination of hypoxia and demersal destructive fisheries. Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a kernel analysis (**D**).\label{fig1}](./figure1.png)

\newpage

![Figure 2. Example of multi-scale management of the Estuary and Gulf of St. Lawrence using time-series data from the drivers platform. Regional (**B**, **C**) and local (**D**, **E**) scale decision-making are contrasted to show how management processes operating at different scales can influence (**A**, **F**). The data used for the period before management actions (**A**, **B**, **C**) is available on the platform, while the data used for the period after management actions (**D**, **E**, **F**) was simulated to correspond to their likely effects on drivers distribution and intensity. The effects of port development on shipping were simulated by increasing shipping intensity from the St. Lawrence Estuary through the Saguenay River to the projected marine terminal by the expected 2.33-fold increase in number of ships [**D**; @dfo2018]. The effects of the static fixed engine fisheries closure in the Gulf of St. Lawrence in response to North Atlantic Right Whale (*Eubalaena glacialis*) 2017 casualties were simulated by redistributing the mean total annual biomass yield from the closure area to neighbouring fishing zones (**F**). Effects of local and regional management actions were then used to update the overall distribution and intensity of all drivers in the St. Lawrence (**F**). \label{fig2}](./figure2.png)

\newpage

# Reference

### 2019-03-22 eDriversMSv2.rmd
genesis

--- 2019-03-22 eDriversMS.Rmd
@@
 title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
-author: David Beauchesne^1,2,\*^, Rémi M.
-Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^5^, Simon Bélanger^4^, Pascal Bernatchez^4^, Hugo Bourdages^7^, Clément Chion^6^, Peter S.
+author: David Beauchesne^1,2,\*^, Rémi Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^5^, Simon Bélanger^4^, Pascal Bernatchez^4^, Clément Chion^6^, Peter S.
 Galbraith^7^, Benjamin Halpern^8,9,10^, Camille Lavoie^2^, Christopher W.
-McKindsey^7^, Alfonso Mucci^11^, Simon Pineault^12^, Michel Starr^7^, Anne-Sophie Ste-Marie^5^, Philippe Archambault^2^
-date: 2019-02-06
+McKindsey^7^, Alfonso Mucci^11^, Michel Starr^7^, Anne-Sophie Ste-Marie^5^, Philippe Archambault^2^
+date: 2018-11-19
 fontsize: 12pt
 output:
+  html_document:
+    toc: true
+    toc_float:
+      collapsed: false
+    fig_width: 3
   pdf_document:
     toc: false
-    number_sections: true
   word_document:
@@
    # - \posttitle{\end{flushleft}}
-   - \usepackage{lscape}
-   - \newcommand{\blandscape}{\begin{landscape}}
-   - \newcommand{\elandscape}{\end{landscape}}
 relativeurls: true
-bibliography: eDriversMS.bib
+bibliography: platformMS.bib
 csl: frontiers.csl
@@
 
-<!--
-cd phdproj/eDrivers/eDriversMS/
-rmarkdown::render('./eDriversMS.rmd', 'pdf_document')
-rmarkdown::render('./eDriversMS.rmd', 'word_document')
--->
-
-<!--
-<style type="text/css">
+<!-- <style type="text/css">
 /* body{
@@
 
-</style>
--->
+</style> -->
 
-<!-- Setting up parameters for code snippet presented in Box 1 -->
 \lstset{language=R,
@@
 }
+
+<!--
+setwd('/users/davidbeauchesne/dropbox/phd/phd_obj1/stressor_spat_dist/platformMS/')
+rmarkdown::render('./platformMS.rmd', 'html_document')
+rmarkdown::render('./platformMSv2.rmd', 'pdf_document')
+rmarkdown::render('./platformMSv2.rmd', 'word_document')
+-->
 
@@
 ^11^Department of Earth & Planetary Sciences, McGill University, Montréal, QC, Canada \newline
-^12^Ministère Environnement et Lutte contre les changements climatiques, Québec, QC, Canada  \newline
 
@@
 David Beauchesne \newline
-david.beauchesne@uqar.com \newline
+david.beauchesne@hotmail.com \newline
 
-**Number of figures**: 5 \newline
+**Number of figures**: 3 \newline
 **Number of tables**: 1 \newline
@@
 # Abstract
+200 words
 
-Gathering environmental data for large scale, systematic initiatives such as
-ecosystem-based management is a very challenging process. Building such
-holistic datasets is made increasingly possible through initiatives like the
-Ocean Biogeographic Information System (OBIS) and Bio-ORACLE. Data on
-environmental drivers (*e.g.* climate change and fisheries) are however largely
-missing even though their characterization is crucial to a better understanding
-of the effects of multiple environmental threats. In this paper we discuss how
-we are addressing this issue by launching an open data platform called *eDrivers*, focused
-on characterizing and sharing data on the distribution and intensity of drivers
-and using the Estuary and Gulf of St. Lawrence in eastern Canada as a case study.
-The core objective underlying this initiative is to create a community-led
-project built on a series of guiding principles aiming at upholding the highest
-standards of data management and open science. The platform currently shares data for 22 drivers
-including coastal, climate, fisheries and marine traffic related layers. Using
-the currently available data, we show that few areas are free of cumulative
-exposure in the St. Lawrence. The Estuary, the Anticosti Gyre and coastal areas
-are particularly exposed, especially in the vicinity of urban centers. We also
-identify 7 distinct areas of similar cumulative exposure regime (*i.e.* threat
-complexes) that show that numerous drivers typically co-occur in different
-regions of the St. Lawrence and that coastal areas are exposed to all driver
-types. These observations are destined to improve as *eDrivers* evolves
-through time to address gaps in knowledge, incorporates additional driver
-layers and refines the characterization of current driver layers. Ultimately,
-we believe that *eDrivers* represents a much needed solution that could
-radically influence broad scale research and management practices by increasing
-data accessibility and interoperability and by increasing research and
-decision-making efficiency. All that is required is a commitment to open data,
-adaptive monitoring and an integrated vision of ecosystem management.
+**Keywords: open data, St. Lawrence, environmental drivers, holistic, ** up to 8, 5 minimum
 
-**Keywords**: ocean observing systems, St. Lawrence, environmental drivers,
-cumulative exposure, threat complex, multiple stressors, global change
-
-# Introduction
+# Context
 
@@
 information overload may reduce ability to take decision based on scientific
-information, cause massive effort duplication, disproportionately appropriate
-research funds for certain sectors, and obscure knowledge gaps amid a sea of information
+information, massive effort duplication, disproportionate appropriation of
+research funds, and it may obscure knowledge gaps amid a sea of information
 [@eppler2004]. On the other hand, crucial data are lacking and remain largely
@@
 
-Moreover, the current digital infrastructure is highly
+Additionally, the current digital infrastructure is highly
 decentralized and the practices are heterogeneous, preventing us from maximizing
@@
 and the World Register of Marine Species [WoRMS; @worms2017]. Abiotic data can
-also be accessed through WorldClim [@hijmans2005], Bio-ORACLE
+similarly be accessed through WorldClim [@hijmans2005], Bio-ORACLE
 [@tyberghein2012], and MARSPEC [@sbrocco2013]. Initiatives focused on sharing
@@
 and the Global Ocean Observing System (GOOS; http://www.goosocean.org).
-However, environmental drivers, *i.e.* any externality that affects environmental processes
-and disturbs natural systems and often referred to as stressors or pressures, remain
-conspicuously missing from most of these initiatives [but see @halpern2015].
-This is in spite of the need for high resolution data on their distribution and
-intensity at a scale suitable for environmental management and impact assessments.
 
-Drivers may be natural (e.g. sea surface
-temperature anomalies and hypoxia) or anthropogenic (e.g. fisheries and marine
-pollution) in origin.
-The potential for complex interactions between drivers remains the largest
-uncertainty when studying or predicting environmental change
-[@darling2008; @cote2016]. The effects of multiple drivers can combine
-non-linearly and result in effects that are greater (synergistic effect) or
-lower (antagonistic effect) than the sum of individual effects
-[@crain2008; @darling2008; @cote2016]. The uncertainty associated with complex
-driver interactions must therefore be taken into account [@cote2016], yet most
-research on driver effects in marine environments remains overwhelmingly focused
-on single driver assessments [@obrien2019]. This is partly explained by the
-complexity of studying multiple drivers simultaneously [@cote2016], but also
-by the need for standardized data management practices and tools tailored to
-the study of the effects of multiple drivers [@dafforn2016; @stock2018].
+Environmental drivers are however conspicuously missing from most of these
+initiatives [but see @halpern2015], even though high resolution data on their
+distribution and intensity is essential to environmental management and impact
+assessments. For this paper, we define environmental drivers (hereafter referred
+to as drivers) as any process that may disturb a natural system. Drivers may be
+natural (e.g. sea surface temperature anomalies and hypoxia) or anthropogenic
+(e.g. fisheries and marine pollution) in origin.
 
-To address these issues, we are launching an initiative called *eDrivers*, an
+To address this issue, we are launching an initiative called *eDrivers*, an
 open data platform focused on the description of the spatial distribution and
-intensity of drivers and using the Estuary and Gulf of St. Lawrence in Canada
-as a case study. The overarching goal of this platform is to create an expert
-community committed to structuring, sharing and standardizing drivers data in
-support to science and management.
-
-The objective of this paper is twofold. The first is to present the guiding
-principles underlying the inception and structure of *eDrivers*. We also wish
-this paper to serve as a call for collaboration to any person, group or
-organization who might be interested in contributing to this initiative. The
-second objective focuses using the platform content to describe the intensity and distribution of drivers
-in the St. Lawrence system and aims at 1) identifying areas of high cumulative exposure,
-2) at characterizing areas with similar cumulative exposure regimes, dubbed threat
-complexes [@bowler2019] and 3) at identifying drivers that are likely to interact
-in the St. Lawrence.
+intensity of drivers in the Estuary and Gulf of St. Lawrence in Canada, a
+unique and complex social-ecological system. The overarching goal of this
+platform is to create an expert community committed to structuring, sharing
+and standardizing drivers data in support to science and management. The
+objective of this paper is to present the guiding principles underlying the
+inception *eDrivers*, the structure of the platform and potential applications.
+Most importantly, we wish to initiate a call for collaboration to any person,
+group or organization who might be interested in contributing to this
+initiative.
 
@@
 
-The St. Lawrence system, formed by one of the largest estuaries in the world and a
-vast interior sea, is an complex social-ecological system characterized by highly
+The St. Lawrence, formed by one of the largest estuaries in North America and a
+vast interior sea, is an ecologically complex system characterized by highly
 variable environmental conditions and oceanographic processes, both in space
@@
 access to more than 40 ports, has a booming tourism industry and an expanding
-aquaculture production, develops emerging activities and has a yet untapped
-hydrocarbon potential [@beauchesne2016; @schloss2017]. With major
-investments recently made and more forthcoming in economic and infrastructure
-development and research [*e.g.* @goq2015; @rqm2018], an intensification of the
-human footprint is expected in the St. Lawrence system.
+aquaculture production, develops emerging activities and a yet untapped
+hydrocarbon potential [@beauchesne2016]. With major investments recently made
+and more forthcoming in economic and infrastructure development and research
+[*e.g.* @goq2015; @rqm2018], an intensification of the human footprint is
+expected in the St. Lawrence.
 
@@
 whole. Yet, jurisdiction over the St. Lawrence is divided between the Canadian
-federal government and five provincial governments (Newfoundland and
-Labrador, New Brunswick, Nova Scotia, Prince Edward Island, and Québec) with
-each governmental entity further divided into multiple departments.
-Researchers from dozens of academic institutions and
+federal government and five provincial governments, i.e. Newfoundland and
+Labrador, New Brunswick, Nova Scotia, Prince Edward Island, and Québec, with
+each governmental entity further divided into multiple departments, each with
+different mandates. Researchers from dozens of academic institutions and
 countless local organizations also share an interest in the St. Lawrence. This
-wide range of actors results in highly scattered data management
+wide range of actors and interests results in highly scattered data management
 and sharing processes, which we wish to address through this open data platform
@@
 
-Five guiding principles motivate the development of the platform and shape its
-structure (Figure \ref{diag}).
+Four guiding principles motivate the development of the platform and shaped how
+we structured the platform (Figure \ref{diag}).
 
@@
 requires a vast and diverse expertise that cannot possibly be possessed by any
-one individual or organization. Consequently, we envision an initiative that
-seeks to mobilize all individuals and entities with relevant expertise.
+one individual or organization. Consequently, we envision a crowd science-like
+initiative [see @franzoni2014] that seeks to mobilize all entities
+with relevant expertise.
 
-***How***: By promoting, consolidating and working with experts involved in
-existing and highly valuable environmental initiatives already in place in the
-St. Lawrence. Notable examples of environmental initiatives are the annual
-review of physical [@galbraith2018], chemical, and biological [@blais2018]
-oceanographic conditions in the St. Lawrence, the fisheries monitoring program
-[@dfo2016], the annual groundfish and shrimp multidisciplinary survey
-[@bourdages2018], the characterisation of benthic [@dutil2011], epipelagic and
-coastal [@dutil2012] habitats of the St. Lawrence, and Canada's shoreline
-classification [@eccc2018]. There are also nascent efforts to make available
-information on several human activities in the St. Lawrence such as the Marine
-Spatial Data Infrastructure portal, which provides data on zoning, shipping,
-port activities, and other human activities in Canadian waters, including the
-St. Lawrence system [@goc2016].
-
-By working with existing data portals whose objective is to share
-environmental data. We are thus collaborating actively with the St. Lawrence
-Global Observatory (SLGO) to develop the initiative and to host the
-platform on their web portal (https://ogsl.ca/en). The mission of SLGO is to
-promote and facilitate the accessibility, dissemination and exchange of official
-and quality data and information on the St. Lawrence ecosystem through the
-networking of organisations and data holders to meet their needs and those of
-users, to improve knowledge and to assist decision-making in areas such as
-public safety, climate change, transportation, resources and biodiversity
-conservation. The SLGO is also one of three regional association spearheading
-the Canadian Integrated Ocean Observing System
-(CIOOS; http://meopar.ca/research/cioos-call-for-proposals/),
-which will focus on integrating oceanographic data from multiple sources to
-make them accessible to end-users and to enable the national coordination of
-ocean observing efforts by integrating isolated or inaccessible data, and by
-identifying gaps or duplications in observations and research efforts.
-We are also developing collaborations with the Portal on water knowledge
-(http://www.environnement.gouv.qc.ca/eau/portail/), an initiative from the
-Québec provincial government. This portal aims at collecting and sharing
-accurate, complete and updated ressources on water and aquatic ecosystems to
-support the mandate of relevant actors and stakeholders working in
-water and aquatic ecosystems management.
-
+***How***:
 By actively inviting, seeking, and developing collaborations as well as
@@
 By inviting external community contributions (Figure \ref{diag}).
-External researchers or entities wishing to submit marine data will be able to
-do so through the SLGO web portal (https://ogsl.ca/en). Submissions through
-other data portals will also be accepted either through the development of data
-sharing agreements or with the caveat that shared data are under an open source
-license and that they adhere to the platform data standards.
+External researchers or entities wishing to submit data will be invited to do
+so mainly through the SLGO web portal. However, submissions through other data
+portals will be accepted either through the development of data sharing
+agreements or with the caveat that shared data are under an open source license
+and that they adhere to the platform data standards.
+
+By promoting and consolidating existing and highly
+valuable environmental initiatives already in place in the St. Lawrence. Notable
+examples are the annual review of physical [@galbraith2018], chemical, and
+biological [@blais2018] oceanographic conditions in the St. Lawrence, the
+fisheries monitoring program [@dfo2016], the annual groundfish and shrimp
+multidisciplinary survey [@bourdages2017], the characterisation of benthic
+[@dutil2011], epipelagic and coastal [@dutil2012] habitats of the St. Lawrence,
+and Canada's shoreline classification [@eccc2018]. There are also nascent
+efforts to make available information on several human activities in the
+St. Lawrence such as the Marine Spatial Data Infrastructure portal, which
+provides data on zoning, shipping, port activities, and other human activities
+in Canadian waters, including the St. Lawrence [@goc2018].
+<!-- include MDDELCC? -->
 
 ### Findability, accessibility, interoperability and reusability
+<!-- Original: Openness, transparency and replicability -->
 
@@
 research initiatives [@reichman2011; @poisot2013; @lowndes2017]. As such, many
-scientific journals (*e.g.* Frontiers, PLoS, Nature, Science), governmental agencies <!-- mettre nom de la revue ou on soumet -->
+scientific journals (*e.g.* PLoS, Nature, Science), governmental agencies
 [*e.g.* @goc2016] and other institutions are adopting open
@@
 
+By hosting on and developing the platform collaboratively with the St. Lawrence
+Global Observatory [SLGO; https://ogsl.ca/en]. The mission of SLGO is to promote
+and facilitate the accessibility, dissemination and exchange of official and
+quality data and information on the St. Lawrence ecosystem through the networking
+of organisations and data holders to meet their needs and those of users, to
+improve knowledge and to assist decision-making in areas such as public safety,
+climate change, transportation, resources and biodiversity conservation.
+The SLGO is also one of three regional association spearheading the Canadian
+Integreated Ocean Observing System
+(CIOOS; http://meopar.ca/research/cioos-call-for-proposals/),
+which will focus on integrating oceanographic data from multiple sources to
+make them accessible to end-usersand to enable the national coordination of
+ocean observing efforts by integrating isolated or inaccessible data, and by
+identifying gaps or duplications in observations and research efforts.
+
 By making data and associated tools accessible through a variety of ways, namely
-the SLGO web portal (https://ogsl.ca/en), two R packages called *eDrivers*
-(https://github.com/orgs/eDrivers/eDrivers) and *eDriversEx*
-(https://github.com/orgs/eDrivers/eDriversEx) to access the data through SLGO's
-API and to provide analytical tools to explore data, respectively, and a Shiny
-application (https://david-beauchesne.shinyapps.io/eDriversApp/) that allows
-users to explore drivers data interactively (Figure \ref{diag}).
+the SLGO web portal, two R packages called *eDrivers* (https://github.com/orgs/eDrivers/eDrivers)
+and *eDriversEx* (https://github.com/orgs/eDrivers/eDriversEx) to access the data trough SLGO's API and to provide
+analytical tools to explore data, respectively, and a Shiny application
+(https://david-beauchesne.shinyapps.io/platformapp/) that allows users to
+explore drivers data interactively (Figure \ref{diag}).
 Note that the data are currently contained within and accessible through the
-*eDrivers* R package only, as we are actively working to allow users to
-download selected layers from SLGO's web portal and geoserver. The functions
-available in *eDrivers* to access the data have however been developed to
-ensure forward compatibility once the data migrate to SLGO's geoserver.
+*eDrivers* R package only, as we are actively structuring the project on SLGO's
+web portal and geoserver. The functions available in *eDrivers* to access the
+data have however been developed to ensure forward compatiblity once the data
+migrate to SLGO's geoserver.
 
@@
 
-By committing to provide an increasingly FAIR infrastructure as we continue
-developing it.
+By commiting to provide an increasingly FAIR infrastructure (*i.e.* a FAIRport)
+as we continue developing it.
 
@@
 been discussed in the context of multi-drivers and cumulative impact assessments
-[@halpern2015; @beauchesne2016; @cote2016; @schloss2017]. Adaptive management can only be
+[@halpern2015; @beauchesne2016; @cote2016]. Adaptive management can only be
 truly achieved through a commitment to adaptive monitoring and data reporting
@@
 management requires the development of adaptive monitoring tools and
-infrastructures, which we seek to address through a continuously-evolving platform.
+infrastructures, which we seek to address for the St. Lawrence through a
+continuously-evolving platform.
 
@@
 
-### Validation
+### Peer-review and data credit
 
-***Why***: Using high-quality data is crucial to ensure robust analyses and
-decision-making [@cai2015; @veiga2017]. Data quality assessment (*i.e.* quality
-measurement, validation and classification) and management (*i.e.* quality
-control and assurance) are therefore crucial steps in data sharing [@veiga2017].
-
-***How***: By selecting a framework to assess and manage data quality and by
-providing a data quality report with for all driver layers shared on the
-platform.
-
-### Recognition
-
-***Why***: Like peer-reviewed publications, data must also be given its due
+***Why***: Peer-review is a fundamental tenet of science that ensures robustness
+and quality of published research and lends credibility to communicated
+scientific outputs [@ware2008; @rennie2016]. In an effort to
+uphold the highest standards of scientific research, we wish to provide data
+layers that have been reviewed, ideally through peer-reviewed
+journals. Like peer-reviewed publications, data must also be given its due
 importance in scientific endeavors and thus be considered as legitimate citable
@@
 
-***How***: By including the persistent identifier, when possible, and full
-citation of all data layers used and shared on the platform.
+***How***:
+By clearly identifying peer-reviewed material and associated citation.
+
+By including the persistent identifier, when possible, and full citation of of
+all data layers used and shared on the platform.
 
@@
 on the platform will be identified by a persistent identifier (*e.g.* DOI),
-possess a license outlining conditions of use and a clear *how to cite* label
+posseess a license outlining conditions of use and a clear *how to cite* label
 including the authors, the data version and year, an informative title, the
@@
 
-# St. Lawrence drivers
+## Content
 
-## Data
-
-The list of drivers for which we sought data was informed by global cumulative
-impacts assessments initiatives [@halpern2008; @halpern2015], regional holistic
+We first identified a list of 34 drivers (Table 1) known to potentially impact
+the St. Lawrence system. This list was informed by global cumulative impacts
+assessments initiatives [@halpern2008; @halpern2015], regional holistic
 evaluation of the state of the St. Lawrence [@dufour2007; @benoit2012], and
@@
 anthropogenic (*e.g.* fisheries) to natural (*e.g.* temperature anomalies).
-Through the data gathering process, we developed and continue developing
-collaborations with regional experts and data holders.
-We also use global data from seminal works on cumulative impacts on global
-habitats [@halpern2008; @halpern2015] and available on the National Center for
-Ecological Analysis and Synthesis (NCEAS) online data repository
-[Table 1; @halpern2015b]. We selected global data that
-were unavailable at the regional scale and that were available at a resolution
-adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).
 
-The platform currently provides access to 22 driver layers (Table 1) and we are
-actively working on updating or developing additional driver layers.
-We divided drivers into 4 groups: coastal, climate, fisheries and marine traffic
-related drivers (Table 1). All data layers and methodologies used to produce
-driver layers are described in the supplementary materials of this paper and on
-the *eDrivers* GitHub organization (https://github.com/orgs/eDrivers/).
-Also note that the specific
-methodologies underlying the figures presented in the paper are described in
-the figure captions to ease the flow of text.
+We have developed and continue developing collaborations with regional experts
+to gather data and characterize the distribution and intensity of these drivers
+in the St. Lawrence system. We have thus far characterized 12 drivers and are
+actively working on the characterization of 12 more, with their development at
+various stages of completion (Table 1). We are also actively seeking
+collaborations or data for 10 drivers (Table 1). We also use global data from
+the National Center for Ecological Analysis and Synthesis (NCEAS) online data
+repository [Table 1; @halpern2015b], as certain layers provide valuable
+information at a resolution adequate for use at regional scales or currently
+constitute the best available knowledge for the St. Lawrence (*e.g.* marine
+pollution).
+<!-- Review with the final list of data that I will indeed be able to share -->
 
-## Cumulative exposure
+# Applications
 
-Apart from the northeastern Gulf, cumulative exposure (*i.e.* sum of normalized
-driver intensity) is ubiquitous in the St. Lawrence (Figure \ref{footprint}).
-Cumulative exposure is generally highest along the coast (Figure \ref{footprint}),
-with most intense hotposts (*i.e.* areas of overlapping drivers at high relative
-intensity) located in the vicinity of coastal cities (Figure \ref{hotspot}).
-In general, offshore areas are less exposed to cumulative drivers, with the
-Estuary and the Anticosti Gyre being notable exceptions (Figures \ref{footprint}
-and \ref{hotspot}). That is however not to say that offshore areas are free of
-exposure, as most of the St. Lawrence is exposed to multiple drivers
-(Figures \ref{footprint} and \ref{hotspot}).
+Assembling all drivers in a single platform will provide a wide array of
+visualization and analytical possibilities that can be useful for both decision
+making and scientific research. Here we describe two such potential
+applications, *i.e.* multi-drivers assessment and multi-scale management.
+Furthermore, we demonstrate the ease with which those applications will be
+possible through the platform by providing the code to replicate the elements
+presented on \ref{fig1} for multi-drivers assessment in box 1.
 
-These results are consistent with observations elsewhere in the world, where
-cumulative exposure notably arises from and markedly intensifies close to
-coastal cities and at the mouth of rivers draining highly populated areas
-[e.g. @halpern2015; @feist2016; @mach2017; @stock2018]. These are areas
-where human uses (*e.g.* coastal development and shipping) and footprint
-(pollution runoff) are the most intense [@feist2016], and on which is
-overlaid a background of natural disturbances [@micheli2016]. They are also the
-areas in which the most dramatic increases in exposure are expected, with
-populations increasing more rapidly along the coast than inland [@feist2016].
+## Multi-drivers assessment
 
-In the St. Lawrence, large coastal cities are mostly located along the Estuary
-and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
-or home to small coastal towns. Moreover, the Estuary, along with the
-St. Lawrence river, provide access to and serve as the primary drainage
-outflow of the Great Lakes Basin, which is the most densely populated region
-in Canada [@statcan2017].
-
-## Driver interactions
-
-By providing data on driver intensity and distribution, the platform will
-greatly facilitate the study of the effects of multiple drivers, their interaction
-likelihood and the areas where they may interact.
-For instance, drivers like hypoxia (Figure \ref{kernel}A) and demersal destructive
-fisheries (Figure \ref{kernel}B) are known to occur mainly in deeper areas of the
-St. Lawrence, and hence an interaction between the effects of the two drivers
-could be anticipated. Fisheries in the St. Lawrence have historically affected
-biodiversity distribution and habitat quality [@moritz2015]. Hypoxia, meanwhile,
-decreases overall habitat quality but has variable effects that are
-species-dependent, ranging from well-adapted [*e.g.* northern shrimp
-*Pandalus borealis* and Greenland halibut *Reinhardtius hippoglossoides*;
-@pillet2016] to reduced growth rates [@dupont2013] and avoidance of oxygen
-depleted habitats [*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased
-mortality [*e.g.* sessile benthic invertebrates; @eby2005; @gilbert2007; @belley2010].
-Certain species may thus be negatively affected by fisheries and withstand hypoxia
-but still experience a decrease in prey availability, while others may be
-negatively affected by the compounded effect of both drivers [@deleo2017].
-
-By combining both drivers (Figure \ref{kernel}C), we can evaluate
-which regions in the St. Lawrence are more likely to be affected by their
-simultaneous presence and evaluate the relative intensity at which they tend to
-co-occur (Figure \ref{kernel}D). Doing this, we observe that hypoxia and bottom
+This platform initiative will enable usual, single driver impact assessments,
+but it will also pave the way towards holistic assessments of the state of the
+St. Lawrence by enabling the study of combinations of multiple drivers. For
+instance, drivers like hypoxia (Figure \ref{fig1}A) and demersal destructive
+fisheries (Figure \ref{fig1}B) are known to occur mainly in deeper areas of the
+St. Lawrence, and hence a synergy between the two drivers could be anticipated.
+By combining them (Figure \ref{fig1}C), we can evaluate which
+regions in the St. Lawrence are more likely to be affected by both drivers
+simultaneously and evaluate the relative intensity at which they tend to
+co-occur (Figure \ref{fig1}D). Doing this, we observe that hypoxia and bottom
 fisheries tend to co-occur in the St. Lawrence and that they do so at relatively
@@
 low levels of oxygen neighboring the limits of hypoxia (*i.e.* ~0.7 mg*L^-1^,
-corresponding to ~30% saturation). We demonstrate the ease with which this
-can be accomplished using *eDrivers* by providing the code necessary to reproduce
-Figure \ref{kernel} in box 1.
+corresponding to ~30% saturation).
 
-## Threat complexes
+Fisheries in the St. Lawrence have historically affected biodiversity distribution
+and habitat quality [@moritz2015]. Hypoxia, meanwhile, decreases overall
+habitat quality but has variable effects that are species-dependent, ranging
+from well-adapted (*e.g.* northern shrimp *Pandalus borealis* and Greenland
+halibut *Reinhardtius hippoglossoides*; [@pillet2016] to reduced growth rates
+[@dupont2013] and avoidance of oxygen depleted habitats
+[*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality
+[*e.g.* sessile benthic invertebrates; @eby2005; @belley2010]. Certain species
+may thus be negatively affected by fisheries and withstand hypoxia but still
+experience a decrease in prey availability, while others may be negatively
+affected by the compounded effect of both drivers. By providing data on multiple
+drivers and their spatial interaction, the platform will yield valuable
+insights for the integrated management of the St. Lawrence and the ability to
+tackle complex problems.
 
-The number of driver overlapping in the St. Lawrence increases with cumulative
-exposure (Figure SX). Areas with high exposure such as the Estuary, the
-Anticosti Gyre and the southwestern Gulf coastline (Figure \ref{footprint} and
-\ref{hotspot}) are thus areas in the St. Lawrence where driver interactions are
-most likely, and where they can arise between a host of different drivers.
-While we cannot ascertain that high exposure areas are also the most impacted,
-we can safely predict that these are the areas where studying ecosystem state
-will be the most complex due to the uncertainty associated with driver
-interactions, an uncertainty certain to increase rapidly with the increasing
-number of interacting drivers [@cote2016]. It is however important to note
-that there are very few areas of the St. Lawrence that do not experience driver
-overlap (Figure \ref{hotspot}) and that drivers do not necessarily need to be
-co-occurring at high intensities to cause significant adverse effects
-[*e.g.* @relyea2009; @liess2016].
+## Multi-scale management
 
-In an effort to describe areas where interactions are most likely between
-multiple and typical combinations of drivers, we identified areas with similar cumulative
-exposure regimes, which we refer to as threat complexes after @bowler2019. Seven
-distinct threat complexes were identified in the St. Lawrence, which can be
-divided into 4 offshore and 3 coastal complexes (Figure \ref{cluster}). Refer
-to the supplementary information for the figures related to the following
-description of threat complexes.
+Management is typically distributed across multiple jurisdictional, spatial and
+temporal scales. For example, in the St. Lawrence, commercial fisheries are
+managed at the federal level, with three regional jurisdictions. Ports and shipping,
+meanwhile, are managed by a combination of local, provincial and federal
+institutions. Such divisions, while necessary, isolate decision-making into
+contextualized boxes even though decisions may have consequences reaching far
+beyond its immediate context. With the appropriate time-series data provided by
+cyclic updates of platform content and a commitment to adaptive monitoring, the
+platform will allow us to visualize and evaluate management outcomes and
+efficiency across scales. We draw on two topical issues in
+the St. Lawrence and use simulated data to illustrate this.
 
-Coastal threat complexes (2, 4 and 6; Figure \ref{cluster}) are differentiated
-from offshore complexes by the inclusion of
-all coastal drivers. They however also include all other types of drivers except
-hypoxia. They are also the most exposed threat complexes, both in terms of
-driver overlap and intensity. Among coastal complexes, complex 2 is
-differentiated by the presence of aquaculture sites. Threat complex 4, which is
-widespread through the coastal Estuary and Gulf, includes areas associated with
-port activities at higher relative intensity. Finally, complex 6, in the
-southeastern and northernmost Gulf, is characterized by negative bottom
-temperature anomalies.
+The first issue is the case of the North Atlantic Right Whale
+(NARW; *Eubalaena glacialis*), an endangered species that received a lot of media,
+political and scientific attention in the summer of 2017 when 12 dead NARW and
+5 live-entanglements were documented in the St. Lawrence [@daoust2017]. It was
+established that the most likely causes of death were vessel collisions and
+fishing activities [@daoust2017], and a series of management actions were
+implemented to limit impacts to NARW [@dfo2018; @tc2018]. One such action is a
+static fixed engine fisheries closure in the Gulf of St. Lawrence in an area
+encompassing 90% of 2017 NARW observations. With the platform, we can visualize
+the fisheries intensity before (Figure \ref{fig2}B) and eventually after the
+management action (Figure \ref{fig2}D) to evaluate the effects of management
+actions on fixed engine fisheries.
 
-Offshore threat complexes (1, 3, 5 and 7; Figure \ref{cluster}) are generally
-characterized by climate and marine traffic related drivers. Threat complex 1
-mostly incorporates the shallow areas of the northern Gulf and is characterized
-by lower marine traffic intensity and with higher intensity surface temperature
-anomalies. Threat complex 3 encompasses most of the southern Gulf and is
-characterized by medium to high marine traffic intensity and higher aragonite
-stress. Threat complex 5, meanwhile, is mostly delineated by the deep channels
-of the northern Gulf and is characterized by positive bottom temperature
-anomalies. Finally, complex 7 corresponds to the Laurentian channel, which is a
-deep channel connecting the Estuary to the Atlantic. This threat complex is
-characterized by high intensity hypoxia, marine traffic and pollution, and
-demersal destructive fisheries.
+The second issue is the case of port development initiatives in the
+St. Lawrence. Five local port projects scattered through the St. Lawrence have
+been submitted for an environmental assessment review at the Canadian
+Environmental Assessment Agency [@ssl2018]. Of those projects, we focus on the
+marine terminal project on the north shore of the Saguenay River, for which a
+draft environment assessment report was published [@ceaa2018] as well as a
+science response by the Canadian Science Advisory Secretariat [@dfo2018b]. It is
+reported that the establishment of this port alone would lead to a 2.33-fold
+increase in the number of ships in the Saguenay River by 2030, with vessels
+transiting through the St. Lawrence to reach international markets
+[@ceaa2018; @dfo2018b]. With the platform, we can visualize shipping intensity
+before (Figure \ref{fig2}C) and eventually after (Figure \ref{fig2}E) the
+establishment of the marine terminal to evaluate the effects of this project
+on local marine traffic, but along the whole river and the estuary.
 
-Fisheries are generally distributed among all offshore threat complexes and
-coastal complex 4. This widespread distribution among clusters causes fisheries
-to have limited contribution to differences between complexes. There are however
-certain threat complexes characterized by higher proportion of fishing
-activities. Threat complex 3 holds higher intensities of pelagic low-bycatch and
-demersal high-bycatch fisheries. Threat complex 5, meanwhile, has higher intensities of
-demersal destructive fisheries. Finally, threat complex 7 is characterized by higher intensities
-of demersal destructure and pelagic high-bycatch fisheries.
-
-Describing these threat complexes allows us to see that each of those areas
-are characterized by distinct overlapping suites of drivers. This holds
-significant importance for the species, habitats and ecosystems located
-within or moving through those threat complexes, as they are likely to be
-affected by the joint effects of all drivers typical to those
-complexes [@bowler2019].
+Critically, the platform will allow users to explore the compounded effects of
+management outcomes by providing the ability to explore effects on the
+integrative distribution and intensity of drivers by considering them all
+simultaneously (Figure \ref{fig2}A, F), or using any combination of drivers desired.
+Similarly, the effects of multiple decision-making processes could be evaluated
+in combination to see if their outcomes overlapped. This could provide an invaluable
+tool for management and to justify and encourage a strong commitment to adaptive
+monitoring.
 
@@
 
-Understanding how ecosystem state will be affected by global change necessitates
-a comprehensive understanding of how threats are distributed and interacting in
-space and time, which hinges on appropriate data tailored to multi-drivers
-studies [@dafforn2016; @stock2018; @bowler2019]. In the St. Lawrence,
-we found that few areas are free of cumulative exposure and that the whole
-Estuary, the Anticosti Gyre and coastal southwestern Gulf are particularly
-exposed to cumulative drivers, especially close to urban areas. We also
-identified seven geographically distinct threat complexes that show similar
-cumulative exposure regimes. These complexes also show that coastal areas are
-particularly exposed to all types of drivers and that multiple drivers are
-typically co-occurring in space. The *eDrivers* initiative allows us to
-efficiently identify these areas in need of heightened scrutiny. It also ensures
-an iterative improvement of cumulative exposure assessments and threat complexes
-identification as gaps in knowledge are addressed and approaches to describe
-drivers are refined.
+In this perspective paper we present an open data platform focused on
+describing the distribution and intensity of environmental drivers in the
+Estuary and Gulf of St. Lawrence, Canada. The core objectives underlying this
+initiative are to create a community-led project based on transparency and
+inclusivity that will increase accessibility to and interoperability of critical
+data for scientific inquiries and decision-making.
 
-Arguably the most meaningful benefit anticipated from *eDrivers* is the
-significant gain in data accessibility through the creation of a
-cross-jurisdictional collaborative tool. This will effectively curtail the need
-to reach dozens of experts across multiple organizations and over extensive
-periods of time to assemble the data needed for systematic research. These
-benefits could pay quick scientific and management dividends by drawing on the
-knowledge and efforts of a wide range of contributors, by expanding avenues of
-scientific inquiry, by decreasing overall effort duplication and research costs,
-and by increasing research efficiency [@franzoni2014]. Critically - and we
+Arguably the most meaningful benefits stemming from this initiative are the
+creation of a cross-jurisdictional collaborative tool and the gains in data
+accessibility and interoperability, effectively curtailing the need to reach
+dozens of experts across multiple departments and over extensive periods of
+time to assemble the data needed for systematic research. These benefits could
+pay quick scientific and management dividends by drawing on the knowledge and
+efforts of a wide range of contributors, by expanding avenues of scientific
+inquiry, by decreasing overall research costs, by increasing research speed
+and by decreasing effort duplication [@franzoni2014]. Critically - and we
 emphasize this point - the platform will allow the scientific and governmental
 communities to identify key knowledge gaps that will assist in prioritizing
-and optimizing research efforts. For example, the platform could assist in
-designing targeted experiments to study the effects of multiple drivers by
-identifying sampling sites capturing appropriate exposure gradients
-[@stock2018; @bowler2019].
+and optimizing research efforts.
 
-Another key issue with data sharing that will be addressed by the platform is
-data interoperability. By using and enforcing (meta)data standards the platform
-will strongly encourage the standardization of future data collection protocols
-to maintain compatibility. Standardized protocols will ease the regional
-aggregation of local datasets that are currently highly fragmented in
-availability, quality and experimental approaches [@reichman2011]. It will also
-enable interoperability with other open data platforms (*e.g.* OBIS). This
-standardization could also be accompanied by an initiative to extend
+Another key issue with data sharing that will be addressed by the platform is,
+as discussed previously, interoperability. By using and enforcing (meta)data
+standards the platform will strongly encourage the standardization of future
+data collection protocols to maintain compatibility. Standardized protocols
+will ease the regional aggregation of local datasets that are currently highly
+fragmented in availability, quality and experimental approaches [@reichman2011].
+It will also enable interoperability with other open data platform (*e.g.
+OBIS). This standardization could also be accompanied by an initiative to extend
 the Essential Ocean Variables (http://www.goosocean.org) to include a list of
@@
 
-We also believe that the *eDrivers* initiative will operationalize
-evidence-based decision-making by streamlining data management and research,
-allowing science output to be available and interpretable on a time scale
-relevant to management. The platform will thus greatly facilitate the
-application of broad scale, holistic research and management approaches such as
-ecosystem-based management, strategic environmental assessments and
-social-ecological metanetworks [*e.g.* @halpern2015; @jones2016; @dee2017].
+We believe that this project will operationalize evidence-based decision-making
+by streamlining data management and research, allowing science output to be
+available and interpretable on a time scale relevant to management.
+The platform will thus greatly facilitate the application of broad scale,
+holistic research and management approaches such as ecosystem-based management,
+strategic environmental assessments and social-ecological metanetworks
+[*e.g.* @halpern2015; @jones2016; @dee2017].
 
@@
 that list is to maintain our efforts to foster collaborations, develop platform
-content and identify key knowledge gaps. A fair and efficient organizational
-structure will also be developed in order to manage the platform as a community.
+content and identify key knowledge gaps. Terrestrial and coastal environments
+must also be incorporated, as the sources of stress within those habitats extend
+to marine environments. A fair and efficient organizational structure will also
+be developed in order to manage the platform as a community.
 Appropriate funding must also be secured to continue building this community and
 ensure the long-term viability of the initiative, although the partnership with
-SLGO partly addresses this issue.
+SLGO partly addresses this issue. Nonetheless, we are well on our way and hopeful
+that this initiative will be a very successful one.
 
-Finally, terrestrial and coastal environments must also be incorporated, as
-sources of stress within those habitats extend to the marine environments.
-Moreover, despite coastal areas being recognized as the most exposed to
-environmental threats we continue delineating
-terrestrial and marine realms, using coastlines as we would an impermeable
-barrier. While there is a sensible rationale for this division, we must strive
-to eliminate this barrier if we are to appropriately study and predict the
-impacts of global change [*e.g.* see @bowler2019].
-
-Despite the work still ahead, we are hopeful that this initiative will be a very
-successful one. Ultimately, this platform represents a much needed solution to
-address important issues in data management that could radically shift broad
-scale research and management practices towards efficient, adaptive and holistic
-ecosystem-based management in the St. Lawrence and elsewhere in the world. All
-it requires to be successful is for the scientific and political communities to
-fully commit to open data, adaptive monitoring and, most of all, an integrated
-vision of ecosystem management.
-
-# Acknowledgements
-
-We thank the Fond de Recherche Québécois Nature et Technologie (FRQNT) and the
-Natural Science and Engineering Council of Canada (CRSNG) for financial support.
-This project is also supported by Québec Océan, the Quebec Centre for
-Biodiversity Science (QCBS), Takuvik, and the Notre Golfe networks. This
-research is also sponsored by the NSERC Canadian Healthy Oceans Network and its
-Partners: Department of Fisheries and Oceans Canada and INREST (representing the
-Port of Sept-Îles and City of Sept-Îles).
-
-# Author contributions statement
-
-TO WRITE
-
-# Conflict of interest statement
-
-The authors declare that the submitted work was carried out in the absence of any personal, professional or financial relationships that could potentially be construed as a conflict of interest.
+Ultimately, this platform represents a much needed solution to address important
+issues in data management that could radically shift broad scale research and
+management practices towards efficient, adaptive and holistic ecosystem-based
+management in the St. Lawrence and elsewhere in the world. All it requires to
+be successful is for the scientific and political communities to fully commit
+to open data, adaptive monitoring and, most of all, an integrated vision of
+ecosystem management.
 
@@
 
-Box 1. Code snippet demonstrating how to use the *eDrivers* to reproduce
-figure \ref{kernel} in R.
+## Box 1
 
 \begin{lstlisting}[frame=single]
-# Install and load eDrivers package
+# Code to access data and generate figure 2
+# Install eDrivers package
 devtools::install_github('eDrivers/eDrivers')
-library(eDrivers)
 
 # Load data
+library(eDrivers)
 drivers <- fetchDrivers(drivers = c('hypoxia','fishDD'))
@@
 All data is then hosted on SLGO's geoserver and accessible through their API.
-We developed a R package called *eDrivers* to access the driver layers through R and we are actively developing a second R package called *eDriversEx* that includes analytical tools to explore drivers data.
-Finally, we have developed a Shiny app that allows users to explore drivers data interactively (https://david-beauchesne.shinyapps.io/eDriversApp/).
+We developped a R package called *eDrivers* to access the driver layers through R and we are actively developping a second R package called *eDriversEx* that includes analytical tools to explore drivers data.
+Finally, we have developped a Shiny app that allows users to explore drivers data interactively (https://david-beauchesne.shinyapps.io/platformapp/).
 All R components of the project are also hosted on the *eDrivers* GitHub organization.
@@
 
-\newpage
-
-![Distribution of driver footprint in the Estuary and Gulf of St.
-Lawrence.
-The footprint is measured by summing the relative intensity of all drivers for each grid cell: $F_d = \sum_{i=1}^{n} D_i$.
-Driver layers were also log-transformed when their frequency distribution was non-normal.
-All driver layers were also normalized between 0 and 1 using the 99th quantile to allow for direct comparison of relative intensities between drivers and control for extreme values.
-\label{footprint}](./figures/footprint.png)
-
-\newpage
-
-![Distribution of hotspots of driver footprint in the Estuary and Gulf od St.
-Lawrence.
-Hotspots were identified using the sum of all driver's 80th intensity quantile.
-This results in the distribution of areas with overlapping drivers at high intensity.
-\label{hotspot}](./figures/hotspot.png)
-
-\newpage
-
 ![Example of platform content using the spatial distribution and the intensity of hypoxia and demersal destructive fisheries in the St.
 Lawrence.
-An index of hypoxia (**A**) was created using bottom oxygen concentrations from a newly instated survey during the months of August and September [@starr2019].
-We interpolated the survey data using cokriging with depth as a covariable [@dutil2011].
-A hypoxic index was then measured using a hypoxic threshold of 2 $mlL^{-1}$ bottom oxygen concentration [@diaz1995].
-An inverted logistic curve was used to create an index of hypoxic stress ranging from close to hypoxic (*i.e.* 0) to anoxic (*i.e.* 1), under the assumption that hypoxic stress increases following a logistic curve as it approaches the hypoxic threshold.
+An index of hypoxia (**A**) was measured by normalizing values of bottom oxygen saturation percentage [@dutil2011] between 0 and 1 so that values close to 0 are fully saturated in $O_2$, while values close to 1 are hypoxic.
+Areas with less than 30% $O_2$ saturation (index $>$ 0.7) are generally considered hypoxic.
 Demersal destructive fisheries (**B**) correspond to fishing activities using trawl and dredges.
@@
 Values were then normalized to allow comparison with hypoxia values.
-Drivers intensity was summed (**C**) to visualize the combination of hypoxia and demersal destructive fisheries.
-Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a kernel analysis (**D**).\label{kernel}](./figures/kernel.png)
+drivers intensity was summed (**C**) to visualize the combination of hypoxia and demersal destructive fisheries.
+Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a kernel analysis (**D**).\label{fig1}](./figures/figure1.png)
 
@@
 
-![Distribution of threat complexes in the Estuary and Gulf of St.
-Lawrence.
-Threat complexes (a term coined by [@bowler2019]) are areas with similar cumulative driver exposure regimes.
-Threat complexes were identified using a partional *k-medoids* clustering algorithm, CLARA [CLustering for Large Applications; @kaufman1990].
-The appropriate number of clusters $k$ was tested using a range of values and validated by selecting the number of clusters that maximized the average silhouette width [@kaufman1990] and minimized the total within-cluster sum of squares (WSS).
-We also validated the clustering by comparing *k-medoids* clustering with *k-means* clustering with the *Lloyd* algorithm [@lloyd1982] and found a 66% cell correspondence between both approaches, with most of the differences attributable to a single undivided cluster in the *k-means* approach.
-Seven distinct threat complexes were identified in the St.
-Lawrence.
-The difference between clusters was explored by measuring the total inter-cluster dissimilarity and the contribution of each driver to the total inter-cluster dissimilarity through a similarity percentage analysis (SIMPER) with Bray-Curtis dissimilarity [@clarke1993].
-As the drivers data is too large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting 5% of each cluster to run the analysis and repeating the process over 300 iterations.
-A similar bootstrapping procedure was used to evaluate the intra-cluster similarity with the Bray-Curtis similarity using 300 iterations using 5% of each cluster data when the total number of observations in cluster exceeded 10000 observations.
-The *k-medoid*, *k-means* and SIMPER analyses were performed using the *cluster*, *stats* and *vegan* R packages, respectively [@maechler2018; @oksanen2018; @rcoreteam2018].
-Refer to the Supplementary Materials for more details.
-\label{cluster}](./figures/cluster.png)
+![Example of multi-scale management of the Estuary and Gulf of St.
+Lawrence using time-series data from the drivers platform.
+Regional (**B**, **C**) and local (**D**, **E**) scale decision-making are contrasted to show how management processes operating at different scales can influence (**A**, **F**).
+The data used for the period before management actions (**A**, **B**, **C**) is available on the platform, while the data used for the period after management actions (**D**, **E**, **F**) was simulated to correspond to their likely effects on drivers distribution and intensity.
+The effects of port development on shipping were simulated by increasing shipping intensity from the St.
+Lawrence Estuary through the Saguenay River to the projected marine terminal by the expected 2.33-fold increase in number of ships [**D**; @dfo2018b].
+The effects of the static fixed engine fisheries closure in the Gulf of St.
+Lawrence in response to North Atlantic Right Whale (*Eubalaena glacialis*) 2017 casualties were simulated by redistributing the mean total annual biomass yield from the closure area to neighbouring fishing zones (**F**).
+Effects of local and regional management actions were then used to update the overall distribution and intensity of all drivers in the St.
+Lawrence (**F**).
+\label{fig2}](./figures/figure2.png)
 
 \newpage
-\blandscape
 
@@
 
-Table 1. List of drivers currently available on *eDrivers* and used for the analyses presented in this paper.
+Table 1.
+List of drivers listed for inclusion in the multi-drivers platform.
+Their development status is detailed as such: data unavailable: no data is known to be available; seeking collaboration: we are actively seeking collaborations and expertise to develop this drivers layer; in development: we are in the process of building the driver layer, at varying levels of progress; complete: driver layer is complete for this version of the platform.
+Note that drivers identified as complete but identified with an * were completed using data from [@halpern2015b] and that development of these drivers using regional data may still be pursued.
+Finally, a relative evaluation of progress made towards completion of each driver is presented.
+Black and grey boxes are for regional and global data development, respectively.
 
 ```{r table1, echo=F, fontsize=4}
-load('./tables/table1.RData')
-table1
+readRDS('./tables/table1.RDS')
 ```
 
-\elandscape
 \newpage
 
-# References
+# Acknowledgements
+
+TO WRITE
+
+# Author contributions statement
+
+# Conflict of interest statement
+
+The authors declare that the submitted work was carried out in the absence of any personal, professional or financial relationships that could potentially be construed as a conflict of interest.
+
+# Reference
 
@@
  -->
+
+ <!-- It may also be coupled with or serve
+ as a means of operationalizing a network of sampling stations for real-time
+ environmental monitoring [*e.g.* @onc2018]. -->
+
+<!-- management scenarios -->

### 2019-03-22 Code/1-drData.R

@@ -1,0 +1,9 @@
+# This script is simply to load a version of the drivers data that will not
+# change through time, as will be the case with the data accessed through the
+# package. This will ensure that all the analyses and figures can be replicated.
+library(slmeta)
+data(drivers)
+save(drivers, file = './data/drivers.RData')
+
+data(driversRaw)
+save(driversRaw, file = './data/driversRaw.RData')

### 2019-03-22 Code/10-validateCluster.R

@@ -1,0 +1,65 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+k <- 2:10
+
+# ~~~~~~~~~~~~~~~~~~~ Data ~~~~~~~~~~~~~~~~~~~ #
+load('./data/aswMedoid.RData')
+load('./data/wssMedoid.RData')
+load('./data/aswMean2.RData')
+load('./data/wssMean.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ Figure ~~~~~~~~~~~~~~~~~~~ #
+png('./figures/validation.png', width = 1280, height = 920, res = 200, pointsize = 7)
+layout(matrix(1:4, ncol = 2))
+par(mar = c(5,5,2,1))
+# k-medoid
+  # Silhouette
+  plot(aswMedoid ~ k,
+       pch = 20,
+       ylab = 'Average Silhouette Width',
+       type = 'b',
+       frame = F,
+       xlab = '',
+       xlim = c(0,10),
+       cex = 1.5,
+       cex.axis = .8)
+  mtext(side = 3, 'k-medoids clustering', font = 2, line = .75)
+
+  # WSS
+  plot(x = c(1,k),
+       y = wssMedoid,
+       type="b",
+       frame = F,
+       xlab="Number of Clusters (k)",
+       ylab="Total within-clusters sum of squares",
+       pch = 20,
+       xlim = c(0,10),
+       cex = 1.5,
+       cex.axis = .8)
+
+# k-means
+  # Silhouette
+  plot(aswMean2 ~ k,
+       pch = 20,
+       frame = F,
+       xlim = c(0,10),
+       xlab = '',
+       ylab = 'Average Silhouette Width',
+       type = 'b',
+       cex = 1.5,
+       cex.axis = .8)
+  mtext(side = 3, 'k-means clustering', font = 2, line = .75)
+
+  # WSS
+  plot(x = c(1,k),
+       y = wssMean,
+       type="b",
+       frame = F,
+       xlim = c(0,10),
+       xlab="Number of Clusters (k)",
+       ylab="Total within-clusters sum of squares",
+       pch = 20,
+       cex = 1.5,
+       cex.axis = .8)
+dev.off()

### 2019-03-22 Code/11-selectCluster.R

@@ -1,0 +1,17 @@
+# Simple script to select the number of clusters in light of the
+# clustering analyses and the validation, and export the simplified data
+# Number of clusters
+k <- 7
+
+# Load full datasets
+load('./data/kmedoid.RData')
+load('./data/kmeans.RData')
+
+# Object with only selected clustering results
+clMed <- kmedoid[[k-1]]$clustering
+clMean <- kMeans[[k-1]]$cluster
+
+# Export simplified dataset
+save(k, file = './data/k.RData')
+save(clMed, file = './data/clMed.RData')
+save(clMean, file = './data/clMean.RData')

### 2019-03-22 Code/12-compareCluster.R

@@ -1,0 +1,68 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./data/k.RData')
+load('./data/clMed.RData')
+load('./data/clMean.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ METHOD COMPARISON ~~~~~~~~~~~~~~~~~~~ #
+# Make clusters similar
+# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+# may need to be redone depending on the analyses results
+# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+id1 <- clMean == 1
+id2 <- clMean == 2
+id3 <- clMean == 3
+id4 <- clMean == 4
+id5 <- clMean == 5
+id6 <- clMean == 6
+id7 <- clMean == 7
+clMean[id1] <- 6
+clMean[id2] <- 3
+clMean[id3] <- 2
+clMean[id4] <- 7
+clMean[id5] <- 4
+clMean[id6] <- 5
+clMean[id7] <- 1
+
+# Differences
+cl12 <- as.numeric(clMed == clMean) + 1
+corr <- round((sum(clMed == clMean)/length(clMed)*100))
+
+png('./figures/clusterComp.png', width = 2560, height = 1840, res = 300, pointsize = 6)
+par(mfrow = c(2,2))
+
+# K-medoids clustering
+cols <- pal(k)[clMed]
+plotEGSL('egslSimple')
+plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
+plot(st_geometry(egslSimple), col = 'transparent', border = '#25364A', lwd = .75, add = T)
+text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 75000, 'k-medoid clustering', cex = 2, font = 2)
+
+# K-means clustering
+cols <- pal(k)[clMean]
+plotEGSL('egslSimple')
+plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
[…]

### 2019-03-22 Code/13-cluster.R

@@ -1,0 +1,48 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./data/k.RData')
+load('./data/clMed.RData')
+nCl <- length(unique(clMed))
+
+
+# ~~~~~~~~~~~~~~~~~~~ CLUSTER ~~~~~~~~~~~~~~~~~~~ #
+png('./figures/cluster.png', width = 1280, height = 920, res = 300, pointsize = 6)
+cols <- pal(nCl)[clMed]
+plotEGSL(layers     = c('egslSimple', 'canada','usa'),
+         cols       = c('#00000000',off,off),
+         borders    = c('#00000000',focus,focus),
+         lwds       = c(.75,.75,.75),
+         mar        = c(2,2,2,2),
+         box        = T,
+         axes       = 1:4,
+         northArrow = T)
+plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
+plot(st_geometry(egslSimple),
+     col = 'transparent',
+     border = focus,
+     lwd = .75,
+     add = T)
+
+# Legend
+cols <- pal(nCl)
+x <- par('usr')[1:2]
+y <- par('usr')[3:4]
+yH <- 20000
+xW <- 50000
+gap <- 50000
+gap2 <- 40000
+text(x = x[1]+gap, y = y[2]-gap, 'Threat complexes', adj = c(0,.5), font = 2)
+# Polygons
+for(i in 1:7) {
+  xmn <- x[1]+gap+20000
+  xmx <- x[1]+gap+20000+xW
+  ymn <- y[2]-gap-i*gap2-yH/2
+  ymx <- y[2]-gap-i*gap2+yH/2
+  polygon(x = c(xmn, xmx, xmx, xmn), y = c(ymn,ymn,ymx,ymx), col = cols[i], lwd = .5)
+  text(x = xmx+10000, y = mean(c(ymn,ymx)), labels = paste('Threat complex', i), cex = .6, adj = c(0, .5))
+}
+
+dev.off()

### 2019-03-22 Code/14-interDissimilarity.R

@@ -1,0 +1,71 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./data/k.RData')
+load('./data/clMed.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./code/bootstrapSimper.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ SIMPER ~~~~~~~~~~~~~~~~~~~ #
+# Run iterative simper analysis
+iter <- 200
+prop <- .05
+bootSimper <- bootstrapSimper(dr, clMed, iter = iter, prop = prop)
+
+# # This is to run it multiple times simultaneously
+# bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '1')
+# bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '2')
+# load('./data/bootSimper1.RData')
+# x1 <- bootSimper
+# load('./data/bootSimper2.RData')
+# x2 <- bootSimper
+# bootSimper <- c(x1,x2)
+# id <- which(unlist(lapply(bootSimper, is.null)))
+# for(i in rev(id)) bootSimper[[i]] <- NULL
+# save(bootSimper, file = './data/bootSimper.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ SIMPER SUMMARY ~~~~~~~~~~~~~~~~~~~ #
+# Summarize in array format
+iter <- length(bootSimper) # To delete
+nList <- length(bootSimper[[1]])
+
+# Name of variables to extract from simper analyses
+varNames <- c('contr','sd','ava','avb')
+
+# Empty list to store summaries
+simperSummary <- vector('list', nList)
+names(simperSummary) <- names(bootSimper[[1]])
+
+# Empty arrays to store inter-cluster dissimilarity results
+for(i in 1:nList) {
+  simperSummary[[i]] <- array(data = 0,
+                              dim = c(ncol(dr), length(varNames), iter),
+                              dimnames = list(drNames$accr, varNames))
+}
+
+# Extract summary for simper analyses interations
+for(i in 1:iter) {
[…]

### 2019-03-22 Code/15-intraSimilarity.R

@@ -1,0 +1,57 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./data/k.RData')
+load('./data/clMed.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./code/intraSimilarity.R')
+
+# ~~~~~~~~~~~~~~~~~~~ INTRA-CLUSTER SIMILARITY ~~~~~~~~~~~~~~~~~~~ #
+# As with the simper analysis, the dataset is mostly too large to use
+# regular function in packages available in R. Also, while the software
+# PRIMER automatically measures the intra-cluster similarity using the
+# Bray-Curtis similarity, `vegan::simper` does not do it.
+# I therefore create a custom function that allows me to extract the
+# intra-cluster Bray-Curtis similarity and that does it using a bootstrap
+# procedure when the cluster has too many observations.
+#
+sim <- vector('list', k)
+for(i in 1:k) {
+  cat("   k: ", i, "\r")
+  id <- clMed == i
+  sim[[i]] <- bootIntraSim(dr[id, ], iter = 50, samp = 1200)
+  save(sim, file = './data/intraSimilarity.RData')
+}
+
+# Extract species, average and sd
+# Summarize in array format
+iter <- length(sim[[1]])
+varNames <- c('average')
+simSummary <- vector('list', k)
+for(i in 1:k) simSummary[[i]] <- array(data = 0, dim = c(ncol(dr), 1, iter), dimnames = list(drNames$accr, varNames))
+
+for(i in 1:k) {
+  # Check if there are multiple iterations or not.
+  # If length(sim[[i]]) == 2 it means that there were no iterations and the whole cluster was parsed through the simper2 function
+  # If not, then sim[[i]] should be equal to the number of iterations
+  if(length(sim[[i]]) == 2) {
+    simSummary[[i]] <- matrix(ncol = 1,
+                              data = sim[[i]]$similarityContribution$average,
+                              dimnames = list(drNames$accr, 'average'))
+  } else {
+    for(j in 1:iter) {
[…]

### 2019-03-22 Code/16-contributionDissimilarity.R

@@ -1,0 +1,145 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ EXTRA LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+library(gridExtra)
+library(ggplot2)
+library(ggplotify)
+library(grid)
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./data/k.RData')
+load('./data/clMed.RData')
+load('./data/simperMean.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./code/circularPlot.R')
+source('./code/butterfly.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ INDIVIDUAL CONTRIBUTION TO INTER-CLUSTER DISSIMILARITY ~~~~~~~~~~~~~~~~~~~ #
+# Driver percent contribution to dissimilarity
+contDiss <- lapply(simperMean, function(x) x[,'contr'] / sum(x[,'contr'])) %>%
+            as.data.frame()
+colnames(contDiss) <- names(simperMean)
+
+# Total dissimilarity
+totDiss <- unlist(lapply(simperMean, function(x) sum(x[, 'contr'])))
+
+# Driver intensity in compared clusters
+drInt1 <- lapply(simperMean, function(x) x[,'ava']) %>%
+          as.data.frame()
+drInt2 <- lapply(simperMean, function(x) x[,'avb']) %>%
+          as.data.frame()
+
+# Mean value in each cluster
+totMean <- numeric(k)
+for(i in 1:k) {
+  id <- clMed == i
+  totMean[i] <- mean(as.matrix(dr[id, ]))
+}
+
+# Fish
+fish <- matrix(ncol = 5, nrow = 7, dimnames = list(c(), drNames$var[5:9]))
+for(i in 1:k) {
+  id <- clMed == i
+  for(j in colnames(fish)) {
+    fish[i, j]<- sum((dr[id, j]))
+  }
+}
+
+fish <- apply(fish, 2, function(x) round(x / sum(x) * 100))
+
+# Set figure layout view
+mat <- matrix(0,k,k)
[…]

### 2019-03-22 Code/17-contributionSimilarity.R

@@ -1,0 +1,105 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ EXTRA LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+library(gridExtra)
+library(ggplot2)
+library(ggplotify)
+library(grid)
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./data/k.RData')
+load('./data/clMed.RData')
+load('./data/similarityMean.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./code/circularPlot.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ SIMILARITY CONTRIBUTION ~~~~~~~~~~~~~~~~~~~ #
+# Driver percent contribution to dissimilarity
+contSim <- lapply(similarityMean, function(x) x[,'average'] / sum(x[,'average'])) %>%
+            as.data.frame()
+colnames(contSim) <- names(similarityMean)
+
+# Total similarity
+totSim <- lapply(similarityMean, function(x) sum(x[,'average'])) %>%
+          as.data.frame()
+
+# Circulat plots
+p1 <- vector('list', ncol(contSim))
+for(i in 1:length(contSim)) {
+  temp <- data.frame(individual = drNames$accr,
+                     group = drNames$group,
+                     value = round(contSim[, i]*100))
+  rownames(temp) <- NULL
+  temp2 <- round(totSim[i],2)
+  p1[[i]] <- circularPlot(temp, temp2)
+}
+
+# Individual clusters
+egsl <- st_centroid(egslGrid)
+p2 <- vector('list', k)
+for(i in 1:length(p2)) {
+  # Identify observations in cluster i
+  id <- clMed == i
+
+  # Select only grid cells in cluster i
+  temp <- egsl %>%
+          filter(clMed == i) %>%
+          st_coordinates() %>%
+          as.data.frame()
+
+  p2[[i]] <- ggplot() +
+             geom_sf(data = egslSimple, fill = NA) +
+             geom_point(aes(x = X, y = Y), colour = '#187962', size = .01, data = temp) +
+             coord_sf(crs = st_crs(egslSimple), datum = NA) +
[…]

### 2019-03-22 Code/2-param.R

@@ -1,0 +1,78 @@
+# cd phdproj/platformms
+# Libraries
+library(graphicsutils)
+library(MASS)
+library(sf)
+library(magrittr)
+library(vegan)
+library(raster)
+library(cluster)
+library(slmeta)
+
+# Data
+data(egslGrid, package = 'slmeta')
+data(egslSimple, package = 'slmeta')
+data(egslCoast, package = 'slmeta')
+load('./data/drivers.RData')
+dr <- drivers[, -1]
+
+# Parameters
+nDr <- ncol(dr) # Number of drivers
+
+# Driver names data.frame
+drNames <- data.frame(var = character(),
+                      full = character(),
+                      accr = character(),
+                      group = character(),
+                      stringsAsFactors = F)
+
+drNames[1, ] <- c("Aragonite","Aragonite","ARAG","Cl")
+drNames[2, ] <- c("aquacultureInv","Aquaculture","AQUA","C")
+drNames[3, ] <- c("coastDev","Coastal development","CD","C")
+drNames[4, ] <- c("dirHumImpact","Direct human impact","DHI","C")
+drNames[5, ] <- c("fisheriesDD","Demersal destructive fisheries","DD","F")
+drNames[6, ] <- c("fisheriesDNH","Demersal, non-destructive, high-bycatch fisheries","DNH","F")
+drNames[7, ] <- c("fisheriesDNL","Demersal, non-destructive, low-bycatch fisheries","DNL","F")
+drNames[8, ] <- c("fisheriesPHB","Pelagic, high-bycatch fisheries","PHB","F")
+drNames[9, ] <- c("fisheriesPLB","Pelagic, low-bycatch fisheries","PLB","F")
+drNames[10, ] <- c("Hypoxia","Hypoxia","HYP","Cl")
+drNames[11, ] <- c("inorgPol","Inorganic pollution","IP","C")
+drNames[12, ] <- c("invasives", "Invasives species","INV","T")
+drNames[13, ] <- c("marinePol","Marine pollution","MP","T")
+drNames[14, ] <- c("nutrientInput","Nutrient input","NI","C")
+drNames[15, ] <- c("orgPol","Organic pollution","OP","C")
+drNames[16, ] <- c("negSBT","Negative sea bottom temperature anomalies","SBT-","Cl")
[…]

### 2019-03-22 Code/3-drivers.R

@@ -1,0 +1,53 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+load('./data/driversRaw.RData')
+drRaw <- driversRaw[, -1]
+
+# ~~~~~~~~~~~~~~~~~~~ HISTOGRAMS ~~~~~~~~~~~~~~~~~~~ #
+# Value for `par` parameters for figures
+parVal <- c(ceiling(nDr / 4), 4)
+
+# Let's take a look with 0 removed from raw data
+# This is especially important for the coastal drivers, as including zeros
+# would include offshore cells, which vastly outnumber coastal cells and
+# to which coastal drivers to not extend
+png('./figures/drHist.png', width = 1280, height = 1500, res = 200)
+par(mfrow = parVal, mar = c(4,4,1,1))
+for(i in 1:nDr) {
+  id <- drRaw[, i] > 0
+  drID <- drNames$accr[drNames$var == colnames(drRaw)[i]]
+  hist(drRaw[id, i], prob = T, main = '', xlab = drID, col = focus)
+  lines(density(drRaw[id, i], from = 0, to = 1), col = uni, lwd=2)
+}
+dev.off()
+
+
+# ~~~~~~~~~~~~~~~~~~~ DRIVERS ~~~~~~~~~~~~~~~~~~~ #
+png('./figures/drivers.png', width = 1280, height = 1500, res = 200)
+par(mfrow = parVal, mar = c(4,4,1,1))
+# Drivers
+for(i in 1:nDr) {
+  drData <- dr[, i]
+  drID <- drNames$accr[drNames$var == colnames(dr)[i]]
+  id <- drData > 0
+  cols <- pal(101)[((drData[id]/max(drData[id]))*100)+1]
+  plotEGSL('egslSimple')
+  plot(st_geometry(egslGrid[id, ]), add = T, col = cols, border = cols, lwd = .1)
+  text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 70000, labels = drID)
+}
+
+# Scale
+colRamp <- pal(100)
+scale = (length(colRamp) - 1)/(1 - 0)
+ticks <- seq(0, 1, len = 11)
+par(mar = c(.75,0,.75,11))
+plot(c(0, 10), c(0, 1), type = "n", bty = "n", xaxt = "n",
[…]

### 2019-03-22 Code/4-footprint.R

@@ -1,0 +1,41 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+Fd <- rowSums(dr)
+
+
+# ~~~~~~~~~~~~~~~~~~~ FOOTPRINT ~~~~~~~~~~~~~~~~~~~ #
+# Layout
+# Figure 2
+cols <- pal(101)[((Fd/max(Fd, na.rm = T))*100)+1]
+cexMain <- 1
+cexMain2 <- .75
+png('./figures/footprint.png', width = 1280, height = 920, res = 300, pointsize = 6)
+# par(bg = colBack)
+plotEGSL(layers     = c('egslSimple', 'canada','usa'),
+         cols       = c('#00000000',off,off),
+         borders    = c('#00000000',focus,focus),
+         lwds       = c(.75,.75,.75),
+         mar        = c(2,2,2,2),
+         box        = T,
+         axes       = 1:4,
+         northArrow = T)
+plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
+plot(st_geometry(egslSimple),
+     col = 'transparent',
+     border = focus,
+     lwd = .75,
+     add = T)
+
+# Legend
+legendEGSL(range = c(min(Fd), max(Fd)),
+           mainTitle = 'Cumulative exposure',
+           subTitle = '',
+           cexMain = .9,
+           cexSub = .65,
+           n = 6)
+
+# City names
+cityEGSL()
+dev.off()

### 2019-03-22 Code/5-marimekko.R

@@ -1,0 +1,166 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+Fd <- rowSums(dr)
+
+# Create vector of bins
+nBins <- 512 # Just because
+binData <- seq(0,max(Fd), length.out = 513)
+bins <- data.frame(lower = binData[1:512],
+                   upper = binData[2:513])
+bins$mid <- (bins$upper + bins$lower) / 2
+
+# Sum of fd values within each bin
+bins$fdTot <- numeric(nBins)
+for(i in 1:nBins) {
+  id <- Fd > bins$lower[i] & Fd <= bins$upper[i]
+  bins$fdTot[i] <- sum(Fd[id])
+}
+
+# Fd values associated with each driver
+binDr <- matrix(nrow = nrow(bins),
+                ncol = ncol(dr),
+                data = 0,
+                dimnames = list(c(), colnames(dr)))
+
+for(j in 1:ncol(binDr)) {
+  for(i in 1:nBins) {
+    id <- Fd > bins$lower[i] & Fd <= bins$upper[i]
+    binDr[i, j] <- sum(dr[id, j])
+  }
+}
+
+# Transform into a marimekko graph
+# Begin by transforming the data into proportions of total Fd
+binDrProp <- binDr / bins$fdTot
+
+# Sort drivers by group
+sortNames <- as.factor(drNames$group) %>%
+             order() %>%
+             drNames[., ]
+sortNames$group <- as.factor(sortNames$group)
+
+# Add colors
+trans <- c('FF','DD','BB','99','77','55','33','11')
+sortNames$col <- gg_color_hue(nGroup)[as.numeric(sortNames$group)]
+for(i in levels(sortNames$group)) {
+  id <- which(sortNames$group == i)
+  nId <- length(id)
+  for(j in 1:nId) {
+    sortNames$col[id[j]] <- paste0(substr(sortNames$col[id[j]],1,7), trans[j])
+  }
+}
+
+# Sort by group
+binDrProp <- binDrProp[, sortNames$var]
+
+# # Sort drivers according to extent
[…]

### 2019-03-22 Code/6-hotspot.R

@@ -1,0 +1,61 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# Quantile
+# 80th quantile
+# These will be transformed as binary values
+# Then sum the drivers
+# Then look at distribution
+
+# Empty matrix to store results
+hot <- matrix(ncol = ncol(dr),
+              nrow = nrow(dr),
+              data = NA,
+              dimnames = list(c(), drNames$var))
+
+# Evaluate hotspots
+for(i in 1:nDr) {
+  # ID = 0
+  id0 <- dr[, i] > 0
+  # 80th quantile threshold
+  th <- quantile(dr[id0, i], .8)
+  # Values over hotspot threshold
+  hot[, i] <- dr[, i] > th
+}
+
+# Cumulative hotspots
+cumulHotspot <- rowSums(hot)
+cumulHotspot[cumulHotspot == 0] <- NA
+nOver <- max(cumulHotspot, na.rm = T)
+
+
+# Graph
+png('./figures/hotspot.png', width = 1280, height = 920, res = 300, pointsize = 6)
+cols <- pal(nOver)[cumulHotspot]
+plotEGSL(layers     = c('egslSimple', 'canada','usa'),
+         cols       = c('#00000000',off,off),
+         borders    = c('#00000000',focus,focus),
+         lwds       = c(.75,.75,.75),
+         mar        = c(2,2,2,2),
+         box        = T,
+         axes       = 1:4,
+         northArrow = T)
+plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
+plot(st_geometry(egslSimple),
+     col = 'transparent',
+     border = focus,
+     lwd = .75,
+     add = T)
+
+# Legend
+legendEGSL(range = c(1, nOver),
+           mainTitle = 'Number of drivers',
+           subTitle = expression(80^th~quantile),
+           cexMain = .9,
+           cexSub = .65,
+           n = 6)
+
+# City names
+cityEGSL()
+dev.off()

### 2019-03-22 Code/7-kernel.R

@@ -1,0 +1,141 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# -----
+# Figure
+png('./figures/kernel.png', width = 1920, height = 1280, res = 300, pointsize = 6)
+
+# Layout
+mat <- matrix(0, 28, 32)
+mat[1:12, 1:16] <- 1 # traffic
+mat[15:26, 1:16] <- 2 # demersal
+mat[1:12, 17:32] <- 3 # traffic + demersal
+mat[18:28, 17:29] <- 4 # kernel
+mat[15:17, 17:29] <- 5 # dens traffic (x)
+mat[18:28, 30:32] <- 6 # dens demersal (y)
+mat[13:14, 6:11] <- 7 # scale traffic
+mat[27:28, 6:11] <- 8 # scale demersal
+mat[13:14, 22:27] <- 9 # scale combine
+mat[1:2, 15:16] <- 10 # panel A
+mat[15:16, 15:16] <- 11 # panel B
+mat[1:2, 31:32] <- 12 # panel C
+mat[15:16, 31:32] <- 13 # panel D
+layout(mat)
+# layout.show(13)
+
+# Plot 1 - Hypoxia
+  cols <- rbPal(101)[(drivers$Hypoxia * 100)+1]
+  plotEGSL(layers = 'egslSimple', borders = '#C7CBCE66')
+  plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
+  plot(st_geometry(egslSimple), border = '#00000077', lwd = 0.25, add = T)
+
+# Plot 2 - Demersal destructive
+  cols <- rbPal(101)[(drivers$fisheriesDD * 100)+1]
+  plotEGSL(layers = 'egslSimple', borders = '#C7CBCE66')
+  plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
+  plot(st_geometry(egslSimple), border = '#00000077', lwd = 0.25, add = T)
+
+# Plot 3 - Added layers - hypoxia + demersal destructive
+   drivers$combine <- drivers$Hypoxia + drivers$fisheriesDD
+   drivers$combine <- drivers$combine / max(drivers$combine)# Normalize it x / max(x))
+   cols <- rbPal(101)[(drivers$combine * 100)+1]
+   plotEGSL(layers = 'egslSimple', borders = '#C7CBCE66')
[…]

### 2019-03-22 Code/8-kmedoids.R

@@ -1,0 +1,48 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ Data prep ~~~~~~~~~~~~~~~~~~~ #
+# Data needs to be scaled for clustering
+dr <- scale(dr)
+
+# ~~~~~~~~~~~~~~~~~~~ k-medoids clustering ~~~~~~~~~~~~~~~~~~~ #
+# Cluster analysis
+nSample <- 100
+sampleSize <- 10000
+k <- 2:10
+kmedoid <- list('list', length(k))
+for(i in k) {
+  cat("   k: ", i, "\r")
+  kmedoid[[i]] <- clara(dr,
+                        k = i,
+                        sampsize = sampleSize,
+                        samples = nSample,
+                        metric="manhattan",
+                        stand = TRUE,
+                        keep.data = FALSE)
+}
+kmedoid[[1]] <- NULL
+save(kmedoid, file = './data/kmedoid.RData')
+
+# ~~~~~~~~~~~~~~~~~~~ Average silhouette width ~~~~~~~~~~~~~~~~~~~ #
+aswMedoid <- unlist(lapply(kmedoid, function(x) x$silinfo$avg.width))
+save(aswMedoid, file = './data/aswMedoid.RData')
+
+# ~~~~~~~~~~~~~~~~~~~ Elbow method ~~~~~~~~~~~~~~~~~~~ #
+# Verify number of clusters by using the elbow method,
+# i.e. the inflexion point in the sum of squared error (SSE) scree plot
+# Extract within cluster sum of squared error (WSS) from each clustering run
+# Wss function
+wss <- function(x) (nrow(x)-1) * sum(apply(x,2,var))
+
+# Measure for each k and each cluster
+wssMedoid <- wss(dr)
+for(i in 1:length(kmedoid)) { # For each k value tested
+  wssTemp <- numeric()
+  for(j in 1:k[i]) { # For each cluster
+    id <- kmedoid[[i]]$clustering == j # Identify observations in cluster j
+    wssTemp[j] <- wss(dr[id, ])
+  }
+  wssMedoid[k[i]] <- sum(wssTemp)
+}
+save(wssMedoid, file = './data/wssMedoid.RData')

### 2019-03-22 Code/9-kmeans.R

@@ -1,0 +1,111 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ Data prep ~~~~~~~~~~~~~~~~~~~ #
+# Data needs to be scaled for clustering
+dr <- scale(dr)
+
+# ~~~~~~~~~~~~~~~~~~~ k-means clustering ~~~~~~~~~~~~~~~~~~~ #
+# Cluster analysis
+k <- 2:10
+kMeans <- list('list', length(k))
+for(i in k) {
+  cat("   k: ", i, "\r")
+  kMeans[[i]] <- kmeans(x = dr,
+                      centers = i,
+                      iter.max = 1000,
+                      nstart = 25,
+                      algorithm = 'Lloyd')
+}
+kMeans[[1]] <- NULL
+save(kMeans, file = './data/kmeans.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ Average silhouette width ~~~~~~~~~~~~~~~~~~~ #
+# Silhouette needs to be computed manually here, because the general function crashes
+# This will make a very inefficient function, but one that will work
+sil <- function(x, cl) {
+  # Euclidian distance function between a single point and another point or a string of points
+  euc <- function(x, y) {
+    if(is.null(dim(y))) {
+      sqrt(sum((x - y) ^ 2))
+    } else {
+      sqrt(colSums((x - t(y)) ^ 2))
+    }
+  }
+
+  # Number of rows in dataset x
+  nr <- length(cl)
+
+  # Clusters in dataset
+  k <- unique(cl)
+
+  # Empty data.frame to store silhouette information
+  df <- data.frame(cluster = numeric(nr),
+                   neighbor = numeric(nr),
+                   sil_width = numeric(nr),
+                   stringsAsFactors = F)
+  # Calculate silhouette for all data points in dataset x
+  for(i in 1:nr) {
+    # Cluster of data point i
+    df$cluster[i] <- cl[i]
+
+    # ID of points in cluster of point i
+    id <- cl == df$cluster[i]
+
+    # a = mean distance of point i to all points within its cluster
+      a <- euc(x[i, ], x[id, ]) %>%
+           mean()
+
+    # b =  bi is the smallest average distance of i to all points in any other
+    #      cluster of which i is not a member.
+      # Clusters of which i is not a member
[…]

### 2019-03-22 Code/bootstrapSimper.R

@@ -1,0 +1,28 @@
+# iter <- 10
+# prop <- .02 # .05: ~500s/iter | .01: ~20s/iter | .02: ~80s/iter
+# The analysis being very long, we will proceed using a
+# bootstrap procedure, i.e. iterative simper analysis
+bootstrapSimper <- function(x, cl, iter = 100, prop = .1, accr = '') {
+  # List to store simper results for post-processing
+  bootSimper <- vector('list', iter)
+
+  # Determine the number of data points to sample from each cluster
+    clSub <- ceiling(table(cl) * prop)
+
+  # Repeat the process for the desired number of iterations
+  for(i in 1:iter) {
+    cat("   i: ", i, ' of ', iter, "\r")
+
+    # Subsampled dataset
+    samp <- vector('list', length(clSub))
+    for(j in 1:length(clSub)) samp[[j]] <- sample(x = which(cl == j), size = clSub[j])
+    samp <- unlist(samp)
+
+    # Run simper analysis on subset of data
+    bootSimper[[i]] <- simper(x[samp, ], cl[samp])
+
+    # Save file
+    save(bootSimper, file = paste0('./data/bootSimper', accr, '.RData'))
+  }
+  return(bootSimper)
+}

### 2019-03-22 Code/intraSimilarity.R

@@ -1,0 +1,92 @@
+# library(slmeta)
+# data(drivers, package = 'slmeta')
+# dr <- drivers[, !colnames(drivers) %in% c('UVpositive','UVnegative','ID')]
+#
+# x <- dr[1:500, ]
+# system.time(sim <- simper2(x))
+#
+
+intraSim <- function(x, samp = NULL, iter = NULL) {
+  # ~~~~~~~~~~~~~~~~~~ PARAMETERS & DATA SETUP ~~~~~~~~~~~~~~~~~~
+  # Transform data as matrix
+  x <- as.matrix(x)
+
+  # Create empty list to store results
+    sim <- vector('list', 2)
+    names(sim) <- c('totalSimilarity', 'similarityContribution')
+    class(sim) <- 'simper2'
+
+  # ~~~~~~~~~~~~~~~~~~ MEASUREMENTS FOR SPECIES CONTRIBUTION ~~~~~~~~~~~~~~~~~~
+  # Here is the function that we will be applying for each species i between pairs of observations j, k
+    # Numerator: 2 * min(A_{i,j}, A_{i,k})
+    # Denominator: sum(j, k)
+
+  # Denominator: The species contribution will be divided by the sum of all abundances in all pairs of observations
+    # Sum of rows of community matrix
+    xs <- rowSums(x)
+
+    # Sum of all pairs of observations
+    denom <- combn(xs, m = 2, FUN = sum)
+
+  # Numerator: 2 * min(id1, id2)
+    # Minimum
+    sp <- apply(x, 2, function(x) combn(x, m = 2, FUN = min))
+
+    # Minimum value * 2
+    sp <- 2 * sp
+
+    # Divide by denominator
+    sp <- apply(sp, 2, function(x) x/denom)
+
+  # ~~~~~~~~~~~~~~~~~~ TOTAL CONTRIBUTION ~~~~~~~~~~~~~~~~~~
+  # Now from the sp matrix, we can obtain the intra-cluster average similarity
+  # by taking the mean of row sums
+    # Sum of individual species contributions
+    simTot <- rowSums(sp)
+
+    # Average and standard deviation result for intra-cluster similarity
+    sim$totalSimilarity <- c(mean(simTot), sd(simTot))
+    names(sim$totalSimilarity) <- c('Mean','SD')
+
+  # ~~~~~~~~~~~~~~~~~~ SPECIES CONTRIBUTION ~~~~~~~~~~~~~~~~~~
+  # Data frame with summary results
[…]

### 2019-03-22 Tables/datatable.R

@@ -1,0 +1,61 @@
+library(knitr)
+library(kableExtra)
+
+# Table 1
+df <- data.frame(Groups = character(0),
+                 Drivers = character(0),
+                 SpatRes = character(0),
+                 TempRes = character(0),
+                 Years = character(0),
+                 Units = character(0),
+                 Source = character(0),
+                 Accronym = character(0),
+                 stringsAsFactors = F)
+
+df[1,] <- c("Climate", "Aragonite", 'Lat/long', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ARAG')
+df[2,] <- c("Coastal", "Aquaculture", 'Lat/long', '-','Variable, between 1990-2016', '$presence-absence$','***TBD***','AQUA')
+df[3,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ negative anomalies', '[@galbraith2018]','SBT-')
+df[4,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ positie anomalies', '[@galbraith2018]','SBT+')
+df[5,] <- c("Climate",  "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ negative anomalies', '[@galbraith2018]','SST-')
+df[6,] <- c("Climate", "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ positive anomalies', '[@galbraith2018]','SST+')
+df[7,] <- c("Climate", "Sea level rise", 'Modeled 0.25 degree', '10 days', '1992-2012', '$mm$','[@halpern2015b]','SLR')
+df[8,] <- c("Marine traffic", "Shipping", '0.1 degree', 'Event based', '2003-2011', '$n$ lanes', '[@halpern2015b]','SHP')
+df[9,] <- c("Fisheries", "Pelagic, low-bycatch", 'Lat/long', 'Event based', '2010-2015', '$kg$', '[@dfo2016]','PLB')
+df[10,] <- c("Fisheries", "Pelagic, high-bycatch", 'Lat/long', 'Event based', '2010-2015', '$kg$', '[@dfo2016]','PHB')
[…]

### 2019-03-29 eDriversMS-v2.Rmd
add updated introduction

---
title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
author: David Beauchesne^1,2,\*^, Rémi M. Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^5^, Simon Bélanger^4^, Pascal Bernatchez^4^, Hugo Bourdages^7^, Clément Chion^6^, Peter S. Galbraith^7^, Benjamin Halpern^8,9^, Camille Lavoie^2^, Christopher W. McKindsey^7^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^7^, Anne-Sophie Ste-Marie^5^, Philippe Archambault^2^
date: 2019-02-06
fontsize: 12pt
output:
  pdf_document:
    toc: false
    number_sections: true
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   # - \usepackage{titling}
   # - \pretitle{\begin{flushleft}}
   # - \posttitle{\end{flushleft}}
   - \usepackage{lscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
relativeurls: true
bibliography: eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
cd phdproj/eDrivers/eDriversMS/
rmarkdown::render('./eDriversMS.rmd', 'pdf_document')
rmarkdown::render('./eDriversMS.rmd', 'word_document')
-->

<!-- Setting up parameters for code snippet presented in Box 1 -->
\lstset{language=R,
    basicstyle=\small\ttfamily,
    stringstyle=\color{DarkGreen},
    otherkeywords={fetchDrivers, install_github, getData, kde2d},
    morekeywords={TRUE,FALSE},
    deletekeywords={data,frame,length,as,character},
  keywordstyle=\color{blue},
  commentstyle=\color{DarkGreen},
}

^1^Institut des sciences de la mer, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^2^Québec Océan, Département de biologie, Université Laval, Québec, QC, Canada \newline
^3^Département de biologie, Université de Sherbrooke, Sherbrooke, QC, Canada \newline
^4^Département de Biologie, Chimie et Géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
^5^St. Lawrence Global Observatory, Rimouski, QC, Canada \newline
^6^Département des Sciences naturelles, Université du Québec en Outaouais, Gatineau, QC, Canada \newline
^7^Fisheries and Oceans Canada, Maurice Lamontagne Institute, Mont-Joli, QC, Canada \newline
^8^National Center for Ecological Analysis and Synthesis, University of California, Santa Barbara, CA, United States \newline
^9^Bren School of Environmental Science and Management, University of California, Santa Barbara, CA, United States \newline
^10^Department of Earth & Planetary Sciences, McGill University, Montréal, QC, Canada \newline
^11^Ministère Environnement et Lutte contre les changements climatiques, Québec, QC, Canada  \newline

**Correspondence**: \newline
David Beauchesne \newline
david.beauchesne@uqar.com \newline

**Number of figures**: 5 \newline
**Number of tables**: 1 \newline

\newpage

# Abstract

# Introduction

The St. Lawrence System, formed by one of the largest estuaries in the world and
a vast interior sea, is a complex social-ecological system characterized by
highly variable environmental conditions and oceanographic processes, both in
space and time [@el-sabh1990; @white1997; @dufour2007]. It thus offers a unique
and heterogeneous array of habitats suited for the establishment of diverse and
productive ecological communities [@savenkoff2000]. As a result, the St.
Lawrence System provides a wealth of ecosystem services that have historically
and contemporarily benefited the Canadian economy. It sustains a rich fisheries
industry targeting more than 50 species, serves as the gateway to eastern
North-America by granting access to more than 40 ports and the most densely
populated Canadian region, hosts a booming tourism industry and an expanding
aquaculture production, fosters emerging activities and boasts a yet untapped
hydrocarbon potential [@beauchesne2016; @schloss2017 @archambault2017]. With
major investments recently made and more forthcoming in economic and
infrastructure development and research [*e.g.* @goq2015; @rqm2018], an
intensification of the human footprint is expected in the St. Lawrence System.

As elsewhere in the world [see @halpern2015], this intensifying human footprint
will likely result in increasingly intricate environmental exposure regimes,
*i.e.* suites of overlapping and potentially interacting environmental drivers
threatening ecosystems, habitats or ecological communities. Drivers, often
referred to as stressors or pressures, are any externalities that affect
environmental processes and disturb natural systems. Drivers may originate from
natural or human-induced biophysical processes (*e.g.* sea surface temperature
anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
fisheries and marine pollution). The potential for complex interactions between
drivers is the largest uncertainty when studying or predicting environmental
change [@darling2008; @cote2016]. The effects of multiple drivers can combine
non-linearly and result in effects that are greater (synergistic effect) or
lower (antagonistic effect) than the sum of individual effects
[@crain2008; @darling2008; @cote2016].  <!-- For example...??? -->

The uncertainty associated with complex driver interactions must therefore be
taken into account when investigating environmental impacts [@cote2016], yet
most research on driver effects in marine environments remains overwhelmingly
focused on single driver assessments [@obrien2019]. Arising from the expected
increase in environmental exposure and the experiences of past and
current ecological tragedies such as the collapse of cod fisheries
[@frank2005; @dempsey2018] and the decline of the beluga and right whale
populations [@plourde2014] is a need to characterize the distribution,
intensity and overlap between drivers in the St. Lawrence System. This will
provide critical information on areas most exposed to cumulative drivers and on
the interaction potential of drivers in the St. Lawrence. It is also a necessary
step towards the holistic and integrated management of the St. Lawrence System.

Gathering environmental data for large scale, systematic initiatives can,
however, be a very challenging and time consuming - not to say painful - process.
On one hand, there is an overwhelming and expanding wealth of data available.
Such information overload may inhibit our ability to make decisions based on
scientific information, promote massive effort duplication, disproportionately
appropriate research funds to certain sectors, and obscure knowledge gaps amid
a sea of information [@eppler2004]. On the other hand, crucial data are lacking
and remain largely unavailable or inaccessible for a variety of reasons,
including proprietary rights, lack of organizational time, capacity and
training, and in some rare cases unwillingness to share, curtailing our ability
for appropriate decision-making.

The current digital infrastructure is highly decentralized and the data
management and sharing practices are highly heterogeneous, preventing us from
maximizing benefits from research investments [@wilkinson2016].
Yet there now exists multiple initiatives that address this issue by assembling,
organizing and sharing vast arrays of environmental data.
Biotic data can be accessed through web portals such as the Ocean Biogeographic
Information System [OBIS; @obis2018], the Global Biodiversity Information Facility
[GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
and the World Register of Marine Species [WoRMS; @worms2017]. Abiotic data can
also be accessed through WorldClim [@hijmans2005], Bio-ORACLE
[@tyberghein2012], and MARSPEC [@sbrocco2013]. Initiatives focused on sharing
environmental data for specific areas also exist, such as the St. Lawrence
Global Observatory (SLGO; https://ogsl.ca/en), the European Marine Observation
and Data Network (EMODnet; http://www.emodnet.eu/) and the U.S. Integrated Ocean
Observing System (IOOS; https://ioos.noaa.gov/). Essential environmental
parameters are also organized, coordinated and acquired through global
initiatives like the Group on Earth Observations Biodiversity Observation
Network [GEO BON; @scholes2012], the Census of Marine Life [CoML; @coml2010]
and the Global Ocean Observing System (GOOS; http://www.goosocean.org).

Tools collating data and knowledge on multiple drivers in a comparable and
interoperable way are however conspicuously missing from these initiatives
[but see @halpern2015]. This is in spite of integerated management and assessment
approaches needing efficient data reporting, standardized data management
practices and tools tailored to the study of the effects of multiple drivers
[@dafforn2016; @stock2018]. An additional objective thus emerged in the process of
addressing our initial goals:
sharing the knowledge gathered through the process of characterizing drivers in the St. Lawrence.
We are launching *eDrivers*,
an open knowledge platform focused on sharing knowledge on the distribution and
intensity of drivers in the St. Lawrence System and gathering a community of
experts committed to structuring, standardizing and sharing knowledge on drivers
in support of science and management.

This manuscript thus serves to present two distinct but related goals. The first
is to characterize the distribution and intensity of drivers in the St. Lawrence
System in order to: 1) identify areas of high cumulative exposure,
2) characterize areas with similar cumulative exposure regimes, dubbed threat
complexes [@bowler2019] and 3) identify drivers that are likely to interact
in the St. Lawrence. The second is to present the guiding
principles underlying the inception and structure of *eDrivers* and use
this paper as a call for collaboration to any person, group or
organization that might be interested in contributing to this initiative.

### 2019-04-23 Code/8-kmedoids.R
update analyses

@@ -10,7 +10,7 @@ dr <- scale(dr)
 nSample <- 100
 sampleSize <- 10000
 k <- 2:10
-kmedoid <- list('list', length(k))
+kmedoid <- vector('list', length(k)+1)
 for(i in k) {
   cat("   k: ", i, "\r")
   kmedoid[[i]] <- clara(dr,
@@ -20,6 +20,7 @@ for(i in k) {
                         metric="manhattan",
                         stand = TRUE,
                         keep.data = FALSE)
+save(kmedoid, file = './data/kmedoid.RData')
 }
 kmedoid[[1]] <- NULL
 save(kmedoid, file = './data/kmedoid.RData')

### 2019-04-23 Code/1-data.R
update analyses

@@ -1,0 +1,18 @@
+# This script is simply to load a version of the drivers and spatial objects data that will not
+# change through time, as will be the case with the data accessed through the
+# package. This will ensure that all the analyses and figures can be replicated.
+# Also, I transform the spatial objects to decimal degrees for the figures, as requested by co-authors
+library(slmeta)
+data(drivers, package = 'slmeta')
+save(drivers, file = './data/drivers.RData')
+
+data(driversRaw)
+save(driversRaw, file = './data/driversRaw.RData')
+
+data(egslGrid, package = 'slmeta')
+egslGrid <- st_transform(egslGrid, crs = 4326)
+save(egslGrid, file = './data/egslGrid.RData')
+
+data(egslSimple, package = 'slmeta')
+egslSimple <- st_transform(egslSimple, crs = 4326)
+save(egslSimple, file = './data/egslSimple.RData')

### 2019-04-23 Code/3-drivers.R
adjust names on drivers figure

@@ -33,7 +33,7 @@ for(i in 1:nDr) {
   cols <- pal(101)[((drData[id]/max(drData[id]))*100)+1]
   plotEGSL('egslSimple', prj = slmetaPrj('world'), extent = 'egslSimple')
   plot(st_geometry(egslGrid[id, ]), add = T, col = cols, border = cols, lwd = .1)
-  text(x = mean(par('usr')[1:2]), y = par('usr')[4] - 70000, labels = drID)
+  text(x = mean(par('usr')[1:2]), y = par('usr')[4]-.5, labels = drID)
 }
 
 # Scale

### 2019-04-24 eDriversMS-SI-v2.rmd
latest version of the manuscript

--- 2019-03-22 eDriversMS-SI.rmd
@@
     number_sections: true
+header-includes:
+    - \usepackage{float}
+    - \usepackage{setspace}
+    - \onehalfspacing
 relativeurls: true
@@
 cd phdproj/edrivers/edriversms
-rmarkdown::render('./eDriversMS-SI.rmd', 'pdf_document')
+rmarkdown::render('./eDriversMS-SI-v2.rmd', 'pdf_document')
 rmarkdown::render('./eDriversMS-SI.rmd', 'html_document')
@@
 
-### Aragonite
+### Acidification
 
-***Describe data***
+Oceans are the largest reservoir and sink of atmospheric carbon dioxide ($CO_2$). Its uptake
+increases seawater acidity and the lowers saturation state with respect to calcite
+($\Omega_C$) and aragonite ($\Omega_A$), the two most common $CaCO_3$ polymorphs
+constituting the shells and skeleton of many marine organisms [@mucci2017].
+This can have deleterious effects on carbonate-secreting organisms (*e.g.* mollusks and crustaceans)
+and certain physiological processes in non-calcifying organisms
+[@fabry2008; @kroeker2013].
 
-We interpolated aragonite distribution using the exponential kriging model.
-We then built an index of aragonite level stress. When $\Omega$ aragonite levels
-decrease below 1, the water becomes corrosive to calcifying organisms such as
-echinoderms, corals and mollusks larvae. These organisms thus have to
-continually recalcify their shells to maintain their structure, which becomes
-harder to do as $\Omega$ aragonite levels further decrease towards 0. We therefore
-considered stress associated to decreasing $\Omega$ aragonite levels to increase
-linearly between values of 1 and 0. The index of stress therefore has a value of
-0 and 1 for values of $\Omega$ aragonite equal to 1 and 0, respectively.
+When $\Omega_C$ or $\Omega_A$ decrease below 1, water
+becomes undersaturated and corrosive to the skeletal minerals of carbonate-secreting
+organisms. Note that if $\Omega_A$ $< 1$, the waters
+may still be supersaturated with respect to calcite since it is 50% more soluble
+than aragonite. However, organisms
+have wide ranging responses to $\Omega_C$ and $\Omega_A$ saturation state.
+For example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019], while other organisms
+may have adapted to precipitate $CaCO_3$ even when seawater is undersaturated
+with respect to calcite [$\Omega_C$ $< 1$ or $\Omega_A$ $< 0.65$; e.g. @uthicke2016].
+
+We used $\Omega_{A}$ saturation state to characterize ocean acidification in the
+bottom waters of the St. Lawrence. The data were collected [...]. $\Omega_{A}$,
+pH, dissolved oxygen ($O_2$) and pressure were sampled at 117 stations in the
+Estuary and Gulf of St. Lawrence in the summer of 2018. The majority of sampling
+stations were located in the Southern Gulf ($n = 84$), compared to the Estuary
+and Northern Gulf ($n = 33$).
+
+To account for low sample size in the Northern Gulf, we used the
+correlation between $O_2$ and $CO_2$, which are linked through the stoichiometry
+of the respiration reaction in waters that are isolated from the atmosphere [@mucci2011].
+Metabolic $CO_2$ increases in
+water through biotic processes such microbial respiration of organic matter
+[@mucci2011]. In deep waters, variations of $\Omega_{A}$ are largely governed by the addition of metabolic
+$CO_2$, whereas near the air-sea interface $O_2$ is replenished faster than $CO_2$ can
+escape [@zeebe2001]. The impact of respiration on carbonate chemistry is highest in hypoxic
+regions, where metabolic $CO_2$ accumulates to high concentrations [@mucci2011].
+
+We explored the correlation between $\Omega_{A}$ and $O_2$ in the data and
+found no significant correlation at the scale of the whole St. Lawrence (*p-value*:
+0.08; $R^2$: 0.03). Nevertheless, a correlation
+between $\Omega_A$ and $O_2$ exists in the Northern St. Lawrence (*p-value*:
+< 0.01; $R^2$: 0.49), but not in the Southern St. Lawrence (*p-value*: 0.85;
+$R^2$: < 0.01). Furthermore, the correlation observed in the north between $\Omega_A$
+and $O_2$ is especially strong when deep Channels (*Deep* model: *p-value*:
+< 0.01; $O_2$: 0.20; $R^2$: 0.97) are separated from shallower regions
+(*Shallow* model: *p-value*: < 0.01; $O_2$: 0.07; $R^2$: 0.96), with $\Omega_A$
+increasing more steeply with increasing $O_2$ in deep Channels. This stronger
+correlation is likely caused by the lack of gas exchange with the atmosphere
+or the surface mixed layer and the importance of metabolic $CO_2$ in deep
+waters, some of which may have mean ages of up to 20 years [@mucci2011]. In the
+Northern Gulf, the acidification is following
+depth and distance from the Cabot Strait along the Laurentian, Anticosti and
+Esquiman Channels due to the progressive oxygen depletion and metabolic carbon
+dioxide accumulation [@mucci2011]. There are likely different dynamics at play
+in the Southern Gulf, such as the accumulation of metabolic carbon dioxide
+produced by rapid oxic degradation of organic matter in the sandy and permeable
+sediments of the Magdalen Shelf [*personal communication?*]. <!-- see with Michel -->
+
+We thus divided the analyses between the Southern and Northern St. Lawrence.
+The St. Lawrence was divided using the 350m isobath on the southern
+slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé
+Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modelled using the exponential
+kriging model.
+
+We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. Whereas
+there are few published datasets on $\Omega_{A}$ in the St.
+Lawrence, there is a robust time-series on oxygen concentration in the water
+column from the Department of Fisheries and Oceans’ (DFO) Atlantic Zone
+Monitoring Program [AZMP; @galbraith2018]. We used oxygen concentration data
+collected between 2013 and 2017
+[@blais2018] and interpolated oxygen % saturation using cokriging with depth as a
+covariable, as done in [@dutil2011]. Bathymetric data come from @dutil2011
+and have a 100 $km^2$ resolution. For more details, refer to the hypoxia
+description. The *Deep* model was used to transform $O_2$ values to $\Omega_A$
+in grid cells with depth > 350m, while the *Shallow* model was
+applied to cells with depth < 350m.
+
+Based on the wide range of $\Omega_A$ levels at which organism are negatively
+affected (see above), we built an index of acidification stress ($A_s$)
+ranging between 0 and 1 that slowly begins increasing as $\Omega_A$ $\approx$
+$2$, increases more rapidly close to $\Omega_A$ $= 1$ and reaches its peak at
+$\Omega_A$ $\approx$ $0.5$:
+
+$$A_s = \frac{-3}{.99 + e^{-2 * \Omega_A}} + 3$$
+
+Hence, the higher the acidification stress, the higher the index of
+acidification becomes.
 
@@
 
-***Describe data***
+The data used to characterize hypoxia uses oxygen saturation in % and come from
+the Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
+[AZMP; @galbraith2018] surveys in late spring and fall, as well as from annual
+DFO’s multispecies surveys for the northern Gulf in August and for the Magdalen
+Shallows in September. We provide a brief summary of data and methods to
+characterize hypoxia in this document. For more details, refer to @blais2018.
 
-We interpolated the dissolved oxygen using cokriging with depth as a covariable,
-as done in [@dutil2011]. According to @diaz1995, severe hypoxia can be observed
-when oxygen saturation falls below 2 $ml$ $L^{-1}$ or 62.5 $\mu mol$ $L^{-1}$, which is considered
-the level necessary to maintain most animal life. This threshold is equal to 1.4
-$ml$ $L^{-1}$ (1 $mg$ $L^{-1}$ = 0.7 $ml$ $L^{-1}$).
-We used this threshold to create an index of hypoxia using an inverted logistic curve
-to transform the dissolved oxygen values as an index of hypoxic stress $H_s$ ranging
-from 0 to 1, under the assumption that hypoxic stress will increase following
-a logistic curve as it approaches the hypoxic threshold:
+Oxygen concentration is measured at every station using an oxygen probe
+(Sea-Bird SBE43) mounted on the CTD. The probe is calibrated against seawater
+samples that are analyzed by Winkler titration on every cast
+[see @blais2018 for calibration procedure]. The data used is the last depth
+sampled on the CTD profile, which typically corresponds to ~10m above the
+bottom. Oxygen concentration ($\mu M$ $L^{-1}$) is converted to oxygen
+saturation taking into account the salinity and temperature at the selected
+depth.
 
-$$H_s = \frac{-1}{200 * e^{-2.5 * DO_2}} + 1$$
+We interpolated oxygen saturation using cokriging with depth as a covariable,
+as done in [@dutil2011]. Bathymetric data come from @dutil2011 and have a
+100 $km^2$ resolution.   According to @diaz1995, severe hypoxia can be observed
+when dissolved oxygen falls below 2 $ml$ $L^{-1}$, which corresponds to 62.5
+$\mu mol$ $L^{-1}$ and ~20% saturation. This is considered as the level
+necessary to maintain most animal life. Similarly, @chabot2008 studied the
+effects of hypoxia on the energy budget of cod in the St. Lawrence and found
+that behavioural effects began manifesting below 70% oxygen saturation and that
+survival becomes jeopardised below ~20% saturation. We used these observations
+to create an index of hypoxia using an inverted logistic curve that slowly
+increases below 70% oxygen saturation, increases acutely as it approaches the
+20% oxygen saturation threshold and reaches its peak intensity between 20 and
+30% oxygen saturation:
+
+$$H_s = \frac{-1}{.99 + 200 * e^{-0.15 * O_2}} + 1$$
+
+Hence, the higher the hypoxic stress, the higher the index of hypoxia becomes.
 
@@
 Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
-[AZMP; @galbraith2018]. We provide a brief summary of data and methods
-to characterize surface temperature climatology and anomalies in this document.
+[AZMP; @galbraith2018]. We provide a brief summary of data and methods to
+characterize the bottom temperature climatology and anomalies in this document.
 For more details, refer to @galbraith2018.
@@
 multispecies surveys for the northern Gulf in August and for the Magdalen
-Shallows in September. Using this sampling survey, temperatures are interpolated
-at each 1 m depth layer on a 2 km resolution grid. Bottom temperature are then
-extracted from this grid by using a bathymetry layer from the Canadian
-Hydrographic Survey [@dutil2012].
+Shallows in September. Using this sampling survey, temperatures are horizontally
+interpolated at each 1 m depth layer on a 2 km resolution grid. Bottom
+temperatures are then extracted by using a bathymetry layer from the Canadian
+Hydrographic Survey [@dutil2012] and selecting the interpolated temperature
+from the layer corresponding to the bottom depth at each grid point.
 
-We used temperature anomalies, *i.e.* deviations from long-term normal conditions,
-to measure an annual index of stress associated with extreme temperatures between
-2013 and 2017. Temperature anomalies were calculated using the difference between
-grid cell monthly climatologies to the associated long-term averages generated
-from the reference period between 1981 and 2010. Grid cells whose monthly value
-exceeded ±0.5 standard deviation (SD) from the long-term average were considered
-as anomalous [@galbraith2018].
-
-Outliers in the data were considered as those that fell beyond the
-interquartile range (IQR) * 3, identified as extreme ouliers by @tukey1977.
-Outlier values were capped to correspond to the 5th and 95th percentiles.
-Anomalies were divided into positive and negative anomalies and
-summed for the available months of September and August.
-By adding the monthly anomaly values, we essentially considered the summed
-deviation from the mean as an indicator of the annual intensity of surface
-temperature anomalies in the St. Lawrence.
+We used temperature anomalies, *i.e.* deviations from long-term normal
+conditions, to measure an annual index of stress associated with extreme
+temperatures between 2013 and 2017. Temperature anomalies were calculated using
+the difference between grid cell values with 1981-2010 climatological averages.
+Anomaly time series were normalized by their standard deviation (SD) to allow
+comparison between areas of the St. Lawrence with different temperature ranges.
+For example, temperatures observed in deep channels are less variable than in
+shallower regions of the St. Lawrence. Hence, if anomalies were expressed in
+degrees Celsius, it would underestimate the relative importance of anomalies
+in deep channels when compared to shallower regions. Grid cells whose monthly
+value exceeded ±0.5 standard deviation (SD) from the long-term average were
+considered as anomalous [@galbraith2018]. Outliers in the data were defined as
+those that fell beyond the interquartile range * 3, identified as extreme
+outliers by @tukey1977. Outlier values were capped to correspond to the 5th and
+95th percentiles. Anomalies were divided into positive and negative anomalies
+and the absolute value of anomalies were used as an indicator of the intensity
+of bottom temperature anomalies. The mean anomaly intensity between 2013 and
+2017 for each grid cell was used to generate the final index of bottom
+temperature anomalies.
 
@@
 Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
-[AZMP; @galbraith2018]. We provide a brief summary of data and methods
-to characterize surface temperature climatology and anomalies in this document.
+[AZMP; @galbraith2018]. We provide a brief summary of data and methods to
+characterize the surface temperature climatology and anomalies in this document.
 For more details, refer to @galbraith2018.
 
-The surface layer is characterized using a variety of methods:
+The surface layer is characterized using sea surface temperature (SST) monthly
+composites from Advanced Very High Resolution Radiometer (AVHRR) satellite
+images obtained from the National Oceanic and Atmospheric Administration (NOAA)
+and European Organisation for the Exploitation of Meteorological Satellites
+(EUMETSAT). Images used are from DFO’s Maurice Lamontagne Institute at a 1 km
+resolution from 1985-2013 and from DFO’s Bedford Institute of Oceanography (BIO)
+Operational Remote Sensing group at a 1.5 km resolution since 2014. Monthly
+anomalies were constructed as the difference between monthly averages and the
+1985-2010 climatological mean for each month.
 
-1. Temperature-salinity sensors installed on various sihps and forming the
-shipboard thermosalinograph network,
-2. Moored instruments recording water temperature every 30 minutes and forming
-the thermograph network,
-3. Sea surface temperature (SST) monthly composite climatologies using Advanced
-Very High Resolution Radiometer (AVHRR) satellite images from the National
-Oceanic and Atmospheric Administration (NOAA) and European Organisation for the
-Exploitation of Meteorological Satellites (EUMETSAT) available from the Maurice
-Lamontagne Institute sea surface temperature processing facility at a 1km
-resolution from 1985-2013 and from the Bedford Institute of Oceanography (BIO)
-Operational Remote Sensing group at a 1.5km resolution since 2014.
-
-Positive and negative surface temperature anomalies were identified following
-the same approach used for bottom temperature anomalies. Compared to the
-bottom temperature anomalies, however, data is available thoughout the year.
-Only the months of April to November were included to avoid biases associated
-with anomalies measurements due to the presence of ice cover.
+Surface temperature anomalies were characterized following the same method used
+for bottom temperature anomalies. Only the months of May to November were
+included to avoid biases associated with the presence of ice cover. Monthly
+anomalies from May to November values were summed to obtain an indicator of
+annual surface temperature anomaly intensity in each grid cell. The mean
+anomaly intensity between 2013 and 2017 for each grid cell was used to generate
+the final index of surface temperature anomalies.
 
@@
 - Organic pollution
-- Sea bottom temperature anomalies
+- Sea bottom temperature anomalies <!-- verify -->
 - Shipping
@@
 
-![Frequency distribution of the untransformed data for all driver layers. \label{hist}](./figures/drHist.png)
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/drHist.png}
+\caption{Frequency distribution of the untransformed data for all driver layers.}
+\label{hist}
+\end{figure}
 
-![Distribution and intensity of transformed and normalized drivers in the Estuary and Gulf of St. Lawrence available on *eDrivers*. \label{drivers}](./figures/drivers.png)
-
-\newpage
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/drivers.png}
+\caption{Distribution and intensity of transformed and normalized drivers in the Estuary and Gulf of St. Lawrence available on *eDrivers*.}
+\label{drivers}
+\end{figure}
 
@@
 
-![Frequency distribution of cumulative exposure (*i.e.* sum of normalized driver intensity in each grid cell) and percent contribution of each driver to the frequency distribution of cumulative exposure in the Estuary and Gulf of St.
-Lawrence \label{marimekko}](./figures/marimekko.png)
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/marimekko.png}
+\caption{Frequency distribution of cumulative exposure (*i.e.* sum of normalized driver intensity in each grid cell) and percent contribution of each driver to the frequency distribution of cumulative exposure in the Estuary and Gulf of St.
+Lawrence.}
+\label{marimekko}
+\end{figure}
 
@@
 
-We identified regions with similar cumulative exposure regimes, referred to as
-threat complexes after [@bowler2019], using a partional *k-medoids* clustering
-algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
-designed to use with large datasets. The CLARA algorithm uses the PAM
-(Partition Around Medoids) algorithm on a sample from the original dataset
-to identify a set of $k$ objects that are representative of all other objects,
-*i.e.* medoids (also referred to as examplars) and that are central to the
-cluster they represent. The goal of the algorithm is to iteratively minimize
-dissimilarity between objects in the cluster and the representative medoid.
-Sets of $k$ medoids are thus selected and replaced in order to iteratively
-minimize the intra-cluster dissimilarity. With the CLARA algorithm, this
-process is repeated and iterations are compared on the basis of their average
-dissimilarity between objects and cluster medoids to select the optimal set of
-$k$ medoids that minimizes average dissimilarity. We performed the analyses for
-a total of 100 iterations using samples of 10000 observations, representing ~5%
-of the drivers dataset. Analyses were performed using the *cluster* R package
-[@maechler2018].
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/validation.png}
+\caption{Validation procedure for the *k-medoids* and *k-means* clustering algorithms based on the number of cluster that maximizes average silhouette width [upper panels; @kaufman1990] and minimizes the total within-cluster sum of squares (WSS; lower panels).}
+\label{valid}
+\end{figure}
 
-Because the algorithm identifies clusters by minimizing dissimilarity around a
-set of *k* medoids, it requires a pre-defined number of clusters. The
-appropriate number of clusters $k$ was thus tested using values ranging from
-2 to 10 and validated by selecting the number of clusters that maximized the average
-silhouette width [@kaufman1990] and minimized the total within-cluster sum of
-squares (WSS; Figure S\ref{valid}).
-
-We also validated the clustering by comparing *k-medoids* clustering with
-*k-means* clustering with the *Lloyd* algorithm [@lloyd1982]. The *k-means*
-approach is similar to the *k-medoids*, but identifies observations belonging
-to a cluster iteratively by minimizing the mean intra-cluster squared distance
-until it converges to an optimal solution. We used 25 random sets and set a
-maximum of 1000 iterations for the analysis. Analyses were performed using
-the *stats* R package [@rcoreteam2018]. We used the same validation procedure
-to select the optimal number of clusters $k$ than with the *k-medoids*
-clustering (Figure S\ref{valid}).
-
-Both the *k-medoids* and *k-means* clustering validation procedure suggest that
-there are 7 distinc threat complexes in the St. Lawrence (Figure S\ref{valid}).
-While the comparison of of the spatial distribution of clusters results in only
-a 66% cell correspondance between the *k-medoids* and *k-means* approaches, the
-bulk of the difference is attributable to a single undivided cluster that
-combines most of the southern and norther Gulf in a single cluster in favor of
-a more localised cluster by the *k-means* algorithm (Figure S\ref{comparison}).
-While *k-means* algorithms are much faster than *k-medoids* since it does not
-need a pairwise dissimilarity measurement of all observations, it is also more
-sensitive to outliers since it focuses on the mean rather than a centroid. We
-therefore opted the use of *k-medoids* clustering.
-
-![Validation procedure for the *k-medoids* and *k-means* clustering algorithms based on the number of cluster that maximizes average silhouette width [upper panels; @kaufman1990] and minimizes the total within-cluster sum of squares (WSS; lower panels).
-\label{valid}](./figures/validation.png)
-
-![Comparison of the distribution of the 7 clusters identified using the *k-medoids* and *k-means* clustering algoriths and the percent grid cell correspondance between the two approaches \label{comparison}](./figures/clusterComp.png)
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/clusterComp.png}
+\caption{Comparison of the distribution of the 6 clusters identified using the *k-medoids* and *k-means* clustering algoriths and the percent grid cell correspondance between the two approaches.}
+\label{comparison}
+\end{figure}
 
@@
 
-The difference between clusters was explored by measuring the total inter-cluster
-dissimilarity and the contribution of each driver to the total inter-cluster
-dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis
-dissimilarity [Figure S\ref{inter}; @clarke1993]. As the drivers dataset is too
-large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting
-5% of each cluster to run the analysis and repeating the process over 300 iterations.
-We also compared the mean intensity of each driver within each cluster to
-better capture the inter-cluster dissimilarity. Analyses were performed using
-the *vegan* R package [@oksanen2018].
-
-![Evaluation of inter-cluster dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis dissimilarity [@clarke1993].
-The figure diagonal presents the distribution of the 7 clusters identified using the *k-medoids* clustering algorithm.
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/interDissimilarity.png}
+\caption{Evaluation of inter-cluster dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis dissimilarity [@clarke1993].
+The figure diagonal presents the distribution of the 6 clusters identified using the *k-medoids* clustering algorithm.
 The lower triangle shows all combinations of inter-cluster dissimilarity with circular barplots showing the percent contribution to total dissimilarity of each driver and with the total inter-cluster dissimilarity in the center of the barplots.
-The upper triangle shows the average relative intensity of each driver for all driver combinations, with barplots to the left and the right representing the row and columns clusters, respectively.
-\label{inter}](./figures/interDissimilarity.png)
+The upper triangle shows the average relative intensity of each driver for all driver combinations, with barplots to the left and the right representing the row and columns clusters, respectively.}
+\label{inter}
+\end{figure}
 
@@
 
-Intra-cluster similarity was evaluated using the Bray-Curtis similarity index
-(Figure S\ref{intra}). As with the inter-cluster dissimilarity, we used a
-bootstrap procedure for the intra-cluster similarity, randomly selecting 5% of
-each cluster observation to run the analysis and repeating the process over 300
-iterations. We however did not use the bootstraping procedure for clusters 2
-and 6, as they both contain less than 10000 observations, making the calculation
-time more manageable.
-
-![Evaluation of intra-cluster similarity using the Bray-Curtis similarity index.
-The distribution of all 7 clusters is presented along with circular barplots showing the percent contribution to total similarity of each driver and with the total intra-cluster similarity in the center of the barplots.
-\label{intra}](./figures/intraSimilarity.png)
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/intraSimilarity.png}
+\caption{Evaluation of intra-cluster similarity using the Bray-Curtis similarity index.
+The distribution of all 7 clusters is presented along with circular barplots showing the percent contribution to total similarity of each driver and with the total intra-cluster similarity in the center of the barplots.}
+\label{intra}
+\end{figure}
 

### 2019-04-24 eDriversMS-v2.Rmd
latest version of the manuscript

@@
 author: David Beauchesne^1,2,\*^, Rémi M.
-Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^5^, Simon Bélanger^4^, Pascal Bernatchez^4^, Hugo Bourdages^7^, Clément Chion^6^, Peter S.
-Galbraith^7^, Benjamin Halpern^8,9^, Camille Lavoie^2^, Christopher W.
-McKindsey^7^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^7^, Anne-Sophie Ste-Marie^5^, Philippe Archambault^2^
-date: 2019-02-06
+Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^4^, Simon Bélanger^5^, Pascal Bernatchez^5^, Marjolaine Blais ^6^, Hugo Bourdages^6^, Clément Chion^7^, Peter S.
+Galbraith^6^, Benjamin Halpern^8,9^, Camille Lavoie^2^, Christopher W.
+McKindsey^6^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
+date: 2019-04-23
 fontsize: 12pt
@@
    - \newcommand{\elandscape}{\end{landscape}}
+   - \usepackage{bbm}
+   - \usepackage{setspace}
+   - \onehalfspacing
+   - \usepackage{float}
 relativeurls: true
@@
 cd phdproj/eDrivers/eDriversMS/
-rmarkdown::render('./eDriversMS.rmd', 'pdf_document')
+rmarkdown::render('./eDriversMS-v2.rmd', 'pdf_document')
 rmarkdown::render('./eDriversMS.rmd', 'word_document')
@@
 ^3^Département de biologie, Université de Sherbrooke, Sherbrooke, QC, Canada \newline
-^4^Département de Biologie, Chimie et Géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
-^5^St. Lawrence Global Observatory, Rimouski, QC, Canada \newline
+^4^St. Lawrence Global Observatory, Rimouski, QC, Canada \newline
+^5^Département de Biologie, Chimie et Géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
 ^6^Département des Sciences naturelles, Université du Québec en Outaouais, Gatineau, QC, Canada \newline
@@
 
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
 # Abstract
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+<!-- 350 words -->
 
+The St. Lawrence System is a vast and complex social-ecological system providing
+a wealth of ecosystem services sustaining numerous economic sectors. Related
+human pressures combine with climate driven environmental changes to produce
+intricate exposure regimes of overlapping and potentially interacting
+environmental drivers (*e.g.* hypoxia and fisheries) threatening ecosystems.
+There is a critical need to systematically characterize the distribution,
+intensity and overlap between drivers to support integrative initiatives such
+as ecosystem-based management. While portals providing knowledge on individual
+drivers exist, platforms collating comparable and interoperable knowledges on
+multiple drivers remain conspicuously missing. This paper presents two distinct,
+but related objectives. The first is to characterize the distribution and
+intensity of drivers in the St. Lawrence System. The second is to launch
+*eDrivers*, an open knowledge platform gathering experts committed to
+structuring, standardizing and sharing knowledges on drivers in support of
+science and management. We gathered data on 22 coastal, climate, fisheries and
+marine traffic drivers through collaborations, existing environmental
+initiatives and open data portals. We show that few areas of the St. Lawrence
+are free of cumulative exposure. The Estuary, the Anticosti Gyre and coastal
+areas are particularly exposed, especially in the vicinity of urban centers.
+We identified 6 areas of distinct cumulative exposure regime that show that
+certain drivers typically co-occur in different regions of the St. Lawrence and
+that coastal areas are exposed to all driver types. Of particular concern are
+two threat complexes capturing most exposure hotspots that show the convergence
+of contrasting exposure regimes at the head of the Laurentian Channel. These
+observations are destined to improve as *eDrivers* evolves through time to
+address knowledge gaps and refine current driver layers. In an effort to share
+the knowledge acquired and to ensure the lasting relevance of the description
+of drivers presented in this manuscript, *eDrivers* was built on a series of
+guiding principles upholding existing data management and open science
+standards. Ultimately, we believe that *eDrivers* represents a much needed
+solution that could radically influence broad scale research and management
+practices by increasing data accessibility and interoperability and by
+increasing research and decision-making efficiency.
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
 # Introduction
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
 
@@
 current ecological tragedies such as the collapse of cod fisheries
-[@frank2005; @dempsey2018] and the decline of the beluga and right whale
+[@frank2005; @dempsey2018] and the decline of the beluga and right whale <!-- right whale ref -->
 populations [@plourde2014] is a need to characterize the distribution,
@@
 
-Tools collating data and knowledge on multiple drivers in a comparable and
-interoperable way are however conspicuously missing from these initiatives
-[but see @halpern2015]. This is in spite of integerated management and assessment
-approaches needing efficient data reporting, standardized data management
-practices and tools tailored to the study of the effects of multiple drivers
-[@dafforn2016; @stock2018]. An additional objective thus emerged in the process of
-addressing our initial goals:
-sharing the knowledge gathered through the process of characterizing drivers in the St. Lawrence.
-We are launching *eDrivers*,
-an open knowledge platform focused on sharing knowledge on the distribution and
-intensity of drivers in the St. Lawrence System and gathering a community of
-experts committed to structuring, standardizing and sharing knowledge on drivers
-in support of science and management.
+However, equivalent platforms for environmental drivers have largely focused on
+single drivers (e.g. Global Fishing Watch) and platforms collating
+data and knowledge on multiple drivers in a comparable and interoperable way
+remain conspicuously missing [but see @halpern2015]. This is in spite of
+integerated management and assessment approaches needing efficient data reporting,
+standardized data management practices and tools tailored to the study of the
+effects of multiple drivers [@dafforn2016; @stock2018]. An additional objective
+thus emerged in the process of addressing our initial goals: sharing the
+knowledge gathered through the description of drivers in the St.
+Lawrence.
 
-This manuscript thus serves to present two distinct but related goals. The first
+This manuscript thus has two distinct, but related, goals. The first
 is to characterize the distribution and intensity of drivers in the St. Lawrence
@@
 complexes [@bowler2019] and 3) identify drivers that are likely to interact
-in the St. Lawrence. The second is to present the guiding
-principles underlying the inception and structure of *eDrivers* and use
-this paper as a call for collaboration to any person, group or
-organization that might be interested in contributing to this initiative.
+in the St. Lawrence. The second goal is to present how we are sharing the
+knowledge acquired by launching an open knowledge platform, *eDrivers*, and use
+this paper as a call for collaboration to any person, group or organization that
+might be interested in contributing to this initiative.
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Materials and Methods
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+## Drivers
+
+The list of drivers for which we sought data was informed by a global cumulative
+impact assessment initiative [@halpern2008; @halpern2015], regional holistic
+evaluations of the state of the St. Lawrence [@dufour2007; @benoit2012], and
+communications with regional experts. Through the data gathering process, we
+developed and continue to develop collaborations with regional experts and data
+holders. We also use global data from the global cumulative impact assessment
+initiative [@halpern2008; @halpern2015] available from the National Center
+for Ecological Analysis and Synthesis (NCEAS) online data repository
+[Table 1; @halpern2015b]. We selected global data that were unavailable at the
+regional scale and that were available at a resolution adequate for use at the
+scale of the St. Lawrence (*e.g.* marine pollution).
+
+We were able to characterize the intensity and distribution of 22 drivers
+(Table 1;) and we are actively working on updating or developing additional
+driver layers. Drivers incorporated in the analyses are varied in origin, *i.e.* from
+terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping), and from
+large scale biophysical processes (*e.g.* temperature anomalies) to localized
+anthropogenic activities (*e.g.* fisheries). Drivers were divided into 4 groups:
+coastal, climate, fisheries and marine traffic (Table 1). All data layers and
+methodologies are described in the supplementary materials.
+
+Drivers with non-normal frequency distributions were log-transformed (Figure S1)
+and all drivers were scaled between 0 and 1 to allow driver comparisons. The
+99th quantile of individual driver distribution was used as the upper bound for
+scaling to control for extreme values and produce maps of individual drivers
+(Figure S2). All drivers were embedded in a regular grid composed of 245604
+$1 km^2$ exagonal cells to construct the integrated dataset used for the
+analyses.
+
+## Cumulative exposure
+
+Areas with high cumulative exposure (objective 1) were identified by evaluating
+the cumulative footprint of combined drivers and by identifying hotspots of
+cumulative footprint. Cumulative footprint ($F$) was defined as the sum of
+the scaled intensity of all drivers in each grid cell:
+
+$$F_x = \sum_{i=1}^{n} D_{i,x}$$
+
+where $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
+driver $i$.
+
+Cumulative hotspots ($H$) were defined as the number of drivers in each grid
+cell with scaled intensity contained over their respective 80th percentile:
+
+<!-- Indicator function -->
+$$H_x = \sum_{i=1}^{n} \mathbbm{1} (D_{i,x} \; \epsilon \; P_{80, D_i})$$
+
+where, $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
+driver $i$ and $P_{80, D_i}$ is the 80th percentile of driver $i$. Hotspots
+thus identify areas where drivers are co-occurring at high relative intensities.
+
+## Driver interactions
+
+The distribution of driver interactions was investigated through the spatial
+overlap of combinations of drivers using the cumulative footprint ($F$) equation
+with pairs of drivers. The intensity at which pairs of drivers co-occur was
+evaluated using a two-dimensional kernel density estimate. As there are 231
+pairwise combinations between 22 drivers, we focus on a single example using
+hypoxia and demersal destructive fisheries, two drivers known to occur mainly
+in deeper areas of the St. Lawrence and, hence, an interaction between the
+effects of the two drivers could be anticipated. Note, however, that *eDrivers*
+will provide the capacity to compare any combinations of drivers.
+
+## Threat complexes
+
+Natural systems are likely to host multiple overlapping drivers. The
+dimensionality of the problem can quickly rise as additional drivers are
+considered. In order to decrease the dimensionality of the integrated dataset,
+we identify threat complexes, *i.e.* regions with similar cumulative exposure
+regimes, using a clustering approach [*e.g.* see @bowler2019].
+
+### Clustering
+
+Threat complexes were identified using a partional *k-medoids* clustering
+algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
+designed for large datasets. The CLARA algorithm uses the PAM (Partition Around
+Medoids) algorithm on a sample from the original dataset to identify a set of
+$k$ objects that are representative of all other objects, *i.e.* medoids and
+that are central to the cluster they represent. The goal of the algorithm is to
+iteratively minimize intra-cluster dissimilarity. Iterations are compared on the
+basis of the average dissimilarity between cluster objects and representative
+medoid to select the optimal set of $k$ medoids that minimizes average
+dissimilarity. We used 100 iterations using samples of 10000 observations
+(*i.e.* ~5% of observations) to identify clusters. Analyses were performed
+using the *cluster* R package [@maechler2018].
+
+Partitional clustering algorithms require a user-defined number of clusters.
+Values of $k$ ranging from 2 to 10 were tested and validated by selecting
+the number of clusters that maximized the average silhouette width [@kaufman1990]
+and minimized the total within-cluster sum of squares (Figure S3).
+
+We also validated the clustering by comparing *k-medoids* clustering with
+*k-means* clustering with the *Lloyd* algorithm [@lloyd1982]. The *k-means*
+approach is similar to the *k-medoids*, but identifies observations belonging
+to a cluster iteratively by minimizing the mean intra-cluster squared distance
+until it converges to an optimal solution. We used 25 random sets and set a
+maximum of 1000 iterations for the analysis. Analyses were performed using
+the *stats* R package [@rcoreteam2018]. We used the same validation procedure
+to select the optimal number of clusters $k$ than with the *k-medoids*
+clustering (Figure S3).
+
+While *k-means* algorithms are more  efficient since they do not compute
+pairwise dissimilarities, it is more sensitive to outliers through the use of
+the mean rather than a centroid. We therefore favored the use of the *k-medoids*
+algorithm, but used the *k-means* to validate clusters.
+
+### Inter-cluster dissimilarity
+
+The difference between clusters was explored by measuring the total inter-cluster
+dissimilarity and the contribution of each driver to the total inter-cluster
+dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis
+dissimilarity [Figure S4; @clarke1993]. As the drivers dataset is too
+large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting
+5% of each cluster to run the analysis and repeating the process over 300
+iterations. We also compared the mean intensity of each driver within each
+cluster to better capture the inter-cluster dissimilarity. Analyses were
+performed using the *vegan* R package [@oksanen2018].
+
+### Intra-cluster similarity
+
+Intra-cluster similarity was evaluated using the Bray-Curtis similarity index
+(Figure S5). As with the inter-cluster dissimilarity, we used a
+bootstrap procedure for the intra-cluster similarity, randomly selecting 5% of
+each cluster observation to run the analysis and repeating the process over 300
+iterations. We however did not use the bootstraping procedure for clusters with
+less than 10000 observations since computation time was manageable.
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Results and discussion
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+## Cumulative exposure
+
+Apart from the northeastern Gulf, the cumulative footprint of drivers is
+ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
+generally highest along the coast (Figure \ref{footprint}), with hotposts
+located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
+offshore areas are less exposed to cumulative drivers, with the Estuary and the
+Anticosti Gyre being notable exceptions (Figures \ref{footprint} and
+\ref{hotspot}). This is not to say that offshore areas are free of exposure,
+as most of the St. Lawrence is exposed to multiple overlapping drivers
+(Figures \ref{footprint} and \ref{hotspot}). For example, it is worthy to note
+high cumulative footprint observed at the heads of the Anticosti and Esquiman
+Channels (Figure \ref{footprint}).
+
+These results are consistent with observations elsewhere in the world, where
+cumulative driver exposure conspicuously arises from and markedly intensifies
+close to coastal cities and at the mouth of rivers draining highly populated
+areas [e.g. @halpern2015; @feist2016; @mach2017; @stock2018]. These are areas
+where human activities (*e.g.* coastal development and shipping) and footprint
+(*e.g.* pollution runoff) are the most intense [@feist2016], and on which is
+overlaid a background of natural disturbances [@micheli2016]. They are also the
+areas in which the most dramatic increases in exposure are expected, with
+populations increasing more rapidly along the coast than inland [@feist2016].
+In the St. Lawrence, large coastal cities are mostly located along the Estuary
+and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
+or home to small coastal communities.
+
+As for offshore exposure, the Estuary, along with the St. Lawrence River,
+provide access to and serve as the primary drainage outflow of the Great Lakes
+Basin, the most densely populated region in Canada [@statcan2017]. Most marine
+traffic thus converges to the Estuary.
+
+While we cannot ascertain that high exposure areas are the most impacted,
+we can safely predict that these are the areas where studying ecosystem state
+will be the most complex due to the uncertainty associated with driver
+interactions, an uncertainty bound to increase rapidly with the
+number of interacting drivers [@cote2016].
+
+## Driver interactions
+
+Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels,
+with the head of the Channels most exposed to hypoxia (Figure \ref{kernel}A).
+Demersal destructive fisheries are located along the Laurentian Channel, the
+heads of the Anticosti and Esquiman Channels and around the Magdalen Islands
+(Figure \ref{kernel}B). By combining both drivers, we can observe that hypoxia
+and demersal destructive fisheries overlap mostly at high relative intensity
+(Figure \ref{kernel}D) in the vicinity of the  Anticosti Gyre and the heads of
+the Esquiman and Anticosti Channels (Figure \ref{kernel}C). The ease with which
+this figure can be created using *eDrivers* is demonstrated in box 1.
+
+Fisheries in the St. Lawrence have historically affected
+biodiversity distribution and habitat quality [@moritz2015]. Concurrently,
+hypoxia decreases overall habitat quality, but triggers species-dependent
+responses ranging from adaptation [*e.g.* northern shrimp *Pandalus borealis*
+and Greenland halibut *Reinhardtius hippoglossoides*; @pillet2016] to reduced
+growth rates [@dupont2013] and avoidance of oxygen-depleted habitats
+[*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality
+[*e.g.* sessile benthic invertebrates; @eby2005; @gilbert2007; @belley2010].
+Certain species may thus be adversely affected by fisheries and withstand
+hypoxia but still experience a decrease in prey availability, while others may
+be deleteriously affected by the compounded effect of both drivers [@deleo2017].
+
+## Threat complexes
+
+While informative, the hypoxia-fisheries example focuses on a single pair of
+drivers and falls short of the number of drivers overlapping at
+high intensities throughout the St. Lawrence (Figure \ref{hotspot}).
+The number of drivers overlapping in the St. Lawrence increases with cumulative
+exposure (Figure S3). Areas with high exposure such as the Estuary, the
+Anticosti Gyre and the southwestern Gulf coastline (Figure \ref{footprint} and
+\ref{hotspot}) are thus areas where driver interactions are
+most likely, and where they can arise between a host of different drivers.
+
+The identification of threat complexes provides a crucial tool to simplify the
+multi-dimensional complexity of overlapping drivers to areas exposed to similar
+suites of drivers [@bowler2019]. This may prove critical for a  better
+understanding the state of species, habitats and ecosystems located within or
+moving through threat complexes and exposed to the combined effects of all
+drivers typical to those areas.
+
+Six distinct threat complexes were identified in the St. Lawrence using
+the *k-medoids* and *k-means* algorithms (Figures S4, S5). Based on their
+distribution and representative drivers, threat complexes can be divided into
+3 offshore and 3 coastal complexes (Figures \ref{cluster}, S6 and S7).
+Coastal threat complexes (1 to 3; Figure \ref{cluster}) include all types
+of drivers besides hypoxia and are the most exposed threat complexes, both in
+terms of driver overlap and intensity. Threat complex 2 is differentiated from
+other complexes by the presence of aquaculture sites. Threat complex 1
+encompasses the coastline and is characterized by higher direct human impact
+(*i.e.* population density). Threat complex 3 is the most exposed complex and
+has a distribution similar to the most exposed coastal hotspots (Figure
+\ref{hotspot}). This complex is characterized by high intensities of land-based
+drivers (*e.g.* nutrient input), demersal non-destructive high-bycath fisheries
+(*e.g.* trap fishing), climate drivers and marine traffic drivers in
+the vicinity of ports.
+
+Offshore threat complexes (4 to 6; Figure \ref{cluster}) are generally
+characterized by high intensity climate and marine traffic drivers.
+Threat complex 4 is differentiated by demersal non-destructive high-bycath
+fisherie, higher marine traffic drivers compared to complex 5 and generally
+corresponds to the whole Southern Gulf. Threat complex 5 is characterized by
+more fisheries types (*i.e.* demersal destructive and pelagic high-bycatch),
+generally lower intensity marine traffic drivers and is located almost
+exclusively in the Northern Gulf. Finally, threat complex 6 corresponds
+primarily to the Laurentian Channel, a deep (250-500 m) and long (1250 km)
+submerged valley connecting the Estuary to the Atlantic. It also incorporates
+parts of the Esquiman and Anticosti Channels, two deep channels that branch off
+from the the Laurential Channel to the north towards the Arctic and the north
+of Anticosti Island, respectively. This threat complex is the most exposed
+offshore threat complex and includes all offshore hotspots (Figure \ref{hotspot}).
+This complex is characterized by high intensity hypoxia, marine traffic and
+pollution, as well as demersal destructive and pelagic high-bycatch fisheries.
+
+Of particular concern are threat complexes 3 and 6, which are the two most
+exposed complexes in the St. Lawrence and are characterized by distinct
+cumulative exposure regimes Between them, they capture most of the coastal
+and offshore hotspots identified in the St. Lawrence and discussed above. <!-- more here? -->
+
+They also offer some insight into the potential importance of considering
+spatial dynamics in areas intersecting multiple threat complexes. For example,
+threat complexes 3 and 6 meet at the mouth of the River Saguenay. This area
+is particularly dynamic, with deep Atlantic waters advected through estuarine
+circulation mixing with surface waters from the Great Lakes Basin and the
+Saguenay River [@dufour2007]. This results in the convergence of climate
+drivers from the bottom of the Laurentian Channel and marine traffic drivers
+(threat complex 6) with terrestrial run-off from river outflows and direct
+human impacts (threat complex 3). This dynamic area is also highly
+productive and hosts large aggregations of krill exploited by numerous fish,
+marine mammal and marine bird species [@simard2009; @lavoie2016]. It might
+therefore be reasonable to expect highly unpredictable environmental effects in
+this area.
+<!-- I know this is weak, but it's late and I'm tired! -->
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Open Knowledge Platform: *eDrivers*
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+Sharing the knowledge acquired through the description of drivers in the
+St. Lawrence quickly emerged as a priority to curtail the need to reach dozens
+of experts across multiple organizations and over extensive periods of time to
+assemble the data needed to apply integrated research and management. It is also
+a requirement to ensure that this manuscript will not become a quickly outdated
+snapshot of drivers distribution and intensity in the St. Lawrence System,
+but rather serve as a stepping stone towards an adaptive and ever-improving
+collection of knowledges.
+
+As such, we are launching *eDrivers*, an open knowledge platform focused on
+sharing knowledge on the distribution and intensity of drivers and on gathering
+a community of experts committed to structuring, standardizing and sharing
+knowledge on drivers in support of science and management. In launching this
+intiative, our objective is to uphold the highest existing standards of data
+management and open science. We identified four guiding principles to meet this
+objective and that guide the structure of the platform (Figure \ref{diag}).
+
+## Unity and inclusiveness
+
+***Why***: Operating over such large scales in time, space and subject matter
+requires a vast and diverse expertise that cannot possibly be possessed by any
+one individual or organization. Consequently, we envision an initiative that
+seeks to mobilize all individuals and entities with relevant expertise.
+
+***How***: By promoting, consolidating and working with experts involved in
+existing and highly valuable environmental initiatives already in place in the
+St. Lawrence. Notable examples of environmental initiatives are the annual
+review of physical [@galbraith2018], chemical, and biological [@blais2018]
+oceanographic conditions in the St. Lawrence, the fisheries monitoring program
+[@dfo2016], the annual groundfish and shrimp multidisciplinary survey
+[@bourdages2018], the characterisation of benthic [@dutil2011], epipelagic and
+coastal [@dutil2012] habitats of the St. Lawrence, and Canada's shoreline
+classification [@eccc2018]. There are also nascent efforts to share
+information on several human activities in the St. Lawrence such as the Marine
+Spatial Data Infrastructure portal, which provides data on zoning, shipping,
+port activities, and other human activities in Canadian waters, including the
+St. Lawrence system [@goc2016].
+
+By working with existing data portals whose objective is to share
+environmental data. We are thus collaborating actively with the St. Lawrence
+Global Observatory (SLGO) to develop the initiative and to host the
+platform on their web portal (https://ogsl.ca/en). The mission of SLGO is to
+promote and facilitate the accessibility, dissemination and exchange of official
+and quality data and information on the St. Lawrence ecosystem through the
+networking of organisations and data holders to meet their needs and those of
+users, to improve knowledge and to assist decision-making in areas such as
+public safety, climate change, transportation, resources and biodiversity
+conservation. The SLGO is also one of three regional association spearheading
+the Canadian Integrated Ocean Observing System
+(CIOOS; http://meopar.ca/research/cioos-call-for-proposals/),
+which will focus on integrating oceanographic data from multiple sources to
+make them accessible to end-users and to enable the national coordination of
+ocean observing efforts by integrating isolated or inaccessible data, and by
+identifying gaps or duplications in observations and research efforts.
+We are also developing collaborations with the Portal on water knowledge
+(http://www.environnement.gouv.qc.ca/eau/portail/), an initiative from the
+Québec provincial government. This portal aims at collecting and sharing
+accurate, complete and updated ressources on water and aquatic ecosystems to
+support the mandate of relevant actors and stakeholders working in
+water and aquatic ecosystems management.
+
+By actively inviting, seeking, and developing collaborations as well as
+encouraging constructive criticism from the inception and throughout the
+lifetime of the platform.
+
+By inviting external community contributions (Figure \ref{diag}).
+External researchers or entities wishing to submit marine data will be able to
+do so through the SLGO web portal (https://ogsl.ca/en). Submissions through
+other data portals will also be accepted either through the development of data
+sharing agreements or with the caveat that shared data are under an open source
+license and that they adhere to the platform data standards.
+
+## Findability, accessibility, interoperability and reusability
+
+***Why***: Open data has been propelled to the forefront of scientific research
+in an era of open, collaborative and reproducible science. By moving towards
+large scale, cross-disciplinary research and management projects, there is a
+growing need to increase the efficiency of data discovery, access,
+interoperability and analysis [@reichman2011; @wilkinson2016]. Our goal is to
+foster efficient and functional open science by creating a fully open,
+transparent and replicable open knowledge platform.
+
+***How***:
+By building an infrastructure adhering to the FAIR Data Principles,
+which states that data and metadata must be Findable, Accessible, Interoperable
+and Reusable. These principles focus on the ability of humans and machines to
+automatically find and (re)use data and knowledge [@wilkinson2016].
+
+By making data and associated tools accessible through a variety of ways:
+the SLGO web portal (https://ogsl.ca/en), two R packages called *eDrivers*
+(https://github.com/orgs/eDrivers/eDrivers) and *eDriversEx*
+(https://github.com/orgs/eDrivers/eDriversEx) to access the data through SLGO's
+API and to provide analytical tools to explore data, respectively, and a Shiny
+application (https://david-beauchesne.shinyapps.io/eDriversApp/) to explore
+drivers data interactively (Figure \ref{diag}).
+Note that the data are currently contained within and accessible through the
+*eDrivers* R package only, as we are actively working to allow users to
+download selected layers from SLGO's web portal and geoserver. The functions
+available in *eDrivers* to access the data have however been developed to
+ensure forward compatibility once the data migrate to SLGO's geoserver.
+
+By defining clear data and metadata standards and specifications to support
+the regional standardization of current and future protocols and practices and
+to favour interoperability with national and international initiatives like the
+Essential Ocean Variables (EOV) identified by the Global Ocean Observing System
+[GOOS; http://www.goosocean.org]. As such, we will adopt the metadata standard
+currently targetted for the upcoming CIOOS, *i.e.* the North American Profile
+of ISO 19115:2014 - Geographic information - Metadata, a schema favoured
+for geospatial data in Canada and the United-States.
+
+By providing version control and code access to the workflows set up to generate
+driver layers from raw data, the R packages and the Shiny application through
+a GitHub organization called *eDrivers* (https://github.com/orgs/eDrivers/).
+
+## Adaptiveness
+
+***Why***: In the face of uncertainty and in an effort to address impending
+environmental changes, adaptive management has been identified as the chief
+strategy to guide efficient decision-making
+[*e.g.* @costanza1998; @margules2000; @keith2011; @jones2016] and has already
+been discussed in the context of multi-drivers and cumulative impact assessments
+[@halpern2015; @beauchesne2016; @cote2016; @schloss2017]. Adaptive management can only be
+truly achieved through a commitment to adaptive monitoring and data reporting
+[@margules2000; @halpern2012; @lubchenco2015]. We further contend that adaptive
+management requires the development of adaptive monitoring tools and
+infrastructures, which we seek to address through a continuously-evolving platform.
+
+***How***:
+By setting up mechanisms structuring cyclic reviews of platform content,
+for the integration of new material (*e.g.* data and methods) as it becomes
+available or accessible, and by striving to provide time-series data that are
+crucial to assess temporal trends and potentially early-warning signals of
+ecosystem change (Figure \ref{diag}).
+
+## Recognition
+
+***Why***: Like peer-reviewed publications, data must also be given its due
+importance in scientific endeavors and thus be considered as legitimate citable
+products contributing to the overall scientific output of data providers
+[@codata2013; @force112014]. Appropriate citations should
+therefore be provided for all data layers used and shared by the platform.
+
+***How***: By adhering to the Data Citation Principles [@force112014], which
+focus on citation practices that provide appropriate credit to data products.
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Perspectives
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+Understanding how ecosystem state will be affected by global change requires
+a comprehensive understanding of how threats are distributed and interact in
+space and time, which in turn hinges on appropriate data tailored to
+multi-driver studies [@dafforn2016; @stock2018; @bowler2019]. In the St.
+Lawrence, we found that few areas are free of cumulative exposure and that the
+whole Estuary, the Anticosti Gyre and coastal southwestern Gulf are particularly
+exposed to cumulative drivers, especially close to urban areas. We also
+identified six geographically distinct threat complexes that display similar
+cumulative exposure regimes. These complexes reveal that coastal areas are
+particularly exposed to all types of drivers and that multiple drivers
+typically co-occur in space. These resuts allow us to efficiently identify areas
+in need of heightened scrutiny from a science and management perspective.
+
+Through *eDrivers*, these observations will be iteratively improved towards an
+increasingly robust assessment of cumulative exposure and threat complexes as
+gaps in knowledge are addressed or approaches to describe drivers are refined.
+Arguably, the most meaningful benefit anticipated from *eDrivers* will be the
+gain in efficient access to comparable data-based knowledge on the exposure of
+ecosystems to multiple threats. This could pay quick scientific and management
+dividends by drawing on the knowledge and efforts of a wide range of
+contributors, by expanding avenues of scientific inquiry, by decreasing overall
+effort duplication and research costs, and by increasing research efficiency
+[@franzoni2014].
+
+Critically - and we emphasize this point - *eDrivers* will
+allow the scientific and governmental communities to identify key knowledge
+gaps that will assist in prioritizing and optimizing research efforts.
+Ultimately, we believe that *eDrivers* will operationalize evidence-based
+decision-making by streamlining data management and research, allowing science
+output to be available and interpretable on a time scale relevant to management
+[see @reichman2011]. The platform will thus greatly facilitate the application
+of broad scale, holistic research and management approaches such as
+ecosystem-based management, strategic environmental assessments and
+social-ecological metanetworks [*e.g.* @halpern2015; @jones2016; @dee2017].
+
+Significant effort is still needed to bring our vision to fruition. Foremost
+is to maintain our efforts to foster collaborations, develop platform
+content and identify key knowledge gaps. A fair and efficient organizational
+structure will be developed in order to manage *eDrivers* as a community and
+appropriate funding must be secured to continue building this community and
+ensure the long-term viability of the initiative, although the partnership with
+SLGO partly addresses this issue.
+
+Finally, terrestrial and coastal environments must be incorporated, as
+sources of stress within those habitats extend to the marine environments.
+Moreover, despite coastal areas being recognized as the most exposed to
+environmental threats, we continue to delineate
+terrestrial and marine realms, considering coastlines as an impermeable
+barrier. While there is a sensible rationale for this division, we must strive
+to eliminate it if we are to appropriately study and predict the
+impacts of global change [*e.g.* see @bowler2019].
+
+Despite the challenges and work ahead, we are hopeful that this initiative will
+be very successful. Ultimately, *eDrivers* represents a much needed solution to
+address important issues in data management that could radically shift broad
+scale research and management practices towards efficient, adaptive and holistic
+ecosystem-based management in the St. Lawrence and elsewhere in the world. All
+it requires to be successful is for the scientific and political communities to
+fully commit to open knowledge, adaptive monitoring and, most of all, an
+integrated vision of ecosystem management.
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Acknowledgements
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+We thank the Fond de Recherche Québécois Nature et Technologie (FRQNT) and the
+Natural Science and Engineering Council of Canada (CRSNG) for financial support.
+This project is also supported by Québec Océan, the Quebec Centre for
+Biodiversity Science (QCBS), Takuvik, and the Notre Golfe networks. This
+research is also sponsored by the NSERC Canadian Healthy Oceans Network and its
+Partners: Department of Fisheries and Oceans Canada and INREST (representing the
+Port of Sept-Îles and City of Sept-Îles).
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Author contributions statement
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+TO WRITE
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Conflict of interest statement
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+The authors declare that the submitted work was carried out in the absence of
+any personal, professional or financial relationships that could potentially be
+construed as a conflict of interest.
+
+\newpage
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Listings
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+Box 1. Code snippet demonstrating how to use the *eDrivers* to reproduce
+figure \ref{kernel} in R.
+
+\begin{lstlisting}[frame=single]
+# Install and load eDrivers package
+devtools::install_github('eDrivers/eDrivers')
+library(eDrivers)
+
+# Load data
+drivers <- fetchDrivers(drivers = c('hypoxia','fishDD'))
+
+# Get data from `eDrivers` class object
+driverData <- getData(drivers)
+
+# Normalize data
+driverData <- driverData / cellStats(driverData, 'max')
+
+# Visualize data and combination
+plot(driverData$fishDD)   # Demersal destructive fisheries
+plot(driverData$hypoxia)  # Hypoxia
+plot(sum(driverData))     # Combination
+
+# Identify values > 0 and not NAs
+driverData$fishDD[driverData$fishDD < 0] <- NA
+driverData$fishDNH[driverData$hypoxia < 0] <- NA
+id0 <- !is.na(values(driverData$fishDD)) &
+       !is.na(values(driverData$hypoxia))
+
+# 2D kernel for driver co-intensity
+library(MASS)
+coInt  <- kde2d(x = values(driverData$fishDD)[id0],
+                y = values(driverData$hypoxia)[id0],
+                n = 500, lims = c(0, 1, 0, 1))
+image(coInt, zlim = c(0,max(coInt$z)))
+
+# Driver density distribution
+plot(density(driverData$fishDD[id0]))   # Demersal destructive
+plot(density(driverData$hypoxia[id0]))  # Hypoxia
+\end{lstlisting}
+
+\newpage
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Figures
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/footprint.png}
+\caption{Distribution of driver footprint in the Estuary and Gulf of St.
+Lawrence.
+The footprint is measured by summing the relative intensity of all drivers for each grid cell: $F_d = \sum_{i=1}^{n} D_i$.
+Driver layers were also log-transformed when their frequency distribution was non-normal.
+All driver layers were also normalized between 0 and 1 using the 99th quantile to allow for direct comparison of relative intensities between drivers and control for extreme values.}
+\label{footprint}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/kernel.png}
+\caption{Example of platform content using the spatial distribution and the intensity of hypoxia and demersal destructive fisheries in the St.
+Lawrence.
+An index of hypoxia (**A**) was created using bottom-water dissolved oxygen between 2013 and 2017 [@blais2018].
+Demersal destructive fisheries (*i.e.* trawl and dredges) (**B**) intensity was evaluated from fisheries catch data collected between 2010 and 2015 used to measure annual area weighted total biomass ($kg$) in i $km^2$ grid cells [@dfo2016].
+See supplementary materials for more information on specific methodologies.
+Relative hypoxic stress and demersal destructive fisheries intensity was summed (**C**) to visualize their combined spatial distribution and intensity.
+Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a two-dimensional kernel analysis (**D**).}
+\label{kernel}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/cluster.png}
+\caption{Distribution of threat complexes in the Estuary and Gulf of St.
+Lawrence.
+Threat complexes [a term coined by @bowler2019] are areas with similar cumulative driver exposure regimes.
+Threat complexes were identified using a partitional clustering algorithm [@kaufman1990].
+The appropriate number of clusters ($k$) was tested using a range of values and validated by selecting the number of clusters that maximized the average silhouette width [@kaufman1990] and minimized the total within-cluster sum of squares (WSS).
+Six distinct threat complexes were identified in the St.
+Lawrence.
+The partitional clustering algorithm analysis were performed using the *cluster* R packages [@maechler2018].
+Refer to the Supplementary Materials for more details.}
+\label{cluster}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/structure.png}
+\caption{Diagram of the platform structure.
+Community input in the form of raw data is accessed through the St.
+Lawrence Global Observatory (SLGO); https://ogsl.ca/en) repository - the platform host - or through open access repositories (*e.g.* NASA data).
+The raw data are then processed through a workflow hosted on the *eDrivers* GitHub organization (https://github.com/orgs/eDrivers/).
+Data processing may be as simple as data rescaling (*e.g.* night lights) or make use of more complex methodologies (*e.g.* acidification).
+All data is then hosted on SLGO's geoserver and accessible through their API.
+We developed a R package called *eDrivers* to access the driver layers through R and we are actively developing a second R package called *eDriversEx* that includes analytical tools to explore drivers data.
+Finally, we have developed a Shiny application that allows users to explore drivers data interactively (https://david-beauchesne.shinyapps.io/eDriversApp/).
+All R components of the project are hosted and available on the *eDrivers* GitHub organization.}
+\label{diag}
+\end{figure}
+
+\newpage
+\blandscape
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# Tables
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+Table 1. List of drivers currently available on *eDrivers* and used for the analyses presented in this paper.
+
+```{r table1, echo=F, fontsize=4}
+load('./tables/table1.RData')
+table1
+```
+
+\elandscape
+\newpage
+
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+# References
+<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
+
+<!--
+To check:
+  - Kaufman, L., and Rousseeuw, P. (1990)
+-->

### 2019-04-24 eDriversMS-SI-v2.rmd
latest modifications including co-author comments and author contribution statement

@@
 The data used to characterize sea level rise risk comes from the global
-cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
-available on the NCEAS online data repository [@halpern2015b]. We provide a
+cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+available on the NCEAS online data repository [@halpern2015]. We provide a
 brief summary of data and methods in this document. For more details, refer to
@@
 
-For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
+For the St. Lawrence, we overlaid the raw data layers [@halpern2015] with our
 1 $km^2$ grid cell using weighted area average.
@@
 We used the annual Version 1 Nighttime VIIRS DNB composites between 2015 and
-2016 [@eog2019] to characterize coastal development in coastal areas of the
+2016 [@earthobservationgroup2019] to characterize coastal development in coastal areas of the
 St. Lawrence. As the effects of coastal development are likely acute in its
@@
 
-As in @halpern2008 and @halpern2015, we used the sum of coastal populations as
+As in @halpern2008b and @halpern2015, we used the sum of coastal populations as
 a proxy of direct human impact. We used Statistics Canada dissemination area
 population count from the 2016 census to obtain coastal population size
-around the St. Lawrence [@statscan2017]. Dissemination areas are the smallest
+around the St. Lawrence [@statisticscanada2017]. Dissemination areas are the smallest
 standard geographic area in which census data are disseminated and they combine
@@
 The data used to characterize inorganic pollution comes from the global
-cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
-available on the NCEAS online data repository [@halpern2015b]. We provide a
+cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+available on the NCEAS online data repository [@halpern2015]. We provide a
 brief summary of data and methods in this document. For more details, refer to
@@
 The data used to characterize nutrient pollution comes from the global
-cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
-available on the NCEAS online data repository [@halpern2015b]. We provide a
+cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+available on the NCEAS online data repository [@halpern2015]. We provide a
 brief summary of data and methods in this document. For more details, refer to
@@
 The data used to characterize organic pollution comes from the global
-cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
-available on the NCEAS online data repository [@halpern2015b]. We provide a
+cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+available on the NCEAS online data repository [@halpern2015]. We provide a
 brief summary of data and methods in this document. For more details, refer to
@@
 For example, traps and trawls have very different effects on a system. Gear
-classification was done using the classification presented in @halpern2008 and
-@halpern2015b and is broken down into 5 distinct classes: demersal destructive
+classification was done using the classification presented in @halpern2008b and
+@halpern2015 and is broken down into 5 distinct classes: demersal destructive
 (DD), demersal, non-destructive, low-bycatch (DNL), demersal, non-destructive,
@@
 The data used to characterize shipping comes from the global cumulative impacts
-assessment on habitats [@halpern2008; @halpern2015] and available on the NCEAS
-online data repository [@halpern2015b]. We provide a brief summary of data and
+assessment on habitats [@halpern2008b; @halpern2015a] and available on the NCEAS
+online data repository [@halpern2015]. We provide a brief summary of data and
 methods in this document. For more details, refer to @halpern2015.
@@
 gathered as part of the World Meteorological Organization Voluntary Observing
-Ships' (VOS) scheme [@jcomm2018]. Ships participating in the program gather
+Ships' (VOS) scheme. Ships participating in the program gather
 meteorological data along with observation location as part of an open-ocean
@@
 an initiative launched in 2002 that sought to improve marine safety by providing
-mariners with real-time vessel traffic [@tetreault2002]. Through the
+mariners with real-time vessel traffic. Through the
 International Maritime Organization SOLAS agreement, all vessels of over 300
@@
 The data used to characterize invasive species risk comes from the global
-cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
-available on the NCEAS online data repository [@halpern2015b]. We provide a
+cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+available on the NCEAS online data repository [@halpern2015]. We provide a
 brief summary of data and methods in this document. For more details, refer to
-supplementary materials provided in @halpern2008 and @halpern2015.
+supplementary materials provided in @halpern2008b and @halpern2015.
 
@@
 The data used to characterize marine pollution risk comes from the global
-cumulative impacts assessment on habitats [@halpern2008; @halpern2015] and
-available on the NCEAS online data repository [@halpern2015b].
+cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+available on the NCEAS online data repository [@halpern2015].
 Marine pollution was considered to be mainly driver by the shipping industry.
@@
 invasive. For more details, refer to supplementary materials provided in
-@halpern2008 and @halpern2015.
+@halpern2008b and @halpern2015.
 
-For the St. Lawrence, we overlaid the raw data layers [@halpern2015b] with our
+For the St. Lawrence, we overlaid the raw data layers [@halpern2015] with our
 1 $km^2$ grid cell using weighted area average.

### 2019-04-24 eDriversMS-v2.Rmd
latest modifications including co-author comments and author contribution statement

@@
 
+**Keywords**: ocean observing systems, St. Lawrence, environmental drivers,
+cumulative exposure, threat complex, multiple stressors, global change
+
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
@@
 highly variable environmental conditions and oceanographic processes, both in
-space and time [@el-sabh1990; @white1997; @dufour2007]. It thus offers a unique
+space and time [@el-sabh1990a; @white1997; @dufour2007]. It thus offers a unique
 and heterogeneous array of habitats suited for the establishment of diverse and
@@
 major investments recently made and more forthcoming in economic and
-infrastructure development and research [*e.g.* @goq2015; @rqm2018], an
+infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018], an
 intensification of the human footprint is expected in the St. Lawrence System.
 
-As elsewhere in the world [see @halpern2015], this intensifying human footprint
+As elsewhere in the world [see @halpern2015a], this intensifying human footprint
 will likely result in increasingly intricate environmental exposure regimes,
@@
 [GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
-and the World Register of Marine Species [WoRMS; @worms2017]. Abiotic data can
+and the World Register of Marine Species [WoRMS; @wormseditorialboard2017]. Abiotic data can
 also be accessed through WorldClim [@hijmans2005], Bio-ORACLE
@@
 The list of drivers for which we sought data was informed by a global cumulative
-impact assessment initiative [@halpern2008; @halpern2015], regional holistic
+impact assessment initiative [@halpern2008b; @halpern2015a], regional holistic
 evaluations of the state of the St. Lawrence [@dufour2007; @benoit2012], and
@@
 holders. We also use global data from the global cumulative impact assessment
-initiative [@halpern2008; @halpern2015] available from the National Center
+initiative [@halpern2008b; @halpern2015a] available from the National Center
 for Ecological Analysis and Synthesis (NCEAS) online data repository
-[Table 1; @halpern2015b]. We selected global data that were unavailable at the
+[Table 1; @halpern2015]. We selected global data that were unavailable at the
 regional scale and that were available at a resolution adequate for use at the
@@
 close to coastal cities and at the mouth of rivers draining highly populated
-areas [e.g. @halpern2015; @feist2016; @mach2017; @stock2018]. These are areas
+areas [e.g. @halpern2015a; @feist2016; @mach2017; @stock2018]. These are areas
 where human activities (*e.g.* coastal development and shipping) and footprint
@@
 provide access to and serve as the primary drainage outflow of the Great Lakes
-Basin, the most densely populated region in Canada [@statcan2017]. Most marine
+Basin, the most densely populated region in Canada [@statisticscanada2017]. Most marine
 traffic thus converges to the Estuary.
@@
 and Greenland halibut *Reinhardtius hippoglossoides*; @pillet2016] to reduced
-growth rates [@dupont2013] and avoidance of oxygen-depleted habitats
+growth rates [@dupont-prinet2013] and avoidance of oxygen-depleted habitats
 [*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality
@@
 port activities, and other human activities in Canadian waters, including the
-St. Lawrence system [@goc2016].
+St. Lawrence system [@governmentofcanada2018].
 
@@
 public safety, climate change, transportation, resources and biodiversity
-conservation. The SLGO is also one of three regional association spearheading
-the Canadian Integrated Ocean Observing System
-(CIOOS; http://meopar.ca/research/cioos-call-for-proposals/),
+conservation. SLGO is also one of three regional associations spearheading
+the Canadian Integrated Ocean Observing System (CIOOS; https://cioos.ca),
 which will focus on integrating oceanographic data from multiple sources to
@@
 External researchers or entities wishing to submit marine data will be able to
-do so through the SLGO web portal (https://ogsl.ca/en). Submissions through
+do so through SLGO web portal (https://ogsl.ca/en). Submissions through
 other data portals will also be accepted either through the development of data
@@
 
-***Why***: Open data has been propelled to the forefront of scientific research
+***Why***: Open knowledge has been propelled to the forefront of scientific research
 in an era of open, collaborative and reproducible science. By moving towards
@@
 [GOOS; http://www.goosocean.org]. As such, we will adopt the metadata standard
-currently targetted for the upcoming CIOOS, *i.e.* the North American Profile
+currently targetted for CIOOS, *i.e.* the North American Profile
 of ISO 19115:2014 - Geographic information - Metadata, a schema favoured
-for geospatial data in Canada and the United-States.
+for geospatial data in Canada and the United States.
 
@@
 strategy to guide efficient decision-making
-[*e.g.* @costanza1998; @margules2000; @keith2011; @jones2016] and has already
+[*e.g.* @margules2000; @keith2011; @jones2016; @chion2018] and has already
 been discussed in the context of multi-drivers and cumulative impact assessments
-[@halpern2015; @beauchesne2016; @cote2016; @schloss2017]. Adaptive management can only be
+[@halpern2015a; @beauchesne2016; @cote2016; @schloss2017]. Adaptive management can only be
 truly achieved through a commitment to adaptive monitoring and data reporting
@@
 output to be available and interpretable on a time scale relevant to management
-[see @reichman2011]. The platform will thus greatly facilitate the application
+[see @sutherland2004; @reichman2011]. The platform will thus greatly facilitate the application
 of broad scale, holistic research and management approaches such as
 ecosystem-based management, strategic environmental assessments and
-social-ecological metanetworks [*e.g.* @halpern2015; @jones2016; @dee2017].
+social-ecological metanetworks [*e.g.* @halpern2015a; @jones2016; @dee2017].
 
@@
 
-TO WRITE
+DB, RD, DG and PA conceived the manuscript and the underlying objective. DB
+prepared/formatted the data, performed the analyses, was in charge of technical
+developments and lead the drafting of the manuscript. All co-authors contributed
+to data, analases and writing based on their respective expertise and
+contributed to the revision of the manuscript.
 

### 2019-04-24 Code/4-footprint.R

@@ -39,5 +39,5 @@ legendEGSL(range = c(min(Fd), max(Fd)),
            n = 6)
 
 # City names
-cityEGSL()
+cityEGSL(prj = 4326)
 dev.off()

### 2019-04-24 Tables/datatable.R

@@ -12,27 +12,27 @@ df <- data.frame(Groups = character(0),
                  Accronym = character(0),
                  stringsAsFactors = F)
 
-df[1,] <- c("Climate", "Aragonite", 'Lat/long', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ARAG')
+df[1,] <- c("Climate", "Aragonite", 'Lat/long', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ACID')
 df[2,] <- c("Coastal", "Aquaculture", 'Lat/long', '-','Variable, between 1990-2016', '$presence-absence$','***TBD***','AQUA')
 df[3,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ negative anomalies', '[@galbraith2018]','SBT-')
 df[4,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ positie anomalies', '[@galbraith2018]','SBT+')
 df[5,] <- c("Climate",  "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ negative anomalies', '[@galbraith2018]','SST-')
 df[6,] <- c("Climate", "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ positive anomalies', '[@galbraith2018]','SST+')
-df[7,] <- c("Climate", "Sea level rise", 'Modeled 0.25 degree', '10 days', '1992-2012', '$mm$','[@halpern2015b]','SLR')
-df[8,] <- c("Marine traffic", "Shipping", '0.1 degree', 'Event based', '2003-2011', '$n$ lanes', '[@halpern2015b]','SHP')
+df[7,] <- c("Climate", "Sea level rise", 'Modeled 0.25 degree', '10 days', '1992-2012', '$mm$','[@halpern2015]','SLR')
+df[8,] <- c("Marine traffic", "Shipping", '0.1 degree', 'Event based', '2003-2011', '$n$ lanes', '[@halpern2015]','SHP')
[…]

### 2019-04-24 eDriversMS-v2.Rmd
urls as footnotes

@@
 environmental data for specific areas also exist, such as the St. Lawrence
-Global Observatory (SLGO; https://ogsl.ca/en), the European Marine Observation
-and Data Network (EMODnet; http://www.emodnet.eu/) and the U.S. Integrated Ocean
-Observing System (IOOS; https://ioos.noaa.gov/). Essential environmental
+Global Observatory (SLGO[^slgo]), the European Marine Observation
+and Data Network (EMODnet[^emod]) and the U.S. Integrated Ocean
+Observing System (IOOS[^ioos]). Essential environmental
 parameters are also organized, coordinated and acquired through global
@@
 Network [GEO BON; @scholes2012], the Census of Marine Life [CoML; @coml2010]
-and the Global Ocean Observing System (GOOS; http://www.goosocean.org).
+and the Global Ocean Observing System (GOOS[^goos]).
+
+[^slgo]: https://ogsl.ca/en
+[^emod]: http://www.emodnet.eu/
+[^ioos]: https://ioos.noaa.gov/
+[^goos]: http://www.goosocean.org
 
@@
 Global Observatory (SLGO) to develop the initiative and to host the
-platform on their web portal (https://ogsl.ca/en). The mission of SLGO is to
+platform on their web portal. The mission of SLGO is to
 promote and facilitate the accessibility, dissemination and exchange of official
@@
 conservation. SLGO is also one of three regional associations spearheading
-the Canadian Integrated Ocean Observing System (CIOOS; https://cioos.ca),
+the Canadian Integrated Ocean Observing System (CIOOS[^cioos]),
 which will focus on integrating oceanographic data from multiple sources to
@@
 identifying gaps or duplications in observations and research efforts.
-We are also developing collaborations with the Portal on water knowledge
-(http://www.environnement.gouv.qc.ca/eau/portail/), an initiative from the
-Québec provincial government. This portal aims at collecting and sharing
-accurate, complete and updated ressources on water and aquatic ecosystems to
-support the mandate of relevant actors and stakeholders working in
-water and aquatic ecosystems management.
+We are also developing collaborations with the Portal on water knowledge[^melcc],
+an initiative from the Québec provincial government. This portal aims at
+collecting and sharing accurate, complete and updated ressources on water and
+aquatic ecosystems to support the mandate of relevant actors and stakeholders
+working in water and aquatic ecosystems management.
+
+[^cioos]: https://cioos.ca
+[^melcc]: http://www.environnement.gouv.qc.ca/eau/portail/
 
@@
 External researchers or entities wishing to submit marine data will be able to
-do so through SLGO web portal (https://ogsl.ca/en). Submissions through
+do so through SLGO web portal. Submissions through
 other data portals will also be accepted either through the development of data
@@
 By making data and associated tools accessible through a variety of ways:
-the SLGO web portal (https://ogsl.ca/en), two R packages called *eDrivers*
-(https://github.com/orgs/eDrivers/eDrivers) and *eDriversEx*
-(https://github.com/orgs/eDrivers/eDriversEx) to access the data through SLGO's
-API and to provide analytical tools to explore data, respectively, and a Shiny
-application (https://david-beauchesne.shinyapps.io/eDriversApp/) to explore
-drivers data interactively (Figure \ref{diag}).
-Note that the data are currently contained within and accessible through the
-*eDrivers* R package only, as we are actively working to allow users to
-download selected layers from SLGO's web portal and geoserver. The functions
-available in *eDrivers* to access the data have however been developed to
-ensure forward compatibility once the data migrate to SLGO's geoserver.
+the SLGO web portal, two R packages called *eDrivers*[^eDR] and *eDriversEx*[^eDX]
+to access the data through SLGO's API and to provide analytical tools to explore
+data, respectively, and a Shiny application[^eDapp] to explore drivers data
+interactively (Figure \ref{diag}). Note that the data are currently contained
+within and accessible through the *eDrivers* R package only, as we are actively
+working to allow users to download selected layers from SLGO's web portal and
+geoserver. The functions available in *eDrivers* to access the data have however
+been developed to ensure forward compatibility once the data migrate to SLGO's
+geoserver.
+
+[^eDR]: https://github.com/orgs/eDrivers/eDrivers
+[^eDX]: https://github.com/orgs/eDrivers/eDriversEx
+[^eDapp]: https://david-beauchesne.shinyapps.io/eDriversApp/
 
@@
 Essential Ocean Variables (EOV) identified by the Global Ocean Observing System
-[GOOS; http://www.goosocean.org]. As such, we will adopt the metadata standard
+(GOOS). As such, we will adopt the metadata standard
 currently targetted for CIOOS, *i.e.* the North American Profile
@@
 driver layers from raw data, the R packages and the Shiny application through
-a GitHub organization called *eDrivers* (https://github.com/orgs/eDrivers/).
+a GitHub organization called *eDrivers*[^eD].
+
+[^eD]: https://github.com/orgs/eDrivers/
 

### 2019-04-25 eDriversMS-v2.Rmd
super minor comments on MS

@@
 as ecosystem-based management. While portals providing knowledge on individual
-drivers exist, platforms collating comparable and interoperable knowledges on
+drivers exist, platforms collating comparable and interoperable knowledges
+<!-- this is the first time I've ever seen the plural of knowledge... technically correct, but seems weird!--> on
 multiple drivers remain conspicuously missing. This paper presents two distinct,
@@
 aquaculture production, fosters emerging activities and boasts a yet untapped
-hydrocarbon potential [@beauchesne2016; @schloss2017 @archambault2017]. With
+hydrocarbon potential [@beauchesne2016; @schloss2017 @archambault2017].<!-- @archambault doesn't render? --> With
 major investments recently made and more forthcoming in economic and
@@
 generally highest along the coast (Figure \ref{footprint}), with hotposts
-located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
+located in the vicinity of coastal cities (Figure \ref{hotspot})<!-- this figure isn't loaded at the bottom -->. In general,
 offshore areas are less exposed to cumulative drivers, with the Estuary and the

### 2019-04-25 eDriversMS-v2.Rmd
begin adjusting references

--- 2019-04-24 eDriversMS-v2.Rmd
@@
 fontsize: 12pt
+font-family: Times New Roman
 output:
@@
    - \usepackage{bbm}
-   - \usepackage{setspace}
-   - \onehalfspacing
    - \usepackage{float}
@@
 David Beauchesne \newline
-david.beauchesne@uqar.com \newline
+david.beauchesne@uqar.ca \newline
 

### 2019-04-28 eDriversMS-v2.Rmd
address comments and change Bray-Curtis dissimilarity for Manhattan distance

@@
 as ecosystem-based management. While portals providing knowledge on individual
-drivers exist, platforms collating comparable and interoperable knowledges
-<!-- this is the first time I've ever seen the plural of knowledge... technically correct, but seems weird!--> on
+drivers exist, platforms collating comparable and interoperable knowledge on
 multiple drivers remain conspicuously missing. This paper presents two distinct,
@@
 *eDrivers*, an open knowledge platform gathering experts committed to
-structuring, standardizing and sharing knowledges on drivers in support of
+structuring, standardizing and sharing knowledge on drivers in support of
 science and management. We gathered data on 22 coastal, climate, fisheries and
@@
 
+<!-- AQUACULTURE:
+QC: MAPAQ (2016) Description des sites maricoles de la province de Québec.
+Sous-ministériat aux pêches et à l’aquaculture commerciales, Direction régionale des Îles-de-la-Madeleine.
+Ministère de l'Agriculture, des Pêcheries et de l'Alimentation du Québec.
+Québec, Canada
+PEI: DFO (2016) PEI Aquaculture Leases.
+Aquaculture Division.
+Department of Fisheries and Oceans Canada.
+Prince-Edward-Island, Canada.
+Available at: http://www.arcgis.com/home/item.html?id=16aa8830c7084a8a92ce066b525978b4
+NB: Department of Agriculture, Aquaculture and Fisheries (2016) Marine Aquaculture Site Mapping Program (MASMP).
+Government of New-Brunswick, New-Brunswick, Canada.
+Available at: https://www2.gnb.ca/content/gnb/en/departments/10/aquaculture/content/masmp.html
+NS: Department of Fisheries and Aquaculture (2016) Aquaculture Site Mapping Tool.
+Department of Agriculture, Government of Nova-Scotia.
+Nova-Scotia, Canada.
+Available at: https://novascotia.ca/fish/aquaculture/site-mapping-tool/
+NL: Department of Fisheries, Forestry and Agrifoods (2016) Description of aquaculture sites in Newfoundland. Government of Newfoundland and Labrador. Newfoundland, Canada.
+-->
+
 The list of drivers for which we sought data was informed by a global cumulative
@@
 medoid to select the optimal set of $k$ medoids that minimizes average
-dissimilarity. We used 100 iterations using samples of 10000 observations
+dissimilarity. We used the clustering algorithm with the manhattan distance
+since this measure is less affected by extreme values [@ref]. <!-- find reference, legendre et legendre? -->
+We used 100 iterations using samples of 10000 observations
 (*i.e.* ~5% of observations) to identify clusters. Analyses were performed
@@
 dissimilarity and the contribution of each driver to the total inter-cluster
-dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis
-dissimilarity [Figure S4; @clarke1993]. As the drivers dataset is too
-large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting
+dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan
+distance [Figure S4]. The Manhattan distance was again preferred for continuity
+with the clustering analysis and to ensure that outliers did not have a strong
+influence the analysis. As the drivers dataset is large (~ 250000 observations),
+we used a bootstrap procedure for the SIMPER analysis, randomly selecting
 5% of each cluster to run the analysis and repeating the process over 300
 iterations. We also compared the mean intensity of each driver within each
-cluster to better capture the inter-cluster dissimilarity. Analyses were
-performed using the *vegan* R package [@oksanen2018].
+cluster to better capture the inter-cluster dissimilarity.
 
@@
 
-Intra-cluster similarity was evaluated using the Bray-Curtis similarity index
-(Figure S5). As with the inter-cluster dissimilarity, we used a
-bootstrap procedure for the intra-cluster similarity, randomly selecting 5% of
-each cluster observation to run the analysis and repeating the process over 300
-iterations. We however did not use the bootstraping procedure for clusters with
-less than 10000 observations since computation time was manageable.
+Intra-cluster similarity was evaluated calculating the intra-cluster Manhattan
+distance and by transforming the mean contribution to distance ($M_c$) of each
+driver by $.1 / (.1 + M_c)$ to obtain a similarity measure for each driver
+($S_d$). The total similarity is the sum of all $S_d$. (Figure S5).
+As with the inter-cluster dissimilarity, we used a bootstrap procedure for the
+intra-cluster similarity, randomly selecting 25% of each cluster observation to
+run the analysis and repeating the process over 50 iterations. We only used the
+bootstraping procedure for clusters with less than 40000 observations since
+computation time was manageable.
 
@@
 generally highest along the coast (Figure \ref{footprint}), with hotposts
-located in the vicinity of coastal cities (Figure \ref{hotspot})<!-- this figure isn't loaded at the bottom -->. In general,
+located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
 offshore areas are less exposed to cumulative drivers, with the Estuary and the
@@
 but rather serve as a stepping stone towards an adaptive and ever-improving
-collection of knowledges.
+collection of knowledge.
 
@@
 \includegraphics{./Figures/footprint.png}
-\caption{Distribution of driver footprint in the Estuary and Gulf of St.
+\caption{Distribution of cumulative footprint in the Estuary and Gulf of St.
 Lawrence.
@@
 \label{footprint}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{./Figures/hotspot.png}
+\caption{Distribution of cumulative hotspots in the Estuary and Gulf of St.
+Lawrence.
+Hotspots are defined as the number of drivers in each grid cell with scaled intensity contained over their respective 80th percentile.}
+\label{hotspot}
 \end{figure}

### 2019-04-28 Code/14-interDissimilarity.R

@@ -9,28 +9,27 @@ load('./data/clMed.RData')
 
 # ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
 source('./code/bootstrapSimper.R')
+source('./code/simperMan.R')
 
 
 # ~~~~~~~~~~~~~~~~~~~ SIMPER ~~~~~~~~~~~~~~~~~~~ #
 # Run iterative simper analysis
 iter <- 300
 prop <- .05
-bootSimper <- bootstrapSimper(dr, clMed, iter = iter, prop = prop)
+system.time(bootSimper <- bootstrapSimper(dr, clMed, iter = iter, prop = prop))
 
 # # This is to run it multiple times simultaneously
 # bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '1')
 # bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '2')
 # bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '3')
-# load('./data/bootSimper1.RData')
-# x1 <- bootSimper
-# load('./data/bootSimper2.RData')
-# x2 <- bootSimper
-# load('./data/bootSimper3.RData')
-# x3 <- bootSimper
-# bootSimper <- c(x1,x2,x3)
-# id <- which(unlist(lapply(bootSimper, is.null)))
-# for(i in rev(id)) bootSimper[[i]] <- NULL
-# save(bootSimper, file = './data/bootSimper.RData')
+load('./data/bootSimper1.RData')
+x1 <- bootSimper
+load('./data/bootSimper2.RData')
+x2 <- bootSimper
+load('./data/bootSimper3.RData')
+x3 <- bootSimper
+bootSimper <- c(x1,x2,x3)
+save(bootSimper, file = './data/bootSimper.RData')
 
 
 # ~~~~~~~~~~~~~~~~~~~ SIMPER SUMMARY ~~~~~~~~~~~~~~~~~~~ #
@@ -54,7 +53,7 @@ for(i in 1:nList) {
 
 # Extract summary for simper analyses interations
 for(i in 1:iter) {
-  temp <- summary(bootSimper[[i]], ordered = F)
+  temp <- bootSimper[[i]]
   for(j in 1:nList) {
     simperSummary[[j]][,,i] <- as.matrix(temp[[j]][, varNames])
   }

### 2019-04-28 Code/15-intraSimilarity.R

@@ -8,7 +8,7 @@ load('./data/clMed.RData')
 
 
 # ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
-source('./code/intraSimilarity.R')
+source('./code/intraSim.R')
 
 # ~~~~~~~~~~~~~~~~~~~ INTRA-CLUSTER SIMILARITY ~~~~~~~~~~~~~~~~~~~ #
 # As with the simper analysis, the dataset is mostly too large to use
@@ -23,14 +23,14 @@ sim <- vector('list', k)
 for(i in 1:k) {
   cat("   k: ", i, "\r")
   id <- clMed == i
-  sim[[i]] <- bootIntraSim(dr[id, ], iter = 50, samp = 1200)
+  sim[[i]] <- bootIntraSim(dr[id, ], iter = 50, samp = .25)
   save(sim, file = './data/intraSimilarity.RData')
 }
 
-# Extract species, average and sd
+# Extract species, contribution and sd
 # Summarize in array format
-iter <- length(sim[[1]])
-varNames <- c('average')
+iter <- length(sim[[4]])
+varNames <- c('contr')
 simSummary <- vector('list', k)
 for(i in 1:k) simSummary[[i]] <- array(data = 0, dim = c(ncol(dr), 1, iter), dimnames = list(drNames$accr, varNames))
 
@@ -38,13 +38,13 @@ for(i in 1:k) {
   # Check if there are multiple iterations or not.
   # If length(sim[[i]]) == 2 it means that there were no iterations and the whole cluster was parsed through the simper2 function
   # If not, then sim[[i]] should be equal to the number of iterations
-  if(length(sim[[i]]) == 2) {
+  if(i %in% c(1,2,3)) {
     simSummary[[i]] <- matrix(ncol = 1,
-                              data = sim[[i]]$similarityContribution$average,
-                              dimnames = list(drNames$accr, 'average'))
+                              data = sim[[i]]$contr,
+                              dimnames = list(drNames$accr, 'contr'))
   } else {
     for(j in 1:iter) {
-      simSummary[[i]][,'average',j] <- sim[[i]][[j]]$similarityContribution$average
+      simSummary[[i]][,'contr',j] <- sim[[i]][[j]]$contr
     }
   }
 }

### 2019-04-28 Code/17-contributionSimilarity.R

@@ -20,12 +20,12 @@ source('./code/circularPlot.R')
 
 # ~~~~~~~~~~~~~~~~~~~ SIMILARITY CONTRIBUTION ~~~~~~~~~~~~~~~~~~~ #
 # Driver percent contribution to dissimilarity
-contSim <- lapply(similarityMean, function(x) x[,'average'] / sum(x[,'average'])) %>%
+contSim <- lapply(similarityMean, function(x) x[,'contr'] / sum(x[,'contr'])) %>%
             as.data.frame()
 colnames(contSim) <- names(similarityMean)
 
 # Total similarity
-totSim <- lapply(similarityMean, function(x) sum(x[,'average'])) %>%
+totSim <- lapply(similarityMean, function(x) sum(x[,'contr'])) %>%
           as.data.frame()
 
 # Circulat plots

### 2019-04-28 Code/bootstrapSimper.R

@@ -19,7 +19,7 @@ bootstrapSimper <- function(x, cl, iter = 100, prop = .1, accr = '') {
     samp <- unlist(samp)
 
     # Run simper analysis on subset of data
-    bootSimper[[i]] <- simper(x[samp, ], cl[samp])
+    bootSimper[[i]] <- simperMan(x[samp, ], cl[samp])
 
     # Save file
     save(bootSimper, file = paste0('./data/bootSimper', accr, '.RData'))

### 2019-05-01 eDriversMS-SI-v2.rmd
latest modifications to manuscript, including frontiers formatting and supervisor comments

@@
 rmarkdown::render('./eDriversMS-SI-v2.rmd', 'pdf_document')
-rmarkdown::render('./eDriversMS-SI.rmd', 'html_document')
+rmarkdown::render('./eDriversMS-SI-v2.rmd', 'word_document')
 -->
@@
 
-Table S1. List of drivers currently available on *eDrivers* along with their respective accronym used in the figures in the supplementary material.
+Table S1. List of drivers currently available on *eDrivers* along with their respective acronym used in the figures in the supplementary material.
 
@@
 
-Oceans are the largest reservoir and sink of atmospheric carbon dioxide ($CO_2$). Its uptake
-increases seawater acidity and the lowers saturation state with respect to calcite
-($\Omega_C$) and aragonite ($\Omega_A$), the two most common $CaCO_3$ polymorphs
-constituting the shells and skeleton of many marine organisms [@mucci2017].
-This can have deleterious effects on carbonate-secreting organisms (*e.g.* mollusks and crustaceans)
-and certain physiological processes in non-calcifying organisms
-[@fabry2008; @kroeker2013].
+Oceans are the largest reservoir and sink of atmospheric carbon dioxide ($CO_2$).
+Its uptake increases seawater acidity and the lowers saturation state with
+respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common
+$CaCO_3$ polymorphs constituting the shells and skeleton of many marine
+organisms [@mucci2017]. This can have deleterious effects on carbonate-secreting
+organisms (*e.g.* mollusks and crustaceans) and certain physiological processes
+in non-calcifying organisms [@fabry2008; @kroeker2013].
 
-When $\Omega_C$ or $\Omega_A$ decrease below 1, water
-becomes undersaturated and corrosive to the skeletal minerals of carbonate-secreting
-organisms. Note that if $\Omega_A$ $< 1$, the waters
-may still be supersaturated with respect to calcite since it is 50% more soluble
-than aragonite. However, organisms
-have wide ranging responses to $\Omega_C$ and $\Omega_A$ saturation state.
-For example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019], while other organisms
-may have adapted to precipitate $CaCO_3$ even when seawater is undersaturated
-with respect to calcite [$\Omega_C$ $< 1$ or $\Omega_A$ $< 0.65$; e.g. @uthicke2016].
+When $\Omega_C$ or $\Omega_A$ decrease below 1, water becomes undersaturated and
+corrosive to the skeletal minerals of carbonate-secreting organisms. Note that
+if $\Omega_A$ $< 1$, the waters may still be supersaturated with respect to
+calcite since it is 50% more soluble than aragonite. However, organisms have
+wide ranging responses to $\Omega_C$ and $\Omega_A$ saturation state. For
+example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019],
+while other organisms may have adapted to precipitate $CaCO_3$ even when
+seawater is undersaturated with respect to calcite [$\Omega_C$ $< 1$ or
+$\Omega_A$ $< 0.65$; e.g. @uthicke2016].
 
 We used $\Omega_{A}$ saturation state to characterize ocean acidification in the
-bottom waters of the St. Lawrence. The data were collected [...]. $\Omega_{A}$,
-pH, dissolved oxygen ($O_2$) and pressure were sampled at 117 stations in the
-Estuary and Gulf of St. Lawrence in the summer of 2018. The majority of sampling
-stations were located in the Southern Gulf ($n = 84$), compared to the Estuary
-and Northern Gulf ($n = 33$).
+bottom waters of the St. Lawrence. The data come from the Department of
+Fisheries and Oceans' (DFO) Atlantic Zone Monitoring Program [AZMP; @galbraith2018]
+surveys and DFO's multispecies surveys [@bourdages2018] collected in
+August-September of 2017 [@starr2019]. The carbonate chemistry was determined
+through pH and TA measurements. Samples for pH and total alkalinity (TA) were
+collected under a bubble free and no head space conditions into 500 $mL$
+borosilicate glass flasks, 250 $\mu L$ of the saturated $HgCl_2$ solution was
+added to sample, and processed following the "Guide to best practices for Ocean
+CO2 Measurements" [@dickson2007].
 
-To account for low sample size in the Northern Gulf, we used the
-correlation between $O_2$ and $CO_2$, which are linked through the stoichiometry
-of the respiration reaction in waters that are isolated from the atmosphere [@mucci2011].
-Metabolic $CO_2$ increases in
-water through biotic processes such microbial respiration of organic matter
-[@mucci2011]. In deep waters, variations of $\Omega_{A}$ are largely governed by the addition of metabolic
-$CO_2$, whereas near the air-sea interface $O_2$ is replenished faster than $CO_2$ can
-escape [@zeebe2001]. The impact of respiration on carbonate chemistry is highest in hypoxic
-regions, where metabolic $CO_2$ accumulates to high concentrations [@mucci2011].
+$pH_T$ was determined spectrophotometrically using the indicator dye m-cresol
+purple (Sigma-Aldrich). Absorbance was measured at 730, 578 and 434 nm before
+and after dye addition in 10 cm quartz cells thermostated at 25 ± $0.05^oC$
+[@dickson2007]. A similar procedure was carried out before each set of sample
+measurements using a TRIS (Tris (hydroxymethyl) -aminomethane) buffer prepared
+at a practical salinity (S) of approximately 30 [@millero1986]. Certified
+Reference Material (CRM) (supplied by Professor Andrew Dickson, Scripps
+Institution of Oceanography, San Diego, USA) was used for quality control of
+our pH TRIS buffer.
 
-We explored the correlation between $\Omega_{A}$ and $O_2$ in the data and
-found no significant correlation at the scale of the whole St. Lawrence (*p-value*:
-0.08; $R^2$: 0.03). Nevertheless, a correlation
-between $\Omega_A$ and $O_2$ exists in the Northern St. Lawrence (*p-value*:
-< 0.01; $R^2$: 0.49), but not in the Southern St. Lawrence (*p-value*: 0.85;
-$R^2$: < 0.01). Furthermore, the correlation observed in the north between $\Omega_A$
-and $O_2$ is especially strong when deep Channels (*Deep* model: *p-value*:
-< 0.01; $O_2$: 0.20; $R^2$: 0.97) are separated from shallower regions
-(*Shallow* model: *p-value*: < 0.01; $O_2$: 0.07; $R^2$: 0.96), with $\Omega_A$
-increasing more steeply with increasing $O_2$ in deep Channels. This stronger
-correlation is likely caused by the lack of gas exchange with the atmosphere
-or the surface mixed layer and the importance of metabolic $CO_2$ in deep
-waters, some of which may have mean ages of up to 20 years [@mucci2011]. In the
-Northern Gulf, the acidification is following
-depth and distance from the Cabot Strait along the Laurentian, Anticosti and
-Esquiman Channels due to the progressive oxygen depletion and metabolic carbon
-dioxide accumulation [@mucci2011]. There are likely different dynamics at play
-in the Southern Gulf, such as the accumulation of metabolic carbon dioxide
-produced by rapid oxic degradation of organic matter in the sandy and permeable
-sediments of the Magdalen Shelf [*personal communication?*]. <!-- see with Michel -->
+TA was determined by potentiometric titration in an open cell using an
+automated Radiometer potentiometric titrator (Titrilab 865) and a pH combination
+electrode (pHC2001) in a continuous titrant addition mode, an algorithm
+specifically designed for shallow end-point detection [@dickson2007]. The dilute
+HCl titrant 0.1M in a solution of NaCl of 0.6M was calibrated using CRM provided
+by Professor Andrew G. Dickson.
+
+The carbonate system parameters (including $\Omega_{A}$) were
+corrected for in situ pressure and temperature using the algorithm CO2SYS
+[@lewis1998] with measured pH, total alkalinity, soluble reactive
+and silicate concentrations as input parameters.
+
+$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
+stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
+majority of sampling stations were located in the Southern Gulf ($n = 84$),
+compared to the Estuary and Northern Gulf ($n = 33$).
+
+To account for low sample size in the Northern Gulf, we used the correlation
+between $O_2$ and $CO_2$, which are linked through the stoichiometry of the
+respiration reaction in waters that are isolated from the atmosphere [@mucci2011].
+Metabolic $CO_2$ increases in water through biotic processes such microbial
+respiration of organic matter [@mucci2011]. In deep waters, variations of
+$\Omega_{A}$ are largely governed by the addition of metabolic $CO_2$, whereas
+near the air-sea interface $O_2$ is replenished faster than $CO_2$ can escape
+[@zeebe2001]. The impact of respiration on carbonate chemistry is highest in
+hypoxic regions, where metabolic $CO_2$ accumulates to high concentrations
+[@mucci2011].
+
+We explored the correlation between $\Omega_{A}$ and $O_2$ in the data and found
+no significant correlation at the scale of the whole St. Lawrence (*p-value*:
+0.08; $R^2$: 0.03). Nevertheless, a correlation between $\Omega_A$ and $O_2$
+exists in the Northern St. Lawrence (*p-value*: < 0.01; $R^2$: 0.49), but not
+in the Southern St. Lawrence (*p-value*: 0.85; $R^2$: < 0.01). Furthermore, the
+correlation observed in the north between $\Omega_A$ and $O_2$ is especially
+strong when deep Channels (*Deep* model: *p-value*: < 0.01; $O_2$: 0.20; $R^2$:
+0.97) are separated from shallower regions (*Shallow* model: *p-value*: < 0.01;
+$O_2$: 0.07; $R^2$: 0.96), with $\Omega_A$ increasing more steeply with
+increasing $O_2$ in deep Channels. This stronger correlation is likely caused
+by the lack of gas exchange with the atmosphere or the surface mixed layer and
+the importance of metabolic $CO_2$ in deep waters, some of which may have mean
+ages of up to 20 years [@mucci2011]. In the Northern Gulf, the acidification is
+following depth and distance from the Cabot Strait along the Laurentian,
+Anticosti and Esquiman Channels due to the progressive oxygen depletion and
+metabolic carbon dioxide accumulation [@mucci2011].
+
+There are likely different dynamics at play in the Southern Gulf, such as
+waters with lowered alkalinity and salinity caused by the St. Lawrence plume.
+Such water masses are have lower buffering capacities to decreasing pH and
+$\Omega_{A}$. Also, the accumulation of metabolic carbon dioxide produced by
+rapid oxic degradation of organic matter in the sandy
+and permeable sediments of the Magdalen Shelf or through the St. Lawrence plume
+could decrease $\Omega_{A}$ in the Southern Gulf (Starr and Chassé, *personal
+communication*).
 
@@
 slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé
-Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modelled using the exponential
-kriging model.
+Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modelled
+using the exponential kriging model.
 
@@
 collected between 2013 and 2017
-[@blais2018] and interpolated oxygen % saturation using cokriging with depth as a
+[@blais2019] and interpolated oxygen % saturation using cokriging with depth as a
 covariable, as done in [@dutil2011]. Bathymetric data come from @dutil2011
@@
 Shallows in September. We provide a brief summary of data and methods to
-characterize hypoxia in this document. For more details, refer to @blais2018.
+describe hypoxia in this document. For more details, refer to @blais2019.
 
@@
 samples that are analyzed by Winkler titration on every cast
-[see @blais2018 for calibration procedure]. The data used is the last depth
+[see @blais2019 for calibration procedure]. The data used is the last depth
 sampled on the CTD profile, which typically corresponds to ~10m above the
@@
 that behavioural effects began manifesting below 70% oxygen saturation and that
-survival becomes jeopardised below ~20% saturation. We used these observations
+survival becomes jeopardized below ~20% saturation. We used these observations
 to create an index of hypoxia using an inverted logistic curve that slowly
@@
 images obtained from the National Oceanic and Atmospheric Administration (NOAA)
-and European Organisation for the Exploitation of Meteorological Satellites
+and European Organization for the Exploitation of Meteorological Satellites
 (EUMETSAT). Images used are from DFO’s Maurice Lamontagne Institute at a 1 km
@@
 The data used to characterize sea level rise risk comes from the global
-cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
 available on the NCEAS online data repository [@halpern2015]. We provide a
@@
 to gather the data on aquaculture sites from the 5 provinces dividing the
-St. Lawrence.
+St. Lawrence [@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016].
 
@@
 Atmospheric Administration's (NOAA) National Centers for Environmental
-Information (NCEI). They use glogally available nighttime data obtained from
+Information (NCEI). They use globally available nighttime data obtained from
 the Visible Infrared Imaging Radiometer Suite (VIIRS) Day/Night Band (DNB) of
@@
 
-As in @halpern2008b and @halpern2015, we used the sum of coastal populations as
+As in @halpern2008a and @halpern2015, we used the sum of coastal populations as
 a proxy of direct human impact. We used Statistics Canada dissemination area
 population count from the 2016 census to obtain coastal population size
-around the St. Lawrence [@statisticscanada2017]. Dissemination areas are the smallest
+around the St. Lawrence [@statistics-canada2017]. Dissemination areas are the smallest
 standard geographic area in which census data are disseminated and they combine
@@
 The data used to characterize inorganic pollution comes from the global
-cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
 available on the NCEAS online data repository [@halpern2015]. We provide a
@@
 The data used to characterize nutrient pollution comes from the global
-cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
 available on the NCEAS online data repository [@halpern2015]. We provide a
@@
 The data used to characterize organic pollution comes from the global
-cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
 available on the NCEAS online data repository [@halpern2015]. We provide a
@@
 methodology as that used to characterized nutrient pollution was then applied
-to organic polllution.
+to organic pollution.
 
@@
 The information provided on this expert map on the 5 toxins [@bates2019].
-was georeferenced and transformed as vecctorized objects.
+was georeferenced and transformed as vectorized objects.
 We calculated a toxic algae risk ($T$) index for each cell ($x$) in the 1 $km^2$
@@
 218323 fishing events (36387 $\pm$ 3147 fishing events per year). There were 31
-targetted species and a total of 53 caught species in the dataset.
+targeted species and a total of 53 caught species in the dataset.
 
@@
 For example, traps and trawls have very different effects on a system. Gear
-classification was done using the classification presented in @halpern2008b and
+classification was done using the classification presented in @halpern2008a and
 @halpern2015 and is broken down into 5 distinct classes: demersal destructive
@@
 In order to characterize the intensity of fishing activities ($FI$), we used
-a biomass yield density index. We multipled the total annual biomass captured
+a biomass yield density index. We multiplied the total annual biomass captured
 in each grid cell *j*, regardless of species, by the proportion of fishing area
@@
 The data used to characterize shipping comes from the global cumulative impacts
-assessment on habitats [@halpern2008b; @halpern2015a] and available on the NCEAS
+assessment on habitats [@halpern2008a; @halpern2015a] and available on the NCEAS
 online data repository [@halpern2015]. We provide a brief summary of data and
@@
 climate dataset. The data spans 20 years and annually covers 10-20% of ships
-worlwide. Data used spanned 2003 to 2011.
+worldwide. Data used spanned 2003 to 2011.
 
@@
 The data used to characterize invasive species risk comes from the global
-cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
 available on the NCEAS online data repository [@halpern2015]. We provide a
 brief summary of data and methods in this document. For more details, refer to
-supplementary materials provided in @halpern2008b and @halpern2015.
+supplementary materials provided in @halpern2008a and @halpern2015.
 
@@
 The plume model was then clipped to areas less than 60 m deep, as
-invasives species are more likely to invade shallow areas.
+invasive species are more likely to invade shallow areas.
 
@@
 The data used to characterize marine pollution risk comes from the global
-cumulative impacts assessment on habitats [@halpern2008b; @halpern2015a] and
+cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
 available on the NCEAS online data repository [@halpern2015].
@@
 As such, the driver layer was constructed by combining the shipping (*i.e.*
-shipping lanes) and invasives species (*i.e.* cargo volume) layers.
+shipping lanes) and invasive species (*i.e.* cargo volume) layers.
 invasive. For more details, refer to supplementary materials provided in
-@halpern2008b and @halpern2015.
+@halpern2008a and @halpern2015.
 
@@
 - Organic pollution
-- Sea bottom temperature anomalies <!-- verify -->
+- Sea bottom temperature anomalies
 - Shipping
@@
 \includegraphics{./Figures/marimekko.png}
-\caption{Frequency distribution of cumulative exposure (*i.e.* sum of normalized driver intensity in each grid cell) and percent contribution of each driver to the frequency distribution of cumulative exposure in the Estuary and Gulf of St.
+\caption{Frequency distribution of cumulative exposure (\textit{i.e.} sum of normalized driver intensity in each grid cell) and percent contribution of each driver to the frequency distribution of cumulative exposure in the Estuary and Gulf of St.
 Lawrence.}
@@
 \includegraphics{./Figures/validation.png}
-\caption{Validation procedure for the *k-medoids* and *k-means* clustering algorithms based on the number of cluster that maximizes average silhouette width [upper panels; @kaufman1990] and minimizes the total within-cluster sum of squares (WSS; lower panels).}
+\caption{Validation procedure for the \textit{k-medoids} and \textit{k-means} clustering algorithms based on the number of cluster that maximizes average silhouette width (upper panels; Kaufman and Rousseeuw, 1990) and minimizes the total within-cluster sum of squares (WSS; lower panels).}
 \label{valid}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics{./Figures/clusterComp.png}
-\caption{Comparison of the distribution of the 6 clusters identified using the *k-medoids* and *k-means* clustering algoriths and the percent grid cell correspondance between the two approaches.}
-\label{comparison}
 \end{figure}
@@
 \includegraphics{./Figures/interDissimilarity.png}
-\caption{Evaluation of inter-cluster dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis dissimilarity [@clarke1993].
-The figure diagonal presents the distribution of the 6 clusters identified using the *k-medoids* clustering algorithm.
+\caption{Evaluation of inter-cluster dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan distance (Clarke, 1993).
+The figure diagonal presents the distribution of the 6 clusters identified using the \textit{k-medoids} clustering algorithm.
 The lower triangle shows all combinations of inter-cluster dissimilarity with circular barplots showing the percent contribution to total dissimilarity of each driver and with the total inter-cluster dissimilarity in the center of the barplots.
@@
 \includegraphics{./Figures/intraSimilarity.png}
-\caption{Evaluation of intra-cluster similarity using the Bray-Curtis similarity index.
-The distribution of all 7 clusters is presented along with circular barplots showing the percent contribution to total similarity of each driver and with the total intra-cluster similarity in the center of the barplots.}
+\caption{Evaluation of intra-cluster similarity using the Manhattan distance transformed to a similarity index.
+The distribution of the 6 clusters is presented along with circular barplots showing the percent contribution to total similarity of each driver and with the total intra-cluster similarity in the center of the barplots.}
 \label{intra}

### 2019-05-01 eDriversMS-v2.Rmd
latest modifications to manuscript, including frontiers formatting and supervisor comments

--- 2019-04-25 eDriversMS-v2.Rmd
@@
 ---
-title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
+title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System
 author: David Beauchesne^1,2,\*^, Rémi M.
@@
 McKindsey^6^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
-date: 2019-04-23
 fontsize: 12pt
@@
 rmarkdown::render('./eDriversMS-v2.rmd', 'pdf_document')
-rmarkdown::render('./eDriversMS.rmd', 'word_document')
+rmarkdown::render('./eDriversMS-v2.rmd', 'word_document')
 -->
@@
 
-**Number of figures**: 5 \newline
+**Number of figures**: 6 \newline
 **Number of tables**: 1 \newline
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-<!-- 350 words -->
 
-The St. Lawrence System is a vast and complex social-ecological system providing
-a wealth of ecosystem services sustaining numerous economic sectors. Related
-human pressures combine with climate driven environmental changes to produce
-intricate exposure regimes of overlapping and potentially interacting
-environmental drivers (*e.g.* hypoxia and fisheries) threatening ecosystems.
-There is a critical need to systematically characterize the distribution,
-intensity and overlap between drivers to support integrative initiatives such
-as ecosystem-based management. While portals providing knowledge on individual
-drivers exist, platforms collating comparable and interoperable knowledges on
-multiple drivers remain conspicuously missing. This paper presents two distinct,
-but related objectives. The first is to characterize the distribution and
-intensity of drivers in the St. Lawrence System. The second is to launch
-*eDrivers*, an open knowledge platform gathering experts committed to
-structuring, standardizing and sharing knowledges on drivers in support of
-science and management. We gathered data on 22 coastal, climate, fisheries and
-marine traffic drivers through collaborations, existing environmental
-initiatives and open data portals. We show that few areas of the St. Lawrence
-are free of cumulative exposure. The Estuary, the Anticosti Gyre and coastal
-areas are particularly exposed, especially in the vicinity of urban centers.
-We identified 6 areas of distinct cumulative exposure regime that show that
-certain drivers typically co-occur in different regions of the St. Lawrence and
-that coastal areas are exposed to all driver types. Of particular concern are
-two threat complexes capturing most exposure hotspots that show the convergence
-of contrasting exposure regimes at the head of the Laurentian Channel. These
-observations are destined to improve as *eDrivers* evolves through time to
-address knowledge gaps and refine current driver layers. In an effort to share
-the knowledge acquired and to ensure the lasting relevance of the description
-of drivers presented in this manuscript, *eDrivers* was built on a series of
-guiding principles upholding existing data management and open science
-standards. Ultimately, we believe that *eDrivers* represents a much needed
-solution that could radically influence broad scale research and management
-practices by increasing data accessibility and interoperability and by
-increasing research and decision-making efficiency.
+The St. Lawrence is a vast and complex socio-ecological system providing a
+wealth of services sustaining numerous economic sectors. These ecosystems are
+subject to significant human pressures that overlap and potentially interact
+with climate driven environmental changes. Our objective in this paper is to
+systematically characterize the distribution and intensity of drivers in the
+St. Lawrence System. To do so, we launch *eDrivers*, an open knowledge
+platform gathering experts committed to structuring, standardizing and sharing
+knowledge on drivers in support of science and management. We gathered data on
+22 coastal, climate, fisheries and marine traffic drivers through
+collaborations, existing environmental initiatives and open data portals. We
+show that few areas of the St. Lawrence are free of cumulative exposure. The
+Estuary, the Anticosti Gyre and coastal areas are particularly exposed,
+especially in the vicinity of urban centers. We identified 6 areas of distinct
+cumulative exposure regime that show that certain drivers typically co-occur
+in different regions of the St. Lawrence and that coastal areas are exposed to
+all driver types. Of particular concern are two threat complexes capturing
+most exposure hotspots that show the convergence of contrasting exposure
+regimes at the head of the Laurentian Channel. *eDrivers* was built on a
+series of guiding principles upholding existing data management and open
+science standards. We therefore expect it to evolve through time to address
+knowledge gaps and refine current driver layers. Ultimately, we believe that
+*eDrivers* represents a much needed solution that could radically influence
+broad scale research and management practices by increasing knowledge accessibility
+and interoperability.
 
@@
 
-The St. Lawrence System, formed by one of the largest estuaries in the world and
-a vast interior sea, is a complex social-ecological system characterized by
-highly variable environmental conditions and oceanographic processes, both in
-space and time [@el-sabh1990a; @white1997; @dufour2007]. It thus offers a unique
+The St. Lawrence System, formed by one of the largest estuaries in the world,
+the St. Lawrence Estuary, and a vast interior sea, the Gulf of St. Lawrence, is
+a complex social-ecological system characterized by highly variable
+environmental conditions and oceanographic processes
+[@el-sabh1990a; @white1997; @dufour2007]. It constitutes a unique
 and heterogeneous array of habitats suited for the establishment of diverse and
@@
 aquaculture production, fosters emerging activities and boasts a yet untapped
-hydrocarbon potential [@beauchesne2016; @schloss2017 @archambault2017]. With
+hydrocarbon potential [@beauchesne2016; @schloss2017; @archambault2017]. With
 major investments recently made and more forthcoming in economic and
-infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018], an
-intensification of the human footprint is expected in the St. Lawrence System.
+infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018],
+an intensification of the human footprint is likely in the St. Lawrence System.
 
@@
 drivers is the largest uncertainty when studying or predicting environmental
-change [@darling2008; @cote2016]. The effects of multiple drivers can combine
+impacts [@darling2008; @cote2016]. Multiple drivers can combine
 non-linearly and result in effects that are greater (synergistic effect) or
 lower (antagonistic effect) than the sum of individual effects
-[@crain2008; @darling2008; @cote2016].  <!-- For example...??? -->
+[@crain2008; @darling2008; @cote2016].
 
@@
 most research on driver effects in marine environments remains overwhelmingly
-focused on single driver assessments [@obrien2019]. Arising from the expected
-increase in environmental exposure and the experiences of past and
-current ecological tragedies such as the collapse of cod fisheries
-[@frank2005; @dempsey2018] and the decline of the beluga and right whale <!-- right whale ref -->
-populations [@plourde2014] is a need to characterize the distribution,
+focused on single driver assessments [@obrien2019]. Increasing exposure and the
+experiences of past ecological tragedies such as the collapse of cod fisheries
+[@frank2005; @dempsey2018] and the decline of the beluga and right whale
+populations [@plourde2014] together urge the need to characterize the distribution,
 intensity and overlap between drivers in the St. Lawrence System. This will
@@
 Gathering environmental data for large scale, systematic initiatives can,
-however, be a very challenging and time consuming - not to say painful - process.
+however, be a very challenging and time consuming – not to say painful – process.
 On one hand, there is an overwhelming and expanding wealth of data available.
@@
 
-The current digital infrastructure is highly decentralized and the data
-management and sharing practices are highly heterogeneous, preventing us from
-maximizing benefits from research investments [@wilkinson2016].
-Yet there now exists multiple initiatives that address this issue by assembling,
-organizing and sharing vast arrays of environmental data.
-Biotic data can be accessed through web portals such as the Ocean Biogeographic
-Information System [OBIS; @obis2018], the Global Biodiversity Information Facility
-[GBIF; @gbif2018], the Global Biotic Interactions platform [GloBI; @poelen2014],
-and the World Register of Marine Species [WoRMS; @wormseditorialboard2017]. Abiotic data can
-also be accessed through WorldClim [@hijmans2005], Bio-ORACLE
-[@tyberghein2012], and MARSPEC [@sbrocco2013]. Initiatives focused on sharing
-environmental data for specific areas also exist, such as the St. Lawrence
-Global Observatory (SLGO[^slgo]), the European Marine Observation
-and Data Network (EMODnet[^emod]) and the U.S. Integrated Ocean
-Observing System (IOOS[^ioos]). Essential environmental
-parameters are also organized, coordinated and acquired through global
-initiatives like the Group on Earth Observations Biodiversity Observation
-Network [GEO BON; @scholes2012], the Census of Marine Life [CoML; @coml2010]
-and the Global Ocean Observing System (GOOS[^goos]).
-
-[^slgo]: https://ogsl.ca/en
-[^emod]: http://www.emodnet.eu/
-[^ioos]: https://ioos.noaa.gov/
-[^goos]: http://www.goosocean.org
-
+There are now initiatives that address this issue by assembling, organizing and
+sharing environmental knowledge, such as the Ocean Biogeographic Information
+System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for
+abiotic data.
 However, equivalent platforms for environmental drivers have largely focused on
@@
 remain conspicuously missing [but see @halpern2015]. This is in spite of
-integerated management and assessment approaches needing efficient data reporting,
+integrated management and assessment approaches needing efficient data reporting,
 standardized data management practices and tools tailored to the study of the
@@
 
-This manuscript thus has two distinct, but related, goals. The first
-is to characterize the distribution and intensity of drivers in the St. Lawrence
-System in order to: 1) identify areas of high cumulative exposure,
-2) characterize areas with similar cumulative exposure regimes, dubbed threat
-complexes [@bowler2019] and 3) identify drivers that are likely to interact
-in the St. Lawrence. The second goal is to present how we are sharing the
-knowledge acquired by launching an open knowledge platform, *eDrivers*, and use
-this paper as a call for collaboration to any person, group or organization that
-might be interested in contributing to this initiative.
+Our main objective in this study is to characterize the distribution and
+intensity of drivers in the St. Lawrence. More specifically, we aim to: 1)
+identify areas of high cumulative exposure, 2) identify drivers that are likely
+to interact in the St. Lawrence and 3) characterize areas with similar
+cumulative exposure regimes.
+Here we focus solely on drivers, one of four elements required to evaluate
+environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.*
+habitats or species) and the exposure and the vulnerability of the ecosystem
+components to drivers. We achieve these objectives with the development of an
+open knowledge platform named *eDrivers*. The platform was designed to
+facilitate collaboration, real-time assessments of cumulative exposure and to
+evolve with the addition of information and threats to the St-Lawrence ecosystems.
 
@@
 
+## Estuary and Gulf of St. Lawrence
+
+The St. Lawrence System is formed by the Estuary and the Gulf of St. Lawrence
+(Figure \ref{egsl}). The Estuary is defined by the seawater limit
+close to Îles d'Orléans to the west and by its connexion to the Gulf
+near Pointe-des-Monts.
+Estuarine circulation characterizes the Estuary. The surface layer is
+composed of a freshwater seaward outflow primarily from the Great Lakes Basin
+and through the St. Lawrence River, upstream of the Estuary. Atlantic waters are
+flowing mainly from the Cabot Strait landward through the Laurentian Channel
+(see below).
+
+The Gulf is an interior sea connected to the Atlantic by the Cabot and Belle-Isle
+Straits to the south and north of Newfoundland, respectively.
+The topology of the Northern Gulf is characterized by three deep channels (250-500 m).
+The Laurentian Channel is the main channel connecting the Estuary to the
+Atlantic through the Cabot Strait. The Esquiman and Anticosti Channels are two
+secondary channels that branch off from the Laurentian Channel to the north
+towards the Arctic and the north of Anticosti Island, respectively.
+The Southern Gulf is characterized by the Magdalen Shallows, a vast area with an
+average depth of ~50 m. The Gulf is composed of a seasonal cold intermediate layer
+that separates the surface and deep layers. Seasonal sea ice affects the circulation
+of the St. Lawrence. Finally, three islands impact the physical dynamics of the
+St. Lawrence: the Anticosti Island to the north, the Magdalen Islands in the
+middle of the Magdalen Shallows and Prince Edward Island to the south. See
+@saucier2003 and @galbraith2018 for more information on the physical oceanography
+of the St. Lawrence.
+
+The St. Lawrence drains over 25% of global freshwater reserves through its
+connexion to the Great Lakes Basin. The Basin is home to over 45 million
+North-Americans, *i.e.* 15 and 30 million in Canada and the United States,
+respectively [@archambault2017]. The St. Lawrence itself boasts a much lower
+population of approximately 1 million Canadians living within 10 km of the
+coast, with populations mainly located in a few coastal cities in the Estuary
+and the Southern Gulf [@statistics-canada2017].
+
 ## Drivers
 
-The list of drivers for which we sought data was informed by a global cumulative
-impact assessment initiative [@halpern2008b; @halpern2015a], regional holistic
-evaluations of the state of the St. Lawrence [@dufour2007; @benoit2012], and
-communications with regional experts. Through the data gathering process, we
-developed and continue to develop collaborations with regional experts and data
-holders. We also use global data from the global cumulative impact assessment
-initiative [@halpern2008b; @halpern2015a] available from the National Center
-for Ecological Analysis and Synthesis (NCEAS) online data repository
-[Table 1; @halpern2015]. We selected global data that were unavailable at the
-regional scale and that were available at a resolution adequate for use at the
-scale of the St. Lawrence (*e.g.* marine pollution).
+The list of drivers for which we sought data was informed by a global
+cumulative impact assessment initiative [@halpern2008a; @halpern2015a],
+regional holistic evaluations of the state of the St. Lawrence [@dufour2007;
+@benoit2012], and communications with regional experts. We integrated datasets
+from regional experts and also use global data from the global cumulative
+impact assessment initiative [@halpern2008a; @halpern2015a] available from the
+National Center for Ecological Analysis and Synthesis (NCEAS) online data
+repository [Table 1; @halpern2015]. We selected global data that were
+unavailable at the regional scale and that were available at a resolution
+adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).
 
-We were able to characterize the intensity and distribution of 22 drivers
-(Table 1;) and we are actively working on updating or developing additional
-driver layers. Drivers incorporated in the analyses are varied in origin, *i.e.* from
-terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping), and from
-large scale biophysical processes (*e.g.* temperature anomalies) to localized
-anthropogenic activities (*e.g.* fisheries). Drivers were divided into 4 groups:
-coastal, climate, fisheries and marine traffic (Table 1). All data layers and
-methodologies are described in the supplementary materials.
+We characterized the intensity and distribution of 22 drivers
+(Table 1;). Drivers incorporated in the analyses are varied in origin,
+*i.e.* from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping),
+and from large scale biophysical processes (*e.g.* temperature anomalies) to
+localized anthropogenic activities (*e.g.* fisheries). Drivers were divided
+into 4 groups: coastal, climate, fisheries and marine traffic (Table 1). All
+data layers and methodologies are described in the supplementary materials.
 
@@
 (Figure S2). All drivers were embedded in a regular grid composed of 245604
-$1 km^2$ exagonal cells to construct the integrated dataset used for the
+$1 km^2$ hexagonal cells to construct the integrated dataset used for the
 analyses.
@@
 
-Areas with high cumulative exposure (objective 1) were identified by evaluating
-the cumulative footprint of combined drivers and by identifying hotspots of
-cumulative footprint. Cumulative footprint ($F$) was defined as the sum of
-the scaled intensity of all drivers in each grid cell:
+Areas with high cumulative exposure (objective 1) were identified by comparing
+areas on the basis of the number and relative intensity of drivers in each
+grid cell. Throughout the text, we use exposure when describing areas
+exposed to drivers and we use footprint when describing the distribution and
+intensity of drivers.
+
+Cumulative footprint ($F$) was defined as the sum of the scaled intensity of all
+drivers in each grid cell:
 
@@
 where $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
-driver $i$.
+driver $i$. The cumulative footprint provides an estimate of the total relative
+footprint in each grid cell. A grid cell with a high $F$ value is either
+characterized by multiple drivers at low relative intensity, limited drivers at
+high relative intensity, or both.
 
-Cumulative hotspots ($H$) were defined as the number of drivers in each grid
-cell with scaled intensity contained over their respective 80th percentile:
+We also identified cumulative hotspots ($H$) to explore the the distribution of
+cumulative exposure in the St. Lawrence (objective 1). Cumulative hotspots ($H$)
+were defined as the number of drivers in each grid cell with scaled intensity
+contained over their respective 80th percentile:
 
@@
 
-The distribution of driver interactions was investigated through the spatial
-overlap of combinations of drivers using the cumulative footprint ($F$) equation
-with pairs of drivers. The intensity at which pairs of drivers co-occur was
-evaluated using a two-dimensional kernel density estimate. As there are 231
-pairwise combinations between 22 drivers, we focus on a single example using
-hypoxia and demersal destructive fisheries, two drivers known to occur mainly
-in deeper areas of the St. Lawrence and, hence, an interaction between the
-effects of the two drivers could be anticipated. Note, however, that *eDrivers*
-will provide the capacity to compare any combinations of drivers.
+Interactions among drivers were investigated using the cumulative footprint
+($F$) between pairs of drivers (objective 2). The intensity at which pairs of
+drivers co-occur was evaluated using a two-dimensional kernel density estimate.
+As there are 231 pairwise combinations between 22 drivers, we focus on a single
+example using hypoxia and demersal destructive fisheries, two drivers known to
+occur mainly in deeper areas of the St. Lawrence and, hence, an interaction
+between the effects of the two drivers could be anticipated.
 
@@
 
-Natural systems are likely to host multiple overlapping drivers. The
-dimensionality of the problem can quickly rise as additional drivers are
-considered. In order to decrease the dimensionality of the integrated dataset,
-we identify threat complexes, *i.e.* regions with similar cumulative exposure
-regimes, using a clustering approach [*e.g.* see @bowler2019].
+In order to better capture potential interactions between drivers (objective 2)
+and to identify areas with similar exposure regimes (objective 3),
+we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
+We use the term clusters in presenting the methods, but use threat complex when
+discussing the results on cumulative exposure regimes.
 
@@
 algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
-designed for large datasets. The CLARA algorithm uses the PAM (Partition Around
-Medoids) algorithm on a sample from the original dataset to identify a set of
-$k$ objects that are representative of all other objects, *i.e.* medoids and
-that are central to the cluster they represent. The goal of the algorithm is to
-iteratively minimize intra-cluster dissimilarity. Iterations are compared on the
-basis of the average dissimilarity between cluster objects and representative
-medoid to select the optimal set of $k$ medoids that minimizes average
-dissimilarity. We used 100 iterations using samples of 10000 observations
+designed for large datasets. The CLARA algorithm uses the PAM (Partition
+Around Medoids) algorithm on a sample from the original dataset to identify a
+set of $k$ objects that are representative of all other objects, *i.e.*
+medoids and that are central to the cluster they represent. The goal of the
+algorithm is to iteratively minimize intra-cluster dissimilarity. Iterations
+are compared on the basis of the average dissimilarity between cluster objects
+and representative medoid to select the optimal set of $k$ medoids that
+minimizes average dissimilarity. We used the clustering algorithm with the
+Manhattan distance since this measure is less affected by extreme values
+[@legendre2012], as is the *k-medoids* clustering algorithm [@kaufman1990].
+We used 100 iterations using samples of 10000 observations
 (*i.e.* ~5% of observations) to identify clusters. Analyses were performed
 using the *cluster* R package [@maechler2018].
-
 Partitional clustering algorithms require a user-defined number of clusters.
@@
 the number of clusters that maximized the average silhouette width [@kaufman1990]
-and minimized the total within-cluster sum of squares (Figure S3).
-
-We also validated the clustering by comparing *k-medoids* clustering with
-*k-means* clustering with the *Lloyd* algorithm [@lloyd1982]. The *k-means*
-approach is similar to the *k-medoids*, but identifies observations belonging
-to a cluster iteratively by minimizing the mean intra-cluster squared distance
-until it converges to an optimal solution. We used 25 random sets and set a
-maximum of 1000 iterations for the analysis. Analyses were performed using
-the *stats* R package [@rcoreteam2018]. We used the same validation procedure
-to select the optimal number of clusters $k$ than with the *k-medoids*
-clustering (Figure S3).
-
-While *k-means* algorithms are more  efficient since they do not compute
-pairwise dissimilarities, it is more sensitive to outliers through the use of
-the mean rather than a centroid. We therefore favored the use of the *k-medoids*
-algorithm, but used the *k-means* to validate clusters.
+and minimized the total within-cluster sum of squares (Figure S4).
 
@@
 dissimilarity and the contribution of each driver to the total inter-cluster
-dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis
-dissimilarity [Figure S4; @clarke1993]. As the drivers dataset is too
-large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting
+dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan
+distance [Figure S5; @clarke1993]. The Manhattan distance was again preferred for continuity
+with the clustering analysis and to ensure that outliers did not have a strong
+influence the analysis. As the drivers dataset is large (~ 250000 observations),
+we used a bootstrap procedure for the SIMPER analysis, randomly selecting
 5% of each cluster to run the analysis and repeating the process over 300
 iterations. We also compared the mean intensity of each driver within each
-cluster to better capture the inter-cluster dissimilarity. Analyses were
-performed using the *vegan* R package [@oksanen2018].
+cluster to better capture the inter-cluster dissimilarity.
 
@@
 
-Intra-cluster similarity was evaluated using the Bray-Curtis similarity index
-(Figure S5). As with the inter-cluster dissimilarity, we used a
-bootstrap procedure for the intra-cluster similarity, randomly selecting 5% of
-each cluster observation to run the analysis and repeating the process over 300
-iterations. We however did not use the bootstraping procedure for clusters with
-less than 10000 observations since computation time was manageable.
+Intra-cluster similarity was evaluated calculating the intra-cluster Manhattan
+distance and by transforming the mean contribution to distance ($M_c$) of each
+driver by $.1 / (.1 + M_c)$ to obtain a similarity measure for each driver
+($S_d$). The total similarity is the sum of all $S_d$. (Figure S6).
+As with the inter-cluster dissimilarity, we used a bootstrap procedure for the
+intra-cluster similarity, randomly selecting 25% of each cluster observation to
+run the analysis and repeating the process over 50 iterations. We only used the
+bootstrapping procedure for clusters with less than 40000 observations since
+computation time was manageable.
 
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-# Results and discussion
+# Results and Discussion
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
@@
 ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
-generally highest along the coast (Figure \ref{footprint}), with hotposts
+generally highest along the coast (Figure \ref{footprint}), with hotspots
 located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
@@
 and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
-or home to small coastal communities.
+or home to small coastal communities. Certain smaller coastal communities with
+high cumulative footprint are characterized by large industries (*e.g.*
+Sept-Îles and Charlottetown).
 
@@
 provide access to and serve as the primary drainage outflow of the Great Lakes
-Basin, the most densely populated region in Canada [@statisticscanada2017]. Most marine
+Basin, which is home to over 45 million North-Americans and is the most densely
+populated region in Canada [@archambault2017; @statistics-canada2017]. Most marine
 traffic thus converges to the Estuary.
@@
 Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels,
-with the head of the Channels most exposed to hypoxia (Figure \ref{kernel}A).
+with the head of the Channels most exposed (Figure \ref{kernel}A).
 Demersal destructive fisheries are located along the Laurentian Channel, the
@@
 (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of
-the Esquiman and Anticosti Channels (Figure \ref{kernel}C). The ease with which
-this figure can be created using *eDrivers* is demonstrated in box 1.
+the Esquiman and Anticosti Channels (Figure \ref{kernel}C, Box 1).
 
@@
 most likely, and where they can arise between a host of different drivers.
-
 The identification of threat complexes provides a crucial tool to simplify the
@@
 Six distinct threat complexes were identified in the St. Lawrence using
-the *k-medoids* and *k-means* algorithms (Figures S4, S5). Based on their
+the *k-medoids* algorithm (Figures S4, S5). Based on their
 distribution and representative drivers, threat complexes can be divided into
@@
 of drivers besides hypoxia and are the most exposed threat complexes, both in
-terms of driver overlap and intensity. Threat complex 2 is differentiated from
-other complexes by the presence of aquaculture sites. Threat complex 1
-encompasses the coastline and is characterized by higher direct human impact
-(*i.e.* population density). Threat complex 3 is the most exposed complex and
+terms of driver overlap and intensity. Threat complex 1 encompasses the
+coastline and is characterized by higher direct human impact (*i.e.* population
+density). Threat complex 2 is differentiated from other complexes by the
+presence of aquaculture sites. Threat complex 3 is the most exposed complex and
 has a distribution similar to the most exposed coastal hotspots (Figure
@@
 Threat complex 4 is differentiated by demersal non-destructive high-bycath
-fisherie, higher marine traffic drivers compared to complex 5 and generally
+fisheries, higher marine traffic drivers compared to complex 5 and generally
 corresponds to the whole Southern Gulf. Threat complex 5 is characterized by
@@
 generally lower intensity marine traffic drivers and is located almost
-exclusively in the Northern Gulf. Finally, threat complex 6 corresponds
-primarily to the Laurentian Channel, a deep (250-500 m) and long (1250 km)
-submerged valley connecting the Estuary to the Atlantic. It also incorporates
-parts of the Esquiman and Anticosti Channels, two deep channels that branch off
-from the the Laurential Channel to the north towards the Arctic and the north
-of Anticosti Island, respectively. This threat complex is the most exposed
+exclusively in the Northern Gulf. Finally, threat complex 6 is the most exposed
 offshore threat complex and includes all offshore hotspots (Figure \ref{hotspot}).
-This complex is characterized by high intensity hypoxia, marine traffic and
+It is characterized by high intensity hypoxia, marine traffic and
 pollution, as well as demersal destructive and pelagic high-bycatch fisheries.
+This threat complex corresponds primarily to the Laurentian Channel and
+incorporates parts of the Esquiman and Anticosti Channels.
 
@@
 cumulative exposure regimes Between them, they capture most of the coastal
-and offshore hotspots identified in the St. Lawrence and discussed above. <!-- more here? -->
-
+and offshore hotspots identified in the St. Lawrence.
 They also offer some insight into the potential importance of considering
@@
 is particularly dynamic, with deep Atlantic waters advected through estuarine
-circulation mixing with surface waters from the Great Lakes Basin and the
+circulation mixing with surface waters from the St. Lawrence River and the
 Saguenay River [@dufour2007]. This results in the convergence of climate
@@
 (threat complex 6) with terrestrial run-off from river outflows and direct
-human impacts (threat complex 3). This dynamic area is also highly
-productive and hosts large aggregations of krill exploited by numerous fish,
-marine mammal and marine bird species [@simard2009; @lavoie2016]. It might
-therefore be reasonable to expect highly unpredictable environmental effects in
-this area.
-<!-- I know this is weak, but it's late and I'm tired! -->
+human impacts (*i.e.* population density; threat complex 3).
 
@@
 Sharing the knowledge acquired through the description of drivers in the
-St. Lawrence quickly emerged as a priority to curtail the need to reach dozens
+St. Lawrence emerged as a priority to curtail the need to reach dozens
 of experts across multiple organizations and over extensive periods of time to
 assemble the data needed to apply integrated research and management. It is also
-a requirement to ensure that this manuscript will not become a quickly outdated
+a requirement to ensure that this manuscript will not quickly become an outdated
 snapshot of drivers distribution and intensity in the St. Lawrence System,
 but rather serve as a stepping stone towards an adaptive and ever-improving
-collection of knowledges.
+collection of knowledge.
 
@@
 knowledge on drivers in support of science and management. In launching this
-intiative, our objective is to uphold the highest existing standards of data
+initiative, our objective is to uphold the highest existing standards of data
 management and open science. We identified four guiding principles to meet this
@@
 St. Lawrence. Notable examples of environmental initiatives are the annual
-review of physical [@galbraith2018], chemical, and biological [@blais2018]
+review of physical [@galbraith2018], chemical, and biological [@blais2019]
 oceanographic conditions in the St. Lawrence, the fisheries monitoring program
 [@dfo2016], the annual groundfish and shrimp multidisciplinary survey
-[@bourdages2018], the characterisation of benthic [@dutil2011], epipelagic and
+[@bourdages2018], the characterization of benthic [@dutil2011], epipelagic and
 coastal [@dutil2012] habitats of the St. Lawrence, and Canada's shoreline
@@
 and quality data and information on the St. Lawrence ecosystem through the
-networking of organisations and data holders to meet their needs and those of
+networking of organizations and data holders to meet their needs and those of
 users, to improve knowledge and to assist decision-making in areas such as
@@
 an initiative from the Québec provincial government. This portal aims at
-collecting and sharing accurate, complete and updated ressources on water and
+collecting and sharing accurate, complete and updated resources on water and
 aquatic ecosystems to support the mandate of relevant actors and stakeholders
@@
 (GOOS). As such, we will adopt the metadata standard
-currently targetted for CIOOS, *i.e.* the North American Profile
+currently targeted for CIOOS, *i.e.* the North American Profile
 of ISO 19115:2014 - Geographic information - Metadata, a schema favoured
@@
 ***Why***: Like peer-reviewed publications, data must also be given its due
-importance in scientific endeavors and thus be considered as legitimate citable
+importance in scientific endeavours and thus be considered as legitimate citable
 products contributing to the overall scientific output of data providers
-[@codata2013; @force112014]. Appropriate citations should
-therefore be provided for all data layers used and shared by the platform.
+[@taskgroupondatacitationstandardsandpractoutofciteoutofmindthecurrentsices2013; @force112014].
+Appropriate citations should therefore be provided for all data layers used and
+shared by the platform.
 
@@
 particularly exposed to all types of drivers and that multiple drivers
-typically co-occur in space. These resuts allow us to efficiently identify areas
-in need of heightened scrutiny from a science and management perspective.
+typically co-occur in space. These results allow us to efficiently identify
+areas in need of heightened scrutiny from a science and management perspective.
 
-Through *eDrivers*, these observations will be iteratively improved towards an
-increasingly robust assessment of cumulative exposure and threat complexes as
-gaps in knowledge are addressed or approaches to describe drivers are refined.
+Through *eDrivers*, these observations will be iteratively improved towards
+an increasingly robust assessment of cumulative exposure and threat complexes
+as gaps in knowledge are addressed or approaches to describe drivers are refined.
 Arguably, the most meaningful benefit anticipated from *eDrivers* will be the
@@
 
-Critically - and we emphasize this point - *eDrivers* will
+Critically, *eDrivers* will
 allow the scientific and governmental communities to identify key knowledge
@@
 output to be available and interpretable on a time scale relevant to management
-[see @sutherland2004; @reichman2011]. The platform will thus greatly facilitate the application
-of broad scale, holistic research and management approaches such as
-ecosystem-based management, strategic environmental assessments and
-social-ecological metanetworks [*e.g.* @halpern2015a; @jones2016; @dee2017].
+[see @sutherland2004; @reichman2011]. The platform will thus greatly facilitate
+the application of broad scale, holistic research and management approaches such
+as marine spatial planning, ecosystem-based management, marine spatial planning
+and strategic environmental assessments [*e.g.* @rice2011; @halpern2015a; @jones2016].
+
+While the focus of our paper has been on the description of drivers in the
+St. Lawrence, drivers are but one of the 4 elements required to properly
+evaluate environmental impacts. This single knowledge node should be weaved with
+other, comparable, nodes required for integrated management approaches such as
+species distribution and marine protected areas. Ultimately, all of these
+knowledge nodes could be weaved together through social-ecological metanetworks
+[@dee2017].
 
@@
 
-DB, RD, DG and PA conceived the manuscript and the underlying objective. DB
+DB, RD, DG and PA conceived the manuscript and the underlying objectives. DB
 prepared/formatted the data, performed the analyses, was in charge of technical
 developments and lead the drafting of the manuscript. All co-authors contributed
-to data, analases and writing based on their respective expertise and
+to data, analyses and writing based on their respective expertise and
 contributed to the revision of the manuscript.
@@
 
-Box 1. Code snippet demonstrating how to use the *eDrivers* to reproduce
-figure \ref{kernel} in R.
+Box 1. Code snippet demonstrating how to use the *eDrivers* R package to
+reproduce figure \ref{kernel}.
 
@@
 \centering
+\includegraphics{./Figures/egsl.png}
+\caption{Description of the Estuary and Gulf of St. Lawrence in Eastern Canada}
+\label{egsl}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
+\centering
 \includegraphics{./Figures/footprint.png}
-\caption{Distribution of driver footprint in the Estuary and Gulf of St.
-Lawrence.
-The footprint is measured by summing the relative intensity of all drivers for each grid cell: $F_d = \sum_{i=1}^{n} D_i$.
-Driver layers were also log-transformed when their frequency distribution was non-normal.
-All driver layers were also normalized between 0 and 1 using the 99th quantile to allow for direct comparison of relative intensities between drivers and control for extreme values.}
+\caption{Distribution of cumulative footprint in the St. Lawrence System.}
 \label{footprint}
@@
 \centering
+\includegraphics{./Figures/hotspot.png}
+\caption{Distribution of cumulative hotspots in the St. Lawrence System.}
+\label{hotspot}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
+\centering
 \includegraphics{./Figures/kernel.png}
-\caption{Example of platform content using the spatial distribution and the intensity of hypoxia and demersal destructive fisheries in the St.
+\caption{Interaction between the intensity of hypoxia and demersal destructive fisheries in the St.
 Lawrence.
-An index of hypoxia (**A**) was created using bottom-water dissolved oxygen between 2013 and 2017 [@blais2018].
-Demersal destructive fisheries (*i.e.* trawl and dredges) (**B**) intensity was evaluated from fisheries catch data collected between 2010 and 2015 used to measure annual area weighted total biomass ($kg$) in i $km^2$ grid cells [@dfo2016].
+An index of hypoxia ($A$) was created using bottom-water dissolved oxygen between 2013 and 2017 (Blais et al., 2018).
+Demersal destructive fisheries ($i.e.$ trawl and dredges) ($B$) intensity was evaluated from fisheries catch data collected between 2010 and 2015 used to measure annual area weighted total biomass ($kg$) in i $km^2$ grid cells (DFO, 2016b).
 See supplementary materials for more information on specific methodologies.
-Relative hypoxic stress and demersal destructive fisheries intensity was summed (**C**) to visualize their combined spatial distribution and intensity.
-Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a two-dimensional kernel analysis (**D**).}
+Relative hypoxic stress and demersal destructive fisheries intensity was summed ($C$) to visualize their combined spatial distribution and intensity.
+Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a two-dimensional kernel analysis ($D$).}
 \label{kernel}
@@
 \caption{Distribution of threat complexes in the Estuary and Gulf of St.
-Lawrence.
-Threat complexes [a term coined by @bowler2019] are areas with similar cumulative driver exposure regimes.
-Threat complexes were identified using a partitional clustering algorithm [@kaufman1990].
-The appropriate number of clusters ($k$) was tested using a range of values and validated by selecting the number of clusters that maximized the average silhouette width [@kaufman1990] and minimized the total within-cluster sum of squares (WSS).
-Six distinct threat complexes were identified in the St.
-Lawrence.
-The partitional clustering algorithm analysis were performed using the *cluster* R packages [@maechler2018].
+Lawrence (upper panel).
+Threat complexes (a term coined by Bowler et al., 2019) are areas with similar cumulative driver exposure regimes.
+Mean intensity of all coastal (red), climate (green), fisheries (blue) and marine traffic (purple) drivers within each threat complex (lower panel).
 Refer to the Supplementary Materials for more details.}
@@
 Community input in the form of raw data is accessed through the St.
-Lawrence Global Observatory (SLGO); https://ogsl.ca/en) repository - the platform host - or through open access repositories (*e.g.* NASA data).
-The raw data are then processed through a workflow hosted on the *eDrivers* GitHub organization (https://github.com/orgs/eDrivers/).
-Data processing may be as simple as data rescaling (*e.g.* night lights) or make use of more complex methodologies (*e.g.* acidification).
+Lawrence Global Observatory (SLGO; https://ogsl.ca/en) repository - the platform host - or through open access repositories ($e.g.$ NASA data).
+The raw data are then processed through a workflow hosted on the $eDrivers$ GitHub organization (https://github.com/orgs/eDrivers/).
+Data processing may be as simple as data rescaling ($e.g.$ night lights) or make use of more complex methodologies ($e.g.$ acidification).
 All data is then hosted on SLGO's geoserver and accessible through their API.
-We developed a R package called *eDrivers* to access the driver layers through R and we are actively developing a second R package called *eDriversEx* that includes analytical tools to explore drivers data.
+We developed a R package called $eDrivers$ to access the driver layers through R and we are actively developing a second R package called $eDriversEx$ that will include analytical tools to explore drivers data.
 Finally, we have developed a Shiny application that allows users to explore drivers data interactively (https://david-beauchesne.shinyapps.io/eDriversApp/).
-All R components of the project are hosted and available on the *eDrivers* GitHub organization.}
+All R components of the project are hosted and available on the $eDrivers$ GitHub organization.}
 \label{diag}
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-
-<!--
-To check:
-  - Kaufman, L., and Rousseeuw, P. (1990)
--->

### 2019-05-01 eDriversMS-v2_DG.Rmd
latest modifications to manuscript, including frontiers formatting and supervisor comments

--- 2019-05-01 eDriversMS-v2.Rmd
@@
 ---
-title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System
+title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence
 author: David Beauchesne^1,2,\*^, Rémi M.
@@
 McKindsey^6^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
+date: 2019-04-23
 fontsize: 12pt
@@
 rmarkdown::render('./eDriversMS-v2.rmd', 'pdf_document')
-rmarkdown::render('./eDriversMS-v2.rmd', 'word_document')
+rmarkdown::render('./eDriversMS.rmd', 'word_document')
 -->
@@
 
-**Number of figures**: 6 \newline
+**Number of figures**: 5 \newline
 **Number of tables**: 1 \newline
@@
 \newpage
-
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-# Abstract
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-
-The St. Lawrence is a vast and complex socio-ecological system providing a
-wealth of services sustaining numerous economic sectors. These ecosystems are
-subject to significant human pressures that overlap and potentially interact
-with climate driven environmental changes. Our objective in this paper is to
-systematically characterize the distribution and intensity of drivers in the
-St. Lawrence System. To do so, we launch *eDrivers*, an open knowledge
-platform gathering experts committed to structuring, standardizing and sharing
-knowledge on drivers in support of science and management. We gathered data on
-22 coastal, climate, fisheries and marine traffic drivers through
-collaborations, existing environmental initiatives and open data portals. We
-show that few areas of the St. Lawrence are free of cumulative exposure. The
-Estuary, the Anticosti Gyre and coastal areas are particularly exposed,
-especially in the vicinity of urban centers. We identified 6 areas of distinct
-cumulative exposure regime that show that certain drivers typically co-occur
-in different regions of the St. Lawrence and that coastal areas are exposed to
-all driver types. Of particular concern are two threat complexes capturing
-most exposure hotspots that show the convergence of contrasting exposure
-regimes at the head of the Laurentian Channel. *eDrivers* was built on a
-series of guiding principles upholding existing data management and open
-science standards. We therefore expect it to evolve through time to address
-knowledge gaps and refine current driver layers. Ultimately, we believe that
-*eDrivers* represents a much needed solution that could radically influence
-broad scale research and management practices by increasing knowledge accessibility
-and interoperability.
-
-**Keywords**: ocean observing systems, St. Lawrence, environmental drivers,
-cumulative exposure, threat complex, multiple stressors, global change
-
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-# Introduction
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-
-The St. Lawrence System, formed by one of the largest estuaries in the world,
-the St. Lawrence Estuary, and a vast interior sea, the Gulf of St. Lawrence, is
-a complex social-ecological system characterized by highly variable
-environmental conditions and oceanographic processes
-[@el-sabh1990a; @white1997; @dufour2007]. It constitutes a unique
-and heterogeneous array of habitats suited for the establishment of diverse and
-productive ecological communities [@savenkoff2000]. As a result, the St.
-Lawrence System provides a wealth of ecosystem services that have historically
-and contemporarily benefited the Canadian economy. It sustains a rich fisheries
-industry targeting more than 50 species, serves as the gateway to eastern
-North-America by granting access to more than 40 ports and the most densely
-populated Canadian region, hosts a booming tourism industry and an expanding
-aquaculture production, fosters emerging activities and boasts a yet untapped
-hydrocarbon potential [@beauchesne2016; @schloss2017; @archambault2017]. With
-major investments recently made and more forthcoming in economic and
-infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018],
-an intensification of the human footprint is likely in the St. Lawrence System.
-
-As elsewhere in the world [see @halpern2015a], this intensifying human footprint
-will likely result in increasingly intricate environmental exposure regimes,
-*i.e.* suites of overlapping and potentially interacting environmental drivers
-threatening ecosystems, habitats or ecological communities. Drivers, often
-referred to as stressors or pressures, are any externalities that affect
-environmental processes and disturb natural systems. Drivers may originate from
-natural or human-induced biophysical processes (*e.g.* sea surface temperature
-anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
-fisheries and marine pollution). The potential for complex interactions between
-drivers is the largest uncertainty when studying or predicting environmental
-impacts [@darling2008; @cote2016]. Multiple drivers can combine
-non-linearly and result in effects that are greater (synergistic effect) or
-lower (antagonistic effect) than the sum of individual effects
-[@crain2008; @darling2008; @cote2016].
-
-The uncertainty associated with complex driver interactions must therefore be
-taken into account when investigating environmental impacts [@cote2016], yet
-most research on driver effects in marine environments remains overwhelmingly
-focused on single driver assessments [@obrien2019]. Increasing exposure and the
-experiences of past ecological tragedies such as the collapse of cod fisheries
-[@frank2005; @dempsey2018] and the decline of the beluga and right whale
-populations [@plourde2014] together urge the need to characterize the distribution,
-intensity and overlap between drivers in the St. Lawrence System. This will
-provide critical information on areas most exposed to cumulative drivers and on
-the interaction potential of drivers in the St. Lawrence. It is also a necessary
-step towards the holistic and integrated management of the St. Lawrence System.
-
-Gathering environmental data for large scale, systematic initiatives can,
-however, be a very challenging and time consuming – not to say painful – process.
-On one hand, there is an overwhelming and expanding wealth of data available.
-Such information overload may inhibit our ability to make decisions based on
-scientific information, promote massive effort duplication, disproportionately
-appropriate research funds to certain sectors, and obscure knowledge gaps amid
-a sea of information [@eppler2004]. On the other hand, crucial data are lacking
-and remain largely unavailable or inaccessible for a variety of reasons,
-including proprietary rights, lack of organizational time, capacity and
-training, and in some rare cases unwillingness to share, curtailing our ability
-for appropriate decision-making.
-
-There are now initiatives that address this issue by assembling, organizing and
-sharing environmental knowledge, such as the Ocean Biogeographic Information
-System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for
-abiotic data.
-However, equivalent platforms for environmental drivers have largely focused on
-single drivers (e.g. Global Fishing Watch) and platforms collating
-data and knowledge on multiple drivers in a comparable and interoperable way
-remain conspicuously missing [but see @halpern2015]. This is in spite of
-integrated management and assessment approaches needing efficient data reporting,
-standardized data management practices and tools tailored to the study of the
-effects of multiple drivers [@dafforn2016; @stock2018]. An additional objective
-thus emerged in the process of addressing our initial goals: sharing the
-knowledge gathered through the description of drivers in the St.
-Lawrence.
-
-Our main objective in this study is to characterize the distribution and
-intensity of drivers in the St. Lawrence. More specifically, we aim to: 1)
-identify areas of high cumulative exposure, 2) identify drivers that are likely
-to interact in the St. Lawrence and 3) characterize areas with similar
-cumulative exposure regimes.
-Here we focus solely on drivers, one of four elements required to evaluate
-environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.*
-habitats or species) and the exposure and the vulnerability of the ecosystem
-components to drivers. We achieve these objectives with the development of an
-open knowledge platform named *eDrivers*. The platform was designed to
-facilitate collaboration, real-time assessments of cumulative exposure and to
-evolve with the addition of information and threats to the St-Lawrence ecosystems.
-
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-# Materials and Methods
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-
-## Estuary and Gulf of St. Lawrence
-
-The St. Lawrence System is formed by the Estuary and the Gulf of St. Lawrence
-(Figure \ref{egsl}). The Estuary is defined by the seawater limit
-close to Îles d'Orléans to the west and by its connexion to the Gulf
-near Pointe-des-Monts.
-Estuarine circulation characterizes the Estuary. The surface layer is
-composed of a freshwater seaward outflow primarily from the Great Lakes Basin
-and through the St. Lawrence River, upstream of the Estuary. Atlantic waters are
-flowing mainly from the Cabot Strait landward through the Laurentian Channel
-(see below).
-
-The Gulf is an interior sea connected to the Atlantic by the Cabot and Belle-Isle
-Straits to the south and north of Newfoundland, respectively.
-The topology of the Northern Gulf is characterized by three deep channels (250-500 m).
-The Laurentian Channel is the main channel connecting the Estuary to the
-Atlantic through the Cabot Strait. The Esquiman and Anticosti Channels are two
-secondary channels that branch off from the Laurentian Channel to the north
-towards the Arctic and the north of Anticosti Island, respectively.
-The Southern Gulf is characterized by the Magdalen Shallows, a vast area with an
-average depth of ~50 m. The Gulf is composed of a seasonal cold intermediate layer
-that separates the surface and deep layers. Seasonal sea ice affects the circulation
-of the St. Lawrence. Finally, three islands impact the physical dynamics of the
-St. Lawrence: the Anticosti Island to the north, the Magdalen Islands in the
-middle of the Magdalen Shallows and Prince Edward Island to the south. See
-@saucier2003 and @galbraith2018 for more information on the physical oceanography
-of the St. Lawrence.
-
-The St. Lawrence drains over 25% of global freshwater reserves through its
-connexion to the Great Lakes Basin. The Basin is home to over 45 million
-North-Americans, *i.e.* 15 and 30 million in Canada and the United States,
-respectively [@archambault2017]. The St. Lawrence itself boasts a much lower
-population of approximately 1 million Canadians living within 10 km of the
-coast, with populations mainly located in a few coastal cities in the Estuary
-and the Southern Gulf [@statistics-canada2017].
-
-## Drivers
-
-The list of drivers for which we sought data was informed by a global
-cumulative impact assessment initiative [@halpern2008a; @halpern2015a],
-regional holistic evaluations of the state of the St. Lawrence [@dufour2007;
-@benoit2012], and communications with regional experts. We integrated datasets
-from regional experts and also use global data from the global cumulative
-impact assessment initiative [@halpern2008a; @halpern2015a] available from the
-National Center for Ecological Analysis and Synthesis (NCEAS) online data
-repository [Table 1; @halpern2015]. We selected global data that were
-unavailable at the regional scale and that were available at a resolution
-adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).
-
-We characterized the intensity and distribution of 22 drivers
-(Table 1;). Drivers incorporated in the analyses are varied in origin,
-*i.e.* from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping),
-and from large scale biophysical processes (*e.g.* temperature anomalies) to
-localized anthropogenic activities (*e.g.* fisheries). Drivers were divided
-into 4 groups: coastal, climate, fisheries and marine traffic (Table 1). All
-data layers and methodologies are described in the supplementary materials.
-
-Drivers with non-normal frequency distributions were log-transformed (Figure S1)
-and all drivers were scaled between 0 and 1 to allow driver comparisons. The
-99th quantile of individual driver distribution was used as the upper bound for
-scaling to control for extreme values and produce maps of individual drivers
-(Figure S2). All drivers were embedded in a regular grid composed of 245604
-$1 km^2$ hexagonal cells to construct the integrated dataset used for the
-analyses.
-
-## Cumulative exposure
-
-Areas with high cumulative exposure (objective 1) were identified by comparing
-areas on the basis of the number and relative intensity of drivers in each
-grid cell. Throughout the text, we use exposure when describing areas
-exposed to drivers and we use footprint when describing the distribution and
-intensity of drivers.
-
-Cumulative footprint ($F$) was defined as the sum of the scaled intensity of all
-drivers in each grid cell:
-
-$$F_x = \sum_{i=1}^{n} D_{i,x}$$
-
-where $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
-driver $i$. The cumulative footprint provides an estimate of the total relative
-footprint in each grid cell. A grid cell with a high $F$ value is either
-characterized by multiple drivers at low relative intensity, limited drivers at
-high relative intensity, or both.
-
-We also identified cumulative hotspots ($H$) to explore the the distribution of
-cumulative exposure in the St. Lawrence (objective 1). Cumulative hotspots ($H$)
-were defined as the number of drivers in each grid cell with scaled intensity
-contained over their respective 80th percentile:
-
-<!-- Indicator function -->
-$$H_x = \sum_{i=1}^{n} \mathbbm{1} (D_{i,x} \; \epsilon \; P_{80, D_i})$$
-
-where, $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
-driver $i$ and $P_{80, D_i}$ is the 80th percentile of driver $i$. Hotspots
-thus identify areas where drivers are co-occurring at high relative intensities.
-
-## Driver interactions
-
-Interactions among drivers were investigated using the cumulative footprint
-($F$) between pairs of drivers (objective 2). The intensity at which pairs of
-drivers co-occur was evaluated using a two-dimensional kernel density estimate.
-As there are 231 pairwise combinations between 22 drivers, we focus on a single
-example using hypoxia and demersal destructive fisheries, two drivers known to
-occur mainly in deeper areas of the St. Lawrence and, hence, an interaction
-between the effects of the two drivers could be anticipated.
-
-## Threat complexes
-
-In order to better capture potential interactions between drivers (objective 2)
-and to identify areas with similar exposure regimes (objective 3),
-we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
-We use the term clusters in presenting the methods, but use threat complex when
-discussing the results on cumulative exposure regimes.
-
-### Clustering
-
-Threat complexes were identified using a partional *k-medoids* clustering
-algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
-designed for large datasets. The CLARA algorithm uses the PAM (Partition
-Around Medoids) algorithm on a sample from the original dataset to identify a
-set of $k$ objects that are representative of all other objects, *i.e.*
-medoids and that are central to the cluster they represent. The goal of the
-algorithm is to iteratively minimize intra-cluster dissimilarity. Iterations
-are compared on the basis of the average dissimilarity between cluster objects
-and representative medoid to select the optimal set of $k$ medoids that
-minimizes average dissimilarity. We used the clustering algorithm with the
-Manhattan distance since this measure is less affected by extreme values
-[@legendre2012], as is the *k-medoids* clustering algorithm [@kaufman1990].
-We used 100 iterations using samples of 10000 observations
-(*i.e.* ~5% of observations) to identify clusters. Analyses were performed
-using the *cluster* R package [@maechler2018].
-Partitional clustering algorithms require a user-defined number of clusters.
-Values of $k$ ranging from 2 to 10 were tested and validated by selecting
-the number of clusters that maximized the average silhouette width [@kaufman1990]
-and minimized the total within-cluster sum of squares (Figure S4).
-
-### Inter-cluster dissimilarity
-
-The difference between clusters was explored by measuring the total inter-cluster
-dissimilarity and the contribution of each driver to the total inter-cluster
-dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan
-distance [Figure S5; @clarke1993]. The Manhattan distance was again preferred for continuity
-with the clustering analysis and to ensure that outliers did not have a strong
-influence the analysis. As the drivers dataset is large (~ 250000 observations),
-we used a bootstrap procedure for the SIMPER analysis, randomly selecting
-5% of each cluster to run the analysis and repeating the process over 300
-iterations. We also compared the mean intensity of each driver within each
-cluster to better capture the inter-cluster dissimilarity.
-
-### Intra-cluster similarity
-
-Intra-cluster similarity was evaluated calculating the intra-cluster Manhattan
-distance and by transforming the mean contribution to distance ($M_c$) of each
-driver by $.1 / (.1 + M_c)$ to obtain a similarity measure for each driver
-($S_d$). The total similarity is the sum of all $S_d$. (Figure S6).
-As with the inter-cluster dissimilarity, we used a bootstrap procedure for the
-intra-cluster similarity, randomly selecting 25% of each cluster observation to
-run the analysis and repeating the process over 50 iterations. We only used the
-bootstrapping procedure for clusters with less than 40000 observations since
-computation time was manageable.
-
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-# Results and Discussion
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-
-## Cumulative exposure
-
-Apart from the northeastern Gulf, the cumulative footprint of drivers is
-ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
-generally highest along the coast (Figure \ref{footprint}), with hotspots
-located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
-offshore areas are less exposed to cumulative drivers, with the Estuary and the
-Anticosti Gyre being notable exceptions (Figures \ref{footprint} and
-\ref{hotspot}). This is not to say that offshore areas are free of exposure,
-as most of the St. Lawrence is exposed to multiple overlapping drivers
-(Figures \ref{footprint} and \ref{hotspot}). For example, it is worthy to note
-high cumulative footprint observed at the heads of the Anticosti and Esquiman
-Channels (Figure \ref{footprint}).
-
-These results are consistent with observations elsewhere in the world, where
-cumulative driver exposure conspicuously arises from and markedly intensifies
-close to coastal cities and at the mouth of rivers draining highly populated
-areas [e.g. @halpern2015a; @feist2016; @mach2017; @stock2018]. These are areas
-where human activities (*e.g.* coastal development and shipping) and footprint
-(*e.g.* pollution runoff) are the most intense [@feist2016], and on which is
-overlaid a background of natural disturbances [@micheli2016]. They are also the
-areas in which the most dramatic increases in exposure are expected, with
-populations increasing more rapidly along the coast than inland [@feist2016].
-In the St. Lawrence, large coastal cities are mostly located along the Estuary
-and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
-or home to small coastal communities. Certain smaller coastal communities with
-high cumulative footprint are characterized by large industries (*e.g.*
-Sept-Îles and Charlottetown).
-
-As for offshore exposure, the Estuary, along with the St. Lawrence River,
-provide access to and serve as the primary drainage outflow of the Great Lakes
-Basin, which is home to over 45 million North-Americans and is the most densely
-populated region in Canada [@archambault2017; @statistics-canada2017]. Most marine
-traffic thus converges to the Estuary.
-
-While we cannot ascertain that high exposure areas are the most impacted,
-we can safely predict that these are the areas where studying ecosystem state
-will be the most complex due to the uncertainty associated with driver
-interactions, an uncertainty bound to increase rapidly with the
-number of interacting drivers [@cote2016].
-
-## Driver interactions
-
-Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels,
-with the head of the Channels most exposed (Figure \ref{kernel}A).
-Demersal destructive fisheries are located along the Laurentian Channel, the
-heads of the Anticosti and Esquiman Channels and around the Magdalen Islands
-(Figure \ref{kernel}B). By combining both drivers, we can observe that hypoxia
-and demersal destructive fisheries overlap mostly at high relative intensity
-(Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of
-the Esquiman and Anticosti Channels (Figure \ref{kernel}C, Box 1).
-
-Fisheries in the St. Lawrence have historically affected
-biodiversity distribution and habitat quality [@moritz2015]. Concurrently,
-hypoxia decreases overall habitat quality, but triggers species-dependent
-responses ranging from adaptation [*e.g.* northern shrimp *Pandalus borealis*
-and Greenland halibut *Reinhardtius hippoglossoides*; @pillet2016] to reduced
-growth rates [@dupont-prinet2013] and avoidance of oxygen-depleted habitats
-[*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality
-[*e.g.* sessile benthic invertebrates; @eby2005; @gilbert2007; @belley2010].
-Certain species may thus be adversely affected by fisheries and withstand
-hypoxia but still experience a decrease in prey availability, while others may
-be deleteriously affected by the compounded effect of both drivers [@deleo2017].
-
-## Threat complexes
-
-While informative, the hypoxia-fisheries example focuses on a single pair of
-drivers and falls short of the number of drivers overlapping at
-high intensities throughout the St. Lawrence (Figure \ref{hotspot}).
-The number of drivers overlapping in the St. Lawrence increases with cumulative
-exposure (Figure S3). Areas with high exposure such as the Estuary, the
-Anticosti Gyre and the southwestern Gulf coastline (Figure \ref{footprint} and
-\ref{hotspot}) are thus areas where driver interactions are
-most likely, and where they can arise between a host of different drivers.
-The identification of threat complexes provides a crucial tool to simplify the
-multi-dimensional complexity of overlapping drivers to areas exposed to similar
-suites of drivers [@bowler2019]. This may prove critical for a better
-understanding the state of species, habitats and ecosystems located within or
-moving through threat complexes and exposed to the combined effects of all
-drivers typical to those areas.
-
-Six distinct threat complexes were identified in the St. Lawrence using
-the *k-medoids* algorithm (Figures S4, S5). Based on their
-distribution and representative drivers, threat complexes can be divided into
-3 offshore and 3 coastal complexes (Figures \ref{cluster}, S6 and S7).
-Coastal threat complexes (1 to 3; Figure \ref{cluster}) include all types
-of drivers besides hypoxia and are the most exposed threat complexes, both in
-terms of driver overlap and intensity. Threat complex 1 encompasses the
-coastline and is characterized by higher direct human impact (*i.e.* population
-density). Threat complex 2 is differentiated from other complexes by the
-presence of aquaculture sites. Threat complex 3 is the most exposed complex and
-has a distribution similar to the most exposed coastal hotspots (Figure
-\ref{hotspot}). This complex is characterized by high intensities of land-based
-drivers (*e.g.* nutrient input), demersal non-destructive high-bycath fisheries
-(*e.g.* trap fishing), climate drivers and marine traffic drivers in
-the vicinity of ports.
-
-Offshore threat complexes (4 to 6; Figure \ref{cluster}) are generally
-characterized by high intensity climate and marine traffic drivers.
-Threat complex 4 is differentiated by demersal non-destructive high-bycath
-fisheries, higher marine traffic drivers compared to complex 5 and generally
-corresponds to the whole Southern Gulf. Threat complex 5 is characterized by
-more fisheries types (*i.e.* demersal destructive and pelagic high-bycatch),
-generally lower intensity marine traffic drivers and is located almost
-exclusively in the Northern Gulf. Finally, threat complex 6 is the most exposed
-offshore threat complex and includes all offshore hotspots (Figure \ref{hotspot}).
-It is characterized by high intensity hypoxia, marine traffic and
-pollution, as well as demersal destructive and pelagic high-bycatch fisheries.
-This threat complex corresponds primarily to the Laurentian Channel and
-incorporates parts of the Esquiman and Anticosti Channels.
-
-Of particular concern are threat complexes 3 and 6, which are the two most
-exposed complexes in the St. Lawrence and are characterized by distinct
-cumulative exposure regimes Between them, they capture most of the coastal
-and offshore hotspots identified in the St. Lawrence.
-They also offer some insight into the potential importance of considering
-spatial dynamics in areas intersecting multiple threat complexes. For example,
-threat complexes 3 and 6 meet at the mouth of the River Saguenay. This area
-is particularly dynamic, with deep Atlantic waters advected through estuarine
-circulation mixing with surface waters from the St. Lawrence River and the
-Saguenay River [@dufour2007]. This results in the convergence of climate
-drivers from the bottom of the Laurentian Channel and marine traffic drivers
-(threat complex 6) with terrestrial run-off from river outflows and direct
-human impacts (*i.e.* population density; threat complex 3).
-
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-# Open Knowledge Platform: *eDrivers*
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-
-Sharing the knowledge acquired through the description of drivers in the
-St. Lawrence emerged as a priority to curtail the need to reach dozens
-of experts across multiple organizations and over extensive periods of time to
-assemble the data needed to apply integrated research and management. It is also
-a requirement to ensure that this manuscript will not quickly become an outdated
-snapshot of drivers distribution and intensity in the St. Lawrence System,
-but rather serve as a stepping stone towards an adaptive and ever-improving
-collection of knowledge.
-
-As such, we are launching *eDrivers*, an open knowledge platform focused on
-sharing knowledge on the distribution and intensity of drivers and on gathering
-a community of experts committed to structuring, standardizing and sharing
-knowledge on drivers in support of science and management. In launching this
-initiative, our objective is to uphold the highest existing standards of data
-management and open science. We identified four guiding principles to meet this
-objective and that guide the structure of the platform (Figure \ref{diag}).
-
-## Unity and inclusiveness
-
-***Why***: Operating over such large scales in time, space and subject matter
-requires a vast and diverse expertise that cannot possibly be possessed by any
-one individual or organization. Consequently, we envision an initiative that
-seeks to mobilize all individuals and entities with relevant expertise.
-
-***How***: By promoting, consolidating and working with experts involved in
-existing and highly valuable environmental initiatives already in place in the
-St. Lawrence. Notable examples of environmental initiatives are the annual
-review of physical [@galbraith2018], chemical, and biological [@blais2019]
-oceanographic conditions in the St. Lawrence, the fisheries monitoring program
-[@dfo2016], the annual groundfish and shrimp multidisciplinary survey
-[@bourdages2018], the characterization of benthic [@dutil2011], epipelagic and
-coastal [@dutil2012] habitats of the St. Lawrence, and Canada's shoreline
-classification [@eccc2018]. There are also nascent efforts to share
-information on several human activities in the St. Lawrence such as the Marine
-Spatial Data Infrastructure portal, which provides data on zoning, shipping,
-port activities, and other human activities in Canadian waters, including the
-St. Lawrence system [@governmentofcanada2018].
-
-By working with existing data portals whose objective is to share
-environmental data. We are thus collaborating actively with the St. Lawrence
-Global Observatory (SLGO) to develop the initiative and to host the
-platform on their web portal. The mission of SLGO is to
-promote and facilitate the accessibility, dissemination and exchange of official
-and quality data and information on the St. Lawrence ecosystem through the
-networking of organizations and data holders to meet their needs and those of
-users, to improve knowledge and to assist decision-making in areas such as
-public safety, climate change, transportation, resources and biodiversity
-conservation. SLGO is also one of three regional associations spearheading
-the Canadian Integrated Ocean Observing System (CIOOS[^cioos]),
-which will focus on integrating oceanographic data from multiple sources to
-make them accessible to end-users and to enable the national coordination of
-ocean observing efforts by integrating isolated or inaccessible data, and by
-identifying gaps or duplications in observations and research efforts.
-We are also developing collaborations with the Portal on water knowledge[^melcc],
-an initiative from the Québec provincial government. This portal aims at
-collecting and sharing accurate, complete and updated resources on water and
-aquatic ecosystems to support the mandate of relevant actors and stakeholders
-working in water and aquatic ecosystems management.
-
-[^cioos]: https://cioos.ca
-[^melcc]: http://www.environnement.gouv.qc.ca/eau/portail/
-
-By actively inviting, seeking, and developing collaborations as well as
-encouraging constructive criticism from the inception and throughout the
-lifetime of the platform.
-
-By inviting external community contributions (Figure \ref{diag}).
-External researchers or entities wishing to submit marine data will be able to
-do so through SLGO web portal. Submissions through
-other data portals will also be accepted either through the development of data
-sharing agreements or with the caveat that shared data are under an open source
-license and that they adhere to the platform data standards.
-
-## Findability, accessibility, interoperability and reusability
-
-***Why***: Open knowledge has been propelled to the forefront of scientific research
-in an era of open, collaborative and reproducible science. By moving towards
-large scale, cross-disciplinary research and management projects, there is a
-growing need to increase the efficiency of data discovery, access,
-interoperability and analysis [@reichman2011; @wilkinson2016]. Our goal is to
-foster efficient and functional open science by creating a fully open,
-transparent and replicable open knowledge platform.
-
-***How***:
-By building an infrastructure adhering to the FAIR Data Principles,
-which states that data and metadata must be Findable, Accessible, Interoperable
-and Reusable. These principles focus on the ability of humans and machines to
-automatically find and (re)use data and knowledge [@wilkinson2016].
-
-By making data and associated tools accessible through a variety of ways:
-the SLGO web portal, two R packages called *eDrivers*[^eDR] and *eDriversEx*[^eDX]
-to access the data through SLGO's API and to provide analytical tools to explore
-data, respectively, and a Shiny application[^eDapp] to explore drivers data
-interactively (Figure \ref{diag}). Note that the data are currently contained
-within and accessible through the *eDrivers* R package only, as we are actively
-working to allow users to download selected layers from SLGO's web portal and
-geoserver. The functions available in *eDrivers* to access the data have however
-been developed to ensure forward compatibility once the data migrate to SLGO's
-geoserver.
-
-[^eDR]: https://github.com/orgs/eDrivers/eDrivers
-[^eDX]: https://github.com/orgs/eDrivers/eDriversEx
-[^eDapp]: https://david-beauchesne.shinyapps.io/eDriversApp/
-
-By defining clear data and metadata standards and specifications to support
-the regional standardization of current and future protocols and practices and
-to favour interoperability with national and international initiatives like the
-Essential Ocean Variables (EOV) identified by the Global Ocean Observing System
-(GOOS). As such, we will adopt the metadata standard
-currently targeted for CIOOS, *i.e.* the North American Profile
-of ISO 19115:2014 - Geographic information - Metadata, a schema favoured
-for geospatial data in Canada and the United States.
-
-By providing version control and code access to the workflows set up to generate
-driver layers from raw data, the R packages and the Shiny application through
-a GitHub organization called *eDrivers*[^eD].
-
-[^eD]: https://github.com/orgs/eDrivers/
-
-## Adaptiveness
-
-***Why***: In the face of uncertainty and in an effort to address impending
-environmental changes, adaptive management has been identified as the chief
-strategy to guide efficient decision-making
-[*e.g.* @margules2000; @keith2011; @jones2016; @chion2018] and has already
-been discussed in the context of multi-drivers and cumulative impact assessments
-[@halpern2015a; @beauchesne2016; @cote2016; @schloss2017]. Adaptive management can only be
-truly achieved through a commitment to adaptive monitoring and data reporting
-[@margules2000; @halpern2012; @lubchenco2015]. We further contend that adaptive
-management requires the development of adaptive monitoring tools and
-infrastructures, which we seek to address through a continuously-evolving platform.
-
-***How***:
-By setting up mechanisms structuring cyclic reviews of platform content,
-for the integration of new material (*e.g.* data and methods) as it becomes
-available or accessible, and by striving to provide time-series data that are
-crucial to assess temporal trends and potentially early-warning signals of
-ecosystem change (Figure \ref{diag}).
-
-## Recognition
-
-***Why***: Like peer-reviewed publications, data must also be given its due
-importance in scientific endeavours and thus be considered as legitimate citable
-products contributing to the overall scientific output of data providers
-[@taskgroupondatacitationstandardsandpractoutofciteoutofmindthecurrentsices2013; @force112014].
-Appropriate citations should therefore be provided for all data layers used and
-shared by the platform.
-
-***How***: By adhering to the Data Citation Principles [@force112014], which
-focus on citation practices that provide appropriate credit to data products.
-
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-# Perspectives
-<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
-
-Understanding how ecosystem state will be affected by global change requires
-a comprehensive understanding of how threats are distributed and interact in
-space and time, which in turn hinges on appropriate data tailored to
-multi-driver studies [@dafforn2016; @stock2018; @bowler2019]. In the St.
-Lawrence, we found that few areas are free of cumulative exposure and that the
-whole Estuary, the Anticosti Gyre and coastal southwestern Gulf are particularly
-exposed to cumulative drivers, especially close to urban areas. We also
-identified six geographically distinct threat complexes that display similar
-cumulative exposure regimes. These complexes reveal that coastal areas are
-particularly exposed to all types of drivers and that multiple drivers
-typically co-occur in space. These results allow us to efficiently identify
-areas in need of heightened scrutiny from a science and management perspective.
-
-Through *eDrivers*, these observations will be iteratively improved towards
-an increasingly robust assessment of cumulative exposure and threat complexes
-as gaps in knowledge are addressed or approaches to describe drivers are refined.
-Arguably, the most meaningful benefit anticipated from *eDrivers* will be the
-gain in efficient access to comparable data-based knowledge on the exposure of
-ecosystems to multiple threats. This could pay quick scientific and management
-dividends by drawing on the knowledge and efforts of a wide range of
-contributors, by expanding avenues of scientific inquiry, by decreasing overall
-effort duplication and research costs, and by increasing research efficiency
-[@franzoni2014].
-
-Critically, *eDrivers* will
-allow the scientific and governmental communities to identify key knowledge
-gaps that will assist in prioritizing and optimizing research efforts.
-Ultimately, we believe that *eDrivers* will operationalize evidence-based
-decision-making by streamlining data management and research, allowing science
-output to be available and interpretable on a time scale relevant to management
-[see @sutherland2004; @reichman2011]. The platform will thus greatly facilitate
-the application of broad scale, holistic research and management approaches such
-as marine spatial planning, ecosystem-based management, marine spatial planning
-and strategic environmental assessments [*e.g.* @rice2011; @halpern2015a; @jones2016].
-
-While the focus of our paper has been on the description of drivers in the
-St. Lawrence, drivers are but one of the 4 elements required to properly
-evaluate environmental impacts. This single knowledge node should be weaved with
-other, comparable, nodes required for integrated management approaches such as
-species distribution and marine protected areas. Ultimately, all of these
-knowledge nodes could be weaved together through social-ecological metanetworks
-[@dee2017].
-
-Significant effort is still needed to bring our vision to fruition. Foremost
-is to maintain our efforts to foster collaborations, develop platform
-content and identify key knowledge gaps. A fair and efficient organizational
-structure will be developed in order to manage *eDrivers* as a community and
-appropriate funding must be secured to continue building this community and
-ensure the long-term viability of the initiative, although the partnership with
-SLGO partly addresses this issue.
-
-Finally, terrestrial and coastal environments must be incorporated, as
-sources of stress within those habitats extend to the marine environments.
-Moreover, despite coastal areas being recognized as the most exposed to
-environmental threats, we continue to delineate
-terrestrial and marine realms, considering coastlines as an impermeable
-barrier. While there is a sensible rationale for this division, we must strive
-to eliminate it if we are to appropriately study and predict the
-impacts of global change [*e.g.* see @bowler2019].
-
-Despite the challenges and work ahead, we are hopeful that this initiative will
-be very successful. Ultimately, *eDrivers* represents a much needed solution to
-address important issues in data management that could radically shift broad
-scale research and management practices towards efficient, adaptive and holistic
-ecosystem-based management in the St. Lawrence and elsewhere in the world. All
-it requires to be successful is for the scientific and political communities to
-fully commit to open knowledge, adaptive monitoring and, most of all, an
-integrated vision of ecosystem management.
 
@@
 
-DB, RD, DG and PA conceived the manuscript and the underlying objectives. DB
+DB, RD, DG and PA conceived the manuscript and the underlying objective. DB
 prepared/formatted the data, performed the analyses, was in charge of technical
 developments and lead the drafting of the manuscript. All co-authors contributed
-to data, analyses and writing based on their respective expertise and
+to data, analases and writing based on their respective expertise and
 contributed to the revision of the manuscript.
@@
 
-Box 1. Code snippet demonstrating how to use the *eDrivers* R package to
-reproduce figure \ref{kernel}.
+Box 1. Code snippet demonstrating how to use the *eDrivers* to reproduce
+figure \ref{kernel} in R.
 
@@
 \centering
-\includegraphics{./Figures/egsl.png}
-\caption{Description of the Estuary and Gulf of St. Lawrence in Eastern Canada}
-\label{egsl}
-\end{figure}
-
-\newpage
-
-\begin{figure}[H]
-\centering
 \includegraphics{./Figures/footprint.png}
-\caption{Distribution of cumulative footprint in the St. Lawrence System.}
+\caption{Distribution of cumulative footprint in the Estuary and Gulf of St.
+Lawrence.
+The footprint is measured by summing the relative intensity of all drivers for each grid cell: $F_d = \sum_{i=1}^{n} D_i$.
+Driver layers were  log-transformed when their frequency distribution was non-normal.
+All driver layers were also normalized between 0 and 1 using the 99th quantile to allow for direct comparison of relative intensities between drivers and control for extreme values.}
 \label{footprint}
@@
 \includegraphics{./Figures/hotspot.png}
-\caption{Distribution of cumulative hotspots in the St. Lawrence System.}
+\caption{Distribution of cumulative hotspots in the Estuary and Gulf of St.
+Lawrence.
+Hotspots are defined as the number of drivers in each grid cell with scaled intensity contained over their respective 80th percentile.}
 \label{hotspot}
@@
 
+**DG** Pas nécessaire de toujours répéter les méthodes dans chacune des légendes
+
 \begin{figure}[H]
@@
 \caption{Distribution of threat complexes in the Estuary and Gulf of St.
-Lawrence (upper panel).
-Threat complexes (a term coined by Bowler et al., 2019) are areas with similar cumulative driver exposure regimes.
-Mean intensity of all coastal (red), climate (green), fisheries (blue) and marine traffic (purple) drivers within each threat complex (lower panel).
+Lawrence.
+Threat complexes are areas with similar cumulative driver exposure regimes (Bowler et al., 2019).
+Threat complexes were identified using a partitional clustering algorithm (Kaufman and Rousseeuw, 1990).
+The appropriate number of clusters ($k$) was tested using a range of values and validated by selecting the number of clusters that maximized the average silhouette width (Kaufman and Rousseeuw, 1990) and minimized the total within-cluster sum of squares (WSS).
+Six distinct threat complexes were identified in the St.
+Lawrence.
+The partitional clustering algorithm analysis were performed using the $cluster$ R packages (Maechler et al., 2018).
 Refer to the Supplementary Materials for more details.}

### 2019-05-01 Code/10-validatekmedoid.R

@@ -1,0 +1,38 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+k <- 2:10
+
+# ~~~~~~~~~~~~~~~~~~~ Data ~~~~~~~~~~~~~~~~~~~ #
+load('./data/aswMedoid.RData')
+load('./data/wssMedoid.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ Figure ~~~~~~~~~~~~~~~~~~~ #
+png('./figures/validation.png', width = 920, height = 920, res = 200, pointsize = 7)
+layout(matrix(1:2, ncol = 1))
+par(mar = c(5,5,2,1))
+# k-medoid
+  # Silhouette
+  plot(aswMedoid ~ k,
+       pch = 20,
+       ylab = 'Average Silhouette Width',
+       type = 'b',
+       frame = F,
+       xlab = '',
+       xlim = c(0,10),
+       cex = 1.5,
+       cex.axis = .8)
+  mtext(side = 3, 'k-medoids clustering', font = 2, line = .75)
+
+  # WSS
+  plot(x = c(1,k),
+       y = wssMedoid,
+       type="b",
+       frame = F,
+       xlab="Number of Clusters (k)",
+       ylab="Total within-clusters sum of squares",
+       pch = 20,
+       xlim = c(0,10),
+       cex = 1.5,
+       cex.axis = .8)
+dev.off()

### 2019-05-01 Code/13-cluster.R

@@ -6,10 +6,30 @@ source('./code/2-param.R')
 load('./data/k.RData')
 load('./data/clMed.RData')
 nCl <- length(unique(clMed))
+load('./data/simperMean.RData')
+source('./code/barPlot.R')
+
+
+# Driver intensity in compared clusters
+drInt1 <- lapply(simperMean, function(x) x[,'ava']) %>%
+          as.data.frame()
+drInt2 <- lapply(simperMean, function(x) x[,'avb']) %>%
+          as.data.frame()
+
+dat <- vector('list', nCl)
+dat[[1]] = data.frame(x = drInt1$X1_2)
+dat[[2]] = data.frame(x = drInt1$X2_3)
+dat[[3]] = data.frame(x = drInt1$X3_4)
+dat[[4]] = data.frame(x = drInt1$X4_5)
+dat[[5]] = data.frame(x = drInt1$X5_6)
+dat[[6]] = data.frame(x = drInt2$X5_6)
 
 
 # ~~~~~~~~~~~~~~~~~~~ CLUSTER ~~~~~~~~~~~~~~~~~~~ #
-png('./figures/cluster.png', width = 1280, height = 920, res = 300, pointsize = 6)
+png('./figures/cluster.png', width = 1280, height = 1280, res = 300, pointsize = 6)
+# jpeg('./figures/cluster.jpg', width = 1280, height = 1280, res = 300, pointsize = 6)
+mat <- matrix(c(1,1,1,1,1,1,2,3,4,5,6,7), nrow = 2, byrow = T)
+layout(mat, heights = c(1, .4))
 cols <- pal(nCl)[clMed]
 plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          cols       = c('#00000000',off,off),
@@ -20,7 +40,7 @@ plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          axes       = 1:4,
          northArrow = F,
          prj        = slmetaPrj('world'),
-         extent     = 'egslSimple')
+         extent     = extFig)
 plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
 
 # Legend
@@ -44,4 +64,13 @@ for(i in 1:k) {
   text(x = xmx+.007*xR, y = mean(c(ymn,ymx)), labels = paste('Threat complex', i), cex = .6, adj = c(0, .5))
 }
 
+# Add citie
+cityEGSL(prj = 4326)
+
[…]

### 2019-05-01 Code/18-egsl.R

@@ -1,0 +1,121 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/2-param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./data/bathy.RData') # Bathymetry
+data(canada)
+data(usa)
+canada <- st_transform(canada, crs = 4326)
+usa <- st_transform(usa, crs = 4326)
+bathy <- st_transform(bathy, crs = 4326)
+deph <- bathy$DEPH
+
+
+
+
+
+
+
+# ~~~~~~~~~~~~~~~~~~~ FOOTPRINT ~~~~~~~~~~~~~~~~~~~ #
+# Layout
+pal <- colorRampPalette(c('#E2FCF6', '#1B7CAA'))
+cols <- pal(101)[((deph/max(deph, na.rm = T))*100)+1]
+
+cexMain <- 1
+cexMain2 <- .75
+png('./figures/egsl.png', width = 1280, height = 920, res = 300, pointsize = 6)
+# jpeg('./figures/egsl.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
+# par(bg = colBack)
+plotEGSL(layers     = c('egslSimple', 'canada','usa'),
+         cols       = c('#00000000',off,off),
+         borders    = c('#00000000',focus,focus),
+         lwds       = c(.75,.75,.75),
+         mar        = c(2,2,2,2),
+         box        = T,
+         axes       = 1:4,
+         northArrow = F,
+         prj        = slmetaPrj('world'),
+         extent     = extFig)
+plot(st_geometry(bathy), add = T, col = cols, border = cols, lwd = .1)
+plot(st_geometry(egslSimple), col = 'transparent', border = focus, lwd = .75, add = T)
+plot(st_geometry(canada), col = off, border = focus, lwd = .75, add = T)
+plot(st_geometry(usa), col = off, border = focus, lwd = .75, add = T)
+box2("1234", lwd = 3)
+# Legend
+legendEGSL(range = c(0, 525),
+           pal = pal,
+           mainTitle = 'Bathymetry',
+           subTitle = 'meters',
+           cexMain = .9,
+           cexSub = .65,
+           n = 6)
+
+
[…]

### 2019-05-01 Code/2-param.R

@@ -75,3 +75,6 @@ gg_color_hue <- function(n) {
   hues = seq(15, 375, length = n + 1)
   hcl(h = hues, l = 65, c = 100, alpha = .5)[1:n]
 }
+
+# Extent 
+extFig <- as.numeric(st_bbox(egslSimple)) + c(-1.5,-.5,.5,.5)

### 2019-05-01 Code/4-footprint.R

@@ -11,7 +11,8 @@ Fd <- rowSums(dr)
 cols <- pal(101)[((Fd/max(Fd, na.rm = T))*100)+1]
 cexMain <- 1
 cexMain2 <- .75
-png('./figures/footprint.png', width = 1280, height = 920, res = 300, pointsize = 6)
+# png('./figures/footprint.png', width = 1280, height = 920, res = 300, pointsize = 6)
+jpeg('./figures/footprint.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
 # par(bg = colBack)
 plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          cols       = c('#00000000',off,off),
@@ -22,7 +23,7 @@ plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          axes       = 1:4,
          northArrow = F,
          prj        = slmetaPrj('world'),
-         extent     = 'egslSimple')
+         extent     = extFig)
 plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
 plot(st_geometry(egslSimple),
      col = 'transparent',

### 2019-05-01 Code/6-hotspot.R

@@ -32,6 +32,7 @@ nOver <- max(cumulHotspot, na.rm = T)
 
 # Graph
 png('./figures/hotspot.png', width = 1280, height = 920, res = 300, pointsize = 6)
+# jpeg('./figures/hotspot.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
 cols <- pal(nOver)[cumulHotspot]
 plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          cols       = c('#00000000',off,off),
@@ -42,7 +43,7 @@ plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          axes       = 1:4,
          northArrow = F,
          prj        = slmetaPrj('world'),
-         extent     = 'egslSimple')
+         extent     = extFig)
 plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
 plot(st_geometry(egslSimple),
      col = 'transparent',
@@ -59,5 +60,5 @@ legendEGSL(range = c(1, nOver),
            n = 6)
 
 # City names
-cityEGSL()
+cityEGSL(prj = 4326)
 dev.off()

### 2019-05-01 Code/7-kernel.R

@@ -4,6 +4,8 @@ source('./code/2-param.R')
 # -----
 # Figure
 png('./figures/kernel.png', width = 1920, height = 1280, res = 300, pointsize = 6)
+# jpeg('./figures/footprint.jpg', width = 1920, height = 1280, res = 300, pointsize = 6)
+
 
 # Layout
 mat <- matrix(0, 28, 32)

### 2019-05-01 Tables/datatable.R

@@ -13,11 +13,11 @@ df <- data.frame(Groups = character(0),
                  stringsAsFactors = F)
 
 df[1,] <- c("Climate", "Aragonite", 'Lat/long', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ACID')
-df[2,] <- c("Coastal", "Aquaculture", 'Lat/long', '-','Variable, between 1990-2016', '$presence-absence$','***TBD***','AQUA')
-df[3,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ negative anomalies', '[@galbraith2018]','SBT-')
-df[4,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ positie anomalies', '[@galbraith2018]','SBT+')
-df[5,] <- c("Climate",  "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ negative anomalies', '[@galbraith2018]','SST-')
-df[6,] <- c("Climate", "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', '$n$ positive anomalies', '[@galbraith2018]','SST+')
+df[2,] <- c("Coastal", "Aquaculture", 'Lat/long', '-','Variable, between 1990-2016', '$presence-absence$','[@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016]','AQUA')
+df[3,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SBT-')
+df[4,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'positie anomalies', '[@galbraith2018]','SBT+')
+df[5,] <- c("Climate",  "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SST-')
[…]

### 2020-02-05 Submission/Beauchesne-eDriversMS.Rmd
commit all changes that were made for paper submission

--- 2019-05-01 eDriversMS-v2.Rmd
@@
     number_sections: true
+    keep_tex: true
   word_document:
@@
 Mean intensity of all coastal (red), climate (green), fisheries (blue) and marine traffic (purple) drivers within each threat complex (lower panel).
-Refer to the Supplementary Materials for more details.}
+Refer tp Table 1 for acronym meaning and to the Supplementary Materials for more details.}
 \label{cluster}

### 2020-02-05 Tables/datatable.R

@@ -9,7 +9,7 @@ df <- data.frame(Groups = character(0),
                  Years = character(0),
                  Units = character(0),
                  Source = character(0),
-                 Accronym = character(0),
+                 Acronym = character(0),
                  stringsAsFactors = F)
 
 df[1,] <- c("Climate", "Aragonite", 'Lat/long', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ACID')
@@ -41,18 +41,18 @@ df[22,] <- c("Coastal", "Toxic algae", '-', '-', '-', 'Expert based', '[@bates20
 df <- df[order(df$Groups, df$Drivers), ]
 
 # Table 1
-table1 <- kable(x = df[, c("Groups", "Drivers", 'SpatRes', "TempRes", 'Years', "Units", "Source")],
-      col.names = c("Groups", "Drivers", 'Spatial resolution', "Temporal resolution", 'Years', "Units", "Source"),
+table1 <- kable(x = df[, c("Groups", "Drivers", 'Acronym', 'SpatRes', "TempRes", 'Years', "Units", "Source")],
+      col.names = c("Groups", "Drivers", 'Acronym', 'Spatial resolution', "Temporal resolution", 'Years', "Units", "Source"),
       row.names = F,
-      align = c('l','l','l','l','l','l','l'),
+      align = c('l','l','l','l','l','l','l','l'),
       format = 'markdown'
      )
-table1[[2]] <- "|:--------|:---------------|:----------|:----------|:----------|:----------|:--------------------|"
+table1[[2]] <- "|:--------|:-------------|:-------|:----------|:----------|:----------|:----------|:--------------------|"
 save(table1, file = './tables/table1.RData')
 
 # Table S1
-tableS1 <- kable(x = df[, c("Groups", "Drivers", "Accronym","Source")],
-      col.names = c("Groups", "Drivers", "Accronym","Source"),
+tableS1 <- kable(x = df[, c("Groups", "Drivers", "Acronym","Source")],
+      col.names = c("Groups", "Drivers", "Acronym","Source"),
       row.names = F,
       align = c('l','l','l','l'),
       format = 'markdown'

### 2020-02-06 Revisions/Beauchesne-eDriversMS-SI.rmd
setup files for revisions

--- 2019-05-01 eDriversMS-SI-v2.rmd
@@
     number_sections: true
+    keep_tex: true
   html_document:

### 2020-02-06 Revisions/Beauchesne-eDriversMS.Rmd
setup files for revisions
first draft of answers to reviewer comments and set to do list to modify manuscript accordingly

--- 2019-05-01 eDriversMS-v2.Rmd
@@
    - \usepackage{float}
+   - \usepackage{ulem}
+   - \definecolor{blue}{HTML}{20cad5}
 relativeurls: true
-bibliography: eDriversMS.bib
-csl: frontiers.csl
+bibliography: Beauchesne-eDriversMS.bib
+csl: ../frontiers.csl
 link-citations: yes
@@
 <!--
-cd phdproj/eDrivers/eDriversMS/
-rmarkdown::render('./eDriversMS-v2.rmd', 'pdf_document')
+rmarkdown::render('./Revisions/Beauchesne-eDriversMS.Rmd', 'pdf_document')
 rmarkdown::render('./eDriversMS-v2.rmd', 'word_document')
@@
 with the clustering analysis and to ensure that outliers did not have a strong
-influence the analysis. As the drivers dataset is large (~ 250000 observations),
+influence \textcolor{blue}{on} the analysis. As the drivers dataset is large (~ 250000 observations),
 we used a bootstrap procedure for the SIMPER analysis, randomly selecting
@@
 
-Of particular concern are threat complexes 3 and 6, which are the two most
-exposed complexes in the St. Lawrence and are characterized by distinct
-cumulative exposure regimes Between them, they capture most of the coastal
-and offshore hotspots identified in the St. Lawrence.
-They also offer some insight into the potential importance of considering
+\textcolor{red}{\sout{Of particular concern are threat complexes 3 and 6, which are the two most exposed complexes in the St.
+Lawrence and are characterized by distinct cumulative exposure regimes Between them, they capture most of the coastal and offshore hotspots identified in the St.
+Lawrence.}}
+\textcolor{blue}{Threat complexes 3 and 6 capture most coastal and offshore hotspots and are the two most exposed complexes in the St. Lawrence.}
+They \textcolor{red}{\sout{also}} offer some insight into the potential importance of considering
 spatial dynamics in areas intersecting multiple threat complexes. For example,
@@
 \centering
-\includegraphics{./Figures/egsl.png}
+\includegraphics{../Figures/egsl.png}
 \caption{Description of the Estuary and Gulf of St. Lawrence in Eastern Canada}
@@
 \centering
-\includegraphics{./Figures/footprint.png}
+\includegraphics{../Figures/footprint.png}
 \caption{Distribution of cumulative footprint in the St. Lawrence System.}
@@
 \centering
-\includegraphics{./Figures/hotspot.png}
+\includegraphics{../Figures/hotspot.png}
 \caption{Distribution of cumulative hotspots in the St. Lawrence System.}
@@
 \centering
-\includegraphics{./Figures/kernel.png}
+\includegraphics{../Figures/kernel.png}
 \caption{Interaction between the intensity of hypoxia and demersal destructive fisheries in the St.
@@
 \centering
-\includegraphics{./Figures/cluster.png}
+\includegraphics{../Figures/cluster.png}
 \caption{Distribution of threat complexes in the Estuary and Gulf of St.
@@
 Mean intensity of all coastal (red), climate (green), fisheries (blue) and marine traffic (purple) drivers within each threat complex (lower panel).
-Refer to the Supplementary Materials for more details.}
+Refer \textcolor{red}{\sout{tp}} \textcolor{blue}{to} Table 1 for acronym meaning and to the Supplementary Materials for more details.}
 \label{cluster}
@@
 \centering
-\includegraphics{./Figures/structure.png}
+\includegraphics{../Figures/structure.png}
 \caption{Diagram of the platform structure.
@@
 ```{r table1, echo=F, fontsize=4}
-load('./tables/table1.RData')
+load('../tables/table1.RData')
 table1

### 2020-02-14 Revisions/Rebuttal.md
first complete draft of rebuttal

---
title: "Rebuttal"
fontfamily: fourier
linestretch: 1
fontsize: 10pt
output:
  pdf_document:
    latex_engine: xelatex
---

<!--
rmarkdown::render('./Revisions/Rebuttal.md')
-->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Letter to the editor
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

Dear Dr. Otto,

We are submitting a revised version of the manuscript "Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System" (ID: 469411) for publication in Frontiers in Marine Science. We would like to thank the reviewer for their very constructive comments. We feel that we were able to address all of their comments and are now submitting an improved manuscript. Below we explain how we have revised our manuscript based on the reviews.

We thank you for considering our revised manuscript and I am looking forward to hearing from you.

Best regards,

David Beauchesne

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Question 1
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

***Please summarize the main findings of the study***

*This article presents 22 environmental drivers and uses these environmental drivers to identify areas of high cumulative exposure and characterize areas with similar cumulative exposure in the St. Lawrence System. In particular, the article introduces an open knowledge platform named eDrivers, though it doesn’t seem to be working at this moment (https://github.com/eDrivers - this site contains zero repositories, zero people, zero projects; the R package dDrivers is not available on CRAN). When this R package and the website become available, I believe the article and its related sources will be beneficial to the scientific and management communities. Despite its weaknesses (shown below), I believe this work is a good start for assessing and managing cumulative effects in the St. Lawrence System and can serve as a good example for other marine ecosystems.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
**Answer:**

> - The material has yet to be released publicly on GitHub as we were waiting for the review process to be completed. We can however provide access to the repositories in the interim.

> - As for the package, we will not be releasing a version on CRAN just yet. The material is not yet implemented on the SLGO servers (our partners on this project) and the data is currently contained directly in the package. Our goal is to publish the package and submit it for peer-review once the data is available and accessible through SLGO's API.
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Question 2
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

***Please highlight the limitations and strengths***

*Weaknesses (and potential solutions) are listed below:*

***Comment 1:***

*Authors need to make it clear up in the front that this work is only focusing on one of the four elements required to evaluate environmental impacts. Besides moving Lines 119-122 to around Line 77, it will be helpful to introduce the best practices in cumulative effects assessment and management: (1) selecting valued (ecosystem) components, (2) selecting indicators, (3) setting temporal and spatial scales, (4) addressing uncertainties, and (5) public participation. It will also be helpful to differentiate the two different indicators: effects-based and stressor-based. Effects-based indicators are direct measurements of the valued components (such as species abundance, biodiversity, etc.) and inherently encompass cumulative effects of multiple drivers. However, because of the poorly understood cause-effect linkages between stressors and valued components, it is currently hard to be used for decision-making. In comparison, stressor-based indicators (i.e., driver-based as this article defines and the type that this article uses) are usually better understood and can be proactively linked to management actions. However, these indicators do not capture the effects of all stressors and are not able to discern non-additive cumulative effects (e.g. synergism, antagonism). This point leads to the next weakness of this article.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - We agree with reviewer 1 on the importance of clearly stating that we are solely focusing on drivers in our manuscript. The goal of our manuscript and the *eDrivers* initiative is to facilitate the widespread accessibility of such data for environmental assessments and integrative management. It is therefore clear to us that what we are presenting is but one element of a fully operational impact assessment. Our goal is however not to discuss cumulative effects assessments in particular. It is rather to provide knowledge on an element common to most holistic approaches, *i.e.* drivers characterization. We have strived to make this as clear as possible throughout the manuscript, namely in the introduction and in the perspectives sections, as noted by reviewer 1. As such, and considering the focus of the manuscript, we are hesitant to accept the suggestion of reviewer 1 to discuss cumulative effects assessment, best practices and indicator types at the beginning of the introduction. We feel that, considering the objective of this manuscript, the beginning of the introduction should rather focus on drivers specifically.

> - We nonetheless agree with reviewer 1's suggestions. We are therefore proposing the following modifications to the manuscript that we believe consolidates the essence of the suggestions made by reviewer 1:

  > 1. We modified the introduction to broadly state the major steps undertaken in holistic environmental approaches in general (lines 91 to 96).

  > 2. We modified the statement of the objectives of the manuscript at the end of the introduction to clearly state that we are focusing on drivers (lines 119 to 124).

  > 3. We modified the pre-existing paragraph presenting the importance of valued components in the perspective sections (lines 487 to 509). It is now presenting *eDrivers* as a suite of stressor-based indicators that hold certain advantages, but fail to capture effects on valued components and non-additive effects. We present effects-based indicators and their importance for understanding cumulative effects. We then argue that both are jointly required to properly diagnose causes of ecosystem change (Dube 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

***Comment 2:***

*Because of this inherent weakness of stressor-based indicators, the dealing and conclusion of this article related to driver interactions (Sections 3.4 and 4.2) are inadequate. How different stressors (or drivers) interact to create nonlinear, non-additive cumulative effects are dependent on a specific valued component of interest. This article simply assumes the summed intensity of two drivers as interactions, which is inaccurate and misleading. Consequently, the results based on the two drivers (hypoxia and fisheries) are not convincing (Section 4.2). How to create indicators that truly reflect negative impacts is essential and can be specific to a specific value component. For instance, temperature positive anomalies may have negative impact on certain species, it may not necessarily degrade ecosystem production. While the calculation of drivers intensity in Appendix is only one way of doing it, it is important for the computer package to provide users with flexibility to define their own indicators (i.e., one potential solution to the weakness).*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - We begin by thanking the reviewer for pointing out this obvious weakness in the manuscript. We realize now that, even though we did not want to mislead readers in thinking we were evaluating driver interactions, we are in fact doing just that. As mentionned in reviewer 1 comments, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects. Under the assumption that driver co-occurrence is necessary for driver interactions to exist, we believe it is reasonable to predict that driver interactions and non-additive effects will most likely be observed in areas where drivers overlap at high relative intensities. These areas indeed delineate where the effects of both drivers are most likely to impact valued components simultaneously. In presenting sections 3.4 and 4.2, our goal was to provide an illustration of how a simple two-drivers case could be used to identify areas of high relative intensity driver overlap. This should be used as segway to the more integrative cumulative footprint and hotspot indicators. As rightfully identified by reviewer 1, however, the manuscript is indeed misleading on the topic of driver interactions. We have therefore made the following modifications to the manuscript:

  > 1. We removed objective 2 from the manuscript and mentions of driver interactions, except in the context of discussing the likelihood on observing interactions in areas of driver co-occurrence.

  > 2. The methodology and results/discussions sections on driver interactions were moved to sections 3.3 and 4.1, respectively, and renamed *Driver co-occurrence*. We then modified the section as a simplified example of driver co-occurrence (lines XXX to XXX). We however did not modify the results/discussion section. The only mention of driver interactions was in the title itself. In the discussion, we merely discuss the areas of high intensity co-occurrence between the two drivers and present adverse effects that these drivers may have on species individually or simultaneously. We did however add a closing statement to the section stating that individual and interactive effects of both drivers is species- and community-dependent and that the co-occurrence of drivers in itself is not an evidence of driver interaction. We then state that assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities (lines XXX to XXX)

> - Ultimately, this portion of the manuscript should serve as a simple illustration to better understand the complete cumulative footprint indicator. In this sense, if the reviewer still feels that proposed changes are not sufficient, we propose removing these sections from the manuscript.

> - As for providing flexibility in indicator definition, it is a feature that we wish to provide in future iterations of the package and the platform. We would however rather not include such enhancements at this point, as they are more technical in nature and should not preclude us from publishing this work. We however added a statement to the manuscript that presents this idea as a perspective and future work to be accomplished (lines XXX to XXX in perspectives).
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

***Comment 3:***

*Because of the paramount importance of defining valued components in the process of cumulative effects assessment and management, the absence of this element is the third weakness of this article. This weakness can potentially be overcome by adding “Future work”. In the future, similar work can be carried out to select and mapping valued components, which will allow explicit linkages between drivers and value components to be explored for better understanding cumulative effects.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - We agree with reviewer 1 on the importance of valued components for cumulative effects assessments. Presenting a full cumulative impacts assessments and defining valued components is however not the objective of this manuscript, nor is it to discuss cumulative impacts in particular. The choices of valued components are fully dependent on the objective of a particular assessment or management scheme. In our experience, time and money is spent on recurring initiatives that seek to characterize sources of stress in a system for impacts assessments and integrative management. This is, in our view, unacceptable. Our goal was therefore to build an initiative that would limit this and provide access to formatted and comparable knowledge on sources of stress that are typically used for impact assessments and integrative management. This integrative dataset can then be used and adapted to the need of a user to assess impacts on a single or multiple valued components of interest, or the user can identify gaps in knowledge in the platform and ultimately contribute to the knowledge contained in the platform. We were therefore very deliberate in our decision not to include and discuss specific valued components. This manuscript is in fact part of a broader project that aims at evaluating cumulative impact on the ecological communities of the St. Lawrence. The description of drivers will thus be used on specific valued components in the St. Lawrence System.

> - As proposed by reviewer 1, we had already included a paragraph on the importance of valued components and their vulnerability to soures of stress in the conclusion. We however realize now that it was underwhelming in its ability to provide perspective and direction to how eDrivers may be used in the future. We therefore extended this paragragh, also in response to reviewer 1's first comment, to better discuss the significance of valued components and the different types of indicators for cumulative effects assessment (lines 487 to 509).
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Question 3
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

***Please provide your detailed review report to the editor and authors (including any comments on the Q4 Check List):***

*This article is generally well written, presenting 22 environmental drivers. It uses these environmental drivers to identify areas of high cumulative exposure and characterize areas with similar cumulative exposure in the St. Lawrence System. In particular, the article introduces an open knowledge platform named eDrivers, though it doesn’t seem to be working at this moment (https://github.com/eDrivers - this site contains zero repositories, zero people, zero projects; the R package dDrivers is not available on CRAN). When this R package and the website become available, I believe the article and its related sources will be beneficial to the scientific and management communities. Despite its weaknesses (shown below), I believe this work is a good start for assessing and managing cumulative effects in the St. Lawrence System and can serve as a good example for other marine ecosystems.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - See answer to question 1
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

*Weaknesses (and potential solutions) are listed below:*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Comment 1:***

*Authors need to make it clear up in the front that this work is only focusing on one of the four elements required to evaluate environmental impacts. Besides moving Lines 119-122 to around Line 77, it will be helpful to introduce the best practices in cumulative effects assessment and management: (1) selecting valued (ecosystem) components, (2) selecting indicators, (3) setting temporal and spatial scales, (4) addressing uncertainties, and (5) public participation. It will also be helpful to differentiate the two different indicators: effects-based and stressor-based. Effects-based indicators are direct measurements of the valued components (such as species abundance, biodiversity, etc.) and inherently encompass cumulative effects of multiple drivers. However, because of the poorly understood cause-effect linkages between stressors and valued components, it is currently hard to be used for decision-making. In comparison, stressor-based indicators (i.e., driver-based as this article defines and the type that this article uses) are usually better understood and can be proactively linked to management actions. However, these indicators do not capture the effects of all stressors and are not able to discern non-additive cumulative effects (e.g. synergism, antagonism). This point leads to the next weakness of this article.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - See answer to question 2
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

***Comment 2:***

*Because of this inherent weakness of stressor-based indicators, the dealing and conclusion of this article related to driver interactions (Sections 3.4 and 4.2) are inadequate. How different stressors (or drivers) interact to create nonlinear, non-additive cumulative effects are dependent on a specific valued component of interest. This article simply assumes the summed intensity of two drivers as interactions, which is inaccurate and misleading. Consequently, the results based on the two drivers (hypoxia and fisheries) are not convincing (Section 4.2). How to create indicators that truly reflect negative impacts is essential and can be specific to a specific value component. For instance, temperature positive anomalies may have negative impact on certain species, it may not necessarily degrade ecosystem production. While the calculation of drivers intensity in Appendix is only one way of doing it, it is important for the computer package to provide users with flexibility to define their own indicators (i.e., one potential solution to the weakness).*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - See answer to question 2
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

***Comment 3:***

*Because of the paramount importance of defining valued components in the process of cumulative effects assessment and management, the absence of this element is the third weakness of this article. This weakness can potentially be overcome by adding “Future work”. In the future, similar work can be carried out to select and mapping valued components, which will allow explicit linkages between drivers and value components to be explored for better understanding cumulative effects.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - See answer to question 2
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

*Comment on the methods, results and data interpretation: The method on dealing with driver interactions is inadequate, and the resultant conclusions with regard to interactions are not convincing.*

***Answer:***

> - See answer to question 2

***Minor comments:***

*Line 229: add “on” before the analysis.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - This was modified in the text (lines XXX to XXX).
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

*Lines 327-328: fix language issue.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - We have modified the sentences on lines 327-328 (now lines XXX to XXX).
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

*Figure 5 caption: tp becomes to*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - Figure caption was modified accordingly (line XXX).
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

*Please check Appendix carefully for language errors. In addition, Appendix can be tightened up.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - We have performed a thorough review of the material presented in the Appendix.

TODO:

- [ ] review appendix
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Question 4
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

***Please comment on the methods, results and data interpretation. If there are any objective errors, or if the conclusions are not supported, you should detail your concerns.***

*The method on dealing with driver interactions is inadequate, and the resultant conclusions with regard to interactions are not convincing.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - See answer to question 2
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Question 5
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

*Is the English language of sufficient quality?
- Yes*

Is the quality of the figures and tables satisfactory?
- Yes

Does the reference list cover the relevant literature adequately and in an unbiased manner?
- Yes

Are the statistical methods valid and correctly applied? (e.g. sample size, choice of test)
- Not Applicable

Are the methods sufficiently documented to allow replication studies?
- Yes

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - All the code to reproduce the driver layers and the analyses of the paper will be openly available once the GitHub repositories are made public.

TODO:

- [ ] Make GitHub repositories public
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

Are the data underlying the study available in either the article, supplement, or deposited in a repository? (Sequence/expression data, protein/molecule characterizations, annotations, and taxonomy data are required to be deposited in public repositories prior to publication) (yes/no/not applicable)
- No

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - The formatted driver data will be available through the package and the raw data are now archived on Zenodo (link XXX) and accessible upon request to the authors.

TODO:

- [ ] Archive rawdata on Zenodo.
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

Does the study adhere to ethical standards including ethics committee approval and consent procedure?
- Not Applicable

Have standard biosecurity and institutional safety procedures been adhered to?
- Not Applicable

### 2020-02-14 Revisions/Beauchesne-eDriversMS.Rmd
first complete draft of rebuttal

@@
 bibliography: Beauchesne-eDriversMS.bib
-csl: ../frontiers.csl
+csl: frontiers.csl
 link-citations: yes
@@
 major investments recently made and more forthcoming in economic and
-infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018],
-an intensification of the human footprint is likely in the St. Lawrence System.
+infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018], an intensification of the human footprint is likely in the St. Lawrence System.
 
@@
 
-The uncertainty associated with complex driver interactions must therefore be
-taken into account when investigating environmental impacts [@cote2016], yet
-most research on driver effects in marine environments remains overwhelmingly
-focused on single driver assessments [@obrien2019]. Increasing exposure and the
-experiences of past ecological tragedies such as the collapse of cod fisheries
-[@frank2005; @dempsey2018] and the decline of the beluga and right whale
-populations [@plourde2014] together urge the need to characterize the distribution,
-intensity and overlap between drivers in the St. Lawrence System. This will
-provide critical information on areas most exposed to cumulative drivers and on
-the interaction potential of drivers in the St. Lawrence. It is also a necessary
-step towards the holistic and integrated management of the St. Lawrence System.
+The uncertainty associated with complex driver interactions must therefore be taken into account when investigating environmental impacts [@cote2016], yet most research on driver effects in marine environments remains overwhelmingly focused on single driver assessments [@obrien2019].
+Increasing exposure and the experiences of past ecological tragedies such as the collapse of cod fisheries [@frank2005; @dempsey2018] and the decline of the beluga and right whale populations [@plourde2014] together urge the need to characterize the distribution, intensity and overlap between drivers in the St.
+Lawrence System.
+This will provide critical information on areas most exposed to \textcolor{red}{\sout{cumulative}} drivers \textcolor{red}{\sout{and on the interaction potential of drivers}} in the St.
+Lawrence.
+It is also a necessary step towards the holistic and integrated management of the St.
+Lawrence System.
 
-Gathering environmental data for large scale, systematic initiatives can,
-however, be a very challenging and time consuming – not to say painful – process.
+\textcolor{blue}{Gathering knowledge is a major component of any holistic approach, which typically involve the selection and description of valued ecosystem component (\textit{e.g.} habitats or species) and driver indicators, the assessment of the exposure and vulnerability of valued components to drivers, the selection of a proper spatio-temporal scale, site monitoring, and public and stakeholder participation (Dubé and Munkittrick 2001).} <!-- [@dube2001] jones2016 ????
+Noble2013???
+-->
+Gathering environmental \textcolor{red}{\sout{data}} \textcolor{blue}{knowledge} for large scale, \textcolor{red}{\sout{systematic}} \textcolor{blue}{holistic} initiatives can, however, be a very challenging and time consuming – not to say painful – process.
 On one hand, there is an overwhelming and expanding wealth of data available.
-Such information overload may inhibit our ability to make decisions based on
-scientific information, promote massive effort duplication, disproportionately
-appropriate research funds to certain sectors, and obscure knowledge gaps amid
-a sea of information [@eppler2004]. On the other hand, crucial data are lacking
-and remain largely unavailable or inaccessible for a variety of reasons,
-including proprietary rights, lack of organizational time, capacity and
-training, and in some rare cases unwillingness to share, curtailing our ability
-for appropriate decision-making.
+Such information overload may inhibit our ability to make decisions based on scientific information, promote massive effort duplication, disproportionately appropriate research funds to certain sectors, and obscure knowledge gaps amid a sea of information [@eppler2004].
+On the other hand, crucial data are lacking and remain largely unavailable or inaccessible for a variety of reasons, including proprietary rights, lack of organizational time, capacity and training, and in some rare cases unwillingness to share, curtailing our ability for appropriate decision-making.
 
@@
 intensity of drivers in the St. Lawrence. More specifically, we aim to: 1)
-identify areas of high cumulative exposure, 2) identify drivers that are likely
-to interact in the St. Lawrence and 3) characterize areas with similar
-cumulative exposure regimes.
-Here we focus solely on drivers, one of four elements required to evaluate
-environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.*
-habitats or species) and the exposure and the vulnerability of the ecosystem
-components to drivers. We achieve these objectives with the development of an
+identify areas of high cumulative exposure\textcolor{red}{\sout{, 2) identify drivers that are likely to interact in the St.
+Lawrence}} and \textcolor{red}{\sout{3)}} \textcolor{blue}{2)} characterize areas with similar cumulative exposure regimes.
+\textcolor{red}{\sout{Here we focus solely on drivers, one of four elements required to evaluate environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.* habitats or species) and the exposure and the vulnerability of the ecosystem components to drivers.}}
+\textcolor{blue}{Here we focus solely on drivers, aiming at better understanding their distribution in the St.
+Lawrence and at facilitating the widespread availability of their characterization for holistic assessments and management approaches.}
+We achieve these objectives with the development of an
 open knowledge platform named *eDrivers*. The platform was designed to
@@
 
+## \textcolor{red}{\sout{Driver interactions}} \textcolor{blue}{Driver co-occurrence}
+
+\textcolor{red}{\sout{Interactions among drivers were investigated using the cumulative footprint ($F$) between pairs of drivers (objective 2).
+The intensity at which pairs of drivers co-occur was evaluated using a two-dimensional kernel density estimate.
+As there are 231 pairwise combinations between 22 drivers, we focus on a single example using hypoxia and demersal destructive fisheries, two drivers known to occur mainly in deeper areas of the St.
+Lawrence and, hence, an interaction between the effects of the two drivers could be anticipated.}}
+\textcolor{blue}{As an introduction to indicators of cumulative exposure (objective 1), we begin by presenting a simplifed 2-driver example focusing on hypoxia and demersal destructive fisheries, two drivers known to occur mainly in deeper areas of the St.
+Lawrence.
+Driver co-ccurrence was evaluated spatially by summing the scaled intensity of drivers in each grid cell.
+The intensity at which pairs of drivers co-occur was evaluated using a two-dimensional kernel density estimate to investigate the relative intensity at which these two drivers typically co-occur.}
+
 ## Cumulative exposure
@@
 
-## Driver interactions
-
-Interactions among drivers were investigated using the cumulative footprint
-($F$) between pairs of drivers (objective 2). The intensity at which pairs of
-drivers co-occur was evaluated using a two-dimensional kernel density estimate.
-As there are 231 pairwise combinations between 22 drivers, we focus on a single
-example using hypoxia and demersal destructive fisheries, two drivers known to
-occur mainly in deeper areas of the St. Lawrence and, hence, an interaction
-between the effects of the two drivers could be anticipated.
-
 ## Threat complexes
 
-In order to better capture potential interactions between drivers (objective 2)
-and to identify areas with similar exposure regimes (objective 3),
-we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
-We use the term clusters in presenting the methods, but use threat complex when
-discussing the results on cumulative exposure regimes.
+In order to \textcolor{red}{\sout{better capture potential interactions between drivers (objective 2) and to}}identify areas with similar exposure regimes (objective 3), we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
+We use the term clusters in presenting the methods, but use threat complex when discussing the results on cumulative exposure regimes.
 
@@
 
+## \textcolor{red}{\sout{Driver interactions}} \textcolor{blue}{Driver co-occurrence}
+
+Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels, with the head of the Channels most exposed (Figure \ref{kernel}A).
+Demersal destructive fisheries are located along the Laurentian Channel, the heads of the Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B).
+By combining both drivers, we can observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C, Box 1).
+
+Fisheries in the St.
+Lawrence have historically affected biodiversity distribution and habitat quality [@moritz2015].
+Concurrently, hypoxia decreases overall habitat quality, but triggers species-dependent responses ranging from adaptation [*e.g.* northern shrimp *Pandalus borealis* and Greenland halibut *Reinhardtius hippoglossoides*; @pillet2016] to reduced growth rates [@dupont-prinet2013] and avoidance of oxygen-depleted habitats [*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality [*e.g.* sessile benthic invertebrates; @eby2005; @gilbert2007; @belley2010].
+Certain species may thus be adversely affected by fisheries and withstand hypoxia but still experience a decrease in prey availability, while others may be deleteriously affected by the compounded effect of both drivers [@deleo2017].
+
+\textcolor{blue}{The individual or interactive effect of both drivers is thus species- and community-dependent and the co-occurrence of drivers in itself is not an evidence of driver interaction.
+Assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities.
+}
+
 ## Cumulative exposure
@@
 
-While we cannot ascertain that high exposure areas are the most impacted,
-we can safely predict that these are the areas where studying ecosystem state
-will be the most complex due to the uncertainty associated with driver
-interactions, an uncertainty bound to increase rapidly with the
-number of interacting drivers [@cote2016].
-
-## Driver interactions
-
-Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels,
-with the head of the Channels most exposed (Figure \ref{kernel}A).
-Demersal destructive fisheries are located along the Laurentian Channel, the
-heads of the Anticosti and Esquiman Channels and around the Magdalen Islands
-(Figure \ref{kernel}B). By combining both drivers, we can observe that hypoxia
-and demersal destructive fisheries overlap mostly at high relative intensity
-(Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of
-the Esquiman and Anticosti Channels (Figure \ref{kernel}C, Box 1).
-
-Fisheries in the St. Lawrence have historically affected
-biodiversity distribution and habitat quality [@moritz2015]. Concurrently,
-hypoxia decreases overall habitat quality, but triggers species-dependent
-responses ranging from adaptation [*e.g.* northern shrimp *Pandalus borealis*
-and Greenland halibut *Reinhardtius hippoglossoides*; @pillet2016] to reduced
-growth rates [@dupont-prinet2013] and avoidance of oxygen-depleted habitats
-[*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality
-[*e.g.* sessile benthic invertebrates; @eby2005; @gilbert2007; @belley2010].
-Certain species may thus be adversely affected by fisheries and withstand
-hypoxia but still experience a decrease in prey availability, while others may
-be deleteriously affected by the compounded effect of both drivers [@deleo2017].
+While we cannot ascertain that high exposure areas are the most impacted, we can safely predict that these are the areas where studying ecosystem state will be the most complex due to the uncertainty associated with driver \textcolor{red}{\sout{interactions}} \textcolor{blue}{co-occurrence}, an uncertainty bound to increase rapidly with the number of \textcolor{red}{\sout{interacting}} \textcolor{blue}{co-occurring} drivers [@cote2016].
 
@@
 
-Through *eDrivers*, these observations will be iteratively improved towards
-an increasingly robust assessment of cumulative exposure and threat complexes
-as gaps in knowledge are addressed or approaches to describe drivers are refined.
-Arguably, the most meaningful benefit anticipated from *eDrivers* will be the
-gain in efficient access to comparable data-based knowledge on the exposure of
-ecosystems to multiple threats. This could pay quick scientific and management
-dividends by drawing on the knowledge and efforts of a wide range of
-contributors, by expanding avenues of scientific inquiry, by decreasing overall
-effort duplication and research costs, and by increasing research efficiency
-[@franzoni2014].
+Through *eDrivers*, these observations will be iteratively improved towards an increasingly robust assessment of cumulative exposure and threat complexes as gaps in knowledge are addressed or approaches to describe drivers are refined.
+Arguably, the most meaningful benefit anticipated from *eDrivers* will be the gain in efficient access to comparable data-based knowledge on the exposure of ecosystems to multiple threats.
+This could pay quick scientific and management dividends by drawing on the knowledge and efforts of a wide range of contributors, by expanding avenues of scientific inquiry, by decreasing overall effort duplication and research costs, and by increasing research efficiency [@franzoni2014].
 
-Critically, *eDrivers* will
-allow the scientific and governmental communities to identify key knowledge
-gaps that will assist in prioritizing and optimizing research efforts.
-Ultimately, we believe that *eDrivers* will operationalize evidence-based
-decision-making by streamlining data management and research, allowing science
-output to be available and interpretable on a time scale relevant to management
-[see @sutherland2004; @reichman2011]. The platform will thus greatly facilitate
-the application of broad scale, holistic research and management approaches such
-as marine spatial planning, ecosystem-based management, marine spatial planning
-and strategic environmental assessments [*e.g.* @rice2011; @halpern2015a; @jones2016].
+Critically, *eDrivers* will allow the scientific and governmental communities to identify key knowledge gaps that will assist in prioritizing and optimizing research efforts.
+Ultimately, we believe that *eDrivers* will operationalize evidence-based decision-making by streamlining data management and research, allowing science output to be available and interpretable on a time scale relevant to management [see @sutherland2004; @reichman2011].
+The platform will thus greatly facilitate the application of broad scale, holistic research and management approaches such as marine spatial planning, ecosystem-based management, marine spatial planning and strategic environmental assessments [*e.g.* @rice2011; @halpern2015a; @jones2016].
 
-While the focus of our paper has been on the description of drivers in the
-St. Lawrence, drivers are but one of the 4 elements required to properly
-evaluate environmental impacts. This single knowledge node should be weaved with
-other, comparable, nodes required for integrated management approaches such as
-species distribution and marine protected areas. Ultimately, all of these
-knowledge nodes could be weaved together through social-ecological metanetworks
-[@dee2017].
+\textcolor{red}{\sout{While the focus of our paper has been on the description of drivers in the St.
+Lawrence, drivers are but one of the 4 elements required to properly evaluate environmental impacts.
+This single knowledge node should be weaved with other, comparable, nodes required for integrated management approaches such as species distribution and marine protected areas.
+Ultimately, all of these knowledge nodes could be weaved together through social-ecological metanetworks (Dee et al.
+2017).}}
+
+\textcolor{blue}{A caveat to our initiative is that it focuses on a single element required for fully operational impact assessments.
+Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators (Dube and Munkittrick 2001; Dubé 2003; Jones 2016).
+These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers and fail to diagnose impacts on valued components or non-additive effects (Dube and Munkittrick 2001; Dubé 2003).
+In contrast, effect-based indicators are direct measurements of valued components (*e.g.* species abundance and biodiversity) and inherently capture the effects of multiple drivers (Dube and Munkittrick 2001; Dubé 2003).
+While effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers (Dube and Munkittrick 2001; Dubé 2003).
+Stressor-based and effect-based indicators are therefore both required to diagnose causes of ecosystem change (Dube 2003; Jones 2016).
+As a knowledge node on stressor-based indicators, *eDrivers* should be weaved with other, comparable, nodes describing valued ecosystem components that can be linked to drivers and allow a better understanding of cumulative impacts.
+Ultimately, multiple knowledge nodes could be weaved together through social-ecological metanetworks analyses (Dee et al.
+2016).} <!-- [@dube2001; @dube2003; @jones2016; @dee2017] -->
 
@@
 SLGO partly addresses this issue.
+\textcolor{blue}{We also wish to provide users with enhanced capabilities and flexibility in using the interactive tool and R package.
+This could include creating automatic reports and more flexibility for user-defined driver-based indicators.}
 
@@
 \centering
+\includegraphics{../Figures/kernel.png}
+\caption{\textcolor{red}{\sout{Interaction}} \textcolor{blue}{Co-occurrence} between \textcolor{red}{\sout{the intensity of }}hypoxia and demersal destructive fisheries in the St.
+Lawrence.
+An index of hypoxia ($A$) was created using bottom-water dissolved oxygen between 2013 and 2017 (Blais et al., 2018).
+Demersal destructive fisheries ($i.e.$ trawl and dredges) ($B$) intensity was evaluated from fisheries catch data collected between 2010 and 2015 used to measure annual area weighted total biomass ($kg$) in i $km^2$ grid cells (DFO, 2016b).
+See supplementary materials for more information on specific methodologies.
+Relative hypoxic stress and demersal destructive fisheries intensity was summed ($C$) to visualize their combined spatial distribution and intensity.
+Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a two-dimensional kernel analysis ($D$).}
+\label{kernel}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
+\centering
 \includegraphics{../Figures/footprint.png}
@@
 \label{hotspot}
-\end{figure}
-
-\newpage
-
-\begin{figure}[H]
-\centering
-\includegraphics{../Figures/kernel.png}
-\caption{Interaction between the intensity of hypoxia and demersal destructive fisheries in the St.
-Lawrence.
-An index of hypoxia ($A$) was created using bottom-water dissolved oxygen between 2013 and 2017 (Blais et al., 2018).
-Demersal destructive fisheries ($i.e.$ trawl and dredges) ($B$) intensity was evaluated from fisheries catch data collected between 2010 and 2015 used to measure annual area weighted total biomass ($kg$) in i $km^2$ grid cells (DFO, 2016b).
-See supplementary materials for more information on specific methodologies.
-Relative hypoxic stress and demersal destructive fisheries intensity was summed ($C$) to visualize their combined spatial distribution and intensity.
-Finally, individual density and the co-intensity of hypoxia and demersal destructive fisheries was investigated with a two-dimensional kernel analysis ($D$).}
-\label{kernel}
 \end{figure}

### 2020-02-18 Revisions/Rebuttal.md
minor rebuttal revisions

@@
 Our goal is to publish the package and submit it for peer-review once the data is available and accessible through SLGO's API.
+However, the package will/can?
+be available on github in the interim.
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 The goal of our manuscript and the *eDrivers* initiative is to facilitate the widespread accessibility of such data for environmental assessments and integrative management.
-It is therefore clear to us that what we are presenting is but one element of a fully operational impact assessment.
-Our goal is however not to discuss cumulative effects assessments in particular.
+We believe it is clear that we are presenting but one element of a fully operational impact assessment given the title of the manuscript and the 3rd sentence in the abstract: "Our objective in this paper is to systematically characterize the distribution and intensity of drivers in the St.
+Lawrence System.".
+Our goal is  not to discuss cumulative effects assessments in particular.
 It is rather to provide knowledge on an element common to most holistic approaches, *i.e.* drivers characterization.

### 2020-02-18 Revisions/Rebuttal.md
review comments from Remi and review whole document

@@
 Lawrence System" (ID: 469411) for publication in Frontiers in Marine Science.
-We would like to thank the reviewer for their very constructive comments.
-We feel that we were able to address all of their comments and are now submitting an improved manuscript.
+We would like to thank the reviewer for very constructive comments.
+We feel that we were able to address all the comments and are now submitting an improved manuscript.
 Below we explain how we have revised our manuscript based on the reviews.
 
-We thank you for considering our revised manuscript and I am looking forward to hearing from you.
+We thank you for considering our revised manuscript and we are looking forward to hearing from you.
 
@@
 > - As for the package, we will not be releasing a version on CRAN just yet.
-The material is not yet implemented on the SLGO servers (our partners on this project) and the data is currently contained directly in the package.
+The material is not yet implemented on the SLGO servers (our partners on this project) and the data is currently contained directly in the R package.
 Our goal is to publish the package and submit it for peer-review once the data is available and accessible through SLGO's API.
-However, the package will/can?
-be available on github in the interim.
+However, the package and the data will be available on github in the interim.
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 We present effects-based indicators and their importance for understanding cumulative effects.
-We then argue that both are jointly required to properly diagnose causes of ecosystem change (Dube 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).
+We then argue that both are jointly required to properly diagnose and attribute causes of ecosystem change to stressors (Dube 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).
 
@@
 We realize now that, even though we did not want to mislead readers in thinking we were evaluating driver interactions, we are in fact doing just that.
-As mentionned in reviewer 1 comments, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects.
+As mentionned by reviewer 1, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects.
 Under the assumption that driver co-occurrence is necessary for driver interactions to exist, we believe it is reasonable to predict that driver interactions and non-additive effects will most likely be observed in areas where drivers overlap at high relative intensities.
 These areas indeed delineate where the effects of both drivers are most likely to impact valued components simultaneously.
-In presenting sections 3.4 and 4.2, our goal was to provide an illustration of how a simple two-drivers case could be used to identify areas of high relative intensity driver overlap.
-This should be used as segway to the more integrative cumulative footprint and hotspot indicators.
+In presenting sections 3.4 and 4.2, our goal was to provide an illustration of how a simple two-drivers case could be used to identify areas of such high relative intensity driver overlap.
+This should be used as a segway to the more integrative cumulative footprint and hotspot indicators.
 As rightfully identified by reviewer 1, however, the manuscript is indeed misleading on the topic of driver interactions.
@@
 
-  > 1. We removed objective 2 from the manuscript and mentions of driver interactions, except in the context of discussing the likelihood on observing interactions in areas of driver co-occurrence.
+  > 1. We removed objective 2 from the manuscript and mentions of driver interactions, except in the context of discussing the likelihood of observing interactions in areas of driver co-occurrence.
 
@@
 The methodology and results/discussions sections on driver interactions were moved to sections 3.3 and 4.1, respectively, and renamed *Driver co-occurrence*.
-We then modified the section as a simplified example of driver co-occurrence (lines XXX to XXX).
+We then modified the methods section as a simplified example of driver co-occurrence (lines 178 to 190).
 We however did not modify the results/discussion section.
 The only mention of driver interactions was in the title itself.
-In the discussion, we merely discuss the areas of high intensity co-occurrence between the two drivers and present adverse effects that these drivers may have on species individually or simultaneously.
+In the discussion, we merely discuss the areas of high intensity co-occurrence between the two drivers and present adverse effects that these drivers may have on species individually and simultaneously.
 We did however add a closing statement to the section stating that individual and interactive effects of both drivers is species- and community-dependent and that the co-occurrence of drivers in itself is not an evidence of driver interaction.
-We then state that assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities (lines XXX to XXX)
+We then state that assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities (lines 269 to 272)
 
-> - Ultimately, this portion of the manuscript should serve as a simple illustration to better understand the complete cumulative footprint indicator.
-In this sense, if the reviewer still feels that proposed changes are not sufficient, we propose removing these sections from the manuscript.
+> - Ultimately, this portion of the manuscript should serve as a simple illustration to better understand the complete cumulative footprint indicators.
+In this sense, if the reviewer still feels that proposed changes are not sufficient, we propose removing these sections from the manuscript entirely.
 
 > - As for providing flexibility in indicator definition, it is a feature that we wish to provide in future iterations of the package and the platform.
-We would however rather not include such enhancements at this point, as they are more technical in nature and should not preclude us from publishing this work.
-We however added a statement to the manuscript that presents this idea as a perspective and future work to be accomplished (lines XXX to XXX in perspectives).
+We would however rather not include such enhancements at this point, as they are more technical in nature and we feel that they should not prevent us from publishing this work.
+We however added a statement that presents this idea as a perspective and future work to be accomplished (lines 515 to 517).
+
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 Presenting a full cumulative impacts assessments and defining valued components is however not the objective of this manuscript, nor is it to discuss cumulative impacts in particular.
-The choices of valued components are fully dependent on the objective of a particular assessment or management scheme.
+By themselves, we believe stressor-based indicators offer valuable insights in the dynamics of a system which can help managers in prioritizing decision-making and scientists in targetting specific areas for *in situ* measurements for hypothesis testing, especially when data on valued components are lacking.
+Furthermore, the choices of valued components are fully dependent on the objective of a particular assessment or management scheme.
 In our experience, time and money is spent on recurring initiatives that seek to characterize sources of stress in a system for impacts assessments and integrative management.
@@
 > - As proposed by reviewer 1, we had already included a paragraph on the importance of valued components and their vulnerability to soures of stress in the conclusion.
-We however realize now that it was underwhelming in its ability to provide perspective and direction to how eDrivers may be used in the future.
+We however realize now that it was underwhelming in its ability to provide perspective and direction as to how *eDrivers* may be used in the future.
 We therefore extended this paragragh, also in response to reviewer 1's first comment, to better discuss the significance of valued components and the different types of indicators for cumulative effects assessment (lines 487 to 509).
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
-> - This was modified in the text (lines XXX to XXX).
+> - This was modified in the text.
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
-> - We have modified the sentences on lines 327-328 (now lines XXX to XXX).
+> - We have modified the sentences on lines 327-328 (now lines 339 to 343).
+"Of particular concern are threat complexes 3 and 6, which are the two most exposed complexes in the St.
+Lawrence and are characterized by distinct cumulative exposure regimes Between them, they capture most of the coastal and offshore hotspots identified in the St.
+Lawrence." now becomes "Threat complexes 3 and 6 capture most coastal and offshore hotspots and are the two most exposed complexes in the St.
+Lawrence."
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
-> - Figure caption was modified accordingly (line XXX).
+> - Figure caption was modified accordingly.
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 
-TODO:
-
-- [ ] review appendix
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 
-Is the quality of the figures and tables satisfactory?
-- Yes
+*Is the quality of the figures and tables satisfactory?
+- Yes*
 
-Does the reference list cover the relevant literature adequately and in an unbiased manner?
-- Yes
+*Does the reference list cover the relevant literature adequately and in an unbiased manner?
+- Yes*
 
-Are the statistical methods valid and correctly applied? (e.g. sample size, choice of test)
-- Not Applicable
+*Are the statistical methods valid and correctly applied? (e.g. sample size, choice of test)
+- Not Applicable*
 
-Are the methods sufficiently documented to allow replication studies?
-- Yes
+*Are the methods sufficiently documented to allow replication studies?
+- Yes*
 
@@
 
-TODO:
-
-- [ ] Make GitHub repositories public
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
 
-Are the data underlying the study available in either the article, supplement, or deposited in a repository?
+*Are the data underlying the study available in either the article, supplement, or deposited in a repository?
 (Sequence/expression data, protein/molecule characterizations, annotations, and taxonomy data are required to be deposited in public repositories prior to publication) (yes/no/not applicable)
-- No
+- No*
 
@@
 
-> - The formatted driver data will be available through the package and the raw data are now archived on Zenodo (link XXX) and accessible upon request to the authors.
+> - The formatted driver data will be available through the package and the raw data will be archived on Zenodo and accessible upon request to the authors.
+The archive is ready to be published, but we are waiting for the publication process to be completed so that the proper citation can be referenced with the material.
+Similarly, the workflos used for the first iteration of the platform will be released and archived on Zenodo with the paper publication.
 
-TODO:
-
-- [ ] Archive rawdata on Zenodo.
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
 
-Does the study adhere to ethical standards including ethics committee approval and consent procedure?
-- Not Applicable
+*Does the study adhere to ethical standards including ethics committee approval and consent procedure?
+- Not Applicable*
 
-Have standard biosecurity and institutional safety procedures been adhered to?
-- Not Applicable
+*Have standard biosecurity and institutional safety procedures been adhered to?
+- Not Applicable*

### 2020-02-26 Revisions/Rebuttal.md
include final revisions to the manuscript before sending to co-authors

@@
 rmarkdown::render('./Revisions/Rebuttal.md')
+rmarkdown::render('./Revisions/Rebuttal.md', 'word_document')
 -->
@@
 Lawrence System.
-In particular, the article introduces an open knowledge platform named eDrivers, though it doesn’t seem to be working at this moment (https://github.com/eDrivers - this site contains zero repositories, zero people, zero projects; the R package dDrivers is not available on CRAN).
+In particular, the article introduces an open knowledge platform named eDrivers, though it doesn’t seem to be working at this moment (https://github.com/eDrivers - this site contains zero repositories, zero people, zero projects; the R package eDrivers is not available on CRAN).
 When this R package and the website become available, I believe the article and its related sources will be beneficial to the scientific and management communities.
@@
 
-> - The material has yet to be released publicly on GitHub as we were waiting for the review process to be completed. We can however provide access to the repositories in the interim.
+> - The material has yet to be released publicly on GitHub as we were waiting for the review process to be completed. We can, however, provide access to the repositories in the interim.
 
 > - As for the package, we will not be releasing a version on CRAN just yet.
-The material is not yet implemented on the SLGO servers (our partners on this project) and the data is currently contained directly in the R package.
+The material is not yet implemented on the St.
+Lawrence Global Observatory (SLGO; https://ogsl.ca/en) servers (our partners on this project) and the data is currently contained directly in the R package.
 Our goal is to publish the package and submit it for peer-review once the data is available and accessible through SLGO's API.
-However, the package and the data will be available on github in the interim.
+However, the package and the data will be available on Github in the interim.
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 We believe it is clear that we are presenting but one element of a fully operational impact assessment given the title of the manuscript and the 3rd sentence in the abstract: "Our objective in this paper is to systematically characterize the distribution and intensity of drivers in the St.
-Lawrence System.".
+Lawrence System." (lines 38-39).
 Our goal is not to discuss cumulative effects assessments in particular.
@@
 > - We nonetheless agree with reviewer 1's suggestions.
-We are therefore proposing the following modifications to the manuscript that we believe consolidates the essence of the suggestions made by reviewer 1:
+We are therefore proposing the following modifications to the manuscript that we hope consolidates the essence of the suggestions made by reviewer 1:
 
-  > 1. We modified the introduction to broadly state the major steps undertaken in holistic environmental approaches in general (lines 91 to 96).
+  > 1. We modified the introduction to broadly state the major steps undertaken in holistic environmental approaches in general (lines 96-101).
 
-  > 2. We modified the statement of the objectives of the manuscript at the end of the introduction to clearly state that we are focusing on drivers (lines 119 to 124).
+  > 2. We modified the statement of the objectives of the manuscript at the end of the introduction to clearly state that we are focusing on drivers (lines 123-129).
 
   > 3.
-We modified the pre-existing paragraph presenting the importance of valued components in the perspective sections (lines 487 to 509).
+We modified the pre-existing paragraph presenting the importance of valued components in the perspective sections (lines 492-514).
 It is now presenting *eDrivers* as a suite of stressor-based indicators that hold certain advantages, but fail to capture effects on valued components and non-additive effects.
 We present effects-based indicators and their importance for understanding cumulative effects.
-We then argue that both are jointly required to properly diagnose and attribute causes of ecosystem change to stressors (Dube 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).
+We then argue that both are jointly required to properly diagnose and attribute causes of ecosystem change to stressors (Dubé 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).
 
@@
 We realize now that, even though we did not want to mislead readers in thinking we were evaluating driver interactions, we are in fact doing just that.
-As mentionned by reviewer 1, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects.
+As mentioned by reviewer 1, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects.
 Under the assumption that driver co-occurrence is necessary for driver interactions to exist, we believe it is reasonable to predict that driver interactions and non-additive effects will most likely be observed in areas where drivers overlap at high relative intensities.
@@
 In presenting sections 3.4 and 4.2, our goal was to provide an illustration of how a simple two-drivers case could be used to identify areas of such high relative intensity driver overlap.
-This should be used as a segway to the more integrative cumulative footprint and hotspot indicators.
+This should be used as a transition to the more integrative cumulative footprint and hotspot indicators.
 As rightfully identified by reviewer 1, however, the manuscript is indeed misleading on the topic of driver interactions.
@@
 The methodology and results/discussions sections on driver interactions were moved to sections 3.3 and 4.1, respectively, and renamed *Driver co-occurrence*.
-We then modified the methods section as a simplified example of driver co-occurrence (lines 178 to 190).
-We however did not modify the results/discussion section.
+We then modified the methods section as a simplified example of driver co-occurrence (lines 184-195).
+However, we did not modify the results/discussion section.
 The only mention of driver interactions was in the title itself.
 In the discussion, we merely discuss the areas of high intensity co-occurrence between the two drivers and present adverse effects that these drivers may have on species individually and simultaneously.
-We did however add a closing statement to the section stating that individual and interactive effects of both drivers is species- and community-dependent and that the co-occurrence of drivers in itself is not an evidence of driver interaction.
-We then state that assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities (lines 269 to 272)
+We did add a closing statement to the section stating that individual and interactive effects of both drivers are species- and community-dependent and that the co-occurrence of drivers in itself is not an evidence of driver interaction.
+We then state that assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities (lines 274-277).
 
@@
 > - As for providing flexibility in indicator definition, it is a feature that we wish to provide in future iterations of the package and the platform.
-We would however rather not include such enhancements at this point, as they are more technical in nature and we feel that they should not prevent us from publishing this work.
-We however added a statement that presents this idea as a perspective and future work to be accomplished (lines 515 to 517).
+We would, however, rather not include such enhancements at this point, as they are more technical in nature and we feel that they should not prevent us from publishing this work.
+Instead, we added a statement that presents this idea as a perspective and future work to be accomplished (lines 520-522), as suggested by reviewer 1.
 
@@
 > - We agree with reviewer 1 on the importance of valued components for cumulative effects assessments.
-Presenting a full cumulative impacts assessments and defining valued components is however not the objective of this manuscript, nor is it to discuss cumulative impacts in particular.
-By themselves, we believe stressor-based indicators offer valuable insights in the dynamics of a system which can help managers in prioritizing decision-making and scientists in targetting specific areas for *in situ* measurements for hypothesis testing, especially when data on valued components are lacking.
+Presenting a full cumulative impact assessments and defining valued components are, however, not the objectives of this manuscript, nor is it to discuss cumulative impacts in particular.
+By themselves, we believe stressor-based indicators offer valuable insights in the dynamics of a system which can help managers in prioritizing decision-making and scientists in targeting specific areas for *in situ* measurements for hypothesis testing, especially when data on valued components are lacking.
 Furthermore, the choices of valued components are fully dependent on the objective of a particular assessment or management scheme.
 In our experience, time and money is spent on recurring initiatives that seek to characterize sources of stress in a system for impacts assessments and integrative management.
-This is, in our view, unacceptable.
 Our goal was therefore to build an initiative that would limit this and provide access to formatted and comparable knowledge on sources of stress that are typically used for impact assessments and integrative management.
@@
 
-> - As proposed by reviewer 1, we had already included a paragraph on the importance of valued components and their vulnerability to soures of stress in the conclusion.
+> - As proposed by reviewer 1, we had already included a paragraph on the importance of valued components and their vulnerability to sources of stress in the conclusion.
 We however realize now that it was underwhelming in its ability to provide perspective and direction as to how *eDrivers* may be used in the future.
-We therefore extended this paragragh, also in response to reviewer 1's first comment, to better discuss the significance of valued components and the different types of indicators for cumulative effects assessment (lines 487 to 509).
+We therefore extended this paragraph, also in response to reviewer 1's first comment, to better discuss the significance of valued components and the different types of indicators for cumulative effects assessment (lines 492-514).
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
-> - We have modified the sentences on lines 327-328 (now lines 339 to 343).
+> - We have modified the sentences on lines 327-328 (now lines 344-348).
 "Of particular concern are threat complexes 3 and 6, which are the two most exposed complexes in the St.
-Lawrence and are characterized by distinct cumulative exposure regimes Between them, they capture most of the coastal and offshore hotspots identified in the St.
+Lawrence and are characterized by distinct cumulative exposure regimes between them, they capture most of the coastal and offshore hotspots identified in the St.
 Lawrence." now becomes "Threat complexes 3 and 6 capture most coastal and offshore hotspots and are the two most exposed complexes in the St.
@@
 The archive is ready to be published, but we are waiting for the publication process to be completed so that the proper citation can be referenced with the material.
-Similarly, the workflos used for the first iteration of the platform will be released and archived on Zenodo with the paper publication.
+Similarly, the workflows used for the first iteration of the platform will be released and archived on Zenodo with the paper publication.
 

### 2020-02-26 Revisions/Beauchesne-eDriversMS-SI.rmd
include final revisions to the manuscript before sending to co-authors

--- 2019-05-01 eDriversMS-SI-v2.rmd
@@
     - \onehalfspacing
+    - \usepackage[svgnames]{xcolor}
+    - \usepackage[normalem]{ulem}
+    - \definecolor{blue}{HTML}{20cad5}
 relativeurls: true
-bibliography: eDriversMS.bib
+bibliography: Beauchesne-eDriversMS.bib
 csl: frontiers.csl
@@
 <!--
-cd phdproj/edrivers/edriversms
-rmarkdown::render('./eDriversMS-SI-v2.rmd', 'pdf_document')
-rmarkdown::render('./eDriversMS-SI-v2.rmd', 'word_document')
+rmarkdown::render('./Revisions/Beauchesne-eDriversMS-SI.rmd')
 -->
@@
 ```{r tableS1, echo=F, fontsize=4}
-load('./tables/tableS1.RData')
+load('../tables/tableS1.RData')
 tableS1
@@
 
-Oceans are the largest reservoir and sink of atmospheric carbon dioxide ($CO_2$).
-Its uptake increases seawater acidity and the lowers saturation state with
+Oceans are the largest reservoir\textcolor{blue}{s} and sink\textcolor{blue}{s} of atmospheric carbon dioxide ($CO_2$).
+Its uptake increases seawater acidity and \textcolor{red}{\sout{the}} lowers saturation state with
 respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common
-$CaCO_3$ polymorphs constituting the shells and skeleton of many marine
+$CaCO_3$ polymorphs constituting the shells and skeleton\textcolor{blue}{s} of many marine
 organisms [@mucci2017]. This can have deleterious effects on carbonate-secreting
@@
 surveys and DFO's multispecies surveys [@bourdages2018] collected in
-August-September of 2017 [@starr2019]. The carbonate chemistry was determined
-through pH and TA measurements. Samples for pH and total alkalinity (TA) were
-collected under a bubble free and no head space conditions into 500 $mL$
-borosilicate glass flasks, 250 $\mu L$ of the saturated $HgCl_2$ solution was
+August-September of 2017 [@starr2019].
+\textcolor{blue}{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
+stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
+majority of sampling stations were located in the Southern Gulf ($n = 84$),
+compared to the Estuary and Northern Gulf ($n = 33$).}
+
+The carbonate chemistry was determined
+through pH and \textcolor{blue}{total alkalinity} (TA) measurements. Samples for pH and \textcolor{red}{\sout{total alkalinity}} TA were
+collected under \textcolor{red}{\sout{a}} bubble free and no head space conditions into 500 $mL$
+borosilicate glass flasks. 250 $\mu L$ of \textcolor{red}{\sout{the}} saturated $HgCl_2$ solution was
 added to sample, and processed following the "Guide to best practices for Ocean
@@
 [@dickson2007]. A similar procedure was carried out before each set of sample
-measurements using a TRIS (Tris (hydroxymethyl) -aminomethane) buffer prepared
+measurement\textcolor{red}{\sout{s}} using a TRIS (Tris (hydroxymethyl) -aminomethane) buffer prepared
 at a practical salinity (S) of approximately 30 [@millero1986]. Certified
@@
 electrode (pHC2001) in a continuous titrant addition mode, an algorithm
-specifically designed for shallow end-point detection [@dickson2007]. The dilute
+specifically designed for shallow end-point detection [@dickson2007]. The dilute\textcolor{blue}{d}
 HCl titrant 0.1M in a solution of NaCl of 0.6M was calibrated using CRM provided
@@
 The carbonate system parameters (including $\Omega_{A}$) were
-corrected for in situ pressure and temperature using the algorithm CO2SYS
-[@lewis1998] with measured pH, total alkalinity, soluble reactive
+corrected for *in situ* pressure and temperature using the algorithm CO2SYS
+[@lewis1998] with measured pH, \textcolor{red}{\sout{total alkalinity}} \textcolor{blue}{TA}, soluble reactive
 and silicate concentrations as input parameters.
 
-$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
+\textcolor{red}{\sout{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
 stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
 majority of sampling stations were located in the Southern Gulf ($n = 84$),
-compared to the Estuary and Northern Gulf ($n = 33$).
+compared to the Estuary and Northern Gulf ($n = 33$).}}
 
@@
 waters with lowered alkalinity and salinity caused by the St. Lawrence plume.
-Such water masses are have lower buffering capacities to decreasing pH and
+Such water masses \textcolor{red}{\sout{are}} have lower buffering capacities to decreasing pH and
 $\Omega_{A}$. Also, the accumulation of metabolic carbon dioxide produced by
@@
 slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé
-Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modelled
+Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modeled
 using the exponential kriging model.
 
-We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. Whereas
+We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. \textcolor{red}{\sout{Whereas}} \textcolor{blue}{While}
 there are few published datasets on $\Omega_{A}$ in the St.
@@
 covariable, as done in [@dutil2011]. Bathymetric data come from @dutil2011
-and have a 100 $km^2$ resolution. For more details, refer to the hypoxia
+and have a 100 $km^2$ resolution. For more details, refer to the hypoxia \textcolor{blue}{data layer}
 description. The *Deep* model was used to transform $O_2$ values to $\Omega_A$
@@
 affected (see above), we built an index of acidification stress ($A_s$)
-ranging between 0 and 1 that slowly begins increasing as $\Omega_A$ $\approx$
+ranging between 0 and 1 that slowly \textcolor{red}{\sout{begins increasing}} \textcolor{blue}{increases} as $\Omega_A$ $\approx$
 $2$, increases more rapidly close to $\Omega_A$ $= 1$ and reaches its peak at
@@
 
-The data used to characterize hypoxia uses oxygen saturation in % and come from
+The data used to characterize hypoxia \textcolor{red}{\sout{uses oxygen saturation in % and}} come from
 the Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
-[AZMP; @galbraith2018] surveys in late spring and fall, as well as from annual
-DFO’s multispecies surveys for the northern Gulf in August and for the Magdalen
+[AZMP; @galbraith2018] survey\textcolor{red}{\sout{s}} in late spring and fall, as well as from \textcolor{red}{\sout{annual}}
+DFO’s \textcolor{blue}{annual} multispecies surveys for the northern Gulf in August and for the Magdalen
 Shallows in September. We provide a brief summary of data and methods to
@@
 Oxygen concentration is measured at every station using an oxygen probe
-(Sea-Bird SBE43) mounted on the CTD. The probe is calibrated against seawater
+(Sea-Bird SBE43) mounted on \textcolor{red}{\sout{the}} \textcolor{blue}{a} CTD. The probe is calibrated against seawater
 samples that are analyzed by Winkler titration on every cast
 [see @blais2019 for calibration procedure]. The data used is the last depth
-sampled on the CTD profile, which typically corresponds to ~10m above the
+sampled on the CTD profile, which \textcolor{blue}{is} typically \textcolor{red}{\sout{corresponds to}} ~10m above the
 bottom. Oxygen concentration ($\mu M$ $L^{-1}$) is converted to oxygen
-saturation taking into account the salinity and temperature at the selected
+saturation\textcolor{blue}{,} taking into account the salinity and temperature at the selected
 depth.
@@
 Anomaly time series were normalized by their standard deviation (SD) to allow
-comparison between areas of the St. Lawrence with different temperature ranges.
+comparison\textcolor{blue}{s} between areas of the St. Lawrence with different temperature ranges.
 For example, temperatures observed in deep channels are less variable than in
@@
 outliers by @tukey1977. Outlier values were capped to correspond to the 5th and
-95th percentiles. Anomalies were divided into positive and negative anomalies
+95th percentile\textcolor{red}{\sout{s}} \textcolor{blue}{values}. Anomalies were divided into positive and negative anomalies
 and the absolute value of anomalies were used as an indicator of the intensity
@@
 
-The data used to characterize sea level rise risk comes from the global
+The data used to characterize sea level rise risk come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 
-Aquaculture data comes from a variety of sources in the St. Lawrence because
-aquaculture sites are mostly managed at the provincial level. We therefore had
-to gather the data on aquaculture sites from the 5 provinces dividing the
-St. Lawrence [@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016].
-
+\textcolor{red}{\sout{Aquaculture data comes from a variety of sources in the St. Lawrence because
+aquaculture sites are mostly managed at the provincial level.
+We therefore had to gather the data on aquaculture sites from the 5 provinces dividing the
+St. Lawrence}}
+\textcolor{blue}{As aquaculture sites are managed at the provincial level in the St. Lawrence, data on aquaculture sites come from various sources}
+[@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016].
 Invertebrates aquaculture is especially important in the southern and western
 Gulf. Fish and algae aquaculture, on the other hand, remains marginal.
-Considering this, we only considered invertebrates aquaculture for the
+Considering this, we only \textcolor{red}{\sout{considered}} \textcolor{blue}{included} invertebrates aquaculture for the
 aquaculture driver layer. However, if fish or algae farming were to become more
-important, these driver should be incorporated in future analyses as individual
+important, these driver\textcolor{blue}{s} should be incorporated in future analyses as individual
 layers, as impacts vary between types of aquaculture.
@@
 rarely extend beyond the location of the farms. We therefore only considered
-the actual location of sites to characterize the distribution of this driver.
+the \textcolor{red}{\sout{actual}} location of sites to characterize the distribution of this driver.
 We were unable to characterize site production in terms of biomass farmed, which
 could provide an indication of the intensity of aquaculture activities. As such,
-we considered aquaculture as binary presence-absence data in our analyses.
+we considered aquaculture as binary presence-absence data \textcolor{red}{\sout{in our analyses}}.
 
@@
 
-We used lights at night as a proxy of coastal infrastructure development, as
+\textcolor{red}{\sout{We used lights at night as a proxy of coastal infrastructure development, as
 terrestrial stable lights at night represent light from human settlements and
-industrial sites with electricity.
+industrial sites with electricity.}}
+\textcolor{blue}{Terrestrial stable lights at night mostly represent light from human settlements and industrial sites with electricity.
+We thus used lights at night as a proxy of coastal infrastructure development.}
 
-The data comes from the Nighttime Lights Time Series. Nighttime light products
+The data come from the Nighttime Lights Time Series. Nighttime light products
 are compiled by the Earth Observation Group at the National Oceanic and
@@
 
-The data used to characterize inorganic pollution comes from the global
+The data used to characterize inorganic pollution come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 
-Inorganic pollution was modelled using impervious surface area (*i.e.*
+Inorganic pollution was modeled using impervious surface area (*i.e.*
 artificial surfaces such as paved roads) under the assumption that most of this
@@
 point-sources or in areas lacking paved roads is therefore not captured by
-this layer. The data obtained was aggregated at the watershed scale and spread
-into coastal and marine environments was modelled using a diffusive plume model
+this layer. The data obtained was aggregated at the watershed scale. \textcolor{red}{\sout{and spread}}
+\textcolor{blue}{Spread} into coastal and marine environments was modeled using a diffusive plume model
 from each watershed pourpoints (*e.g.* river mouths).
@@
 
-The data used to characterize nutrient pollution comes from the global
+The data used to characterize nutrient pollution come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 The data used came from the Food and Agriculture Organization of the United
-Nations (FAO). Gaps in data were modelled using a linear regression between
+Nations (FAO). Gaps in data were modeled using a linear regression between
 fertilizer and pesticides or agricultural gross domestic product (GDP).
@@
 
-The data used to characterize organic pollution comes from the global
+The data used to characterize organic pollution come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 Annual pesticide use in tonnes ($t$) was used as a proxy of organic pollution.
-The data used came from the FAO and gaps in data were modelled using a linear
+The data used came from the FAO and gaps in data were modeled using a linear
 regression between pesticides and fertilizers or agricultural GDP. The same
@@
 
-The data we use to describe the risk of toxic algae comes from an expert based
+The data we use to describe the risk of toxic algae come from an expert based
 map delineating the areas where coastal areas are at risk from five different
@@
 for all fisheries in the St. Lawrence, they still provide a very thorough
-overview of the spatial distribution and intensity of fishing activities in the
-St. Lawrence. The data we used spans 6 years from 2010 to 2015 and details
+overview of the spatial distribution and intensity of fishing activities \textcolor{red}{\sout{in the
+St. Lawrence}}. The data we used span 6 years from 2010 to 2015 and detail
 218323 fishing events (36387 $\pm$ 3147 fishing events per year). There were 31
-targeted species and a total of 53 caught species in the dataset.
+targeted species and a total of 53 \textcolor{red}{\sout{caught}} species \textcolor{blue}{caught} in the dataset.
 
@@
 high-bycatch (DNH), pelagic, low-bycatch (PLB) and pelagic, high-bycatch (PHB).
-This categorization therefore divides the fisheries data into 5 distinct
+This categorization \textcolor{red}{\sout{therefore}} divides the fisheries data into 5 distinct
 driver layers characterizing fishing activities.
 
-Gear types can also be further classified into fixed or mobile engines based
+Gear types can \textcolor{red}{\sout{also}} be further classified into fixed or mobile engines based
 on their mobility. We used these two mobility classes to generate a buffer of
 impact around each fishing activity coordinates to consider potential spatial
-uncertainty associated with locations and the fact that mobile engines can be
-tracted over several kilometers during fishing activities and that we do not
+uncertainty associated with locations, \textcolor{red}{\sout{and}} the fact that mobile engines can be
+tracted over several kilometers during fishing activities, and \textcolor{red}{\sout{that}} \textcolor{blue}{because} we do not
 have the beginning and end points of mobile fishing events. Buffer sizes for
@@
 
-In order to characterize the intensity of fishing activities ($FI$), we used
+\textcolor{red}{\sout{In order to}} \textcolor{blue}{To} characterize the intensity of fishing activities ($FI$), we used
 a biomass yield density index. We multiplied the total annual biomass captured
@@
 
-The data used to characterize shipping comes from the global cumulative impacts
+The data used to characterize shipping come from the global cumulative impacts
 assessment on habitats [@halpern2008a; @halpern2015a] and available on the NCEAS
@@
 Positioning System technology to locate vessels every 10 minutes. The data used
-was from November 2010 to December 2011.
-
+\textcolor{red}{\sout{was}} \textcolor{blue}{were} from November 2010 to December 2011.
 Data used come mostly from vessels that move globally (*i.e.* cargo, tanker and
@@
 
-The data used to characterize invasive species risk comes from the global
+The data used to characterize invasive species risk come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 
-The data used to characterize marine pollution risk comes from the global
+The data used to characterize marine pollution risk come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 \centering
-\includegraphics{./Figures/drHist.png}
+\includegraphics{./FigureS1.png}
 \caption{Frequency distribution of the untransformed data for all driver layers.}
@@
 \centering
-\includegraphics{./Figures/drivers.png}
+\includegraphics{./FigureS2.png}
 \caption{Distribution and intensity of transformed and normalized drivers in the Estuary and Gulf of St. Lawrence available on *eDrivers*.}
@@
 \centering
-\includegraphics{./Figures/marimekko.png}
+\includegraphics{./FigureS3.png}
 \caption{Frequency distribution of cumulative exposure (\textit{i.e.} sum of normalized driver intensity in each grid cell) and percent contribution of each driver to the frequency distribution of cumulative exposure in the Estuary and Gulf of St.
@@
 \centering
-\includegraphics{./Figures/validation.png}
+\includegraphics{./FigureS4.png}
 \caption{Validation procedure for the \textit{k-medoids} and \textit{k-means} clustering algorithms based on the number of cluster that maximizes average silhouette width (upper panels; Kaufman and Rousseeuw, 1990) and minimizes the total within-cluster sum of squares (WSS; lower panels).}
@@
 \centering
-\includegraphics{./Figures/interDissimilarity.png}
+\includegraphics{./FigureS5.png}
 \caption{Evaluation of inter-cluster dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan distance (Clarke, 1993).
@@
 \centering
-\includegraphics{./Figures/intraSimilarity.png}
+\includegraphics{./FigureS6.png}
 \caption{Evaluation of intra-cluster similarity using the Manhattan distance transformed to a similarity index.

### 2020-02-26 Revisions/Beauchesne-eDriversMS.Rmd
include final revisions to the manuscript before sending to co-authors

@@
    - \usepackage[svgnames]{xcolor}
-   # - \usepackage{titling}
-   # - \pretitle{\begin{flushleft}}
-   # - \posttitle{\end{flushleft}}
    - \usepackage{lscape}
@@
    - \usepackage{float}
-   - \usepackage{ulem}
+   - \usepackage[normalem]{ulem}
    - \definecolor{blue}{HTML}{20cad5}
@@
 2017).}}
-
 \textcolor{blue}{A caveat to our initiative is that it focuses on a single element required for fully operational impact assessments.
@@
 These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers and fail to diagnose impacts on valued components or non-additive effects (Dube and Munkittrick 2001; Dubé 2003).
-In contrast, effect-based indicators are direct measurements of valued components (*e.g.* species abundance and biodiversity) and inherently capture the effects of multiple drivers (Dube and Munkittrick 2001; Dubé 2003).
+In contrast, effect-based indicators are direct measurements of valued components (\textit{e.g.} species abundance and biodiversity) and inherently capture the effects of multiple drivers (Dube and Munkittrick 2001; Dubé 2003).
 While effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers (Dube and Munkittrick 2001; Dubé 2003).
 Stressor-based and effect-based indicators are therefore both required to diagnose causes of ecosystem change (Dube 2003; Jones 2016).
-As a knowledge node on stressor-based indicators, *eDrivers* should be weaved with other, comparable, nodes describing valued ecosystem components that can be linked to drivers and allow a better understanding of cumulative impacts.
+As a knowledge node on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, nodes describing valued ecosystem components that can be linked to drivers and allow a better understanding of cumulative impacts.
 Ultimately, multiple knowledge nodes could be weaved together through social-ecological metanetworks analyses (Dee et al.
-2016).} <!-- [@dube2001; @dube2003; @jones2016; @dee2017] -->
+2016).}
 
@@
 Port of Sept-Îles and City of Sept-Îles).
+
+<!-- !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! -->
+TODO: Remove [@dube2001; @dube2003; @jones2016]
+<!-- !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! -->
 

### 2020-03-11 Revisions/Rebuttal.md
manuscript revisions to submit to Frontiers

@@
 
-We are submitting a revised version of the manuscript "Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St.
-Lawrence System" (ID: 469411) for publication in Frontiers in Marine Science.
+We are submitting a revised version of the manuscript “Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System” (ID: 469411) for publication in Frontiers in Marine Science.
 We would like to thank the reviewer for very constructive comments.
 We feel that we were able to address all the comments and are now submitting an improved manuscript.
-Below we explain how we have revised our manuscript based on the reviews.
+Below, you will find our responses to the reviewer’s comments and how we have revised our manuscript in view of these.
 
-We thank you for considering our revised manuscript and we are looking forward to hearing from you.
+We thank you for considering our revised manuscript and look forward to hearing back from you.
 
@@
 
+# Responses to the reviewer’s comments
+
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
-# Question 1
+## Question 1
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
@@
 
-> - The material has yet to be released publicly on GitHub as we were waiting for the review process to be completed. We can, however, provide access to the repositories in the interim.
+> - The material has yet to be released publicly on GitHub as we were waiting for the review process to be completed.
+We can, however, provide access to the repositories for the review process if the reviewer wishes to review the contents.
 
-> - As for the package, we will not be releasing a version on CRAN just yet.
+> - As for the package and the data, it will be available on Github, but we will not release a version on CRAN just yet.
 The material is not yet implemented on the St.
-Lawrence Global Observatory (SLGO; https://ogsl.ca/en) servers (our partners on this project) and the data is currently contained directly in the R package.
-Our goal is to publish the package and submit it for peer-review once the data is available and accessible through SLGO's API.
-However, the package and the data will be available on Github in the interim.
+Lawrence Global Observatory (SLGO; https://ogsl.ca/en) servers (our partners on this project) and the data are currently embedded directly in the R package.
+Our goal is to publish the package and submit it for peer-review once the data are available and accessible through SLGO's API.
+
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
-# Question 2
+## Question 2
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
@@
 
-> - We agree with reviewer 1 on the importance of clearly stating that we are solely focusing on drivers in our manuscript.
-The goal of our manuscript and the *eDrivers* initiative is to facilitate the widespread accessibility of such data for environmental assessments and integrative management.
-We believe it is clear that we are presenting but one element of a fully operational impact assessment given the title of the manuscript and the 3rd sentence in the abstract: "Our objective in this paper is to systematically characterize the distribution and intensity of drivers in the St.
-Lawrence System." (lines 38-39).
-Our goal is not to discuss cumulative effects assessments in particular.
-It is rather to provide knowledge on an element common to most holistic approaches, *i.e.* drivers characterization.
-We have strived to make this as clear as possible throughout the manuscript, namely in the introduction and in the perspectives sections, as noted by reviewer 1.
-As such, and considering the focus of the manuscript, we are hesitant to accept the suggestion of reviewer 1 to discuss cumulative effects assessment, best practices and indicator types at the beginning of the introduction.
-We feel that, considering the objective of this manuscript, the beginning of the introduction should rather focus on drivers specifically.
+> - We agree with the reviewer about the importance of clearly stating that this manuscript solely focuses on drivers.
+The goal of this manuscript and the eDrivers initiative is to facilitate the widespread accessibility of such data for environmental assessments and integrative management.
+We believe that we are clearly stating that the goal of our manuscript is focused on a single element of a fully operational impact assessment given the title of the manuscript and the 3rd sentence of the abstract: “Our objective in this paper is to systematically characterize the distribution and intensity of drivers in the St. Lawrence System.” (lines 38-39).
+Our goal is not to address cumulative effects.
+It is rather to provide knowledge on an element common to most holistic approaches, i.e.
+drivers characterization.
+We strived to make this as clear as possible throughout the manuscript, namely in the introduction and in the perspectives sections, as noted by the reviewer.
+As such, and considering the focus of the manuscript, we are hesitant to accept the reviewer’s suggestion to discuss cumulative effect assessment, best practices and indicator types at the beginning of the introduction.
+We feel that, considering the objective of this manuscript, the beginning of the introduction should focus specifically on drivers.
 
-> - We nonetheless agree with reviewer 1's suggestions.
-We are therefore proposing the following modifications to the manuscript that we hope consolidates the essence of the suggestions made by reviewer 1:
+> - We, nonetheless, agree with the reviewer’s suggestions and agree that a broader context is necessary to properly position our manuscript in the environmental assessment literature.
+We are, therefore, proposing the following modifications to the manuscript, as we hope that these will consolidate the essence of the suggestions made by the reviewer:
 
-  > 1. We modified the introduction to broadly state the major steps undertaken in holistic environmental approaches in general (lines 96-101).
+  > 1. We modified the introduction to broadly state the major steps that are generally undertaken in holistic environmental approaches (lines 101-105).
 
-  > 2. We modified the statement of the objectives of the manuscript at the end of the introduction to clearly state that we are focusing on drivers (lines 123-129).
+  > 2. We modified the statement of the objectives that appears at the end of the introduction to clearly state that we are specifically focusing on drivers (lines 130-136).
 
   > 3.
-We modified the pre-existing paragraph presenting the importance of valued components in the perspective sections (lines 492-514).
-It is now presenting *eDrivers* as a suite of stressor-based indicators that hold certain advantages, but fail to capture effects on valued components and non-additive effects.
-We present effects-based indicators and their importance for understanding cumulative effects.
-We then argue that both are jointly required to properly diagnose and attribute causes of ecosystem change to stressors (Dubé 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).
+We modified the paragraph in the perspective sections (lines 496-518) in which the importance of valued components is discussed.
+It now presents eDrivers as a suite of useful, stressor-based indicators that, nevertheless,  fail to capture effects on valued components and non-additive effects.
+We present effects-based indicators and discuss their importance for understanding cumulative effects.
+We then argue that both are required to properly diagnose and attribute causes of ecosystem change to stressors (Dubé 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).
 
@@
 
-> - We begin by thanking the reviewer for pointing out this obvious weakness in the manuscript.
-We realize now that, even though we did not want to mislead readers in thinking we were evaluating driver interactions, we are in fact doing just that.
-As mentioned by reviewer 1, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects.
-Under the assumption that driver co-occurrence is necessary for driver interactions to exist, we believe it is reasonable to predict that driver interactions and non-additive effects will most likely be observed in areas where drivers overlap at high relative intensities.
-These areas indeed delineate where the effects of both drivers are most likely to impact valued components simultaneously.
-In presenting sections 3.4 and 4.2, our goal was to provide an illustration of how a simple two-drivers case could be used to identify areas of such high relative intensity driver overlap.
-This should be used as a transition to the more integrative cumulative footprint and hotspot indicators.
-As rightfully identified by reviewer 1, however, the manuscript is indeed misleading on the topic of driver interactions.
+> - We thank the reviewer for pointing out this obvious weakness of the manuscript.
+We now realize that, even though we did not want to mislead readers in thinking we were evaluating driver interactions, we were doing just that.
+As noted by the reviewer, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects.
+That said, driver co-occurrence is necessary for interactions to exist, and therefore it is expected that interactions and non-additive effects will be most likely observed in areas where drivers overlap at high relative intensities.
+These areas delineate where the effects of both drivers are most likely to impact valued components simultaneously.
+In sections 3.4 and 4.2, we illustrate how a simple two-drivers case could be used to identify areas of such high relative intensity driver overlap.
+This example should serve as a stepping stone to the integrative cumulative footprint and hotspot indicators.
+As rightfully noted by the reviewer, however, the manuscript is misleading on the topic of driver interactions.
 We have therefore made the following modifications to the manuscript:
@@
   > 2.
-The methodology and results/discussions sections on driver interactions were moved to sections 3.3 and 4.1, respectively, and renamed *Driver co-occurrence*.
-We then modified the methods section as a simplified example of driver co-occurrence (lines 184-195).
-However, we did not modify the results/discussion section.
+The methodology and results/discussions sections on driver interactions were moved to sections 3.3 and 4.1, respectively, and renamed Driver co-occurrence.
+We then modified the methods section and provided a simplified example of driver co-occurrence (lines 194-204).
+We did not, however, modify the results/discussion section.
 The only mention of driver interactions was in the title itself.
-In the discussion, we merely discuss the areas of high intensity co-occurrence between the two drivers and present adverse effects that these drivers may have on species individually and simultaneously.
-We did add a closing statement to the section stating that individual and interactive effects of both drivers are species- and community-dependent and that the co-occurrence of drivers in itself is not an evidence of driver interaction.
-We then state that assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities (lines 274-277).
+In the discussion, we merely discuss the areas of high intensity of driver co-occurrence and describe individual and combined adverse effects that these drivers may have on species.
+
+  > 3.
+We clarified our stance on driver co-occurrence and risk of interactions in the introduction, noting that individual and interactive effects of both drivers are species- and community-dependent and that the co-occurrence of drivers in itself is not evidence of driver interaction.
+We then state that, assuming driver co-occurrence is a prerequisite for driver interactions, it expected that interactions will occur in areas where both drivers co-occur simultaneoustly (lines 93-96).
 
 > - Ultimately, this portion of the manuscript should serve as a simple illustration to better understand the complete cumulative footprint indicators.
-In this sense, if the reviewer still feels that proposed changes are not sufficient, we propose removing these sections from the manuscript entirely.
+In this sense, if the reviewer still feels that the proposed changes are insufficient, we are willing to remove these sections from the manuscript.
 
-> - As for providing flexibility in indicator definition, it is a feature that we wish to provide in future iterations of the package and the platform.
-We would, however, rather not include such enhancements at this point, as they are more technical in nature and we feel that they should not prevent us from publishing this work.
-Instead, we added a statement that presents this idea as a perspective and future work to be accomplished (lines 520-522), as suggested by reviewer 1.
+> - As for providing flexibility in indicator definition, it is a feature that we wish to provide in future iterations of the package and platform.
+We would, however, rather not include such enhancements in this manuscript, as they are more technical in nature and we feel that they should not prevent us from publishing this work.
+Instead, we present this idea as a perspective and of future work (lines 524-526), as suggested by the reviewer.
 
@@
 
-> - We agree with reviewer 1 on the importance of valued components for cumulative effects assessments.
-Presenting a full cumulative impact assessments and defining valued components are, however, not the objectives of this manuscript, nor is it to discuss cumulative impacts in particular.
-By themselves, we believe stressor-based indicators offer valuable insights in the dynamics of a system which can help managers in prioritizing decision-making and scientists in targeting specific areas for *in situ* measurements for hypothesis testing, especially when data on valued components are lacking.
+> - We agree with the reviewer on the importance of valued components for cumulative effects assessments.
+Presenting a full cumulative impact assessment and defining valued components are, however, not the objectives of this manuscript or of the platform, nor is it to discuss cumulative impacts in particular.
+By themselves, we believe stressor-based indicators offer valuable insights into the dynamics of a system which can help managers to prioritize decision-making and scientists in target specific areas for in situ measurements for hypothesis testing, especially when data on valued components are lacking.
 Furthermore, the choices of valued components are fully dependent on the objective of a particular assessment or management scheme.
-In our experience, time and money is spent on recurring initiatives that seek to characterize sources of stress in a system for impacts assessments and integrative management.
-Our goal was therefore to build an initiative that would limit this and provide access to formatted and comparable knowledge on sources of stress that are typically used for impact assessments and integrative management.
+In our experience, time and money are spent on recurring initiatives that seek to characterize sources of stress in a system for impacts assessments and integrative management.
+Our goal was therefore to build an initiative that would curb this approach and provide access to formatted and comparable knowledge on sources of stress that are typically used for impact assessments and integrative management.
 This integrative dataset can then be used and adapted to the need of a user to assess impacts on a single or multiple valued components of interest, or the user can identify gaps in knowledge in the platform and ultimately contribute to the knowledge contained in the platform.
 We were therefore very deliberate in our decision not to include and discuss specific valued components.
-This manuscript is in fact part of a broader project that aims at evaluating cumulative impact on the ecological communities of the St.
-Lawrence.
-The description of drivers will thus be used on specific valued components in the St.
-Lawrence System.
+This manuscript is part of a broader project that aims at evaluating cumulative impact on the ecological communities of the St. Lawrence.
+The description of drivers will thus be used on specific valued components in the St. Lawrence System.
 
-> - As proposed by reviewer 1, we had already included a paragraph on the importance of valued components and their vulnerability to sources of stress in the conclusion.
-We however realize now that it was underwhelming in its ability to provide perspective and direction as to how *eDrivers* may be used in the future.
-We therefore extended this paragraph, also in response to reviewer 1's first comment, to better discuss the significance of valued components and the different types of indicators for cumulative effects assessment (lines 492-514).
+> - We already had included a paragraph on the importance of valued components and their vulnerability to sources of stress.
+We now realize, however, that it was ineffectual to provide perspective and direction as to how eDrivers may be used in the future.
+We therefore extended this paragraph, also in response to the reviewer’s first comment, to better address the significance of valued components and the different types of indicators used for cumulative effects assessment (lines 496-518).
+-
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
-# Question 3
+## Question 3
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
-> - We have modified the sentences on lines 327-328 (now lines 344-348).
+> - We have modified the sentences on lines 327-328 (now lines 348-352).
 "Of particular concern are threat complexes 3 and 6, which are the two most exposed complexes in the St.
 Lawrence and are characterized by distinct cumulative exposure regimes between them, they capture most of the coastal and offshore hotspots identified in the St.
-Lawrence." now becomes "Threat complexes 3 and 6 capture most coastal and offshore hotspots and are the two most exposed complexes in the St.
+Lawrence." now becomes "Threat complexes 3 and 6 capture most of the coastal and offshore hotspots and are the two most exposed complexes in the St.
 Lawrence."
@@
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
-> - Figure caption was modified accordingly.
+> - The figure caption was modified accordingly.
 <!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
@@
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
-# Question 4
+## Question 4
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
@@
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
-# Question 5
+## Question 5
 <!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
@@
 The archive is ready to be published, but we are waiting for the publication process to be completed so that the proper citation can be referenced with the material.
-Similarly, the workflows used for the first iteration of the platform will be released and archived on Zenodo with the paper publication.
+Likewise, the workflows used for the first iteration of the platform will be released and archived on Zenodo with the paper publication.
 

### 2020-03-11 Revisions/Beauchesne-eDriversMS-SI.rmd
manuscript revisions to submit to Frontiers

@@
 
-Table S1. List of drivers currently available on *eDrivers* along with their respective acronym used in the figures in the supplementary material.
+Table S1. List of drivers currently available on *eDrivers* along with their respective acronym used in the figures \textcolor{red}{\sout{in}} \textcolor{blue}{of} the supplementary material.
 
@@
 Oceans are the largest reservoir\textcolor{blue}{s} and sink\textcolor{blue}{s} of atmospheric carbon dioxide ($CO_2$).
-Its uptake increases seawater acidity and \textcolor{red}{\sout{the}} lowers saturation state with
-respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common
-$CaCO_3$ polymorphs constituting the shells and skeleton\textcolor{blue}{s} of many marine
-organisms [@mucci2017]. This can have deleterious effects on carbonate-secreting
-organisms (*e.g.* mollusks and crustaceans) and certain physiological processes
-in non-calcifying organisms [@fabry2008; @kroeker2013].
+Its uptake increases seawater acidity and \textcolor{red}{\sout{the}} lowers \textcolor{blue}{the} saturation state \textcolor{blue}{of waters} with respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common $CaCO_3$ polymorphs \textcolor{blue}{that constitute} \textcolor{red}{\sout{constituting}} the shells and skeleton\textcolor{blue}{s} of many marine organisms [@mucci2017].
+\textcolor{blue}{Ocean acidification} \textcolor{red}{\sout{This}} can have deleterious effects on carbonate-secreting organisms (*e.g.* mollusks and crustaceans) and certain physiological processes in non-calcifying organisms [@fabry2008; @kroeker2013].
 
-When $\Omega_C$ or $\Omega_A$ decrease below 1, water becomes undersaturated and
-corrosive to the skeletal minerals of carbonate-secreting organisms. Note that
-if $\Omega_A$ $< 1$, the waters may still be supersaturated with respect to
-calcite since it is 50% more soluble than aragonite. However, organisms have
-wide ranging responses to $\Omega_C$ and $\Omega_A$ saturation state. For
-example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019],
-while other organisms may have adapted to precipitate $CaCO_3$ even when
-seawater is undersaturated with respect to calcite [$\Omega_C$ $< 1$ or
-$\Omega_A$ $< 0.65$; e.g. @uthicke2016].
+When $\Omega_C$ or $\Omega_A$ decrease below 1, water becomes undersaturated and corrosive to the skeletal minerals of carbonate-secreting organisms.
+Note that if $\Omega_A$ $< 1$, the waters may still be supersaturated with respect to calcite since it is 50% more soluble than aragonite.
+However, organisms have wide ranging responses to \textcolor{blue}{changes in $\Omega_C$ and $\Omega_A$} \textcolor{red}{\sout{saturation state}}.
+For example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019], while other organisms may have adapted to precipitate $CaCO_3$ even when seawater is undersaturated with respect to calcite [$\Omega_C$ $< 1$ or $\Omega_A$ $< 0.65$; e.g.
+@uthicke2016].
 
-We used $\Omega_{A}$ saturation state to characterize ocean acidification in the
-bottom waters of the St. Lawrence. The data come from the Department of
-Fisheries and Oceans' (DFO) Atlantic Zone Monitoring Program [AZMP; @galbraith2018]
-surveys and DFO's multispecies surveys [@bourdages2018] collected in
-August-September of 2017 [@starr2019].
-\textcolor{blue}{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
-stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
-majority of sampling stations were located in the Southern Gulf ($n = 84$),
-compared to the Estuary and Northern Gulf ($n = 33$).}
+We used $\Omega_{A}$ \textcolor{red}{\sout{saturation state}} to characterize ocean acidification in the bottom waters of the St.
+Lawrence.
+The data come from the Department of Fisheries and Oceans' (DFO) Atlantic Zone Monitoring Program [AZMP; @galbraith2018] surveys and DFO's multispecies surveys [@bourdages2018] collected in August-September of 2017 [@starr2019].
+\textcolor{blue}{$\Omega_{A}$, pH, and dissolved oxygen ($O_2$) were measured at 117 stations in the Estuary and Gulf of St.
+Lawrence in the summer of 2018.
+The majority of sampling stations were located in the Southern Gulf ($n = 84$), the Estuary and Northern Gulf ($n = 33$).}
 
-The carbonate chemistry was determined
-through pH and \textcolor{blue}{total alkalinity} (TA) measurements. Samples for pH and \textcolor{red}{\sout{total alkalinity}} TA were
-collected under \textcolor{red}{\sout{a}} bubble free and no head space conditions into 500 $mL$
-borosilicate glass flasks. 250 $\mu L$ of \textcolor{red}{\sout{the}} saturated $HgCl_2$ solution was
-added to sample, and processed following the "Guide to best practices for Ocean
+The carbonate \textcolor{red}{\sout{chemistry}} \textcolor{blue}{speciation} was determined through pH and \textcolor{blue}{total alkalinity} (TA) measurements.
+Samples for pH and \textcolor{red}{\sout{total alkalinity}} TA were collected \textcolor{red}{\sout{under a bubble free and}} \textcolor{blue}{with} no head space \textcolor{red}{\sout{conditions}} in\textcolor{red}{\sout{to}} 500 $mL$ borosilicate glass flasks\textcolor{red}{\sout{.}}\textcolor{blue}{, and} 250 $\mu L$ of \textcolor{red}{\sout{the}} saturated $HgCl_2$ solution was
+added to \textcolor{blue}{each} sample\textcolor{red}{\sout{,}} and processed following the "Guide to best practices for Ocean
 CO2 Measurements" [@dickson2007].
@@
 electrode (pHC2001) in a continuous titrant addition mode, an algorithm
-specifically designed for shallow end-point detection [@dickson2007]. The dilute\textcolor{blue}{d}
-HCl titrant 0.1M in a solution of NaCl of 0.6M was calibrated using CRM provided
-by Professor Andrew G. Dickson.
+specifically designed for shallow end-point detection [@dickson2007]. The
+\textcolor{red}{\sout{dilute}} \textcolor{blue}{0.1M} HCl titrant
+\textcolor{red}{\sout{0.1M}} in \textcolor{red}{\sout{a solution of}}
+\textcolor{blue}{0.6M} NaCl \textcolor{red}{\sout{of 0.6M}} was calibrated using CRM provided by Professor Andrew G. Dickson.
 
@@
 corrected for *in situ* pressure and temperature using the algorithm CO2SYS
-[@lewis1998] with measured pH, \textcolor{red}{\sout{total alkalinity}} \textcolor{blue}{TA}, soluble reactive
+[@lewis1998] with measured pH, \textcolor{red}{\sout{total alkalinity}} \textcolor{blue}{TA}, soluble reactive \textcolor{blue}{phosphate}
 and silicate concentrations as input parameters.
-
 \textcolor{red}{\sout{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
@@
 
-To account for low sample size in the Northern Gulf, we used the correlation
+To account for \textcolor{blue}{the} low sample size in the Northern Gulf, we used the correlation
 between $O_2$ and $CO_2$, which are linked through the stoichiometry of the
@@
 Metabolic $CO_2$ increases in water through biotic processes such microbial
-respiration of organic matter [@mucci2011]. In deep waters, variations of
-$\Omega_{A}$ are largely governed by the addition of metabolic $CO_2$, whereas
-near the air-sea interface $O_2$ is replenished faster than $CO_2$ can escape
+respiration of organic matter [@mucci2011]. In deep waters\textcolor{blue}{, found in the Laurentian, Anticosti and Esquiman Channels}, variations of
+$\Omega_{A}$ are largely governed by the addition of metabolic $CO_2$, whereas\textcolor{blue}{,} near the air-sea interface\textcolor{blue}{,} $O_2$ is replenished faster than $CO_2$ can escape
 [@zeebe2001]. The impact of respiration on carbonate chemistry is highest in
@@
 correlation observed in the north between $\Omega_A$ and $O_2$ is especially
-strong when deep Channels (*Deep* model: *p-value*: < 0.01; $O_2$: 0.20; $R^2$:
+strong when deep Channel\textcolor{red}{\sout{s}} \textcolor{blue}{waters} (*Deep* model: *p-value*: < 0.01; $O_2$: 0.20; $R^2$:
 0.97) are separated from shallower regions (*Shallow* model: *p-value*: < 0.01;
 $O_2$: 0.07; $R^2$: 0.96), with $\Omega_A$ increasing more steeply with
-increasing $O_2$ in deep Channels. This stronger correlation is likely caused
-by the lack of gas exchange with the atmosphere or the surface mixed layer and
-the importance of metabolic $CO_2$ in deep waters, some of which may have mean
-ages of up to 20 years [@mucci2011]. In the Northern Gulf, the acidification is
-following depth and distance from the Cabot Strait along the Laurentian,
-Anticosti and Esquiman Channels due to the progressive oxygen depletion and
-metabolic carbon dioxide accumulation [@mucci2011].
+increasing $O_2$ in \textcolor{blue}{the} deep Channels.
+This stronger correlation is likely caused by the lack of gas exchange with the atmosphere or the surface mixed layer and the importance of metabolic $CO_2$ in deep waters, some of which may have mean ages of up to 20 years [@mucci2011].
+In the Northern Gulf, the acidification \textcolor{red}{\sout{is following}} \textcolor{blue}{trends with} depth and distance from the Cabot Strait along the Laurentian, Anticosti and Esquiman Channels due to the progressive oxygen depletion and metabolic carbon dioxide accumulation [@mucci2011].
 
-There are likely different dynamics at play in the Southern Gulf, such as
-waters with lowered alkalinity and salinity caused by the St. Lawrence plume.
-Such water masses \textcolor{red}{\sout{are}} have lower buffering capacities to decreasing pH and
-$\Omega_{A}$. Also, the accumulation of metabolic carbon dioxide produced by
-rapid oxic degradation of organic matter in the sandy
-and permeable sediments of the Magdalen Shelf or through the St. Lawrence plume
-could decrease $\Omega_{A}$ in the Southern Gulf (Starr and Chassé, *personal
-communication*).
+There are likely different dynamics at play in the Southern Gulf, such as waters with lowered alkalinity and salinity caused by the \textcolor{blue}{surface, seaward-flowing Gaspé Current} \textcolor{red}{\sout{St.
+Lawrence plume}}.
+\textcolor{red}{\sout{Such}} \textcolor{blue}{These} water\textcolor{blue}{s} \textcolor{red}{\sout{masses}} \textcolor{red}{\sout{are}} have lower buffering capacities to decreasing pH and $\Omega_{A}$.
+\textcolor{red}{\sout{Also}} \textcolor{blue}{Furthermore}, the accumulation of metabolic carbon dioxide produced by rapid oxic degradation of organic matter in the sandy and permeable sediments of the Magdalen Shelf or \textcolor{red}{\sout{through}} \textcolor{blue}{in} the \textcolor{red}{\sout{St.
+Lawrence plume}} \textcolor{blue}{Gaspé current} could decrease $\Omega_{A}$ in the Southern Gulf (Starr and Chassé, *personal communication*).
 
-We thus divided the analyses between the Southern and Northern St. Lawrence.
-The St. Lawrence was divided using the 350m isobath on the southern
-slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé
-Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modeled
+We thus divided the \textcolor{red}{\sout{analyses}} \textcolor{blue}{and their analysis} between the Southern and Northern St.
+Lawrence.
+The St.
+Lawrence was divided \textcolor{red}{\sout{using}} \textcolor{blue}{along} the 350m isobath on the southern slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé Peninsula.
+\textcolor{blue}{The} $\Omega_A$ distribution in the Southern St.
+Lawrence was modeled
 using the exponential kriging model.
 
-We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. \textcolor{red}{\sout{Whereas}} \textcolor{blue}{While}
+We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. Whereas
 there are few published datasets on $\Omega_{A}$ in the St.
-Lawrence, there is a robust time-series on oxygen concentration in the water
-column from the Department of Fisheries and Oceans’ (DFO) Atlantic Zone
+Lawrence, there is a robust time-series on oxygen concentration\textcolor{blue}{s} in the water column \textcolor{red}{\sout{from}} \textcolor{blue}{available through}
+the Department of Fisheries and Oceans’ (DFO) Atlantic Zone
 Monitoring Program [AZMP; @galbraith2018]. We used oxygen concentration data
 collected between 2013 and 2017
-[@blais2019] and interpolated oxygen % saturation using cokriging with depth as a
-covariable, as done in [@dutil2011]. Bathymetric data come from @dutil2011
+[@blais2019] and interpolated oxygen % saturation using cokriging with depth as a covariable, as done in @dutil2011. Bathymetric data come from @dutil2011
 and have a 100 $km^2$ resolution. For more details, refer to the hypoxia \textcolor{blue}{data layer}
@@
 
-Based on the wide range of $\Omega_A$ levels at which organism are negatively
+Based on the wide range of $\Omega_A$ levels at which organism\textcolor{blue}{s} are negatively
 affected (see above), we built an index of acidification stress ($A_s$)
-ranging between 0 and 1 that slowly \textcolor{red}{\sout{begins increasing}} \textcolor{blue}{increases} as $\Omega_A$ $\approx$
+ranging between 0 and 1 that slowly \textcolor{red}{\sout{begins increasing}} \textcolor{blue}{increases} as $\Omega_A$ \textcolor{blue}{decreases to} $\approx$
 $2$, increases more rapidly close to $\Omega_A$ $= 1$ and reaches its peak at
@@
 
-The data used to characterize hypoxia \textcolor{red}{\sout{uses oxygen saturation in % and}} come from
+The data used to characterize hypoxia
+\textcolor{red}{\sout{uses oxygen saturation in \% and}}
+come from
 the Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
-[AZMP; @galbraith2018] survey\textcolor{red}{\sout{s}} in late spring and fall, as well as from \textcolor{red}{\sout{annual}}
+[AZMP; @galbraith2018] survey\textcolor{red}{\sout{s}} in late spring and fall
+\textcolor{blue}{of 2013 to 2017}, as well as from \textcolor{red}{\sout{annual}}
 DFO’s \textcolor{blue}{annual} multispecies surveys for the northern Gulf in August and for the Magdalen
@@
 
-Oxygen concentration is measured at every station using an oxygen probe
+Oxygen concentration \textcolor{red}{\sout{is}} \textcolor{blue}{were} measured at every station using an oxygen probe
 (Sea-Bird SBE43) mounted on \textcolor{red}{\sout{the}} \textcolor{blue}{a} CTD. The probe is calibrated against seawater
-samples that are analyzed by Winkler titration on every cast
-[see @blais2019 for calibration procedure]. The data used is the last depth
+samples \textcolor{red}{\sout{that are}} \textcolor{blue}{taken directly from the Niskin bottles at every cast and} analyzed by Winkler titration \textcolor{red}{\sout{on every cast}} [see @blais2019 for calibration procedure].
+The data used is the last depth
 sampled on the CTD profile, which \textcolor{blue}{is} typically \textcolor{red}{\sout{corresponds to}} ~10m above the
-bottom. Oxygen concentration ($\mu M$ $L^{-1}$) is converted to oxygen
+bottom.
+\textcolor{blue}{Note that this depth does not necessarily correspond to the depth of the oxygen minimum.
+The oxygen minimum is typically found along the 27.25 isopycnal, often situated as much as 100 m above bottom in the Lower Estuary.} Oxygen concentration ($\mu M$ $L^{-1}$) is converted to oxygen
 saturation\textcolor{blue}{,} taking into account the salinity and temperature at the selected
@@
 We interpolated oxygen saturation using cokriging with depth as a covariable,
-as done in [@dutil2011]. Bathymetric data come from @dutil2011 and have a
-100 $km^2$ resolution. According to @diaz1995, severe hypoxia can be observed
-when dissolved oxygen falls below 2 $ml$ $L^{-1}$, which corresponds to 62.5
-$\mu mol$ $L^{-1}$ and ~20% saturation. This is considered as the level
-necessary to maintain most animal life. Similarly, @chabot2008 studied the
+as done in @dutil2011. Bathymetric data come from @dutil2011 and have a
+100 $km^2$ resolution. According to @diaz1995, severe hypoxia \textcolor{red}{\sout{can be observed}} \textcolor{blue}{occurs}
+when dissolved oxygen falls below 2 $ml$ $L^{-1}$, \textcolor{red}{\sout{which corresponds}} to 62.5 $\mu mol$ $L^{-1}$ \textcolor{red}{\sout{and}} \textcolor{blue}{or} ~20% saturation.
+This is considered as the level
+necessary to maintain most animal life. \textcolor{red}{\sout{Similarly}} \textcolor{blue}{Accordingly}, @chabot2008 studied the
 effects of hypoxia on the energy budget of cod in the St. Lawrence and found
-that behavioural effects began manifesting below 70% oxygen saturation and that
+that behavioural effects began manifesting \textcolor{blue}{themselves} below 70% oxygen saturation and that
 survival becomes jeopardized below ~20% saturation. We used these observations
 to create an index of hypoxia using an inverted logistic curve that slowly
-increases below 70% oxygen saturation, increases acutely as it approaches the
-20% oxygen saturation threshold and reaches its peak intensity between 20 and
-30% oxygen saturation:
+increases below 70% oxygen saturation\textcolor{red}{\sout{,}} \textcolor{blue}{and} increases acutely as it \textcolor{red}{\sout{approaches the 20\% oxygen saturation threshold and}} reaches its peak intensity between 20 and 30% oxygen saturation:
 
@@
 
-### Bottom temperature anomalies
+### Bottom\textcolor{blue}{-water} temperature anomalies
 
-The data used to characterize bottom temperature anomalies come from the
+The data used to characterize bottom\textcolor{blue}{-water} temperature anomalies come from the
 Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
 [AZMP; @galbraith2018]. We provide a brief summary of data and methods to
-characterize the bottom temperature climatology and anomalies in this document.
+characterize the bottom\textcolor{blue}{-water} temperature climatology and anomalies in this document.
 For more details, refer to @galbraith2018.
 
-Bottom temperatures are interpolated in the Gulf using
+Bottom\textcolor{blue}{-water} temperatures are interpolated in the Gulf using
 conductivity-temperature-depth (CTD) sampling performed annually through DFO's
@@
 Shallows in September. Using this sampling survey, temperatures are horizontally
-interpolated at each 1 m depth layer on a 2 km resolution grid. Bottom
+interpolated at each 1 m depth layer on a 2 km resolution grid. Bottom\textcolor{blue}{-water}
 temperatures are then extracted by using a bathymetry layer from the Canadian
@@
 We used temperature anomalies, *i.e.* deviations from long-term normal
-conditions, to measure an annual index of stress associated with extreme
+conditions \textcolor{blue}{between 1981 and 2010}, to measure an annual index of stress associated with extreme
 temperatures between 2013 and 2017. Temperature anomalies were calculated using
@@
 and the absolute value of anomalies were used as an indicator of the intensity
-of bottom temperature anomalies. The mean anomaly intensity between 2013 and
-2017 for each grid cell was used to generate the final index of bottom
+of bottom\textcolor{blue}{-water} temperature anomalies. The mean anomaly intensity between 2013 and
+2017 for each grid cell was used to generate the final index of bottom\textcolor{blue}{-water}
 temperature anomalies.
 
-### Surface temperature anomalies
+### Surface\textcolor{blue}{-water} temperature anomalies
 
-The data used to characterize surface temperature anomalies come from the
+The data used to characterize surface\textcolor{blue}{-water} temperature anomalies come from the
 Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
 [AZMP; @galbraith2018]. We provide a brief summary of data and methods to
-characterize the surface temperature climatology and anomalies in this document.
+characterize the surface\textcolor{blue}{-water} temperature climatology and anomalies in this document.
 For more details, refer to @galbraith2018.
 
-The surface layer is characterized using sea surface temperature (SST) monthly
+The surface\textcolor{blue}{-water} layer is characterized using sea surface\textcolor{blue}{-water} temperature (SST) monthly
 composites from Advanced Very High Resolution Radiometer (AVHRR) satellite
@@
 
-Surface temperature anomalies were characterized following the same method used
-for bottom temperature anomalies. Only the months of May to November were
+Surface\textcolor{blue}{-water} temperature anomalies were characterized following the same method used
+for bottom\textcolor{blue}{-water} temperature anomalies. Only the months of May to November were
 included to avoid biases associated with the presence of ice cover. Monthly
 anomalies from May to November values were summed to obtain an indicator of
-annual surface temperature anomaly intensity in each grid cell. The mean
+annual surface\textcolor{blue}{-water} temperature anomaly intensity in each grid cell. The mean
 anomaly intensity between 2013 and 2017 for each grid cell was used to generate
-the final index of surface temperature anomalies.
+the final index of surface\textcolor{blue}{-water} temperature anomalies.
 
-### Sea level rise
+### Sea\textcolor{blue}{-}level rise
 
-The data used to characterize sea level rise risk come from the global
+The data used to characterize sea\textcolor{blue}{-}level rise risk come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 
-Sea level rise was characterized by @nicholls2010 using NASA's satellite
+Sea\textcolor{blue}{-}level rise was characterized by @nicholls2010 using NASA's satellite
 altimetry data (Topex/Poseidon, Jason-1&2, GFO, ERS-1&2, and Envisat
@@
 
-The rate of sea level rise ($mm/year$) was measured between 1992 and 2012 and
+The rate of sea\textcolor{blue}{-}level rise ($mm/year$) was measured between 1992 and 2012 and
 transformed as a net change value ($mm$) by multiplying by the number of years
 considered. Only positive values were selected under the assumption that
-only positive sea level rise is likely to cause environmental stress.
+only positive sea\textcolor{blue}{-}level rise is likely to cause environmental stress.
 
@@
 - Organic pollution
-- Sea bottom temperature anomalies
+- \textcolor{red}{\sout{Sea}} Bottom\textcolor{blue}{-water} temperature anomalies
 - Shipping

### 2020-03-11 Revisions/Beauchesne-eDriversMS.Rmd
manuscript revisions to submit to Frontiers

@@
 with climate driven environmental changes. Our objective in this paper is to
-systematically characterize the distribution and intensity of drivers in the
+systematically characterize the distribution and intensity of drivers \textcolor{blue}{of change} in the
 St. Lawrence System. To do so, we launch *eDrivers*, an open knowledge
 platform gathering experts committed to structuring, standardizing and sharing
-knowledge on drivers in support of science and management. We gathered data on
+knowledge on drivers \textcolor{blue}{of change} in support of science and management. We gathered data on
 22 coastal, climate, fisheries and marine traffic drivers through
@@
 especially in the vicinity of urban centers. We identified 6 areas of distinct
-cumulative exposure regime that show that certain drivers typically co-occur
+cumulative exposure regime that show that certain drivers \textcolor{blue}{of change} typically co-occur
 in different regions of the St. Lawrence and that coastal areas are exposed to
@@
 
-The St. Lawrence System, formed by one of the largest estuaries in the world,
-the St. Lawrence Estuary, and a vast interior sea, the Gulf of St. Lawrence, is
+The St. Lawrence System, formed by one of the largest estuaries in the world\textcolor{red}{\sout{, the St. Lawrence Estuary,}}
+and a vast interior sea\textcolor{red}{\sout{, the Gulf of St. Lawrence}}, is
 a complex social-ecological system characterized by highly variable
@@
 productive ecological communities [@savenkoff2000]. As a result, the St.
-Lawrence System provides a wealth of ecosystem services that have historically
-and contemporarily benefited the Canadian economy. It sustains a rich fisheries
+Lawrence System provides a wealth of ecosystem services that have \textcolor{red}{\sout{historically and contemporarily}}
+benefited the Canadian economy. It sustains a rich fisheries
 industry targeting more than 50 species, serves as the gateway to eastern
@@
 will likely result in increasingly intricate environmental exposure regimes,
-*i.e.* suites of overlapping and potentially interacting environmental drivers
-threatening ecosystems, habitats or ecological communities. Drivers, often
-referred to as stressors or pressures, are any externalities that affect
+*i.e.* suites of overlapping \textcolor{red}{\sout{and potentially interacting}} environmental drivers \textcolor{blue}{of change}
+threatening ecosystems, habitats or ecological communities.
+Drivers \textcolor{blue}{of change}, often referred to as stressors or pressures \textcolor{blue}{and hereafter called drivers for simplicity}, are any externalities that affect
 environmental processes and disturb natural systems. Drivers may originate from
-natural or human-induced biophysical processes (*e.g.* sea surface temperature
+natural or human-induced biophysical processes (*e.g.* sea surface\textcolor{blue}{-water} temperature
 anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
 fisheries and marine pollution). The potential for complex interactions between
-drivers is the largest uncertainty when studying or predicting environmental
-impacts [@darling2008; @cote2016]. Multiple drivers can combine
-non-linearly and result in effects that are greater (synergistic effect) or
-lower (antagonistic effect) than the sum of individual effects
+\textcolor{blue}{co-occurring} drivers is the largest uncertainty when studying or predicting environmental impacts [@darling2008; @cote2016].
+Multiple drivers can combine non-linearly and result in effects that are greater (synergistic effect) or lower (antagonistic effect) than the sum of individual effects
 [@crain2008; @darling2008; @cote2016].
@@
 The uncertainty associated with complex driver interactions must therefore be taken into account when investigating environmental impacts [@cote2016], yet most research on driver effects in marine environments remains overwhelmingly focused on single driver assessments [@obrien2019].
-Increasing exposure and the experiences of past ecological tragedies such as the collapse of cod fisheries [@frank2005; @dempsey2018] and the decline of the beluga and right whale populations [@plourde2014] together urge the need to characterize the distribution, intensity and overlap between drivers in the St.
+Increasing exposure and the experiences of past ecological tragedies such as the collapse of cod fisheries [@frank2005; @dempsey2018] and the decline of the beluga and right whale populations [@plourde2014] together urge the need to characterize the distribution, intensity and \textcolor{red}{\sout{overlap}} \textcolor{blue}{co-occurrence} between drivers in the St.
 Lawrence System.
-This will provide critical information on areas most exposed to \textcolor{red}{\sout{cumulative}} drivers \textcolor{red}{\sout{and on the interaction potential of drivers}} in the St.
+\textcolor{blue}{The impacts of drivers, whether individually or in combination, depend on species-specific vulnerabilities.
+Driver co-occurrence alone is therefore not sufficient evidence to claim interactions among drivers.
+However, as drivers must co-occur to interact, the likelihood of interactions will increase with the number of drivers co-occurring.} \textcolor{red}{\sout{This will}} \textcolor{blue}{Describing drivers will therefore} provide critical information on areas most exposed to \textcolor{red}{\sout{cumulative}} drivers \textcolor{red}{\sout{and on the interaction potential of drivers}} in the St.
 Lawrence.
@@
 
-\textcolor{blue}{Gathering knowledge is a major component of any holistic approach, which typically involve the selection and description of valued ecosystem component (\textit{e.g.} habitats or species) and driver indicators, the assessment of the exposure and vulnerability of valued components to drivers, the selection of a proper spatio-temporal scale, site monitoring, and public and stakeholder participation (Dubé and Munkittrick 2001).} <!-- [@dube2001] jones2016 ????
+\textcolor{blue}{Holistic approaches typically involve, but are not limited to, the following steps: selecting and describing valued ecosystem components (\textit{e.g.} habitats and species) and drivers (\textit{e.g.} marine traffic and ocean acidification), assessing the exposure and vulnerability of valued components to drivers, selecting a proper spatio-temporal scale, monitoring, and public and stakeholder participation (Dubé and Munkittrick 2001).} <!-- [@dube2001] jones2016 ????
 Noble2013???
 -->
-Gathering environmental \textcolor{red}{\sout{data}} \textcolor{blue}{knowledge} for large scale, \textcolor{red}{\sout{systematic}} \textcolor{blue}{holistic} initiatives can, however, be a very challenging and time consuming – not to say painful – process.
+Gathering environmental
+\textcolor{red}{\sout{data for large scale, systematic}}
+\textcolor{blue}{knowledge for holistic}
+initiatives can, however, be a very challenging and time consuming – not to say painful – process.
 On one hand, there is an overwhelming and expanding wealth of data available.
 Such information overload may inhibit our ability to make decisions based on scientific information, promote massive effort duplication, disproportionately appropriate research funds to certain sectors, and obscure knowledge gaps amid a sea of information [@eppler2004].
-On the other hand, crucial data are lacking and remain largely unavailable or inaccessible for a variety of reasons, including proprietary rights, lack of organizational time, capacity and training, and in some rare cases unwillingness to share, curtailing our ability for appropriate decision-making.
+On the other hand, crucial data are lacking and remain largely unavailable or inaccessible for a variety of reasons, including proprietary rights, lack of organizational time, capacity and training, and in some \textcolor{red}{\sout{rare}} cases unwillingness to share, curtailing our ability for appropriate decision-making.
 
-There are now initiatives that address this issue by assembling, organizing and
-sharing environmental knowledge, such as the Ocean Biogeographic Information
+There are now initiatives that \textcolor{blue}{facilitate the data gathering process} \textcolor{red}{\sout{address this issue}} by assembling, organizing and sharing environmental knowledge, such as the Ocean Biogeographic Information
 System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for
@@
 effects of multiple drivers [@dafforn2016; @stock2018]. An additional objective
-thus emerged in the process of addressing our initial goals: sharing the
-knowledge gathered through the description of drivers in the St.
-Lawrence.
+thus emerged in the process of addressing our initial goals: \textcolor{red}{\sout{sharing the knowledge gathered through the description of drivers in the St.
+Lawrence.}} \textcolor{blue}{sharing information about the distribution and intensity of drivers of change in the St.Lawrence.}
 
@@
 \textcolor{red}{\sout{Here we focus solely on drivers, one of four elements required to evaluate environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.* habitats or species) and the exposure and the vulnerability of the ecosystem components to drivers.}}
-\textcolor{blue}{Here we focus solely on drivers, aiming at better understanding their distribution in the St.
+\textcolor{blue}{Here, we focus solely on drivers, aiming at better understanding their distribution in the St.
 Lawrence and at facilitating the widespread availability of their characterization for holistic assessments and management approaches.}
@@
 The St. Lawrence drains over 25% of global freshwater reserves through its
-connexion to the Great Lakes Basin. The Basin is home to over 45 million
+connexion to the Great Lakes Basin\textcolor{red}{\sout{. The Basin}}\textcolor{blue}{, which }is home to over 45 million
 North-Americans, *i.e.* 15 and 30 million in Canada and the United States,
-respectively [@archambault2017]. The St. Lawrence itself boasts a much lower
+respectively. The St. Lawrence itself boasts a much lower
 population of approximately 1 million Canadians living within 10 km of the
@@
 
-The list of drivers for which we sought data was informed by a global
-cumulative impact assessment initiative [@halpern2008a; @halpern2015a],
+\textcolor{red}{\sout{The list of drivers for which we sought data was informed by}} \textcolor{blue}{Drivers selection was informed by} a global
+cumulative impact assessment initiative [@halpern2008a; @halpern2015a] \textcolor{blue}{and available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository (Halpern et al.
+2015)},
 regional holistic evaluations of the state of the St. Lawrence [@dufour2007;
-@benoit2012], and communications with regional experts. We integrated datasets
-from regional experts and also use global data from the global cumulative
-impact assessment initiative [@halpern2008a; @halpern2015a] available from the
-National Center for Ecological Analysis and Synthesis (NCEAS) online data
-repository [Table 1; @halpern2015]. We selected global data that were
+@benoit2012], and communications with regional experts (Table 1).
+\textcolor{red}{\sout{We integrated datasets from regional experts and also use global data from the global cumulative impact assessment initiative [@halpern2008a; @halpern2015a]  available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository [Table 1; @halpern2015].}} We selected global data that were
 unavailable at the regional scale and that were available at a resolution
@@
 
-Drivers with non-normal frequency distributions were log-transformed (Figure S1)
-and all drivers were scaled between 0 and 1 to allow driver comparisons. The
-99th quantile of individual driver distribution was used as the upper bound for
-scaling to control for extreme values and produce maps of individual drivers
-(Figure S2). All drivers were embedded in a regular grid composed of 245604
-$1 km^2$ hexagonal cells to construct the integrated dataset used for the
-analyses.
+Drivers with non-normal frequency distributions were log-transformed,
+\textcolor{blue}{as in Halpern et al.
+2008, to minimize the impact of extreme outliers that may or may not be real observations} (Figure S1)\textcolor{blue}{,} and all drivers were scaled between 0 and 1 to allow driver comparisons.
+The 99th quantile of individual driver distribution was used as the upper bound for scaling \textcolor{red}{\sout{to control for extreme values and produce maps of individual drivers (Figure S2)}}.
+All drivers were embedded in a regular grid composed of 245604 $1 km^2$ hexagonal cells to construct the integrated dataset used for the
+analyses \textcolor{blue}{(Figure S2)}.
 
@@
 Lawrence and, hence, an interaction between the effects of the two drivers could be anticipated.}}
-\textcolor{blue}{As an introduction to indicators of cumulative exposure (objective 1), we begin by presenting a simplifed 2-driver example focusing on hypoxia and demersal destructive fisheries, two drivers known to occur mainly in deeper areas of the St.
+\textcolor{blue}{As an introduction to indicators of cumulative exposure (objective 1), we begin by presenting a simplifed 2-driver example focusing on hypoxia and demersal destructive fisheries, two drivers known to occur mainly in deeper waters of the St.
 Lawrence.
 Driver co-ccurrence was evaluated spatially by summing the scaled intensity of drivers in each grid cell.
-The intensity at which pairs of drivers co-occur was evaluated using a two-dimensional kernel density estimate to investigate the relative intensity at which these two drivers typically co-occur.}
+The intensity at which pairs of drivers co-occur was evaluated using a two-dimensional kernel density.}
 
@@
 
-In order to \textcolor{red}{\sout{better capture potential interactions between drivers (objective 2) and to}}identify areas with similar exposure regimes (objective 3), we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
+In order to \textcolor{red}{\sout{better capture potential interactions between drivers (objective 2) and to}}identify areas with similar exposure regimes (objective \textcolor{red}{\sout{3}}\textcolor{blue}{2}), we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
 We use the term clusters in presenting the methods, but use threat complex when discussing the results on cumulative exposure regimes.
@@
 Certain species may thus be adversely affected by fisheries and withstand hypoxia but still experience a decrease in prey availability, while others may be deleteriously affected by the compounded effect of both drivers [@deleo2017].
-
-\textcolor{blue}{The individual or interactive effect of both drivers is thus species- and community-dependent and the co-occurrence of drivers in itself is not an evidence of driver interaction.
-Assuming that driver co-occurrence is a prerequisite for driver interactions, however, it is reasonable to expect interactions to occur in areas where both drivers co-occur at high relative intensities.
-}
 
@@
 Lawrence, drivers are but one of the 4 elements required to properly evaluate environmental impacts.
-This single knowledge node should be weaved with other, comparable, nodes required for integrated management approaches such as species distribution and marine protected areas.
-Ultimately, all of these knowledge nodes could be weaved together through social-ecological metanetworks (Dee et al.
+This knowledge node should be weaved with other, comparable, nodes required for integrated management approaches such as species distribution and marine protected areas.
+Ultimately, all of these knowledges could be weaved together through social-ecological metanetworks (Dee et al.
 2017).}}
-\textcolor{blue}{A caveat to our initiative is that it focuses on a single element required for fully operational impact assessments.
+\textcolor{blue}{The next step will be the inclusion of other types of knowledge to our initiative.
+Our focus has been on a single element required for fully operational impact assessments.
 Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators (Dube and Munkittrick 2001; Dubé 2003; Jones 2016).
-These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers and fail to diagnose impacts on valued components or non-additive effects (Dube and Munkittrick 2001; Dubé 2003).
+These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers but fail to diagnose impacts on valued components or non-additive effects (Dube and Munkittrick 2001; Dubé 2003).
 In contrast, effect-based indicators are direct measurements of valued components (\textit{e.g.} species abundance and biodiversity) and inherently capture the effects of multiple drivers (Dube and Munkittrick 2001; Dubé 2003).
-While effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers (Dube and Munkittrick 2001; Dubé 2003).
-Stressor-based and effect-based indicators are therefore both required to diagnose causes of ecosystem change (Dube 2003; Jones 2016).
-As a knowledge node on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, nodes describing valued ecosystem components that can be linked to drivers and allow a better understanding of cumulative impacts.
-Ultimately, multiple knowledge nodes could be weaved together through social-ecological metanetworks analyses (Dee et al.
+Whereas effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers (Dube and Munkittrick 2001; Dubé 2003).
+Stressor-based and effect-based indicators are, therefore, both required to diagnose causes of ecosystem change (Dube 2003; Jones 2016).
+As a collection of knowledge on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, collections of knowledge describing valued ecosystem components that can be linked to drivers and allow for a better understanding of cumulative impacts.
+Ultimately, interdisciplinary colections of knowledge could be weaved together through social-ecological metanetworks analyses (Dee et al.
 2016).}

### 2020-03-11 Tables/datatable.R

@@ -14,11 +14,11 @@ df <- data.frame(Groups = character(0),
 
 df[1,] <- c("Climate", "Aragonite", 'Lat/long', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ACID')
 df[2,] <- c("Coastal", "Aquaculture", 'Lat/long', '-','Variable, between 1990-2016', '$presence-absence$','[@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016]','AQUA')
-df[3,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SBT-')
-df[4,] <- c("Climate", "Sea bottom temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'positie anomalies', '[@galbraith2018]','SBT+')
-df[5,] <- c("Climate",  "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SST-')
-df[6,] <- c("Climate", "Sea surface temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'positive anomalies', '[@galbraith2018]','SST+')
-df[7,] <- c("Climate", "Sea level rise", 'Modeled 0.25 degree', '10 days', '1992-2012', '$mm$','[@halpern2015]','SLR')
+df[3,] <- c("Climate", "Bottom-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SBT-')
+df[4,] <- c("Climate", "Bottom-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'positie anomalies', '[@galbraith2018]','SBT+')
+df[5,] <- c("Climate",  "Surface-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SST-')
+df[6,] <- c("Climate", "Surface-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'positive anomalies', '[@galbraith2018]','SST+')
[…]

### 2020-03-11 Revisions/Beauchesne-eDriversMS.Rmd
update references

@@
 
-\textcolor{blue}{Holistic approaches typically involve, but are not limited to, the following steps: selecting and describing valued ecosystem components (\textit{e.g.} habitats and species) and drivers (\textit{e.g.} marine traffic and ocean acidification), assessing the exposure and vulnerability of valued components to drivers, selecting a proper spatio-temporal scale, monitoring, and public and stakeholder participation (Dubé and Munkittrick 2001).} <!-- [@dube2001] jones2016 ????
-Noble2013???
--->
+\textcolor{blue}{Holistic approaches typically involve, but are not limited to, the following steps: selecting and describing valued ecosystem components (\textit{e.g.} habitats and species) and drivers (\textit{e.g.} marine traffic and ocean acidification), assessing the exposure and vulnerability of valued components to drivers, selecting a proper spatio-temporal scale, monitoring, and public and stakeholder participation} [@dube2001].
 Gathering environmental
@@
 Our focus has been on a single element required for fully operational impact assessments.
-Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators (Dube and Munkittrick 2001; Dubé 2003; Jones 2016).
-These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers but fail to diagnose impacts on valued components or non-additive effects (Dube and Munkittrick 2001; Dubé 2003).
-In contrast, effect-based indicators are direct measurements of valued components (\textit{e.g.} species abundance and biodiversity) and inherently capture the effects of multiple drivers (Dube and Munkittrick 2001; Dubé 2003).
-Whereas effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers (Dube and Munkittrick 2001; Dubé 2003).
-Stressor-based and effect-based indicators are, therefore, both required to diagnose causes of ecosystem change (Dube 2003; Jones 2016).
-As a collection of knowledge on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, collections of knowledge describing valued ecosystem components that can be linked to drivers and allow for a better understanding of cumulative impacts.
-Ultimately, interdisciplinary colections of knowledge could be weaved together through social-ecological metanetworks analyses (Dee et al.
-2016).}
+Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators} [@dube2001; @dube2003; @jones2016].
+\textcolor{blue}{These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers but fail to diagnose impacts on valued components or non-additive effects} [@dube2001; @dube2003].
+\textcolor{blue}{In contrast, effect-based indicators are direct measurements of valued components (\textit{e.g.} species abundance and biodiversity) and inherently capture the effects of multiple drivers} [@dube2001; @dube2003].
+\textcolor{blue}{Whereas effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers} [@dube2001; @dube2003].
+\textcolor{blue}{Stressor-based and effect-based indicators are, therefore, both required to diagnose causes of ecosystem change} [@dube2003; @jones2016].
+\textcolor{blue}{As a collection of knowledge on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, collections of knowledge describing valued ecosystem components that can be linked to drivers and allow for a better understanding of cumulative impacts.
+Ultimately, interdisciplinary colections of knowledge could be weaved together through social-ecological metanetworks analyses} [@dee2017].
 
@@
 Port of Sept-Îles and City of Sept-Îles).
-
-<!-- !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! -->
-TODO: Remove [@dube2001; @dube2003; @jones2016]
-<!-- !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! -->
 

### 2020-04-18 Revisions2/Beauchesne-eDriversMS.Rmd
prepare 2nd round of review and modify figure 1

--- 2020-03-11 Revisions/Beauchesne-eDriversMS.Rmd
@@
 ---
-title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System
+title: Next Generation Planning -- Characterizing Exposure to and Shared Knowledge on Drivers of Ecological Change in the St. Lawrence System
 author: David Beauchesne^1,2,\*^, Rémi M.
@@
 <!--
-rmarkdown::render('./Revisions/Beauchesne-eDriversMS.Rmd', 'pdf_document')
-rmarkdown::render('./eDriversMS-v2.rmd', 'word_document')
+rmarkdown::render('./Revisions2/Beauchesne-eDriversMS.Rmd', 'pdf_document')
+-->
+
+<!--
+To discuss with co-authors
+
+- New article title
+- Should we present how the platform is used in the analyses or really keep
+  that portion fully separate? If we include it, modify answer to question 2
+  comment 1
+
 -->
@@
 The St. Lawrence is a vast and complex socio-ecological system providing a
-wealth of services sustaining numerous economic sectors. These ecosystems are
+wealth of services sustaining numerous economic sectors. ~~\textcolor{red}{These ecosystems are}~~
+\textcolor{blue}{This ecosystem is}
 subject to significant human pressures that overlap and potentially interact
@@
 systematically characterize the distribution and intensity of drivers \textcolor{blue}{of change} in the
-St. Lawrence System. To do so, we launch *eDrivers*, an open knowledge
+St. Lawrence System. ~~To do so, we launch *eDrivers*, an open knowledge
 platform gathering experts committed to structuring, standardizing and sharing
-knowledge on drivers \textcolor{blue}{of change} in support of science and management. We gathered data on
+knowledge on drivers \textcolor{blue}{of change} in support of science and management.~~ We gathered data on
 22 coastal, climate, fisheries and marine traffic drivers through
@@
 show that few areas of the St. Lawrence are free of cumulative exposure. The
-Estuary, the Anticosti Gyre and coastal areas are particularly exposed,
+Estuary, ~~the~~ Anticosti Gyre and coastal areas are particularly exposed,
 especially in the vicinity of urban centers. We identified 6 areas of distinct
@@
 most exposure hotspots that show the convergence of contrasting exposure
-regimes at the head of the Laurentian Channel. *eDrivers* was built on a
+regimes at the head of the Laurentian Channel.
+\textcolor{blue}{Sharing the knowledge acquired emerged as a priority to minimize future
+environmental assessment efforts. We thus launch *eDrivers*, an open knowledge
+platform gathering experts committed to structuring, standardizing and sharing
+knowledge on drivers of change in support of science and management.}
+*eDrivers* was built on a
 series of guiding principles upholding existing data management and open
@@
 
-As elsewhere in the world [see @halpern2015a], this intensifying human footprint
+As elsewhere in the world [see @halpern2015a; @halpern2019], this intensifying human footprint
 will likely result in increasingly intricate environmental exposure regimes,
@@
 standardized data management practices and tools tailored to the study of the
-effects of multiple drivers [@dafforn2016; @stock2018]. An additional objective
-thus emerged in the process of addressing our initial goals: \textcolor{red}{\sout{sharing the knowledge gathered through the description of drivers in the St.
-Lawrence.}} \textcolor{blue}{sharing information about the distribution and intensity of drivers of change in the St.Lawrence.}
+effects of multiple drivers [@dafforn2016; @stock2018].
+~~An additional objective
+thus emerged in the process of addressing our initial goals:
+sharing information about the
+distribution and intensity of drivers of change in the St.Lawrence.~~
 
-Our main objective in this study is to characterize the distribution and
-intensity of drivers in the St. Lawrence. More specifically, we aim to: 1)
+Our main ~~objective~~ \textcolor{blue}{goal} in this study is to characterize the distribution and
+intensity of drivers in the St. Lawrence. More specifically, ~~we aim~~ \textcolor{blue}{our objectives are} to: 1)
 identify areas of high cumulative exposure\textcolor{red}{\sout{, 2) identify drivers that are likely to interact in the St.
@@
 Lawrence and at facilitating the widespread availability of their characterization for holistic assessments and management approaches.}
-We achieve these objectives with the development of an
-open knowledge platform named *eDrivers*. The platform was designed to
+\textcolor{blue}{An additional objective emerged while addressing the main objective of this
+manuscript: sharing information about the distribution and intensity of drivers of change in the St.Lawrence.}
+We achieve ~~these objectives~~ \textcolor{blue}{this} with the development of an
+open knowledge platform,~~named~~ *eDrivers*,~~. The platform was~~ designed to
 facilitate collaboration, real-time assessments of cumulative exposure and to
-evolve with the addition of information and threats to the St-Lawrence ecosystems.
+evolve ~~with the addition of information and threats to~~
+\textcolor{blue}{as knowledge accumulates and new threats emerge in}
+the St-Lawrence ecosystems.
+\textcolor{blue}{Here, we present its guidng principles and accompanying tools.}
 
@@
 
-## Estuary and Gulf of St. Lawrence
+## ~~Estuary and Gulf of St. Lawrence~~ \textcolor{blue}{St. Lawrence System}
 
-The St. Lawrence System is formed by the Estuary and the Gulf of St. Lawrence
+The St. Lawrence System is formed by the \textcolor{blue}{St. Lawrence} Estuary and the Gulf of St. Lawrence
 (Figure \ref{egsl}). The Estuary is defined by the seawater limit
@@
 \textcolor{red}{\sout{The list of drivers for which we sought data was informed by}} \textcolor{blue}{Drivers selection was informed by} a global
-cumulative impact assessment initiative [@halpern2008a; @halpern2015a] \textcolor{blue}{and available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository (Halpern et al.
+cumulative impact assessment initiative [@halpern2008a; @halpern2015a; @halpern2019] \textcolor{blue}{and available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository (Halpern et al.
 2015)},
@@
 @benoit2012], and communications with regional experts (Table 1).
-\textcolor{red}{\sout{We integrated datasets from regional experts and also use global data from the global cumulative impact assessment initiative [@halpern2008a; @halpern2015a]  available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository [Table 1; @halpern2015].}} We selected global data that were
+\textcolor{red}{\sout{We integrated datasets from regional experts and also use global data from the global cumulative impact assessment initiative [@halpern2008a; @halpern2015a]  available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository [Table 1; @halpern2015].}}
+~~We selected global data that were
 unavailable at the regional scale and that were available at a resolution
-adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).
+adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).~~
+\textcolor{blue}{Where regional data on drivers were unavailable, available global data at a resolution adequate for the scale of the St. Lawrence System were used instead (*e.g.* marine pollution).}
 
 We characterized the intensity and distribution of 22 drivers
-(Table 1;). Drivers incorporated in the analyses are varied in origin,
+(Table 1). Drivers incorporated in the analyses are varied in origin,
 *i.e.* from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping),
@@
 
-Drivers with non-normal frequency distributions were log-transformed,
-\textcolor{blue}{as in Halpern et al.
-2008, to minimize the impact of extreme outliers that may or may not be real observations} (Figure S1)\textcolor{blue}{,} and all drivers were scaled between 0 and 1 to allow driver comparisons.
-The 99th quantile of individual driver distribution was used as the upper bound for scaling \textcolor{red}{\sout{to control for extreme values and produce maps of individual drivers (Figure S2)}}.
-All drivers were embedded in a regular grid composed of 245604 $1 km^2$ hexagonal cells to construct the integrated dataset used for the
-analyses \textcolor{blue}{(Figure S2)}.
+~~Drivers with non-normal frequency distributions were log-transformed, as in
+Halpern et al. 2008, to minimize the impact of extreme outliers that may or may
+not be real observations (Figure S1), and all drivers were scaled between 0 and
+1 to allow driver comparisons. The 99th quantile of individual driver
+distribution was used as the upper bound for scaling.~~
+\textcolor{blue}{As in Halpern et al. 2019, drivers with non-normal frequency distributions were log-transformed
+to avoid underestimating intermediate driver values. All drivers were also scaled
+between 0 and 1 to allow driver comparisons. The 99th quantile of individual
+driver distribution was used as the upper bound for scaling to control for
+extreme values that may or may not be real observations.}
+~~All drivers were embedded in a regular grid composed of 245604 $1 km^2$
+hexagonal cells to construct the integrated dataset used for the analyses
+(Figure S2).~~
+\textcolor{blue}{The St. Lawrence System was divided into a regular grid of
+1 $km^2$ hexagonal cells in which all drivers were integrated (Figure S2)}
 
@@
 Demersal destructive fisheries are located along the Laurentian Channel, the heads of the Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B).
-By combining both drivers, we can observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C, Box 1).
+By combining both drivers, we can observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C~~, Box 1~~).
 
@@
 initiative, our objective is to uphold the highest existing standards of data
-management and open science. We identified four guiding principles to meet this
+management and open science. We identified four guiding principles \textcolor{blue}{(Sections 5.1 to 5.4)} to meet this
 objective and that guide the structure of the platform (Figure \ref{diag}).
@@
 Arguably, the most meaningful benefit anticipated from *eDrivers* will be the gain in efficient access to comparable data-based knowledge on the exposure of ecosystems to multiple threats.
-This could pay quick scientific and management dividends by drawing on the knowledge and efforts of a wide range of contributors, by expanding avenues of scientific inquiry, by decreasing overall effort duplication and research costs, and by increasing research efficiency [@franzoni2014].
+\textcolor{blue}{We provide an example of the ease with which this can be accomplished using the \textit{eDrivers} R package to reproduce figure \ref{kernel} (Box 1).} This could pay quick scientific and management dividends by \textcolor{blue}{efficiently} drawing on the knowledge and efforts of a wide range of contributors, by expanding avenues of scientific inquiry, by decreasing overall effort duplication and research costs, and by increasing research efficiency [@franzoni2014].
 

### 2020-04-18 Code/18-egsl.R

@@ -1,3 +1,4 @@
+# source('./Code/18-egsl.R')
 # ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
 source('./code/2-param.R')
 
@@ -11,21 +12,22 @@ bathy <- st_transform(bathy, crs = 4326)
 deph <- bathy$DEPH
 
 
+library(raster)
+r <- raster(ncol=180, nrow=180)
+extent(r) <- extent(bathy)
+r <- fasterize::fasterize(bathy, r, field = 'DEPH')
+r <- mask(r, egslSimple)
 
-
-
-
-
-# ~~~~~~~~~~~~~~~~~~~ FOOTPRINT ~~~~~~~~~~~~~~~~~~~ #
+# ~~~~~~~~~~~~~~~~~~~ EGSL ~~~~~~~~~~~~~~~~~~~ #
 # Layout
 pal <- colorRampPalette(c('#E2FCF6', '#1B7CAA'))
 cols <- pal(101)[((deph/max(deph, na.rm = T))*100)+1]
 
 cexMain <- 1
 cexMain2 <- .75
-png('./figures/egsl.png', width = 1280, height = 920, res = 300, pointsize = 6)
-# jpeg('./figures/egsl.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
-# par(bg = colBack)
+# png('./figures/egsl.png', width = 1280, height = 920, res = 300, pointsize = 6)
+jpeg('./figures/egsl.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
+par(bg = '#ffffff')
 plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          cols       = c('#00000000',off,off),
          borders    = c('#00000000',focus,focus),
@@ -36,10 +38,13 @@ plotEGSL(layers     = c('egslSimple', 'canada','usa'),
          northArrow = F,
          prj        = slmetaPrj('world'),
          extent     = extFig)
-plot(st_geometry(bathy), add = T, col = cols, border = cols, lwd = .1)
+
+# plot(st_geometry(bathy), add = T, col = cols, border = cols, lwd = .1)
+# plot(st_geometry(bathy), add = T, col = cols, border = 'transparent', lwd = .1)
+image(r, col = pal(100), axes = F, ylab = '', xlab = '', add = T)
[…]

### 2020-04-24 Revisions2/Beauchesne-eDriversMS-SI.rmd
finish answer to reviewer comments

--- 2020-02-26 Revisions/Beauchesne-eDriversMS-SI.rmd
@@
 <!--
-rmarkdown::render('./Revisions/Beauchesne-eDriversMS-SI.rmd')
+rmarkdown::render('./Revisions2/Beauchesne-eDriversMS-SI.rmd')
 -->
@@
 
-Table S1. List of drivers currently available on *eDrivers* along with their respective acronym used in the figures in the supplementary material.
+Table S1. List of drivers currently available on *eDrivers* along with their respective acronym used in the figures of the supplementary material.
 
@@
 
-Oceans are the largest reservoir\textcolor{blue}{s} and sink\textcolor{blue}{s} of atmospheric carbon dioxide ($CO_2$).
-Its uptake increases seawater acidity and \textcolor{red}{\sout{the}} lowers saturation state with
-respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common
-$CaCO_3$ polymorphs constituting the shells and skeleton\textcolor{blue}{s} of many marine
-organisms [@mucci2017]. This can have deleterious effects on carbonate-secreting
-organisms (*e.g.* mollusks and crustaceans) and certain physiological processes
-in non-calcifying organisms [@fabry2008; @kroeker2013].
+Oceans are the largest reservoirs and sinks of atmospheric carbon dioxide ($CO_2$).
+Its uptake increases seawater acidity and lowers the saturation state of waters with respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common $CaCO_3$ polymorphs that constitute the shells and skeletons of many marine organisms [@mucci2017].
+Ocean acidification can have deleterious effects on carbonate-secreting organisms (*e.g.* mollusks and crustaceans) and certain physiological processes in non-calcifying organisms [@fabry2008; @kroeker2013].
 
-When $\Omega_C$ or $\Omega_A$ decrease below 1, water becomes undersaturated and
-corrosive to the skeletal minerals of carbonate-secreting organisms. Note that
-if $\Omega_A$ $< 1$, the waters may still be supersaturated with respect to
-calcite since it is 50% more soluble than aragonite. However, organisms have
-wide ranging responses to $\Omega_C$ and $\Omega_A$ saturation state. For
-example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019],
-while other organisms may have adapted to precipitate $CaCO_3$ even when
-seawater is undersaturated with respect to calcite [$\Omega_C$ $< 1$ or
-$\Omega_A$ $< 0.65$; e.g. @uthicke2016].
+When $\Omega_C$ or $\Omega_A$ decrease below 1, water becomes undersaturated and corrosive to the skeletal minerals of carbonate-secreting organisms.
+Note that if $\Omega_A$ $< 1$, the waters may still be supersaturated with respect to calcite since it is 50% more soluble than aragonite.
+However, organisms have wide ranging responses to changes in $\Omega_C$ and $\Omega_A$.
+For example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019], while other organisms may have adapted to precipitate $CaCO_3$ even when seawater is undersaturated with respect to calcite [$\Omega_C$ $< 1$ or $\Omega_A$ $< 0.65$; e.g.
+@uthicke2016].
 
-We used $\Omega_{A}$ saturation state to characterize ocean acidification in the
-bottom waters of the St. Lawrence. The data come from the Department of
-Fisheries and Oceans' (DFO) Atlantic Zone Monitoring Program [AZMP; @galbraith2018]
-surveys and DFO's multispecies surveys [@bourdages2018] collected in
-August-September of 2017 [@starr2019].
-\textcolor{blue}{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
-stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
-majority of sampling stations were located in the Southern Gulf ($n = 84$),
-compared to the Estuary and Northern Gulf ($n = 33$).}
+We used $\Omega_{A}$ to characterize ocean acidification in the bottom waters of the St.
+Lawrence.
+The data come from the Department of Fisheries and Oceans' (DFO) Atlantic Zone Monitoring Program [AZMP; @galbraith2018] surveys and DFO's multispecies surveys [@bourdages2018] collected in August-September of 2017 [@starr2019].
+$\Omega_{A}$, pH, and dissolved oxygen ($O_2$) were measured at 117 stations in the Estuary and Gulf of St.
+Lawrence in the summer of 2018.
+The majority of sampling stations were located in the Southern Gulf ($n = 84$), the Estuary and Northern Gulf ($n = 33$).
 
-The carbonate chemistry was determined
-through pH and \textcolor{blue}{total alkalinity} (TA) measurements. Samples for pH and \textcolor{red}{\sout{total alkalinity}} TA were
-collected under \textcolor{red}{\sout{a}} bubble free and no head space conditions into 500 $mL$
-borosilicate glass flasks. 250 $\mu L$ of \textcolor{red}{\sout{the}} saturated $HgCl_2$ solution was
-added to sample, and processed following the "Guide to best practices for Ocean
-CO2 Measurements" [@dickson2007].
+The carbonate speciation was determined through pH and total alkalinity (TA) measurements.
+Samples for pH and TA were collected with no head space in 500 $mL$ borosilicate glass flasks, and 250 $\mu L$ of saturated $HgCl_2$ solution was added to each sample and processed following the "Guide to best practices for Ocean CO2 Measurements" [@dickson2007].
 
@@
 [@dickson2007]. A similar procedure was carried out before each set of sample
-measurement\textcolor{red}{\sout{s}} using a TRIS (Tris (hydroxymethyl) -aminomethane) buffer prepared
+measurement using a TRIS (Tris (hydroxymethyl) -aminomethane) buffer prepared
 at a practical salinity (S) of approximately 30 [@millero1986]. Certified
@@
 electrode (pHC2001) in a continuous titrant addition mode, an algorithm
-specifically designed for shallow end-point detection [@dickson2007]. The dilute\textcolor{blue}{d}
-HCl titrant 0.1M in a solution of NaCl of 0.6M was calibrated using CRM provided
-by Professor Andrew G. Dickson.
+specifically designed for shallow end-point detection [@dickson2007]. The
+0.1M HCl titrant in a solution of 0.6M NaCl was calibrated using CRM provided by Professor Andrew G. Dickson.
 
@@
 corrected for *in situ* pressure and temperature using the algorithm CO2SYS
-[@lewis1998] with measured pH, \textcolor{red}{\sout{total alkalinity}} \textcolor{blue}{TA}, soluble reactive
-and silicate concentrations as input parameters.
+[@lewis1998] with measured pH, TA, soluble reactive phosphate and silicate concentrations as input parameters.
 
-\textcolor{red}{\sout{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
-stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
-majority of sampling stations were located in the Southern Gulf ($n = 84$),
-compared to the Estuary and Northern Gulf ($n = 33$).}}
-
-To account for low sample size in the Northern Gulf, we used the correlation
+To account for the low sample size in the Northern Gulf, we used the correlation
 between $O_2$ and $CO_2$, which are linked through the stoichiometry of the
@@
 Metabolic $CO_2$ increases in water through biotic processes such microbial
-respiration of organic matter [@mucci2011]. In deep waters, variations of
-$\Omega_{A}$ are largely governed by the addition of metabolic $CO_2$, whereas
-near the air-sea interface $O_2$ is replenished faster than $CO_2$ can escape
+respiration of organic matter [@mucci2011]. In deep waters, found in the Laurentian, Anticosti and Esquiman Channels, variations of
+$\Omega_{A}$ are largely governed by the addition of metabolic $CO_2$, whereas, near the air-sea interface, $O_2$ is replenished faster than $CO_2$ can escape
 [@zeebe2001]. The impact of respiration on carbonate chemistry is highest in
@@
 correlation observed in the north between $\Omega_A$ and $O_2$ is especially
-strong when deep Channels (*Deep* model: *p-value*: < 0.01; $O_2$: 0.20; $R^2$:
+strong when deep Channel waters (*Deep* model: *p-value*: < 0.01; $O_2$: 0.20; $R^2$:
 0.97) are separated from shallower regions (*Shallow* model: *p-value*: < 0.01;
 $O_2$: 0.07; $R^2$: 0.96), with $\Omega_A$ increasing more steeply with
-increasing $O_2$ in deep Channels. This stronger correlation is likely caused
-by the lack of gas exchange with the atmosphere or the surface mixed layer and
-the importance of metabolic $CO_2$ in deep waters, some of which may have mean
-ages of up to 20 years [@mucci2011]. In the Northern Gulf, the acidification is
-following depth and distance from the Cabot Strait along the Laurentian,
-Anticosti and Esquiman Channels due to the progressive oxygen depletion and
-metabolic carbon dioxide accumulation [@mucci2011].
+increasing $O_2$ in the deep Channels.
+This stronger correlation is likely caused by the lack of gas exchange with the atmosphere or the surface mixed layer and the importance of metabolic $CO_2$ in deep waters, some of which may have mean ages of up to 20 years [@mucci2011].
+In the Northern Gulf, the acidification trends with depth and distance from the Cabot Strait along the Laurentian, Anticosti and Esquiman Channels due to the progressive oxygen depletion and metabolic carbon dioxide accumulation [@mucci2011].
 
-There are likely different dynamics at play in the Southern Gulf, such as
-waters with lowered alkalinity and salinity caused by the St. Lawrence plume.
-Such water masses \textcolor{red}{\sout{are}} have lower buffering capacities to decreasing pH and
-$\Omega_{A}$. Also, the accumulation of metabolic carbon dioxide produced by
-rapid oxic degradation of organic matter in the sandy
-and permeable sediments of the Magdalen Shelf or through the St. Lawrence plume
-could decrease $\Omega_{A}$ in the Southern Gulf (Starr and Chassé, *personal
-communication*).
+There are likely different dynamics at play in the Southern Gulf, such as waters with lowered alkalinity and salinity caused by the surface, seaward-flowing Gaspé Current.
+These waters have lower buffering capacities to decreasing pH and $\Omega_{A}$.
+Furthermore, the accumulation of metabolic carbon dioxide produced by rapid oxic degradation of organic matter in the sandy and permeable sediments of the Magdalen Shelf or in the Gaspé current could decrease $\Omega_{A}$ in the Southern Gulf (Starr and Chassé, *personal communication*).
 
-We thus divided the analyses between the Southern and Northern St. Lawrence.
-The St. Lawrence was divided using the 350m isobath on the southern
-slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé
-Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modeled
+We thus divided the and their analysis between the Southern and Northern St.
+Lawrence.
+The St.
+Lawrence was divided along the 350m isobath on the southern slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé Peninsula.
+The $\Omega_A$ distribution in the Southern St.
+Lawrence was modeled
 using the exponential kriging model.
 
-We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. \textcolor{red}{\sout{Whereas}} \textcolor{blue}{While}
+We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. Whereas
 there are few published datasets on $\Omega_{A}$ in the St.
-Lawrence, there is a robust time-series on oxygen concentration in the water
-column from the Department of Fisheries and Oceans’ (DFO) Atlantic Zone
+Lawrence, there is a robust time-series on oxygen concentrations in the water column available through
+the Department of Fisheries and Oceans’ (DFO) Atlantic Zone
 Monitoring Program [AZMP; @galbraith2018]. We used oxygen concentration data
 collected between 2013 and 2017
-[@blais2019] and interpolated oxygen % saturation using cokriging with depth as a
-covariable, as done in [@dutil2011]. Bathymetric data come from @dutil2011
-and have a 100 $km^2$ resolution. For more details, refer to the hypoxia \textcolor{blue}{data layer}
+[@blais2019] and interpolated oxygen % saturation using cokriging with depth as a covariable, as done in @dutil2011. Bathymetric data come from @dutil2011
+and have a 100 $km^2$ resolution. For more details, refer to the hypoxia data layer
 description. The *Deep* model was used to transform $O_2$ values to $\Omega_A$
@@
 
-Based on the wide range of $\Omega_A$ levels at which organism are negatively
+Based on the wide range of $\Omega_A$ levels at which organisms are negatively
 affected (see above), we built an index of acidification stress ($A_s$)
-ranging between 0 and 1 that slowly \textcolor{red}{\sout{begins increasing}} \textcolor{blue}{increases} as $\Omega_A$ $\approx$
-$2$, increases more rapidly close to $\Omega_A$ $= 1$ and reaches its peak at
+ranging between 0 and 1 that slowly increases as $\Omega_A$ decreases to $\approx$ $2$, increases more rapidly close to $\Omega_A$ $= 1$ and reaches its peak at
 $\Omega_A$ $\approx$ $0.5$:
@@
 
-The data used to characterize hypoxia \textcolor{red}{\sout{uses oxygen saturation in % and}} come from
+The data used to characterize hypoxia come from
 the Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
-[AZMP; @galbraith2018] survey\textcolor{red}{\sout{s}} in late spring and fall, as well as from \textcolor{red}{\sout{annual}}
-DFO’s \textcolor{blue}{annual} multispecies surveys for the northern Gulf in August and for the Magdalen
-Shallows in September. We provide a brief summary of data and methods to
-describe hypoxia in this document. For more details, refer to @blais2019.
+[AZMP; @galbraith2018] survey in late spring and fall of 2013 to 2017, as well as from DFO’s annual multispecies surveys for the northern Gulf in August and for the Magdalen Shallows in September.
+We provide a brief summary of data and methods to describe hypoxia in this document.
+For more details, refer to @blais2019.
 
-Oxygen concentration is measured at every station using an oxygen probe
-(Sea-Bird SBE43) mounted on \textcolor{red}{\sout{the}} \textcolor{blue}{a} CTD. The probe is calibrated against seawater
-samples that are analyzed by Winkler titration on every cast
-[see @blais2019 for calibration procedure]. The data used is the last depth
-sampled on the CTD profile, which \textcolor{blue}{is} typically \textcolor{red}{\sout{corresponds to}} ~10m above the
-bottom. Oxygen concentration ($\mu M$ $L^{-1}$) is converted to oxygen
-saturation\textcolor{blue}{,} taking into account the salinity and temperature at the selected
+Oxygen concentration were measured at every station using an oxygen probe
+(Sea-Bird SBE43) mounted on a CTD. The probe is calibrated against seawater
+samples taken directly from the Niskin bottles at every cast and analyzed by Winkler titration [see @blais2019 for calibration procedure].
+The data used is the last depth sampled on the CTD profile, which is typically ~10m above the
+bottom.
+Note that this depth does not necessarily correspond to the depth of the oxygen minimum.
+The oxygen minimum is typically found along the 27.25 isopycnal, often situated as much as 100 m above bottom in the Lower Estuary.
+Oxygen concentration ($\mu M$ $L^{-1}$) is converted to oxygen
+saturation, taking into account the salinity and temperature at the selected
 depth.
@@
 We interpolated oxygen saturation using cokriging with depth as a covariable,
-as done in [@dutil2011]. Bathymetric data come from @dutil2011 and have a
-100 $km^2$ resolution. According to @diaz1995, severe hypoxia can be observed
-when dissolved oxygen falls below 2 $ml$ $L^{-1}$, which corresponds to 62.5
-$\mu mol$ $L^{-1}$ and ~20% saturation. This is considered as the level
-necessary to maintain most animal life. Similarly, @chabot2008 studied the
+as done in @dutil2011. Bathymetric data come from @dutil2011 and have a
+100 $km^2$ resolution. According to @diaz1995, severe hypoxia occurs
+when dissolved oxygen falls below 2 $ml$ $L^{-1}$, to 62.5 $\mu mol$ $L^{-1}$ or ~20% saturation. This is considered as the level
+necessary to maintain most animal life. Accordingly, @chabot2008 studied the
 effects of hypoxia on the energy budget of cod in the St. Lawrence and found
-that behavioural effects began manifesting below 70% oxygen saturation and that
+that behavioural effects began manifesting themselves below 70% oxygen saturation and that
 survival becomes jeopardized below ~20% saturation. We used these observations
 to create an index of hypoxia using an inverted logistic curve that slowly
-increases below 70% oxygen saturation, increases acutely as it approaches the
-20% oxygen saturation threshold and reaches its peak intensity between 20 and
-30% oxygen saturation:
+increases below 70% oxygen saturation and increases acutely as it reaches its peak intensity between 20 and 30% oxygen saturation:
 
@@
 
-### Bottom temperature anomalies
+### Bottom-water temperature anomalies
 
-The data used to characterize bottom temperature anomalies come from the
+The data used to characterize bottom-water temperature anomalies come from the
 Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
 [AZMP; @galbraith2018]. We provide a brief summary of data and methods to
-characterize the bottom temperature climatology and anomalies in this document.
+characterize the bottom-water temperature climatology and anomalies in this document.
 For more details, refer to @galbraith2018.
 
-Bottom temperatures are interpolated in the Gulf using
+Bottom-water temperatures are interpolated in the Gulf using
 conductivity-temperature-depth (CTD) sampling performed annually through DFO's
@@
 Shallows in September. Using this sampling survey, temperatures are horizontally
-interpolated at each 1 m depth layer on a 2 km resolution grid. Bottom
+interpolated at each 1 m depth layer on a 2 km resolution grid. Bottom-water
 temperatures are then extracted by using a bathymetry layer from the Canadian
@@
 We used temperature anomalies, *i.e.* deviations from long-term normal
-conditions, to measure an annual index of stress associated with extreme
+conditions between 1981 and 2010, to measure an annual index of stress associated with extreme
 temperatures between 2013 and 2017. Temperature anomalies were calculated using
@@
 Anomaly time series were normalized by their standard deviation (SD) to allow
-comparison\textcolor{blue}{s} between areas of the St. Lawrence with different temperature ranges.
+comparisons between areas of the St. Lawrence with different temperature ranges.
 For example, temperatures observed in deep channels are less variable than in
@@
 outliers by @tukey1977. Outlier values were capped to correspond to the 5th and
-95th percentile\textcolor{red}{\sout{s}} \textcolor{blue}{values}. Anomalies were divided into positive and negative anomalies
-and the absolute value of anomalies were used as an indicator of the intensity
-of bottom temperature anomalies. The mean anomaly intensity between 2013 and
-2017 for each grid cell was used to generate the final index of bottom
-temperature anomalies.
+95th percentile values.
+Anomalies were divided into positive and negative anomalies and the absolute value of anomalies were used as an indicator of the intensity of bottom-water temperature anomalies.
+The mean anomaly intensity between 2013 and 2017 for each grid cell was used to generate the final index of bottom-water temperature anomalies.
 
-### Surface temperature anomalies
+### Surface-water temperature anomalies
 
-The data used to characterize surface temperature anomalies come from the
+The data used to characterize surface-water temperature anomalies come from the
 Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
 [AZMP; @galbraith2018]. We provide a brief summary of data and methods to
-characterize the surface temperature climatology and anomalies in this document.
+characterize the surface-water temperature climatology and anomalies in this document.
 For more details, refer to @galbraith2018.
 
-The surface layer is characterized using sea surface temperature (SST) monthly
+The surface-water layer is characterized using sea surface-water temperature (SST) monthly
 composites from Advanced Very High Resolution Radiometer (AVHRR) satellite
@@
 
-Surface temperature anomalies were characterized following the same method used
-for bottom temperature anomalies. Only the months of May to November were
+Surface-water temperature anomalies were characterized following the same method used
+for bottom-water temperature anomalies. Only the months of May to November were
 included to avoid biases associated with the presence of ice cover. Monthly
 anomalies from May to November values were summed to obtain an indicator of
-annual surface temperature anomaly intensity in each grid cell. The mean
+annual surface-water temperature anomaly intensity in each grid cell. The mean
 anomaly intensity between 2013 and 2017 for each grid cell was used to generate
-the final index of surface temperature anomalies.
+the final index of surface-water temperature anomalies.
 
-### Sea level rise
+### Sea-level rise
 
-The data used to characterize sea level rise risk come from the global
+The data used to characterize sea-level rise risk come from the global
 cumulative impacts assessment on habitats [@halpern2008a; @halpern2015a] and
@@
 
-Sea level rise was characterized by @nicholls2010 using NASA's satellite
+Sea-level rise was characterized by @nicholls2010 using NASA's satellite
 altimetry data (Topex/Poseidon, Jason-1&2, GFO, ERS-1&2, and Envisat
@@
 
-The rate of sea level rise ($mm/year$) was measured between 1992 and 2012 and
+The rate of sea-level rise ($mm/year$) was measured between 1992 and 2012 and
 transformed as a net change value ($mm$) by multiplying by the number of years
 considered. Only positive values were selected under the assumption that
-only positive sea level rise is likely to cause environmental stress.
+only positive sea-level rise is likely to cause environmental stress.
 
@@
 
-\textcolor{red}{\sout{Aquaculture data comes from a variety of sources in the St. Lawrence because
-aquaculture sites are mostly managed at the provincial level.
-We therefore had to gather the data on aquaculture sites from the 5 provinces dividing the
-St. Lawrence}}
-\textcolor{blue}{As aquaculture sites are managed at the provincial level in the St. Lawrence, data on aquaculture sites come from various sources}
+As aquaculture sites are managed at the provincial level in the St. Lawrence, data on aquaculture sites come from various sources
 [@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016].
@@
 Gulf. Fish and algae aquaculture, on the other hand, remains marginal.
-Considering this, we only \textcolor{red}{\sout{considered}} \textcolor{blue}{included} invertebrates aquaculture for the
+Considering this, we only included invertebrates aquaculture for the
 aquaculture driver layer. However, if fish or algae farming were to become more
-important, these driver\textcolor{blue}{s} should be incorporated in future analyses as individual
+important, these drivers should be incorporated in future analyses as individual
 layers, as impacts vary between types of aquaculture.
@@
 rarely extend beyond the location of the farms. We therefore only considered
-the \textcolor{red}{\sout{actual}} location of sites to characterize the distribution of this driver.
+the location of sites to characterize the distribution of this driver.
 We were unable to characterize site production in terms of biomass farmed, which
 could provide an indication of the intensity of aquaculture activities. As such,
-we considered aquaculture as binary presence-absence data \textcolor{red}{\sout{in our analyses}}.
+we considered aquaculture as binary presence-absence data.
 
@@
 
-\textcolor{red}{\sout{We used lights at night as a proxy of coastal infrastructure development, as
-terrestrial stable lights at night represent light from human settlements and
-industrial sites with electricity.}}
-\textcolor{blue}{Terrestrial stable lights at night mostly represent light from human settlements and industrial sites with electricity.
-We thus used lights at night as a proxy of coastal infrastructure development.}
+Terrestrial stable lights at night mostly represent light from human settlements and industrial sites with electricity. We thus used lights at night as a proxy of coastal infrastructure development.
 
@@
 around the St. Lawrence [@statistics-canada2017]. Dissemination areas are the smallest
-standard geographic area in which census data are disseminated and they combine
-to cover all of Canada. The census provides population count within the boundary
+standard geographic area in which census data are disseminated. They combine
+to cover all of Canada and are highly variable in shapes and sizes.
+\textcolor{blue}{For example, their sizes range from 0.0003 to over 44000 $km^2$ in our study area.}
+The census provides population count within the boundary
 of each dissemination area, which we used to evaluate total coastal population.
@@
 point-sources or in areas lacking paved roads is therefore not captured by
-this layer. The data obtained was aggregated at the watershed scale. \textcolor{red}{\sout{and spread}}
-\textcolor{blue}{Spread} into coastal and marine environments was modeled using a diffusive plume model
+this layer. The data obtained was aggregated at the watershed scale. Spread into coastal and marine environments was modeled using a diffusive plume model
 from each watershed pourpoints (*e.g.* river mouths).
@@
 for all fisheries in the St. Lawrence, they still provide a very thorough
-overview of the spatial distribution and intensity of fishing activities \textcolor{red}{\sout{in the
-St. Lawrence}}. The data we used span 6 years from 2010 to 2015 and detail
+overview of the spatial distribution and intensity of fishing activities. The data we used span 6 years from 2010 to 2015 and detail
 218323 fishing events (36387 $\pm$ 3147 fishing events per year). There were 31
-targeted species and a total of 53 \textcolor{red}{\sout{caught}} species \textcolor{blue}{caught} in the dataset.
+targeted species and a total of 53 species caught in the dataset.
 
@@
 high-bycatch (DNH), pelagic, low-bycatch (PLB) and pelagic, high-bycatch (PHB).
-This categorization \textcolor{red}{\sout{therefore}} divides the fisheries data into 5 distinct
+This categorization divides the fisheries data into 5 distinct
 driver layers characterizing fishing activities.
 
-Gear types can \textcolor{red}{\sout{also}} be further classified into fixed or mobile engines based
+Gear types can be further classified into fixed or mobile engines based
 on their mobility. We used these two mobility classes to generate a buffer of
 impact around each fishing activity coordinates to consider potential spatial
-uncertainty associated with locations, \textcolor{red}{\sout{and}} the fact that mobile engines can be
-tracted over several kilometers during fishing activities, and \textcolor{red}{\sout{that}} \textcolor{blue}{because} we do not
+uncertainty associated with locations, the fact that mobile engines can be
+tracted over several kilometers during fishing activities, and because we do not
 have the beginning and end points of mobile fishing events. Buffer sizes for
@@
 
-\textcolor{red}{\sout{In order to}} \textcolor{blue}{To} characterize the intensity of fishing activities ($FI$), we used
+To characterize the intensity of fishing activities ($FI$), we used
 a biomass yield density index. We multiplied the total annual biomass captured
@@
 Positioning System technology to locate vessels every 10 minutes. The data used
-\textcolor{red}{\sout{was}} \textcolor{blue}{were} from November 2010 to December 2011.
+were from November 2010 to December 2011.
 Data used come mostly from vessels that move globally (*i.e.* cargo, tanker and
@@
 - Organic pollution
-- Sea bottom temperature anomalies
+- Bottom-water temperature anomalies
 - Shipping
@@
 
-# Threat complexes
+# ~~Threat complexes~~ \textcolor{blue}{Cumulative exposure profiles}
 

### 2020-04-24 Revisions2/Beauchesne-eDriversMS.Rmd
finish answer to reviewer comments

@@
 ---
-title: Next Generation Planning -- Characterizing Exposure to and Shared Knowledge on Drivers of Ecological Change in the St. Lawrence System
+title: Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St. Lawrence System
 author: David Beauchesne^1,2,\*^, Rémi M.
@@
 rmarkdown::render('./Revisions2/Beauchesne-eDriversMS.Rmd', 'pdf_document')
--->
-
-<!--
-To discuss with co-authors
-
-- New article title
-- Should we present how the platform is used in the analyses or really keep
-  that portion fully separate? If we include it, modify answer to question 2
-  comment 1
-
 -->
@@
 The St. Lawrence is a vast and complex socio-ecological system providing a
-wealth of services sustaining numerous economic sectors. ~~\textcolor{red}{These ecosystems are}~~
-\textcolor{blue}{This ecosystem is}
-subject to significant human pressures that overlap and potentially interact
-with climate driven environmental changes. Our objective in this paper is to
-systematically characterize the distribution and intensity of drivers \textcolor{blue}{of change} in the
-St. Lawrence System. ~~To do so, we launch *eDrivers*, an open knowledge
-platform gathering experts committed to structuring, standardizing and sharing
-knowledge on drivers \textcolor{blue}{of change} in support of science and management.~~ We gathered data on
-22 coastal, climate, fisheries and marine traffic drivers through
-collaborations, existing environmental initiatives and open data portals. We
-show that few areas of the St. Lawrence are free of cumulative exposure. The
-Estuary, ~~the~~ Anticosti Gyre and coastal areas are particularly exposed,
-especially in the vicinity of urban centers. We identified 6 areas of distinct
-cumulative exposure regime that show that certain drivers \textcolor{blue}{of change} typically co-occur
-in different regions of the St. Lawrence and that coastal areas are exposed to
-all driver types. Of particular concern are two threat complexes capturing
-most exposure hotspots that show the convergence of contrasting exposure
-regimes at the head of the Laurentian Channel.
-\textcolor{blue}{Sharing the knowledge acquired emerged as a priority to minimize future
-environmental assessment efforts. We thus launch *eDrivers*, an open knowledge
-platform gathering experts committed to structuring, standardizing and sharing
-knowledge on drivers of change in support of science and management.}
-*eDrivers* was built on a
-series of guiding principles upholding existing data management and open
-science standards. We therefore expect it to evolve through time to address
-knowledge gaps and refine current driver layers. Ultimately, we believe that
-*eDrivers* represents a much needed solution that could radically influence
-broad scale research and management practices by increasing knowledge accessibility
-and interoperability.
+wealth of services sustaining numerous economic sectors.
+~~These ecosystems are~~ \textcolor{blue}{This ecosystem is} subject to significant human pressures that overlap and potentially interact with climate driven environmental changes.
+Our objective in this paper is to systematically characterize the distribution and intensity of drivers of \textcolor{blue}{environmental} change \textcolor{blue}{(hereafter drivers)} in the St.
+Lawrence System.
+~~To do so, we launch *eDrivers*, an open knowledge platform gathering experts committed to structuring, standardizing and sharing knowledge on drivers of change in support of science and management.~~ We gathered data\textcolor{blue}{-based indicators for} ~~on~~ 22 coastal, climate, fisheries and marine traffic drivers through collaborations, existing environmental initiatives and open data portals.
+We show that few areas of the St.
+Lawrence are free of cumulative exposure.
+The Estuary, ~~the~~ Anticosti Gyre and coastal areas are particularly exposed, especially in the vicinity of urban centers.
+We identified 6 ~~areas of~~ distinct ~~cumulative exposure regimes~~ \textcolor{blue}{clusters with similar suites of co-occurring drivers} and show that certain driver \textcolor{blue}{combinations are typical of} ~~typically co-occur in~~
+different regions of the St. Lawrence and that coastal areas are exposed to
+all driver types. Of particular concern are two ~~threat complexes~~\textcolor{blue}{clusters} capturing
+most exposure hotspots \textcolor{blue}{and} that show the convergence of contrasting ~~exposure
+regimes~~\textcolor{blue}{cumulative exposure profiles} at the head of the Laurentian Channel.
+\textcolor{blue}{Sharing the knowledge of drivers acquired emerged as a priority to minimize future environmental assessment efforts.
+We thus launch \textit{eDrivers}, an open knowledge platform gathering experts committed to structuring, standardizing and sharing knowledge on drivers of environmental change in support of holistic science and management.} *eDrivers* was built on a series of guiding principles upholding existing data management and open science standards.
+We therefore expect it to evolve through time to address knowledge gaps and refine current driver layers.
+Ultimately, we believe that *eDrivers* represents a much needed solution that could radically influence broad scale research and management practices by increasing knowledge accessibility and interoperability.
 
-**Keywords**: ocean observing systems, St. Lawrence, environmental drivers,
-cumulative exposure, threat complex, multiple stressors, global change
+**Keywords**: ocean observing systems, St. Lawrence,
+cumulative exposure, multiple stressors, global change
 
@@
 
-The St. Lawrence System, formed by one of the largest estuaries in the world\textcolor{red}{\sout{, the St. Lawrence Estuary,}}
-and a vast interior sea\textcolor{red}{\sout{, the Gulf of St. Lawrence}}, is
-a complex social-ecological system characterized by highly variable
-environmental conditions and oceanographic processes
+The St. Lawrence System, formed by one of the largest estuaries in the world
+and a vast interior sea, is a complex social-ecological system characterized by highly variable environmental conditions and oceanographic processes
 [@el-sabh1990a; @white1997; @dufour2007]. It constitutes a unique
@@
 productive ecological communities [@savenkoff2000]. As a result, the St.
-Lawrence System provides a wealth of ecosystem services that have \textcolor{red}{\sout{historically and contemporarily}}
 benefited the Canadian economy. It sustains a rich fisheries
@@
 
-As elsewhere in the world [see @halpern2015a; @halpern2019], this intensifying human footprint
-will likely result in increasingly intricate environmental exposure regimes,
-*i.e.* suites of overlapping \textcolor{red}{\sout{and potentially interacting}} environmental drivers \textcolor{blue}{of change}
-threatening ecosystems, habitats or ecological communities.
-Drivers \textcolor{blue}{of change}, often referred to as stressors or pressures \textcolor{blue}{and hereafter called drivers for simplicity}, are any externalities that affect
-environmental processes and disturb natural systems. Drivers may originate from
-natural or human-induced biophysical processes (*e.g.* sea surface\textcolor{blue}{-water} temperature
-anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
-fisheries and marine pollution). The potential for complex interactions between
-\textcolor{blue}{co-occurring} drivers is the largest uncertainty when studying or predicting environmental impacts [@darling2008; @cote2016].
-Multiple drivers can combine non-linearly and result in effects that are greater (synergistic effect) or lower (antagonistic effect) than the sum of individual effects
-[@crain2008; @darling2008; @cote2016].
+~~As elsewhere in the world [see @halpern2015a; @halpern2019], this intensifying human footprint will likely result in increasingly intricate environmental exposure regimes, *i.e.* suites of overlapping drivers of environmental change threatening ecosystems, habitats and ecological communities.~~
+\textcolor{blue}{As a result, the St.
+Lawrence System is exposed to an increasing number of drivers of environmental change, as is observed across ecosystems globally} [see @halpern2015a; @halpern2019].
+~~Drivers of environmental change, often referred to as stressors or pressures and hereafter called drivers for simplicity, are any externalities that affect
+environmental processes and disturb natural systems.~~
+\textcolor{blue}{We broadly define drivers of environmental change (hereafter drivers) as any externality that affects environmental processes and disturbs natural systems.}
+Drivers may originate from natural or human-induced biophysical processes (*e.g.* sea surface-water temperature anomalies and hypoxia) or directly from anthropogenic activities (*e.g.* fisheries and marine pollution).
+The potential for complex interactions between co-occurring drivers is the largest uncertainty when studying or predicting environmental impacts [@darling2008; @cote2016].
+Multiple drivers can combine non-linearly and result in effects that are greater (synergistic effect) or lower (antagonistic effect) than the sum of individual effects [@crain2008; @darling2008; @cote2016].
 
-The uncertainty associated with complex driver interactions must therefore be taken into account when investigating environmental impacts [@cote2016], yet most research on driver effects in marine environments remains overwhelmingly focused on single driver assessments [@obrien2019].
-Increasing exposure and the experiences of past ecological tragedies such as the collapse of cod fisheries [@frank2005; @dempsey2018] and the decline of the beluga and right whale populations [@plourde2014] together urge the need to characterize the distribution, intensity and \textcolor{red}{\sout{overlap}} \textcolor{blue}{co-occurrence} between drivers in the St.
+~~The uncertainty associated with complex driver interactions must therefore be taken into account when investigating environmental impacts [@cote2016], yet research on the effects of drivers in marine environments remains overwhelmingly focused on single driver assessments} [@obrien2019].~~  Increasing exposure and the experiences of past ecological tragedies such as the collapse of cod fisheries [@frank2005; @dempsey2018] and the decline of the beluga and right whale populations [@plourde2014] together urge the need to characterize the distribution, intensity and co-occurrence of drivers in the St.
 Lawrence System.
-\textcolor{blue}{The impacts of drivers, whether individually or in combination, depend on species-specific vulnerabilities.
-Driver co-occurrence alone is therefore not sufficient evidence to claim interactions among drivers.
-However, as drivers must co-occur to interact, the likelihood of interactions will increase with the number of drivers co-occurring.} \textcolor{red}{\sout{This will}} \textcolor{blue}{Describing drivers will therefore} provide critical information on areas most exposed to \textcolor{red}{\sout{cumulative}} drivers \textcolor{red}{\sout{and on the interaction potential of drivers}} in the St.
-Lawrence.
-It is also a necessary step towards the holistic and integrated management of the St.
-Lawrence System.
+\textcolor{blue}{Research on the effects of drivers in marine environments nonetheless remains overwhelmingly focused on single driver assessments} [@obrien2019].
+\textcolor{blue}{While co-occurring drivers may not interact, driver co-occurrence is a requirement for interactions to exist.
+Knowledge of their co-distribution can therefore identify areas where driver interactions are most likely observed.}
+~~Describing drivers will therefore provide critical information on areas most exposed to drivers in the St. Lawrence. It~~
 
-\textcolor{blue}{Holistic approaches typically involve, but are not limited to, the following steps: selecting and describing valued ecosystem components (\textit{e.g.} habitats and species) and drivers (\textit{e.g.} marine traffic and ocean acidification), assessing the exposure and vulnerability of valued components to drivers, selecting a proper spatio-temporal scale, monitoring, and public and stakeholder participation} [@dube2001].
-Gathering environmental
-\textcolor{red}{\sout{data for large scale, systematic}}
-\textcolor{blue}{knowledge for holistic}
-initiatives can, however, be a very challenging and time consuming – not to say painful – process.
+\textcolor{blue}{Characterizing drivers is also a necessary step for the application of holistic management approaches.} Holistic approaches typically involve, but are not limited to, selecting and describing valued ecosystem components (\textit{e.g.} habitats and species) and drivers (\textit{e.g.} marine traffic and ocean acidification), assessing the exposure and vulnerability of valued components to drivers, selecting a proper spatio-temporal scale, monitoring, and public and stakeholder participation [@dube2001].
+Gathering environmental knowledge for holistic initiatives can, however, be a very challenging and time consuming –- not to say painful –- process.
 On one hand, there is an overwhelming and expanding wealth of data available.
 Such information overload may inhibit our ability to make decisions based on scientific information, promote massive effort duplication, disproportionately appropriate research funds to certain sectors, and obscure knowledge gaps amid a sea of information [@eppler2004].
-On the other hand, crucial data are lacking and remain largely unavailable or inaccessible for a variety of reasons, including proprietary rights, lack of organizational time, capacity and training, and in some \textcolor{red}{\sout{rare}} cases unwillingness to share, curtailing our ability for appropriate decision-making.
+On the other hand, crucial data are lacking and remain largely unavailable or inaccessible for a variety of reasons, including proprietary rights, lack of organizational time, capacity and training, and in some cases unwillingness to share; this curtails our ability for appropriate decision-making.
 
-There are now initiatives that \textcolor{blue}{facilitate the data gathering process} \textcolor{red}{\sout{address this issue}} by assembling, organizing and sharing environmental knowledge, such as the Ocean Biogeographic Information
-System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for
-abiotic data.
-However, equivalent platforms for environmental drivers have largely focused on
-single drivers (e.g. Global Fishing Watch) and platforms collating
-data and knowledge on multiple drivers in a comparable and interoperable way
-remain conspicuously missing [but see @halpern2015]. This is in spite of
-integrated management and assessment approaches needing efficient data reporting,
-standardized data management practices and tools tailored to the study of the
-effects of multiple drivers [@dafforn2016; @stock2018].
-~~An additional objective
-thus emerged in the process of addressing our initial goals:
-sharing information about the
-distribution and intensity of drivers of change in the St.Lawrence.~~
+There are now initiatives that facilitate the data gathering process by assembling, organizing and sharing environmental knowledge, such as the Ocean Biogeographic Information System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for abiotic data.
+However, equivalent platforms for ~~environmental~~ drivers have largely focused on single drivers (e.g.
+Global Fishing Watch) and platforms collating data\textcolor{blue}{-based indicators} and knowledge on multiple drivers in a comparable and interoperable way remain conspicuously missing [but see @halpern2015].
+This is in spite of integrated management and assessment approaches needing efficient data reporting, standardized data management practices and tools tailored to the study of the effects of multiple drivers [@dafforn2016; @stock2018].
+~~An additional objective thus emerged in the process of addressing our initial goals: sharing information about the distribution and intensity of drivers of change in the St.Lawrence.~~
 
@@
 intensity of drivers in the St. Lawrence. More specifically, ~~we aim~~ \textcolor{blue}{our objectives are} to: 1)
-identify areas of high cumulative exposure\textcolor{red}{\sout{, 2) identify drivers that are likely to interact in the St.
-Lawrence}} and \textcolor{red}{\sout{3)}} \textcolor{blue}{2)} characterize areas with similar cumulative exposure regimes.
-\textcolor{red}{\sout{Here we focus solely on drivers, one of four elements required to evaluate environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.* habitats or species) and the exposure and the vulnerability of the ecosystem components to drivers.}}
-\textcolor{blue}{Here, we focus solely on drivers, aiming at better understanding their distribution in the St.
-Lawrence and at facilitating the widespread availability of their characterization for holistic assessments and management approaches.}
-\textcolor{blue}{An additional objective emerged while addressing the main objective of this
-manuscript: sharing information about the distribution and intensity of drivers of change in the St.Lawrence.}
-We achieve ~~these objectives~~ \textcolor{blue}{this} with the development of an
-open knowledge platform,~~named~~ *eDrivers*,~~. The platform was~~ designed to
+identify areas of high cumulative exposure \textcolor{blue}{to drivers} and 2) characterize areas with similar cumulative exposure ~~regimes~~ \textcolor{blue}{profiles, \textit{i.e.} areas exposed to similar suites of co-occurring drivers}.
+\textcolor{blue}{An additional objective emerged while addressing the main goal of this manuscript: sharing information about the distribution and intensity of drivers of environmental change in the St.Lawrence.}
+~~We achieve these objectives with the development of an
+open knowledge platform, named *eDrivers*. The platform was designed to
 facilitate collaboration, real-time assessments of cumulative exposure and to
-evolve ~~with the addition of information and threats to~~
-\textcolor{blue}{as knowledge accumulates and new threats emerge in}
-the St-Lawrence ecosystems.
-\textcolor{blue}{Here, we present its guidng principles and accompanying tools.}
+evolve with the addition of information and threats to the St-Lawrence ecosystems.~~
+\textcolor{blue}{We achieve this with the development of an open knowledge platform, \textit{eDrivers}, designed to facilitate the widespread availability of driver characterization for holistic assessments and management approaches.
+Here, we present its guiding principles and accompanying tools.}
 
@@
 The St. Lawrence drains over 25% of global freshwater reserves through its
-connexion to the Great Lakes Basin\textcolor{red}{\sout{. The Basin}}\textcolor{blue}{, which }is home to over 45 million
+connexion to the Great Lakes Basin, which is home to over 45 million
 North-Americans, *i.e.* 15 and 30 million in Canada and the United States,
@@
 
-\textcolor{red}{\sout{The list of drivers for which we sought data was informed by}} \textcolor{blue}{Drivers selection was informed by} a global
-cumulative impact assessment initiative [@halpern2008a; @halpern2015a; @halpern2019] \textcolor{blue}{and available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository (Halpern et al.
-2015)},
-regional holistic evaluations of the state of the St. Lawrence [@dufour2007;
+\textcolor{blue}{Drivers, as broadly defined in this study, are data-based indicators of environmental conditions and human activities that are often referred to as driving forces, stressors, pressures or states in the scientific and environmental assessment litterature} [*e.g.* @kristensen2004; @halpern2019].
+\textcolor{blue}{Defining such categories, however, can be difficult and is often context- and ecosystem-specific} [@gari2015; @dempsey2018].
+\textcolor{blue}{As such, we refrain from articulating our work around a specific framework or imposing categories on data-based products that may change with a user's objective.
+We rather focus on available data-based indicators that contribute to evaluating an ecosystem's cumulative exposure to multiple threats.}
+
+Drivers selection was informed by a global cumulative impact assessment initiative [@halpern2008a; @halpern2015a; @halpern2019] and available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository [@halpern2015], regional holistic evaluations of the state of the St.
+Lawrence [@dufour2007;
 @benoit2012], and communications with regional experts (Table 1).
-\textcolor{red}{\sout{We integrated datasets from regional experts and also use global data from the global cumulative impact assessment initiative [@halpern2008a; @halpern2015a]  available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository [Table 1; @halpern2015].}}
-~~We selected global data that were
-unavailable at the regional scale and that were available at a resolution
-adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).~~
-\textcolor{blue}{Where regional data on drivers were unavailable, available global data at a resolution adequate for the scale of the St. Lawrence System were used instead (*e.g.* marine pollution).}
+~~We selected global data that were unavailable at the regional scale and that were available at a resolution adequate for use at the scale of the St.
+Lawrence (*e.g.* marine pollution).~~ \textcolor{blue}{Where regional data on drivers were unavailable, available global data at a resolution adequate for the scale of the St.
+Lawrence System were used instead (\textit{e.g.} marine pollution).}
 
@@
 distribution was used as the upper bound for scaling.~~
-\textcolor{blue}{As in Halpern et al. 2019, drivers with non-normal frequency distributions were log-transformed
-to avoid underestimating intermediate driver values. All drivers were also scaled
-between 0 and 1 to allow driver comparisons. The 99th quantile of individual
-driver distribution was used as the upper bound for scaling to control for
-extreme values that may or may not be real observations.}
+\textcolor{blue}{As in Halpern et al.
+2019, drivers with non-normal frequency distributions were log-transformed to avoid underestimating intermediate driver values.
+All drivers were also scaled between 0 and 1 to allow driver comparisons.
+The 99th quantile of individual driver distribution was used as the upper bound for scaling to control for extreme values that may or may not be real observations.}
 ~~All drivers were embedded in a regular grid composed of 245604 $1 km^2$
@@
 \textcolor{blue}{The St. Lawrence System was divided into a regular grid of
-1 $km^2$ hexagonal cells in which all drivers were integrated (Figure S2)}
+1 $km^2$ cells in which all drivers were integrated (Figure S2)}
 
-## \textcolor{red}{\sout{Driver interactions}} \textcolor{blue}{Driver co-occurrence}
+## ~~Driver co-occurrence~~ {-}
 
-\textcolor{red}{\sout{Interactions among drivers were investigated using the cumulative footprint ($F$) between pairs of drivers (objective 2).
-The intensity at which pairs of drivers co-occur was evaluated using a two-dimensional kernel density estimate.
-As there are 231 pairwise combinations between 22 drivers, we focus on a single example using hypoxia and demersal destructive fisheries, two drivers known to occur mainly in deeper areas of the St.
-Lawrence and, hence, an interaction between the effects of the two drivers could be anticipated.}}
-\textcolor{blue}{As an introduction to indicators of cumulative exposure (objective 1), we begin by presenting a simplifed 2-driver example focusing on hypoxia and demersal destructive fisheries, two drivers known to occur mainly in deeper waters of the St.
-Lawrence.
-Driver co-ccurrence was evaluated spatially by summing the scaled intensity of drivers in each grid cell.
-The intensity at which pairs of drivers co-occur was evaluated using a two-dimensional kernel density.}
+~~As an introduction to indicators of cumulative exposure
+(objective 1), we begin by presenting a simplifed 2-driver example focusing
+on hypoxia and demersal destructive fisheries, two drivers known to occur
+mainly in deeper waters of the St. Lawrence. Driver co-ccurrence was evaluated
+spatially by summing the scaled intensity of drivers in each grid cell. The
+intensity at which pairs of drivers co-occur was evaluated using a two-dimensional kernel density.~~
 
@@
 
-Areas with high cumulative exposure (objective 1) were identified by comparing
+~~Areas with high cumulative exposure were identified by comparing
 areas on the basis of the number and relative intensity of drivers in each
-grid cell. Throughout the text, we use exposure when describing areas
-exposed to drivers and we use footprint when describing the distribution and
-intensity of drivers.
+grid cell (objective 1).~~
+\textcolor{blue}{We begin by providing a simplifed 2-driver example focusing on
+the co-occurrence of hypoxia and demersal destructive fisheries, two drivers
+occurring mainly in deeper waters of the St. Lawrence. Driver co-ccurrence was
+evaluated spatially by summing the scaled intensity of drivers in each grid cell.
+The intensity at which pairs of drivers co-occur was evaluated using a
+two-dimensional kernel density. This example demonstrates how driver
+co-occurrence was evaluated and serves as a stepping stone to the
+integrative indicators used after, \textit{i.e.} cumulative exposure and
+cumulative hotspots (objective 1).}
 
-Cumulative footprint ($F$) was defined as the sum of the scaled intensity of all
-drivers in each grid cell:
+~~Cumulative exposure ($E_C$), was defined as the sum of the scaled intensity of all
+drivers in each grid cell:~~
+\textcolor{blue}{We evaluated the cumulative exposure ($E_C$) of each grid
+cell as the sum of scaled driver intensities:}
 
-$$F_x = \sum_{i=1}^{n} D_{i,x}$$
+$$E_{C_x} = \sum_{i=1}^{n} D_{i,x}$$
 
 where $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
-driver $i$. The cumulative footprint provides an estimate of the total relative
-footprint in each grid cell. A grid cell with a high $F$ value is either
-characterized by multiple drivers at low relative intensity, limited drivers at
+driver $i$. ~~The cumulative footprint provides an estimate of the total relative
+footprint in each grid cell.~~ A grid cell with a high $E_C$ value is either
+characterized by multiple drivers at low relative intensity, ~~limited~~ \textcolor{blue}{a few} drivers at
 high relative intensity, or both.
 
-We also identified cumulative hotspots ($H$) to explore the the distribution of
+~~We also identified cumulative hotspots ($H_C$) to explore the distribution of
 cumulative exposure in the St. Lawrence (objective 1). Cumulative hotspots ($H$)
 were defined as the number of drivers in each grid cell with scaled intensity
-contained over their respective 80th percentile:
+contained over their respective 80th percentile:~~
+
+\textcolor{blue}{We also identified cumulative hotspots ($H_C$) -- \textit{i.e.} areas where drivers co-occur at high
+relative intensities -- as the number of drivers in each grid cell with scaled
+intensity contained over their respective 80th percentile:}
 
 <!-- Indicator function -->
-$$H_x = \sum_{i=1}^{n} \mathbbm{1} (D_{i,x} \; \epsilon \; P_{80, D_i})$$
+$$H_{C_x} = \sum_{i=1}^{n} \mathbbm{1} (D_{i,x} \; \epsilon \; P_{80, D_i})$$
 
 where, $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
-driver $i$ and $P_{80, D_i}$ is the 80th percentile of driver $i$. Hotspots
-thus identify areas where drivers are co-occurring at high relative intensities.
+driver $i$ and $P_{80, D_i}$ is the 80th percentile of driver $i$.
+~~Hotspots thus identify areas where drivers are co-occurring at high relative intensities.~~
 
-## Threat complexes
+## ~~Threat complexes~~\textcolor{blue}{Cumulative exposure profiles}
 
-In order to \textcolor{red}{\sout{better capture potential interactions between drivers (objective 2) and to}}identify areas with similar exposure regimes (objective \textcolor{red}{\sout{3}}\textcolor{blue}{2}), we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
-We use the term clusters in presenting the methods, but use threat complex when discussing the results on cumulative exposure regimes.
+~~In order to identify areas with similar exposure regimes (objective 2), we identify threat complexes using a clustering approach [*e.g.* see @bowler2019].
+We use the term clusters in presenting the methods, but use threat complex when discussing the results on cumulative exposure regimes.~~
 
@@
 
-Threat complexes were identified using a partional *k-medoids* clustering
+\textcolor{blue}{We identified areas with similar cumulative exposure profiles using a clustering approach} [@bowler2019].
+~~Threat complexes were identified using~~ \textcolor{blue}{We used} a partional *k-medoids* clustering
 algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
@@
 
-## \textcolor{red}{\sout{Driver interactions}} \textcolor{blue}{Driver co-occurrence}
+## ~~Driver co-occurrence~~ {-}
 
-Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels, with the head of the Channels most exposed (Figure \ref{kernel}A).
-Demersal destructive fisheries are located along the Laurentian Channel, the heads of the Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B).
-By combining both drivers, we can observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C~~, Box 1~~).
+~~Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels,
+with the head of the Channels most exposed (Figure \ref{kernel}A). Demersal
+destructive fisheries are located along the Laurentian Channel, the heads of the
+Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B).
+By combining both drivers, we can observe that hypoxia and demersal destructive
+fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in
+the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti
+Channels (Figure \ref{kernel}C, Box 1).~~
 
-Fisheries in the St.
-Lawrence have historically affected biodiversity distribution and habitat quality [@moritz2015].
-Concurrently, hypoxia decreases overall habitat quality, but triggers species-dependent responses ranging from adaptation [*e.g.* northern shrimp *Pandalus borealis* and Greenland halibut *Reinhardtius hippoglossoides*; @pillet2016] to reduced growth rates [@dupont-prinet2013] and avoidance of oxygen-depleted habitats [*e.g.* Atlantic cod *Gadus morhua*; @chabot2008] to increased mortality [*e.g.* sessile benthic invertebrates; @eby2005; @gilbert2007; @belley2010].
-Certain species may thus be adversely affected by fisheries and withstand hypoxia but still experience a decrease in prey availability, while others may be deleteriously affected by the compounded effect of both drivers [@deleo2017].
+~~Fisheries in the St. Lawrence have historically affected biodiversity
+distribution and habitat quality [@moritz2015]. Concurrently, hypoxia decreases
+overall habitat quality, but triggers species-dependent responses ranging from
+adaptation [*e.g.* northern shrimp *Pandalus borealis* and Greenland halibut
+*Reinhardtius hippoglossoides*; @pillet2016] to reduced growth rates
+[@dupont-prinet2013] and avoidance of oxygen-depleted habitats [*e.g.* Atlantic
+cod *Gadus morhua*; @chabot2008] to increased mortality [*e.g.* sessile benthic
+invertebrates; @eby2005; @gilbert2007; @belley2010]. Certain species may thus be
+adversely affected by fisheries and withstand hypoxia but still experience a
+decrease in prey availability, while others may be deleteriously affected by
+the compounded effect of both drivers [@deleo2017].~~
 
@@
 
-Apart from the northeastern Gulf, the cumulative footprint of drivers is
+\textcolor{blue}{We first present the simplified hypoxia-fisheries example to demonstrate how driver co-occurrence was evaluated.
+Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels, with the head of the Channels most exposed (Figure \ref{kernel}A).
+Demersal destructive fisheries are located along the Laurentian Channel, the heads of the Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B).
+By combining both drivers, we observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C); these are the areas where we might expect interactions between these drivers to be more likely}.
+
+\textcolor{blue}{We now focus on the integrative exposure indicators.}
+Apart from the northeastern Gulf, ~~the~~ cumulative ~~footprint~~\textcolor{blue}{exposure} ~~of drivers~~ is
 ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
@@
 as most of the St. Lawrence is exposed to multiple overlapping drivers
-(Figures \ref{footprint} and \ref{hotspot}). For example, it is worthy to note
-high cumulative footprint observed at the heads of the Anticosti and Esquiman
-Channels (Figure \ref{footprint}).
+(Figures \ref{footprint} and \ref{hotspot}).
+For example, \textcolor{blue}{the heads of the Anticosti and Esquiman Channels are highly exposed to cumulative drivers} ~~it is worthy to note high cumulative exposure observed at the heads of the Anticosti and Esquiman
+Channels~~ (Figure \ref{footprint}).
 
 These results are consistent with observations elsewhere in the world, where
-cumulative driver exposure conspicuously arises from and markedly intensifies
+cumulative ~~driver~~ exposure conspicuously arises from and markedly intensifies
 close to coastal cities and at the mouth of rivers draining highly populated
@@
 or home to small coastal communities. Certain smaller coastal communities with
-high cumulative footprint are characterized by large industries (*e.g.*
-Sept-Îles and Charlottetown).
+high cumulative ~~footprint~~\textcolor{blue}{exposure} are characterized by large industries (*e.g.* Sept-Îles and Charlottetown).
 
@@
 Basin, which is home to over 45 million North-Americans and is the most densely
-populated region in Canada [@archambault2017; @statistics-canada2017]. Most marine
+populated region in Canada [@statistics-canada2017]. Most marine
 traffic thus converges to the Estuary.
 
-While we cannot ascertain that high exposure areas are the most impacted, we can safely predict that these are the areas where studying ecosystem state will be the most complex due to the uncertainty associated with driver \textcolor{red}{\sout{interactions}} \textcolor{blue}{co-occurrence}, an uncertainty bound to increase rapidly with the number of \textcolor{red}{\sout{interacting}} \textcolor{blue}{co-occurring} drivers [@cote2016].
+While we cannot ascertain that high exposure areas are the most impacted, we
+can safely predict that these are the areas where studying ecosystem state will
+be the most complex due to the uncertainty associated with driver
+co-occurrence, an uncertainty bound to increase rapidly with the number of co-occurring drivers [@cote2016].
 
-## Threat complexes
+## ~~Threat complexes~~ \textcolor{blue}{Cumulative exposure profiles}
 
 While informative, the hypoxia-fisheries example focuses on a single pair of
-drivers and falls short of the number of drivers overlapping at
+drivers and falls short of the number of drivers typically overlapping at
 high intensities throughout the St. Lawrence (Figure \ref{hotspot}).
@@
 most likely, and where they can arise between a host of different drivers.
-The identification of threat complexes provides a crucial tool to simplify the
-multi-dimensional complexity of overlapping drivers to areas exposed to similar
-suites of drivers [@bowler2019]. This may prove critical for a better
-understanding the state of species, habitats and ecosystems located within or
-moving through threat complexes and exposed to the combined effects of all
-drivers typical to those areas.
+~~The identification of threat complexes~~ \textcolor{blue}{Identifying areas with similar cumulative exposure profiles} provides a crucial tool to simplify the multi-dimensional complexity of overlapping drivers ~~to areas exposed to similar suites of drivers~~ [@bowler2019].
+This ~~may prove critical for a better understanding~~ \textcolor{blue}{could facilitate assessments of} the state of species, habitats and ecosystems located within or
+moving through ~~threat complexes and exposed to the combined effects of all
+drivers typical to those areas~~ \textcolor{blue}{areas exposed to similar suites of drivers}.
 
-Six distinct threat complexes were identified in the St. Lawrence using
+Six distinct ~~threat complexes~~\textcolor{blue}{clusters} were identified in the St. Lawrence using
 the *k-medoids* algorithm (Figures S4, S5). Based on their
-distribution and representative drivers, threat complexes can be divided into
-3 offshore and 3 coastal complexes (Figures \ref{cluster}, S6 and S7).
-Coastal threat complexes (1 to 3; Figure \ref{cluster}) include all types
-of drivers besides hypoxia and are the most exposed threat complexes, both in
-terms of driver overlap and intensity. Threat complex 1 encompasses the
+distribution and representative drivers, ~~threat complexes~~\textcolor{blue}{clusters} can be divided into
+3 offshore and 3 coastal ~~complexes~~\textcolor{blue}{clusters} (Figures \ref{cluster}, S6 and S7).
+Coastal ~~threat complexes~~\textcolor{blue}{clusters} (1 to 3; Figure \ref{cluster}) include all types
+of drivers besides hypoxia; they are also the most exposed ~~threat complexes~~\textcolor{blue}{clusters}, both in
+terms of driver overlap and intensity. ~~Threat complex~~\textcolor{blue}{Cluster} 1 encompasses the
 coastline and is characterized by higher direct human impact (*i.e.* population
-density). Threat complex 2 is differentiated from other complexes by the
-presence of aquaculture sites. Threat complex 3 is the most exposed complex and
+density). ~~Threat complex~~\textcolor{blue}{Cluster} 2 is differentiated from other ~~complexes~~\textcolor{blue}{clusters} by the
+presence of aquaculture sites. ~~Threat complex~~\textcolor{blue}{Cluster} 3 is the most exposed ~~complex~~ and
 has a distribution similar to the most exposed coastal hotspots (Figure
-\ref{hotspot}). This complex is characterized by high intensities of land-based
+\ref{hotspot}). This ~~complex~~\textcolor{blue}{cluster} is characterized by high intensities of land-based
 drivers (*e.g.* nutrient input), demersal non-destructive high-bycath fisheries
@@
 
-Offshore threat complexes (4 to 6; Figure \ref{cluster}) are generally
+Offshore ~~threat complexes~~\textcolor{blue}{clusters} (4 to 6; Figure \ref{cluster}) are generally
 characterized by high intensity climate and marine traffic drivers.
-Threat complex 4 is differentiated by demersal non-destructive high-bycath
-fisheries, higher marine traffic drivers compared to complex 5 and generally
-corresponds to the whole Southern Gulf. Threat complex 5 is characterized by
+~~Threat complex~~\textcolor{blue}{Cluster} 4 is differentiated by demersal non-destructive high-bycath
+fisheries, higher marine traffic drivers compared to ~~complex~~\textcolor{blue}{cluster} 5 and generally
+corresponds to the whole Southern Gulf. ~~Threat complex~~\textcolor{blue}{Cluster} 5 is characterized by
 more fisheries types (*i.e.* demersal destructive and pelagic high-bycatch),
 generally lower intensity marine traffic drivers and is located almost
-exclusively in the Northern Gulf. Finally, threat complex 6 is the most exposed
-offshore threat complex and includes all offshore hotspots (Figure \ref{hotspot}).
+exclusively in the Northern Gulf. Finally, ~~threat complex~~\textcolor{blue}{cluster} 6 is the most exposed
+offshore ~~threat complex~~\textcolor{blue}{cluster} and includes all offshore hotspots (Figure \ref{hotspot}).
 It is characterized by high intensity hypoxia, marine traffic and
 pollution, as well as demersal destructive and pelagic high-bycatch fisheries.
-This threat complex corresponds primarily to the Laurentian Channel and
+This ~~threat complex~~\textcolor{blue}{cluster} corresponds primarily to the Laurentian Channel and
 incorporates parts of the Esquiman and Anticosti Channels.
 
-\textcolor{red}{\sout{Of particular concern are threat complexes 3 and 6, which are the two most exposed complexes in the St.
-Lawrence and are characterized by distinct cumulative exposure regimes Between them, they capture most of the coastal and offshore hotspots identified in the St.
-Lawrence.}}
-\textcolor{blue}{Threat complexes 3 and 6 capture most coastal and offshore hotspots and are the two most exposed complexes in the St. Lawrence.}
-They \textcolor{red}{\sout{also}} offer some insight into the potential importance of considering
-spatial dynamics in areas intersecting multiple threat complexes. For example,
-threat complexes 3 and 6 meet at the mouth of the River Saguenay. This area
+\textcolor{blue}{Clusters} 3 and 6 capture most coastal and offshore hotspots and are the two most exposed clusters in the St. Lawrence.
+They offer some insight into the potential importance of considering
+spatial dynamics in areas intersecting multiple ~~threat complexes~~\textcolor{blue}{clusters}. For example,
+~~threat complexes~~\textcolor{blue}{clusters} 3 and 6 meet at the mouth of the River Saguenay. This area
 is particularly dynamic, with deep Atlantic waters advected through estuarine
@@
 drivers from the bottom of the Laurentian Channel and marine traffic drivers
-(threat complex 6) with terrestrial run-off from river outflows and direct
-human impacts (*i.e.* population density; threat complex 3).
+(~~threat complex~~\textcolor{blue}{cluster} 6) with terrestrial run-off from river outflows and direct
+human impacts (*i.e.* population density; ~~threat complex~~\textcolor{blue}{cluster} 3).
 
@@
 initiative, our objective is to uphold the highest existing standards of data
-management and open science. We identified four guiding principles \textcolor{blue}{(Sections 5.1 to 5.4)} to meet this
-objective and that guide the structure of the platform (Figure \ref{diag}).
+management and open science. We identified four guiding principles \textcolor{blue}{(Section 5.1)} to meet this
+objective and ~~that guide the~~ structure of the ~~platform~~ \textcolor{blue}{initiative} (Figure \ref{diag}).
 
-## Unity and inclusiveness
+## \textcolor{blue}{Guiding principles}
+
+### \textcolor{blue}{Unity and inclusiveness}
 
@@
 
-## Findability, accessibility, interoperability and reusability
+### \textcolor{blue}{Findability, accessibility, interoperability and reusability}
 
@@
 which states that data and metadata must be Findable, Accessible, Interoperable
-and Reusable. These principles focus on the ability of humans and machines to
+and Reusable. These \textcolor{blue}{(sub)}principles focus on the ability of humans and machines to
 automatically find and (re)use data and knowledge [@wilkinson2016].
+\textcolor{blue}{As the FAIR Data Principles already exist as a unified set of
+principles, we adopt them as a set of subprinciples guiding our initiative.}
 
@@
 
-## Adaptiveness
+### \textcolor{blue}{Adaptiveness}
 
@@
 
-## Recognition
+### \textcolor{blue}{Recognition}
 
@@
 focus on citation practices that provide appropriate credit to data products.
+
+## \textcolor{blue}{Using \textit{eDrivers}}
+
+\textcolor{blue}{Using \textit{eDrivers} is simplified through the tools already
+in place and will be increasingly accessible as the initiative evolves and other
+tools are developped to ease user experience. We provide an example of  the ease
+with which the data can be accessed and used with the \textit{eDrivers} R package
+to reproduce figure \ref{kernel} (Box 1). The code using the \textit{eDrivers}
+R package to reproduce all the analyses and figures in this manuscript is also
+available on GitHub.}[^ms]
+
+[^ms]: https://github.com/orgs/eDrivers/eDriversMS
 
@@
 exposed to cumulative drivers, especially close to urban areas. We also
-identified six geographically distinct threat complexes that display similar
-cumulative exposure regimes. These complexes reveal that coastal areas are
-particularly exposed to all types of drivers and that multiple drivers
-typically co-occur in space. These results allow us to efficiently identify
+identified six geographically distinct ~~threat complexes~~\textcolor{blue}{areas} that display similar
+cumulative exposure ~~regimes~~\textcolor{blue}{profiles}; these ~~complexes~~ reveal that coastal areas are
+particularly exposed to all types of drivers and that ~~multiple~~\textcolor{blue}{certain} driver \textcolor{blue}{combinations are typical of certain regions of the St.
+Lawrence}~~typically co-occur in space~~.
+These results allow us to efficiently identify
 areas in need of heightened scrutiny from a science and management perspective.
 
-Through *eDrivers*, these observations will be iteratively improved towards an increasingly robust assessment of cumulative exposure and threat complexes as gaps in knowledge are addressed or approaches to describe drivers are refined.
-Arguably, the most meaningful benefit anticipated from *eDrivers* will be the gain in efficient access to comparable data-based knowledge on the exposure of ecosystems to multiple threats.
-\textcolor{blue}{We provide an example of the ease with which this can be accomplished using the \textit{eDrivers} R package to reproduce figure \ref{kernel} (Box 1).} This could pay quick scientific and management dividends by \textcolor{blue}{efficiently} drawing on the knowledge and efforts of a wide range of contributors, by expanding avenues of scientific inquiry, by decreasing overall effort duplication and research costs, and by increasing research efficiency [@franzoni2014].
+Through *eDrivers*, these observations will be iteratively improved towards an
+increasingly robust assessment of cumulative exposure and ~~threat complexes~~\textcolor{blue}{areas with similar cumulative exposure profiles} as
+gaps in knowledge are addressed or approaches to describe drivers are refined.
+Arguably, the most meaningful benefit anticipated from *eDrivers* will be the
+gain in efficient access to comparable data-based knowledge on the exposure of
+ecosystems to multiple threats. This could pay quick scientific
+and management dividends by \textcolor{blue}{efficiently} drawing on the
+knowledge and efforts of a wide range of contributors, by expanding avenues
+of scientific inquiry, by decreasing overall effort duplication and research
+costs, and by increasing research efficiency [@franzoni2014].
 
@@
 
-\textcolor{red}{\sout{While the focus of our paper has been on the description of drivers in the St.
-Lawrence, drivers are but one of the 4 elements required to properly evaluate environmental impacts.
-This knowledge node should be weaved with other, comparable, nodes required for integrated management approaches such as species distribution and marine protected areas.
-Ultimately, all of these knowledges could be weaved together through social-ecological metanetworks (Dee et al.
-2017).}}
-\textcolor{blue}{The next step will be the inclusion of other types of knowledge to our initiative.
+The next step will be the inclusion of other types of knowledge to our initiative.
 Our focus has been on a single element required for fully operational impact assessments.
-Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators} [@dube2001; @dube2003; @jones2016].
-\textcolor{blue}{These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers but fail to diagnose impacts on valued components or non-additive effects} [@dube2001; @dube2003].
-\textcolor{blue}{In contrast, effect-based indicators are direct measurements of valued components (\textit{e.g.} species abundance and biodiversity) and inherently capture the effects of multiple drivers} [@dube2001; @dube2003].
-\textcolor{blue}{Whereas effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers} [@dube2001; @dube2003].
-\textcolor{blue}{Stressor-based and effect-based indicators are, therefore, both required to diagnose causes of ecosystem change} [@dube2003; @jones2016].
-\textcolor{blue}{As a collection of knowledge on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, collections of knowledge describing valued ecosystem components that can be linked to drivers and allow for a better understanding of cumulative impacts.
-Ultimately, interdisciplinary colections of knowledge could be weaved together through social-ecological metanetworks analyses} [@dee2017].
+Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators [@dube2001; @dube2003; @jones2016].
+These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers and fail to diagnose impacts on valued components or non-additive effects [@dube2001; @dube2003].
+In contrast, effect-based indicators are direct measurements of valued components (\textit{e.g.} species abundance and biodiversity) and inherently capture the effects of multiple drivers [@dube2001; @dube2003].
+Whereas effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers [@dube2001; @dube2003].
+Stressor-based and effect-based indicators are, therefore, both required to diagnose causes of ecosystem change [@dube2003; @jones2016].
+As a collection of knowledge on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, collections of knowledge describing valued ecosystem components that can be linked to drivers and allow for a better understanding of cumulative impacts.
+Ultimately, interdisciplinary collections of knowledge could be weaved together through social-ecological metanetworks analyses [@dee2017].
+\textcolor{blue}{In turn, these could be used in conceptual frameworks that help in establishing causal relationships between drivers and valued ecosystem components such as the DPSIR (Driving forces -- Pressure -- State -- Impact -- Response) framework} [@kristensen2004; @gari2015].
+\textcolor{blue}{Through such frameworks, data-based indicators provided through \textit{eDrivers} could be categorized as driving forces, pressures or states, depending on the objective and context of a study.}
 
@@
 SLGO partly addresses this issue.
-\textcolor{blue}{We also wish to provide users with enhanced capabilities and flexibility in using the interactive tool and R package.
-This could include creating automatic reports and more flexibility for user-defined driver-based indicators.}
+We also wish to provide users with enhanced capabilities and flexibility in using the interactive tool and R package.
+This could include creating automatic reports and more flexibility for user-defined driver-based indicators.
 
@@
 \centering
-\includegraphics{../Figures/egsl.png}
-\caption{Description of the Estuary and Gulf of St. Lawrence in Eastern Canada}
+\includegraphics{./Figure1.jpg}
+\caption{\sout{Description of the Estuary and Gulf of St.
+Lawrence in Eastern Canada.}\textcolor{blue}{Description of the St.
+Lawrence System in Eastern Canada, formed by the St.
+Lawrence Estuary and the Gulf of St.
+Lawrence.
+The Estuary is defined by the seawater limit close to Îles d'Orléans to the west and by its connexion to the Gulf near Pointe-des-Monts.
+The Gulf is an interior sea connected to the Atlantic by the Cabot and Belle-Isle Straits to the south and north of Newfoundland, respectively.}}
 \label{egsl}
@@
 \centering
-\includegraphics{../Figures/kernel.png}
-\caption{\textcolor{red}{\sout{Interaction}} \textcolor{blue}{Co-occurrence} between \textcolor{red}{\sout{the intensity of }}hypoxia and demersal destructive fisheries in the St.
-Lawrence.
+\includegraphics{./Figure2.jpg}
+\caption{\textcolor{blue}{Simplified 2-driver example of driver co-occurrence between hypoxia and demersal destructive fisheries in the St.
+Lawrence}.
 An index of hypoxia ($A$) was created using bottom-water dissolved oxygen between 2013 and 2017 (Blais et al., 2018).
@@
 \centering
-\includegraphics{../Figures/footprint.png}
-\caption{Distribution of cumulative footprint in the St. Lawrence System.}
+\includegraphics{./Figure3.jpg}
+\caption{Distribution of cumulative \sout{footprint} \textcolor{blue}{exposure} in the St. Lawrence System.}
 \label{footprint}
@@
 \centering
-\includegraphics{../Figures/hotspot.png}
+\includegraphics{./Figure4.jpg}
 \caption{Distribution of cumulative hotspots in the St. Lawrence System.}
@@
 \centering
-\includegraphics{../Figures/cluster.png}
-\caption{Distribution of threat complexes in the Estuary and Gulf of St.
-Lawrence (upper panel).
-Threat complexes (a term coined by Bowler et al., 2019) are areas with similar cumulative driver exposure regimes.
-Mean intensity of all coastal (red), climate (green), fisheries (blue) and marine traffic (purple) drivers within each threat complex (lower panel).
-Refer \textcolor{red}{\sout{tp}} \textcolor{blue}{to} Table 1 for acronym meaning and to the Supplementary Materials for more details.}
+\includegraphics{./Figure5.jpg}
+\caption{Distribution of \sout{threat complexes}\textcolor{blue}{areas with similar cumlative exposure profiles} in the Estuary and Gulf of St.
+Lawrence\textcolor{blue}{, identified through a clustering approach} (upper panel).
+Mean intensity of all coastal (red), climate (green), fisheries (blue) and marine traffic (purple) drivers within each \sout{threat complex}\textcolor{blue}{cluster} (lower panel).
+Refer to Table 1 for acronym meaning and to the Supplementary Materials for more details.}
 \label{cluster}
@@
 \centering
-\includegraphics{../Figures/structure.png}
+\includegraphics{./Figure6.jpg}
 \caption{Diagram of the platform structure.
@@
 The raw data are then processed through a workflow hosted on the $eDrivers$ GitHub organization (https://github.com/orgs/eDrivers/).
-Data processing may be as simple as data rescaling ($e.g.$ night lights) or make use of more complex methodologies ($e.g.$ acidification).
+Data processing may be as simple as data rescaling\sout{ ($e.g.$ night lights)} or make use of more complex methodologies\sout{ ($e.g.$ acidification)}.
 All data is then hosted on SLGO's geoserver and accessible through their API.
 We developed a R package called $eDrivers$ to access the driver layers through R and we are actively developing a second R package called $eDriversEx$ that will include analytical tools to explore drivers data.
-Finally, we have developed a Shiny application that allows users to explore drivers data interactively (https://david-beauchesne.shinyapps.io/eDriversApp/).
-All R components of the project are hosted and available on the $eDrivers$ GitHub organization.}
+Finally, we have developed a Shiny application, \textcolor{blue}{\textit{eDrivers} app,} that allows users to explore drivers data interactively (https://david-beauchesne.shinyapps.io/eDriversApp/).
+All R components of the project are hosted and available on the $eDrivers$ GitHub organization.
+Iterative and adaptive processes are identified by circular arrows.}
 \label{diag}
@@
 
-Table 1. List of drivers currently available on *eDrivers* and used for the analyses presented in this paper.
+Table 1. List of drivers ~~currently~~ available on *eDrivers* and used for the analyses presented in this paper. Further details on methods and data are available in the Supplementary Material.
 

### 2020-04-24 Code/13-cluster.R

@@ -1,3 +1,4 @@
+# source('Code/13-cluster.R')
 # ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
 source('./code/2-param.R')
 
@@ -26,8 +27,8 @@ dat[[6]] = data.frame(x = drInt2$X5_6)
 
 
 # ~~~~~~~~~~~~~~~~~~~ CLUSTER ~~~~~~~~~~~~~~~~~~~ #
-png('./figures/cluster.png', width = 1280, height = 1280, res = 300, pointsize = 6)
-# jpeg('./figures/cluster.jpg', width = 1280, height = 1280, res = 300, pointsize = 6)
+# png('./figures/cluster.png', width = 1280, height = 1280, res = 300, pointsize = 6)
+jpeg('./figures/cluster.jpg', width = 1280, height = 1280, res = 300, pointsize = 6)
 mat <- matrix(c(1,1,1,1,1,1,2,3,4,5,6,7), nrow = 2, byrow = T)
 layout(mat, heights = c(1, .4))
 cols <- pal(nCl)[clMed]
@@ -53,7 +54,7 @@ yH <- .022*yR
 xW <- .034*xR
 gap <- .034*xR
 gap2 <- .045*yR
-text(x = x[1]+gap, y = y[2]-gap, 'Threat complexes', adj = c(0,.5), font = 2)
+text(x = x[1]+gap, y = y[2]-gap, 'Clusters', adj = c(0,.5), font = 2)
 # Polygons
 for(i in 1:k) {
   xmn <- x[1]+gap+.014*xR
@@ -61,7 +62,7 @@ for(i in 1:k) {
   ymn <- y[2]-gap-i*gap2-yH/2
   ymx <- y[2]-gap-i*gap2+yH/2
   polygon(x = c(xmn, xmx, xmx, xmn), y = c(ymn,ymn,ymx,ymx), col = cols[i], lwd = .5)
-  text(x = xmx+.007*xR, y = mean(c(ymn,ymx)), labels = paste('Threat complex', i), cex = .6, adj = c(0, .5))
+  text(x = xmx+.007*xR, y = mean(c(ymn,ymx)), labels = paste('Cluster', i), cex = .6, adj = c(0, .5))
 }
 
 # Add citie
@@ -70,7 +71,7 @@ cityEGSL(prj = 4326)
 # barplots
 for(i in 1:6) {
   barPlot(dat[[i]])
[…]

### 2020-04-24 Code/4-footprint.R

@@ -33,7 +33,7 @@ plot(st_geometry(egslSimple),
 
 # Legend
 legendEGSL(range = c(min(Fd), max(Fd)),
-           mainTitle = 'Cumulative footprint',
+           mainTitle = 'Cumulative exposure',
            subTitle = '',
            cexMain = .9,
            cexSub = .65,

### 2020-04-24 Tables/datatable.R

@@ -12,22 +12,22 @@ df <- data.frame(Groups = character(0),
                  Acronym = character(0),
                  stringsAsFactors = F)
 
-df[1,] <- c("Climate", "Aragonite", 'Lat/long', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ACID')
-df[2,] <- c("Coastal", "Aquaculture", 'Lat/long', '-','Variable, between 1990-2016', '$presence-absence$','[@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016]','AQUA')
-df[3,] <- c("Climate", "Bottom-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SBT-')
-df[4,] <- c("Climate", "Bottom-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'positie anomalies', '[@galbraith2018]','SBT+')
-df[5,] <- c("Climate",  "Surface-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'negative anomalies', '[@galbraith2018]','SST-')
-df[6,] <- c("Climate", "Surface-water temperature", '~2 $km^2$', 'Monthly', '1981-2010 vs. 2013-2017', 'positive anomalies', '[@galbraith2018]','SST+')
+df[1,] <- c("Climate", "Aragonite", 'Lat/Lon', 'August-September', '2018', '$\\Omega$ $Aragonite$', '[@starr2019]','ACID')
+df[2,] <- c("Coastal", "Aquaculture", 'Lat/Lon', '-','Variable, between 1990-2016', '$presence-absence$','[@mapaq2016; @dfo2016a; @aaf2016; @fa2016; @ffa2016]','AQUA')
+df[3,] <- c("Climate", "Bottom-water temperature", '~2 $km^2$', 'Monthly', 'Climatology: 1980-2010; Current: 2013-2017', 'negative anomalies', '[@galbraith2018]','SBT-')
+df[4,] <- c("Climate", "Bottom-water temperature", '~2 $km^2$', 'Monthly', 'Climatology: 1980-2010; Current: 2013-2017', 'positive anomalies', '[@galbraith2018]','SBT+')
[…]

### 2020-04-24 Revisions2/Rebuttal.md
add rebuttal

---
title: "Rebuttal"
fontfamily: fourier
linestretch: 1
fontsize: 10pt
output:
  pdf_document:
    latex_engine: xelatex
---

<!--
rmarkdown::render('./Revisions2/Rebuttal.md')
rmarkdown::render('./Revisions2/Rebuttal.md', 'word_document')
-->

<!--

TODO:
- [ ] At the end
  - [x] Add line numbers
  - [ ] Spell check
  - [ ] Check references

- [ ] Role of eDrivers in the analyses
  - [x] Paragraph in eDrivers section
  - [ ] Restructure repo and make it reproducible with eDrivers package
  - [ ] Maybe release a version of the package so that the paper will remain reproducible in the future
  - [ ] Write README.md to present the paper and provide links to zenodo archives
  - [ ] Document `eDrivers`
  - [ ] In development for `eDriversEx`
  - [ ] Transform everything to rasters

- [ ] Update bowler2019?
-->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Letter to the editor
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

Dear Dr. Otto,

We are submitting a revised version of the manuscript “Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System” (ID: 469411) for publication in Frontiers in Marine Science. After review, we now propose the following updated title for the manuscript: "*Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St. Lawrence System*". We would like to thank the reviewer for very constructive comments. We feel that we were able to address all the comments from the second reviewer and are now submitting an improved manuscript. Below, you will find our responses to the reviewer’s comments and how we have revised our manuscript in view of these.

We thank you for considering our revised manuscript and look forward to hearing back from you.

Best regards,

David Beauchesne

# Responses to the reviewer’s comments

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
## Question 2
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

***Please highlight the limitations and strengths***

***Comment 1:***

*I felt the manuscript was a bit disjointed. I find myself confused whether the eDrivers platform or characterizing cumulative exposures is the star of the manuscript. The abstract and title seem to imply that eDrivers is the focus but that is not the bulk of the body of the manuscript. As a matter of fact, section 5 of the manuscript which describes the eDrivers is mostly their rationale for developing the tool and does not even adequately address how the tool was used in the analysis.*

> - We thank the reviewer for pointing out this disconnect between title, abstract, and main text. Initial versions of this manuscript focused on *eDrivers*, but in the process of writing the focus shifted to characterizing cumulative exposure in the St. Lawrence, with the *eDrivers* initiative emerging from this process. This better reflects the process that led to the creation of the initiative. The title and abstract were, however, not modified accordingly. To address this we propose the following modification to the manuscript:

> 1. We propose an alternate title for the manuscript: *"Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St. Lawrence System"*.

> 2. We modified the abstract to mirror the structure of the paper and introduce the *eDrivers* initiative to conclude the abstract (Lines 53 to 56).

> 3. We modified the statement of the manuscript's objective at the end of the introduction (lines 132 to 147).

> - How *eDrivers* is used is not the focus of the manucript. As such, we would prefer not to expand extensively on this particular point in the main text body. We did, however, extend and conclude the section presenting the initiative with a short paragraph stating how the package was used for the analysis. We also provide a link to the github repository for this manuscript, which includes the code to reproduce all the analyses and figures. We also contextualize the rationale behind presenting Box 1 and removed reference to Box 1 from the results section. The goal of Box 1 is to introduce how the initiative and its accompanying tools can increase the ease with which holistic analyses can be performed with a simple coded and reproducible example (lines 508 to 513).

***Comment 2:***

*My other major criticism of this work is the loose use of terminology. While not explicitly using a DPSIR (Driver - Pressure- State - Impact - Response) framework, much of the cumulative impacts literature is familiar with this framework. Table 1 is actually a collage of drivers, pressures, and states. I also find it a stretch to characterize all of them as “environmental drivers” as many are definitely anthropogenic “drivers”. Another instance of loose terminology is actually pointed out by the authors where they use the term clusters in presenting methods but threat complexes when discussing results. I suggest the authors use one set of terminology throughout the manuscript to help the reader keep things straight.*

> - The reviewer is right about terminology issues: we loosely used certain terms such as *environmental drivers* in place of *drivers of environmental change*, and duplicated the terminology for other aspects of our work, such as *cumulative exposure regimes* and *threat complexes*; this awkwardly conveys our message. As such, we agree with the reviewer's comment and modified the manuscript to clarify the terminology used throughout the manuscript.

> - However, we avoided categorizing drivers of environmental change using the DPSIR framework. While the framework is widely used globally, it has numerous shortcomings (*e.g.* Gari et al. 2015, 10.1016/j.ocecoaman.2014.11.013). Namely, the difficulties in properly defining the components of the framework, the unclear classification of exogenic or natural drivers like climate change, and the context-dependencies of framework categories. For example, Dempsey et al. (2018; 10.1016/j.ecolind.2016.12.011) found it more useful to classify drivers of change as exogenic or endogenic (see Elliot et al. 2011, 10.1016/j.marpolbul.2010.11.033) because the indicators they used could be classified differently depending on the context or ecosystem in which they were studied. As an example, hypoxic environments may be considered a state if the goal is to study the impacts of climate change on ecosystems, or a pressure if the goal is to study the impacts of climate change on benthic communities. Furthermore, the DPSIR framework is used to structure initiatives that evaluate the chain of causal links between driving forces and responses (Kristensen et al. 2014, National Environmental Research Institute, Denmark). We do not do such an analyses; we rather focus on the exposure of natural systems to drivers, intentionally omitting valued components and their responses (see answers to reviewer 1). We do not dispute the usefulness of the DPSIR framework; it has provided clarity and guidance for many case studies both in terrestrial and marine ecosystems, and could be used with the data-based indicators that we provide through the *eDrivers* initiative. We simply wish to avoid generating further confusion by articulating our work around a specific framework, or forcing categories on our layers that may change depending on a user's objective. As such, we propose the following modifications to the manuscript:

> 1. We modified the definition in the introduction to: "*We broadly define drivers of environmental change (hereafter drivers) as any externality that affects environmental processes and disturbs natural systems*" (lines 84 to 87).

> 2. We added a paragraph at the beginning of the drivers section (3.1) providing more details on how the drivers of change we are providing are data-based indicators that may be viewed as drivers, pressures, stressors, or states depending on the context and objective: "*Drivers, as broadly defined in this study, are data-based indicators of environmental conditions and human activities that are often referred to as driving forces, stressors, pressures or states in the scientific and environmental assessment litterature (e.g. Kristensen, 2004; Halpern et al., 2019). Defining such categories, however, can be difficult and is often context- and ecosystem-specific (Gari et al., 2015; Dempsey et al., 2018). As such, we refrain from articulating our work around a specific framework or imposing categories on data-based products that may change with a user's objective. We rather focus on available data-based indicators that contribute to evaluating an ecosystem's cumulative exposure to multiple threats*" (lines 177 to 185).

> 3. We concluded the perspective paragraph presenting valued components by stating that interdisciplinary collections of knowledge could be used in a DPSIR framework to help in establishing causal relationship between drivers and valued ecosystem components: "*In turn, these could be used in conceptual frameworks that help in establishing causal relationships between drivers and valued ecosystem components such as the DPSIR (Driving forces -- Pressure -- State -- Impact -- Response) framework (Kristensen, 2004; Gari et al., 2015). Through such a framework, the data-based indicators provided through eDrivers could be categorized as driving forces, pressures or states, depending on the objective and context of a study*" lines 561 to 566).

> 4. We modified all instances of *environmental drivers* for *drivers of environmental change*

> 5. We removed all mentions of *threat complexes* and *cumulative exposure regimes* in favour of *cumulative exposure profiles*, which, we feel, better captures our intent without adding unecessary terms. We also refer to *threat complexes* identified through the clustering approach as clusters.

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
## Question 3
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

***Please provide your detailed review report to the editor and authors (including any comments on the Q4 Check List):***

*The manuscript 469411 entitled “Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System” is a first cut at assessing the cumulative impacts of multiple “drivers” on a system. The authors provide an example of an analysis conducted through the collection of a massive amount of data through the development of an eDrivers platform. I applaud their efforts as I know that collecting large and often disparate data sets to conduct an assessment can be challenging. That being said, I felt the manuscript was a bit disjointed. I find myself confused whether the eDrivers platform or characterizing cumulative exposures is the star of the manuscript. The abstract and title seem to imply that eDrivers is the focus but that is not the bulk of the body of the manuscript. As a matter of fact, section 5 of the manuscript which describes the eDrivers is mostly their rationale for developing the tool and does not even adequately address how the tool was used in the analysis.*

> - See answer to question 2

*My other major criticism of this work is the loose use of terminology. While not explicitly using a DPSIR (Driver - Pressure- State - Impact - Response) framework, much of the cumulative impacts literature is familiar with this framework. Table 1 is actually a collage of drivers, pressures, and states. I also find it a stretch to characterize all of them as “environmental drivers” as many are definitely anthropogenic “drivers”. Another instance of loose terminology is actually pointed out by the authors where they use the term clusters in presenting methods but threat complexes when discussing results. I suggest the authors use one set of terminology throughout the manuscript to help the reader keep things straight.*

> - See answer to question 2

*Below are a few more specific comments:*

*Line 73 - How does the intensifying human footprint result in increasingly intricate environmental exposure regimes? I think this is confusing when you claim anthropogenic drivers as environmental drivers.*

> - We mean that the expanding and intensifying human footprint results in drivers of change increasingly overlapping at higher intensities, which in turns creates more intricate exposure regimes. We modified the terminology used throughout the manuscsript and do not refer to drivers as "environmental drivers" anymore. We rather use the expression "drivers of environmental change" (see answer to question 2 for more details). We also simplified this sentence to: *As a result, the St. Lawrence System is exposed to an increasing number of drivers of environmental change, as is observed across ecosystems globally (see Halpern et al., 2015b; Halpern et al., 2019)* (lines 79 to 84).

*Line 97 - The sentence “Describing drivers will therefore provide critical information on areas most exposed to drivers in the St. Lawrence.” does not make sense. Please clarify what you mean.*

> - We modified this sentence by: *While co-occurring drivers may not interact, driver co-occurrence is a requirement for interactions to exist. Knowledge of their co-distribution can therefore identify areas where driver interactions are most likely observed.* (lines 104 to 107).

*Line 124 - You never state an initial goal prior to this statement.*

> - We modified the statement of the objectives to address this issue (lines 132 to 147). See answer to question 2 for more details.

*Line 128 - Should your goal be to characterize the distribution and intensity of drivers with two objectives as stated starting on line 129?*

> - We now state: *Our main goal in this study is to characterize the distribution and intensity of drivers in the St. Lawrence. More specifically, our objectives are to: 1) identify areas of high cumulative exposure to drivers and 2) characterize areas with similar cumulative exposure profiles, i.e. areas exposed to similar suites of co-occurring drivers* (lines 135 to 138).

*Line 131 - Are cumulative exposure regimes the same as threat complexes described later in the manuscript?*

> - See answer to question 2.

*Section 3 - No mention of eDrivers or it’s role in the analysis.*

> - See answer to question 2.

*Line 142 - I’m not familiar with the St. Lawrence system. Is it simply called the Estuary or the Estuary of the St. Lawrence? I see it listed as Estuary on the map but there is no label for the Gulf of St. Lawrence.*

> - The proper names are the St. Lawrence Estuary and the Gulf of St. Lawrence. We modified the methods section title and text (line 150) and adjusted the map accordingly (see updated Figure 1).

*Line 172 - Is the reference Halpern et al. 2015 a or b?*

> - This reference refers to the data repository, hence Halpern et al. 2015. We are using the journal's reference style `.csl` file, which formats the references as presented in the manuscript. This is why there is no letter `a` after this reference. Frontiers website states: *"Frontiers referencing style will be applied during typesetting, and is dependent on the journal."* We therefore left the reference as is in the revised manuscript.

*Line 177 - The sentence “We selected global data . . . ” feels like it needs more context. Are you saying all drivers in Table 1 were unavailable at the regional scale?*

> - We mean that we used global datasets with sufficient resolution in instances where regional data on drivers were not available,. We modified the text to clarify this: *"Where regional data on drivers were unavailable, available global data at a resolution adequate for the scale of the St. Lawrence System were used instead (*e.g.* marine pollution)"* (lines 190 to 194).

*Line 192 - The figure of the hexagonal cells seems integral to the analysis and should be included in the main manuscript. Actually as I write this, I realize that the sentence reads 245k of 1 km2 hexagonal cells. My original reading was of much larger size cells. Perhaps rewording to say that the system was divided into a regular grid of 1 km2 hexagonal cells is sufficient. Also just an FYI that the hex cells are not shown in figure S2.*

> - We modified the text accordingly: *"The St. Lawrence System was divided into a regular grid of 1 $km^2$ cells in which all drivers were integrated (Figure S2)"* (lines 209 to 212).

> - As for the cells on figure S2, they are invisible due to the scale of the system. Showing cell borders hides fill colors, especially with so many maps shown on a single figure. We doubt whether it would be relevant to the manuscript, but, if requested, we could provde a finer scale view of the system and show the grid. Incidentally, we have since modified our grid for a grid with square cells, which allows us to work with rasters rather than vectorized polygons, which is more computationally efficient.

*Section 3.3 - I fail to see what this section and subsequent results add to the study.*

> - See answer to question 4

*Line 206 - I was confused by what objective 1 was and hence my comment on Line 128.*

> - We agree with this comment and have adjusted the statement of the objectives in the introduction accordingly (lines 135 to 138). Specific references to objectives in the text should now be clear.

*Line 226 - Don’t use different terms in different parts of the manuscript as that makes it difficult for the reader to follow.*

> - See answer to question 2.

*Line 272 - What is Box 1?*

> - The goal of Box 1 is to introduce how the *eDrivers* initiative and its accompanying tools can increase the ease with which holistic analyses can be performed. We removed reference to Box 1 from the results section and now introduce it in the perspective section. See answer to question 2 for more details.

*Section 4.1 - Does this fit with overall goal and objectives as stated in the intro?*

> - See answer to question 4

*Line 285 - Cumulative exposure is called footprints in figure caption.*

> - We modified the figure and the figure caption.

*Line 373 - What are the four guiding principles? I assume they are section 5.1 - 5.4 but it would not hurt to make that explicit.*

> - We modified the text accordingly (line 412).

*Section 5.2 - Is this a principle of principles?*

> - We are adopting a preexisting and unified set of principles dubbed as the FAIR Data Principles as one of our guiding principles. As this can indeed be confusing, we now state that we adopt the unified FAIR Data Principles as a set of subprinciples guiding our initiative (lines 463 to 465).

*Figure 1 - The figure and caption should be stand alone from the manuscript so it would be helpful to add contextual information presented in the text.*

> - We modified Figure 1 to show that the map presents the St. Lawrence and to clearly identify the Gulf of St. Lawrence and St. Lawrence Estuary. We also modified Figure 1's caption to provide a brief description of the Estuary and Gulf (see updated Figure 1 and accompanying caption).

*Figure 6 - What do you mean by night lights in the caption? Also, what do the dots mean? You can also make the diagram specific with the R packages and shiny app to which you are referring.*

> - Night lights referred to the raw data used for the coastal development driver layer. However, as these examples -- along with the acidification example -- do not confer any clarity to that figure, we simply removed them.

> - The dots were in fact two arrows that meant that those processes should be adaptive and iterative; these were unclear. We modified the figure to clarify it, and added an explanation in the figure caption.

> - As suggested, we made the figure specific to the tools and organizations presented in the manuscript.

*Table 1 - What do you mean by Lat/long spatial resolution? I think you mean point estimates or Lat/Lon. Also, what does Dissemination areas mean? I am also confused by what you mean with the 1981 - 2010 vs 2013 - 2017 under years. Is one of those sets the climatology you used? This should be explicit in the main manuscript even if it is explained further in the supplemental materials. Also, positive is spelled wrong for SBT+.*

> - The Lat/long resolution are either sampling coordinates or activity coordinates such as aquaculture or fishing. We modified "Lat/long" for "Lat/Lon" in the table.

> - Dissemination areas are the smallest standard geographic area in which census data, *i.e.* population count, are disseminated and they combine to cover all of Canada and their size and shape is highly variable. For example, their sizes range from 0.0003 to over 44000 $km^2$ in our study area. We modified the table to include this range by stating "Variable: < 1 to > 40000 $km^2$" and included this information in the Supplementary Material. We also indicated that methods and details for each driver are available in the supplementary materials.

> - The years 1981 to 2010 is the climatology used as a reference and 2013-2017 is the current period. We modified the table to clarify this by stating: "Climatology: 1980-2010; Current: 2013-2017"

> - We corrected the word "positive", thank you for identifying the error.

## Question 4

***Please comment on the methods, results and data interpretation. If there are any objective errors, or if the conclusions are not supported, you should detail your concerns.***

***Comment 1:***

*The abstract and title seem to imply that eDrivers is the focus but that is not the bulk of the body of the manuscript. As a matter of fact, section 5 of the manuscript which describes the eDrivers is mostly their rationale for developing the tool and does not even adequately address how the tool was used in the analysis.*

***Answer:***

> - See answer to question 2

***Comment 2:***

*I wonder is section 3.3 and it's corresponding results section 4.1 is necessary to address their overall goal and stated objectives.*

***Answer:***

> - In sections 3.3 and 4.1, we illustrated how a simple 2-driver case could be used to identify areas of such high relative intensity driver overlap. These sections provide an example that demonstrates how driver co-occurrence is evaluated; it should serve as a stepping stone to the integrative exposure indicators that we use after. As such, we believe that they offer valuable support in interpreting our results. However, we agree with the reviewer that these sections appear superfluous. We therefore modified the manuscript to better introduce this example:

> 1. We removed sections 3.3 and 4.1 and incorporated the example in the section on cumulative exposure (lines 221 to 225 and lines 310 to 318).

> 2. We now clearly state that this is an example provided for better clarity and that it serves as a stepping stone to the integrative indicators of exposure (lines 227 to 229).

> 3. We removed the discussion on species-specific responses to hypoxia and fisheries. We intentionally avoid discussing valued components throughout the manuscript to focus specificaly on drivers. Discussing species responses in presenting the simplified example undoubtedly contributed to the example feeling out of place.

> 4. We modified the figure caption to state that it is a simplified example.

> - Although we believe that the example is valuable, it does not address a specific objective, as rightfully identified by the reviewer. As such, if the reviewer still feels that the proposed changes are insufficient, we are willing to remove it entirely from the manuscript.

### 2020-04-24 Revisions2/Rebuttal.md
minor spelling mistakes

@@
 
-> 2. We modified the abstract to mirror the structure of the paper and introduce the *eDrivers* initiative to conclude the abstract (Lines 53 to 56).
+> 2. We modified the abstract to mirror the structure of the paper and introduced the *eDrivers* initiative to conclude the abstract (Lines 53 to 56).
 
@@
 
-> - How *eDrivers* is used is not the focus of the manucript.
+> - How *eDrivers* is used is not the focus of the manuscript.
 As such, we would prefer not to expand extensively on this particular point in the main text body.
@@
 2014, National Environmental Research Institute, Denmark).
-We do not do such an analyses; we rather focus on the exposure of natural systems to drivers, intentionally omitting valued components and their responses (see answers to reviewer 1).
+We do not do such an analysis; we rather focus on the exposure of natural systems to drivers, intentionally omitting valued components and their responses (see answers to reviewer 1).
 We do not dispute the usefulness of the DPSIR framework; it has provided clarity and guidance for many case studies both in terrestrial and marine ecosystems, and could be used with the data-based indicators that we provide through the *eDrivers* initiative.
@@
 > 5.
-We removed all mentions of *threat complexes* and *cumulative exposure regimes* in favour of *cumulative exposure profiles*, which, we feel, better captures our intent without adding unecessary terms.
+We removed all mentions of *threat complexes* and *cumulative exposure regimes* in favour of *cumulative exposure profiles*, which, we feel, better captures our intent without adding uncecessary terms.
 We also refer to *threat complexes* identified through the clustering approach as clusters.
@@
 > - We mean that the expanding and intensifying human footprint results in drivers of change increasingly overlapping at higher intensities, which in turns creates more intricate exposure regimes.
-We modified the terminology used throughout the manuscsript and do not refer to drivers as "environmental drivers" anymore.
+We modified the terminology used throughout the manuscript and do not refer to drivers as "environmental drivers" anymore.
 We rather use the expression "drivers of environmental change" (see answer to question 2 for more details).
@@
 
-> - We mean that we used global datasets with sufficient resolution in instances where regional data on drivers were not available,.
+> - We mean that we used global datasets with sufficient resolution in instances where regional data on drivers were not available.
 We modified the text to clarify this: *"Where regional data on drivers were unavailable, available global data at a resolution adequate for the scale of the St.
@@
 Showing cell borders hides fill colors, especially with so many maps shown on a single figure.
-We doubt whether it would be relevant to the manuscript, but, if requested, we could provde a finer scale view of the system and show the grid.
+We doubt whether it would be relevant to the manuscript, but, if requested, we could provide a finer scale view of the system and show the grid.
 Incidentally, we have since modified our grid for a grid with square cells, which allows us to work with rasters rather than vectorized polygons, which is more computationally efficient.
@@
 We removed the discussion on species-specific responses to hypoxia and fisheries.
-We intentionally avoid discussing valued components throughout the manuscript to focus specificaly on drivers.
+We intentionally avoid discussing valued components throughout the manuscript to focus specifically on drivers.
 Discussing species responses in presenting the simplified example undoubtedly contributed to the example feeling out of place.

### 2020-05-01 Revisions2/Rebuttal.md
add comments from co-authors

@@
 rmarkdown::render('./Revisions2/Rebuttal.md')
-rmarkdown::render('./Revisions2/Rebuttal.md', 'word_document')
--->
-
-<!--
-
-TODO:
-- [ ] At the end
-  - [x] Add line numbers
-  - [ ] Spell check
-  - [ ] Check references
-
-- [ ] Role of eDrivers in the analyses
-  - [x] Paragraph in eDrivers section
-  - [ ] Restructure repo and make it reproducible with eDrivers package
-  - [ ] Maybe release a version of the package so that the paper will remain reproducible in the future
-  - [ ] Write README.md to present the paper and provide links to zenodo archives
-  - [ ] Document `eDrivers`
-  - [ ] In development for `eDriversEx`
-  - [ ] Transform everything to rasters
-
-- [ ] Update bowler2019?
 -->
@@
 After review, we now propose the following updated title for the manuscript: "*Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St.
-Lawrence System*".
+Lawrence System in Canada*".
 We would like to thank the reviewer for very constructive comments.
@@
 The title and abstract were, however, not modified accordingly.
-To address this we propose the following modification to the manuscript:
+To address this, we propose the following modifications to the manuscript:
 
-> 1. We propose an alternate title for the manuscript: *"Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St. Lawrence System"*.
+> 1. We propose an alternate title for the manuscript: *"Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St. Lawrence System in Canada"*.
 
-> 2. We modified the abstract to mirror the structure of the paper and introduced the *eDrivers* initiative to conclude the abstract (Lines 53 to 56).
+> 2. We modified the abstract to mirror the structure of the paper and introduced the *eDrivers* initiative to conclude the abstract (Lines 52 to 55).
 
-> 3. We modified the statement of the manuscript's objective at the end of the introduction (lines 132 to 147).
+> 3. We modified the statement of the manuscript's objective at the end of the introduction (lines 131 to 146).
 
 > - How *eDrivers* is used is not the focus of the manuscript.
-As such, we would prefer not to expand extensively on this particular point in the main text body.
+As such, we would prefer not to expand extensively on this particular point in the main body of the text.
 We did, however, extend and conclude the section presenting the initiative with a short paragraph stating how the package was used for the analysis.
@@
 We also contextualize the rationale behind presenting Box 1 and removed reference to Box 1 from the results section.
-The goal of Box 1 is to introduce how the initiative and its accompanying tools can increase the ease with which holistic analyses can be performed with a simple coded and reproducible example (lines 508 to 513).
+The goal of Box 1 is to introduce how the initiative and its accompanying tools can increase the ease with which holistic analyses can be performed with a simple coded and reproducible example (lines 502 to 506).
 
@@
 
-> - However, we avoided categorizing drivers of environmental change using the DPSIR framework.
+> - Nonetheless, we avoided categorizing drivers of environmental change using the DPSIR framework.
 While the framework is widely used globally, it has numerous shortcomings (*e.g.* Gari et al.
@@
 2011, 10.1016/j.marpolbul.2010.11.033) because the indicators they used could be classified differently depending on the context or ecosystem in which they were studied.
-As an example, hypoxic environments may be considered a state if the goal is to study the impacts of climate change on ecosystems, or a pressure if the goal is to study the impacts of climate change on benthic communities.
+For example, hypoxic environments may be considered a state if the goal is to study the impacts of climate change on ecosystems, or a pressure if the goal is to study the impacts of climate change on benthic communities.
 Furthermore, the DPSIR framework is used to structure initiatives that evaluate the chain of causal links between driving forces and responses (Kristensen et al.
@@
 > 1.
-We modified the definition in the introduction to: "*We broadly define drivers of environmental change (hereafter drivers) as any externality that affects environmental processes and disturbs natural systems*" (lines 84 to 87).
+We modified the definition in the introduction to: "*We broadly define drivers of environmental change (hereafter, drivers) as any externality that affects environmental processes and disturbs natural systems*" (lines 85 to 86).
 
 > 2.
-We added a paragraph at the beginning of the drivers section (3.1) providing more details on how the drivers of change we are providing are data-based indicators that may be viewed as drivers, pressures, stressors, or states depending on the context and objective: "*Drivers, as broadly defined in this study, are data-based indicators of environmental conditions and human activities that are often referred to as driving forces, stressors, pressures or states in the scientific and environmental assessment litterature (e.g.
+We added a paragraph at the beginning of the drivers section (3.1) to provide more details on how the drivers of change we identify are data-based indicators that may be viewed as drivers, pressures, stressors, or states depending on the context and objective: "*Drivers, as broadly defined in this study, are data-based indicators of environmental conditions and human activities that are often referred to as driving forces, stressors, pressures, or states in the scientific and environmental assessment litterature (e.g.
 Kristensen, 2004; Halpern et al., 2019).
@@
 As such, we refrain from articulating our work around a specific framework or imposing categories on data-based products that may change with a user's objective.
-We rather focus on available data-based indicators that contribute to evaluating an ecosystem's cumulative exposure to multiple threats*" (lines 177 to 185).
+Instead, we focus on available data-based indicators that contribute to evaluate an ecosystem's cumulative exposure to multiple threats*" (lines 174 to 182).
 
 > 3.
-We concluded the perspective paragraph presenting valued components by stating that interdisciplinary collections of knowledge could be used in a DPSIR framework to help in establishing causal relationship between drivers and valued ecosystem components: "*In turn, these could be used in conceptual frameworks that help in establishing causal relationships between drivers and valued ecosystem components such as the DPSIR (Driving forces -- Pressure -- State -- Impact -- Response) framework (Kristensen, 2004; Gari et al., 2015).
-Through such a framework, the data-based indicators provided through eDrivers could be categorized as driving forces, pressures or states, depending on the objective and context of a study*" lines 561 to 566).
+We concluded the perspective paragraph presenting valued components by stating that interdisciplinary collections of knowledge could be used in a DPSIR framework to help establish causal relationships between drivers and valued ecosystem components: "*In turn, these could be used in conceptual frameworks that help to establish causal relationships between drivers and valued ecosystem components such as the DPSIR (Driving forces -- Pressure -- State -- Impact -- Response) framework (Kristensen, 2004; Gari et al., 2015).
+Within such a framework, the data-based indicators provided through eDrivers could be categorized as driving forces, pressures or states, depending on the objective and context of a study*" lines 554 to 559).
 
@@
 
-> - We mean that the expanding and intensifying human footprint results in drivers of change increasingly overlapping at higher intensities, which in turns creates more intricate exposure regimes.
+> - We mean that the expanding and intensifying human footprint results in drivers of change increasingly overlapping at higher intensities, which in turn creates more intricate exposure regimes.
 We modified the terminology used throughout the manuscript and do not refer to drivers as "environmental drivers" anymore.
-We rather use the expression "drivers of environmental change" (see answer to question 2 for more details).
-We also simplified this sentence to: *As a result, the St.
-Lawrence System is exposed to an increasing number of drivers of environmental change, as is observed across ecosystems globally (see Halpern et al., 2015b; Halpern et al., 2019)* (lines 79 to 84).
+Instead, we use the expression "drivers of environmental change" (see answer to question 2 for more details).
+We also simplified this sentence to: *Consequently, the St.
+Lawrence System is exposed to an increasing number of drivers of environmental change, as is observed across ecosystems globally (see Halpern et al., 2015b; Halpern et al., 2019)* (lines 80 to 86).
 
@@
 
-> - We modified this sentence by: *While co-occurring drivers may not interact, driver co-occurrence is a requirement for interactions to exist.
-Knowledge of their co-distribution can therefore identify areas where driver interactions are most likely observed.* (lines 104 to 107).
+> - We modified this sentence to: *Whereas co-occurring drivers may not interact, driver co-occurrence is a requirement for interactions to exist.
+Knowledge of their co-distribution can therefore identify areas where driver interactions are most likely observed.* (lines 103 to 105).
 
@@
 
-> - We modified the statement of the objectives to address this issue (lines 132 to 147). See answer to question 2 for more details.
+> - We modified the statement of the objectives to address this issue (lines 134 to 146). See answer to question 2 for more details.
 
@@
 
-> - We now state: *Our main goal in this study is to characterize the distribution and intensity of drivers in the St.
-Lawrence.
+> - We agree and we now state: *The main goal of this study is to characterize the distribution and intensity of drivers in the St.
+Lawrence System.
 More specifically, our objectives are to: 1) identify areas of high cumulative exposure to drivers and 2) characterize areas with similar cumulative exposure profiles, i.e.
-areas exposed to similar suites of co-occurring drivers* (lines 135 to 138).
+areas exposed to similar suites of co-occurring drivers* (lines 134 to 138).
 
@@
 
-> - See answer to question 2.
+> - We removed all mentions of threat complexes and cumulative exposure regimes in favour of cumulative exposure profiles (see answer to question 2).
 
@@
 
-> - The proper names are the St. Lawrence Estuary and the Gulf of St. Lawrence. We modified the methods section title and text (line 150) and adjusted the map accordingly (see updated Figure 1).
+> - The proper names are the St. Lawrence Estuary and the Gulf of St. Lawrence. We modified the methods section title and text (line 148) and adjusted the map accordingly (see updated Figure 1).
 
@@
 We modified the text to clarify this: *"Where regional data on drivers were unavailable, available global data at a resolution adequate for the scale of the St.
-Lawrence System were used instead (*e.g.* marine pollution)"* (lines 190 to 194).
+Lawrence System were used instead (*e.g.* marine pollution)"* (lines 189 to 191).
 
@@
 
-> - We modified the text accordingly: *"The St. Lawrence System was divided into a regular grid of 1 $km^2$ cells in which all drivers were integrated (Figure S2)"* (lines 209 to 212).
+> - We modified the text accordingly: *"The St. Lawrence System was divided into a regular grid of 1 $km^2$ cells in which all drivers were integrated (Figure S2)"* (lines 207 to 209).
 
@@
 We doubt whether it would be relevant to the manuscript, but, if requested, we could provide a finer scale view of the system and show the grid.
-Incidentally, we have since modified our grid for a grid with square cells, which allows us to work with rasters rather than vectorized polygons, which is more computationally efficient.
+Incidentally, we have since modified our grid for a grid with square cells, which allows us to work with rasters rather than vectorized polygons, which is more efficient computationally.
 
@@
 
-> - We agree with this comment and have adjusted the statement of the objectives in the introduction accordingly (lines 135 to 138). Specific references to objectives in the text should now be clear.
+> - We agree with this comment and have adjusted the statement of the objectives in the introduction accordingly (lines 134 to 138). Specific references to objectives in the text should now be clear.
 
@@
 
-> - We are adopting a preexisting and unified set of principles dubbed as the FAIR Data Principles as one of our guiding principles.
-As this can indeed be confusing, we now state that we adopt the unified FAIR Data Principles as a set of subprinciples guiding our initiative (lines 463 to 465).
+> - We are adopting a preexisting and unified set of principles, dubbed the FAIR Data Principles, as one of our guiding principles.
+As this can be confusing, we now state that we adopt the unified FAIR Data Principles as a set of subprinciples guiding our initiative (lines 457 to 458).
 
@@
 > - We modified Figure 1 to show that the map presents the St.
-Lawrence and to clearly identify the Gulf of St.
+Lawrence System and to clearly identify the Gulf of St.
 Lawrence and St.
@@
 We modified the table to include this range by stating "Variable: < 1 to > 40000 $km^2$" and included this information in the Supplementary Material.
-We also indicated that methods and details for each driver are available in the supplementary materials.
+We also indicated that methods and details for each driver are available in the supplementary materials in the table caption.
 
-> - The years 1981 to 2010 is the climatology used as a reference and 2013-2017 is the current period. We modified the table to clarify this by stating: "Climatology: 1980-2010; Current: 2013-2017"
+> - The years 1981 to 2010 form the climatology used as a reference and 2013-2017 is the current period. We modified the table to clarify this by stating: "Climatology: 1980-2010; Current: 2013-2017"
 
@@
 As such, we believe that they offer valuable support in interpreting our results.
-However, we agree with the reviewer that these sections appear superfluous.
+Nevertheless, we agree with the reviewer that these sections needed to be better contextualized to showcase their relevance.
 We therefore modified the manuscript to better introduce this example:
 
-> 1. We removed sections 3.3 and 4.1 and incorporated the example in the section on cumulative exposure (lines 221 to 225 and lines 310 to 318).
+> 1. We removed sections 3.3 and 4.1 and incorporated the example in the section on cumulative exposures (lines 218 to 225 and lines 305 to 313).
 
-> 2. We now clearly state that this is an example provided for better clarity and that it serves as a stepping stone to the integrative indicators of exposure (lines 227 to 229).
+> 2.
+We now state that this is an example provided to better illustrate how we evaluate exposure from multiple drivers and that it serves as a stepping stone to the integrative indicators of exposure (lines 223 to 225).
 
@@
 > 4. We modified the figure caption to state that it is a simplified example.
-
-> - Although we believe that the example is valuable, it does not address a specific objective, as rightfully identified by the reviewer.
-As such, if the reviewer still feels that the proposed changes are insufficient, we are willing to remove it entirely from the manuscript.

### 2020-05-01 Revisions2/Beauchesne-eDriversMS.Rmd
add comments from co-authors

@@
 ---
-title: Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St. Lawrence System
+title: Characterizing Exposure to and Sharing Knowledge of Drivers of Environmental Change in the St. Lawrence System in Canada
 author: David Beauchesne^1,2,\*^, Rémi M.
@@
 ^1^Institut des sciences de la mer, Université du Québec à Rimouski, Rimouski, QC, Canada \newline
-^2^Québec Océan, Département de biologie, Université Laval, Québec, QC, Canada \newline
+^2^ArcticNet, Québec Océan, Département de biologie, Université Laval, Québec, QC, Canada \newline
 ^3^Département de biologie, Université de Sherbrooke, Sherbrooke, QC, Canada \newline
@@
 The St. Lawrence is a vast and complex socio-ecological system providing a
-wealth of services sustaining numerous economic sectors.
-~~These ecosystems are~~ \textcolor{blue}{This ecosystem is} subject to significant human pressures that overlap and potentially interact with climate driven environmental changes.
-Our objective in this paper is to systematically characterize the distribution and intensity of drivers of \textcolor{blue}{environmental} change \textcolor{blue}{(hereafter drivers)} in the St.
+wealth of services that sustain numerous economic sectors.
+~~These ecosystems are~~ \textcolor{blue}{This ecosystem is} subject to significant human pressures that overlap and potentially interact with climate-driven environmental changes.
+Our objective in this paper is to systematically characterize the distribution and intensity of drivers of \textcolor{blue}{environmental} change \textcolor{blue}{(hereafter, drivers)} in the St.
 Lawrence System.
-~~To do so, we launch *eDrivers*, an open knowledge platform gathering experts committed to structuring, standardizing and sharing knowledge on drivers of change in support of science and management.~~ We gathered data\textcolor{blue}{-based indicators for} ~~on~~ 22 coastal, climate, fisheries and marine traffic drivers through collaborations, existing environmental initiatives and open data portals.
+~~To do so, we launch *eDrivers*, an open knowledge platform gathering experts committed to structuring, standardizing and sharing knowledge on drivers of change in support of science and management.~~ We gathered data\textcolor{blue}{-based indicators for} ~~on~~ 22 coastal, climate, fisheries, and marine traffic drivers through collaborations, existing environmental initiatives and open data portals.
 We show that few areas of the St.
 Lawrence are free of cumulative exposure.
-The Estuary, ~~the~~ Anticosti Gyre and coastal areas are particularly exposed, especially in the vicinity of urban centers.
-We identified 6 ~~areas of~~ distinct ~~cumulative exposure regimes~~ \textcolor{blue}{clusters with similar suites of co-occurring drivers} and show that certain driver \textcolor{blue}{combinations are typical of} ~~typically co-occur in~~
+The Estuary, ~~the~~ Anticosti Gyre, and coastal areas are particularly exposed, especially in the vicinity of urban centers.
+We identified 6 ~~areas of~~ distinct ~~cumulative exposure regimes~~ \textcolor{blue}{clusters with similar suites of co-occurring drivers} and show that certain driver \textcolor{blue}{combinations are inherent to} ~~typically co-occur in~~
 different regions of the St. Lawrence and that coastal areas are exposed to
@@
 most exposure hotspots \textcolor{blue}{and} that show the convergence of contrasting ~~exposure
-regimes~~\textcolor{blue}{cumulative exposure profiles} at the head of the Laurentian Channel.
-\textcolor{blue}{Sharing the knowledge of drivers acquired emerged as a priority to minimize future environmental assessment efforts.
+regimes~~ \textcolor{blue}{cumulative exposure profiles} at the head of the Laurentian Channel.
+\textcolor{blue}{Sharing the knowledge of drivers emerged as a priority to facilitate future environmental assessment efforts.
 We thus launch \textit{eDrivers}, an open knowledge platform gathering experts committed to structuring, standardizing and sharing knowledge on drivers of environmental change in support of holistic science and management.} *eDrivers* was built on a series of guiding principles upholding existing data management and open science standards.
@@
 and heterogeneous array of habitats suited for the establishment of diverse and
-productive ecological communities [@savenkoff2000]. As a result, the St.
+productive ecological communities [@savenkoff2000]. As a result, the St. Lawrence System has
 benefited the Canadian economy. It sustains a rich fisheries
 industry targeting more than 50 species, serves as the gateway to eastern
-North-America by granting access to more than 40 ports and the most densely
+North-America by granting access to more than 40 ports and is the most densely
 populated Canadian region, hosts a booming tourism industry and an expanding
-aquaculture production, fosters emerging activities and boasts a yet untapped
+aquaculture production, fosters emerging activities, and boasts a yet untapped
 hydrocarbon potential [@beauchesne2016; @schloss2017; @archambault2017]. With
@@
 infrastructure development and research [*e.g.* @governmentofquebec2015; @rqm2018], an intensification of the human footprint is likely in the St. Lawrence System.
-
 ~~As elsewhere in the world [see @halpern2015a; @halpern2019], this intensifying human footprint will likely result in increasingly intricate environmental exposure regimes, *i.e.* suites of overlapping drivers of environmental change threatening ecosystems, habitats and ecological communities.~~
-\textcolor{blue}{As a result, the St.
+\textcolor{blue}{Consequently, the St.
 Lawrence System is exposed to an increasing number of drivers of environmental change, as is observed across ecosystems globally} [see @halpern2015a; @halpern2019].
@@
 environmental processes and disturb natural systems.~~
-\textcolor{blue}{We broadly define drivers of environmental change (hereafter drivers) as any externality that affects environmental processes and disturbs natural systems.}
+\textcolor{blue}{We broadly define drivers of environmental change (hereafter, drivers) as any externality that affects environmental processes and disturbs natural systems.}
 Drivers may originate from natural or human-induced biophysical processes (*e.g.* sea surface-water temperature anomalies and hypoxia) or directly from anthropogenic activities (*e.g.* fisheries and marine pollution).
@@
 
-~~The uncertainty associated with complex driver interactions must therefore be taken into account when investigating environmental impacts [@cote2016], yet research on the effects of drivers in marine environments remains overwhelmingly focused on single driver assessments} [@obrien2019].~~  Increasing exposure and the experiences of past ecological tragedies such as the collapse of cod fisheries [@frank2005; @dempsey2018] and the decline of the beluga and right whale populations [@plourde2014] together urge the need to characterize the distribution, intensity and co-occurrence of drivers in the St.
-Lawrence System.
-\textcolor{blue}{Research on the effects of drivers in marine environments nonetheless remains overwhelmingly focused on single driver assessments} [@obrien2019].
-\textcolor{blue}{While co-occurring drivers may not interact, driver co-occurrence is a requirement for interactions to exist.
+~~The uncertainty associated with complex driver interactions must therefore be taken into account when investigating environmental impacts [@cote2016], yet research on the effects of drivers in marine environments remains overwhelmingly focused on single driver assessments} [@obrien2019].~~  Increasing exposure and the experiences of past ecological tragedies in the St.
+Lawrence System such as the collapse of cod fisheries [@frank2005; @dempsey2018] and the decline of the beluga and right whale populations [@plourde2014] together urge the need to characterize the distribution, intensity and co-occurrence of drivers in system.
+\textcolor{blue}{Research on the effects of drivers in marine environments, nonetheless, remains overwhelmingly focused on single driver assessments} [@obrien2019].
+\textcolor{blue}{Whereas co-occurring drivers may not interact, driver co-occurrence is a requirement for interactions to exist.
 Knowledge of their co-distribution can therefore identify areas where driver interactions are most likely observed.}
@@
 On one hand, there is an overwhelming and expanding wealth of data available.
-Such information overload may inhibit our ability to make decisions based on scientific information, promote massive effort duplication, disproportionately appropriate research funds to certain sectors, and obscure knowledge gaps amid a sea of information [@eppler2004].
-On the other hand, crucial data are lacking and remain largely unavailable or inaccessible for a variety of reasons, including proprietary rights, lack of organizational time, capacity and training, and in some cases unwillingness to share; this curtails our ability for appropriate decision-making.
+Such information overload may inhibit our ability to make decisions based on scientific information, promote massive duplication of effort, disproportionately appropriate research funds to certain sectors, and obscure knowledge gaps amid a sea of information [@eppler2004].
+On the other hand, crucial data are lacking and remain largely unavailable or inaccessible for a variety of reasons, including proprietary rights, lack of organizational time, capacity and training, and, in some cases, an unwillingness to share; this curtails our ability for appropriate decision-making.
 
-There are now initiatives that facilitate the data gathering process by assembling, organizing and sharing environmental knowledge, such as the Ocean Biogeographic Information System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for abiotic data.
+Current initiatives facilitate the data gathering process by assembling, organizing and sharing environmental knowledge, such as the Ocean Biogeographic Information System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@assis2018] for abiotic data.
 However, equivalent platforms for ~~environmental~~ drivers have largely focused on single drivers (e.g.
 Global Fishing Watch) and platforms collating data\textcolor{blue}{-based indicators} and knowledge on multiple drivers in a comparable and interoperable way remain conspicuously missing [but see @halpern2015].
-This is in spite of integrated management and assessment approaches needing efficient data reporting, standardized data management practices and tools tailored to the study of the effects of multiple drivers [@dafforn2016; @stock2018].
+This is in spite of integrated management and assessment approaches requiring efficient data reporting, standardized data management practices, and tools tailored to the study of the effects of multiple drivers [@dafforn2016; @stock2018].
 ~~An additional objective thus emerged in the process of addressing our initial goals: sharing information about the distribution and intensity of drivers of change in the St.Lawrence.~~
 
-Our main ~~objective~~ \textcolor{blue}{goal} in this study is to characterize the distribution and
-intensity of drivers in the St. Lawrence. More specifically, ~~we aim~~ \textcolor{blue}{our objectives are} to: 1)
+The main ~~objective~~ \textcolor{blue}{goal} of this study is to characterize the distribution and
+intensity of drivers in the St. Lawrence System. More specifically, ~~we aim~~ \textcolor{blue}{our objectives are} to: 1)
 identify areas of high cumulative exposure \textcolor{blue}{to drivers} and 2) characterize areas with similar cumulative exposure ~~regimes~~ \textcolor{blue}{profiles, \textit{i.e.} areas exposed to similar suites of co-occurring drivers}.
@@
 evolve with the addition of information and threats to the St-Lawrence ecosystems.~~
-\textcolor{blue}{We achieve this with the development of an open knowledge platform, \textit{eDrivers}, designed to facilitate the widespread availability of driver characterization for holistic assessments and management approaches.
+\textcolor{blue}{We achieve this through the development of an open knowledge platform, \textit{eDrivers}, that was designed to facilitate the widespread availability of driver characterization for holistic assessments and management approaches.
 Here, we present its guiding principles and accompanying tools.}
@@
 
-The St. Lawrence System is formed by the \textcolor{blue}{St. Lawrence} Estuary and the Gulf of St. Lawrence
-(Figure \ref{egsl}). The Estuary is defined by the seawater limit
-close to Îles d'Orléans to the west and by its connexion to the Gulf
+The St. Lawrence System is composed of the \textcolor{blue}{St. Lawrence} Estuary and the Gulf of St. Lawrence
+(Figure \ref{egsl}). The Estuary is defined by the limit of seawater intrusion,
+close to Îles d'Orléans, to the west and by its connection to the Gulf
 near Pointe-des-Monts.
-Estuarine circulation characterizes the Estuary. The surface layer is
-composed of a freshwater seaward outflow primarily from the Great Lakes Basin
-and through the St. Lawrence River, upstream of the Estuary. Atlantic waters are
-flowing mainly from the Cabot Strait landward through the Laurentian Channel
-(see below).
+The surface layer is
+composed of freshwater flowing seaward, primarily from the Great Lakes Basin
+through the St. Lawrence River.
+Atlantic waters intrude landwards at depth into the Gulf and Estuary from Cabot Strait, but as well as from the Strait of Belle Isle (see below).
 
-The Gulf is an interior sea connected to the Atlantic by the Cabot and Belle-Isle
-Straits to the south and north of Newfoundland, respectively.
 The topology of the Northern Gulf is characterized by three deep channels (250-500 m).
 The Laurentian Channel is the main channel connecting the Estuary to the
-Atlantic through the Cabot Strait. The Esquiman and Anticosti Channels are two
+Atlantic through Cabot Strait. The Esquiman and Anticosti channels are two
 secondary channels that branch off from the Laurentian Channel to the north
-towards the Arctic and the north of Anticosti Island, respectively.
-The Southern Gulf is characterized by the Magdalen Shallows, a vast area with an
-average depth of ~50 m. The Gulf is composed of a seasonal cold intermediate layer
-that separates the surface and deep layers. Seasonal sea ice affects the circulation
-of the St. Lawrence. Finally, three islands impact the physical dynamics of the
-St. Lawrence: the Anticosti Island to the north, the Magdalen Islands in the
+towards the Strait of Belle Isle and the Labrador and north of Anticosti Island, respectively.
+The Southern Gulf hosts the Magdalen Shallows, a vast area with an
+average depth of ~50 m. The water column in the Gulf and St. Lawrence Estuary includes a seasonal cold intermediate layer
+that separates the surface and deep layers. Seasonal sea ice affects circulation
+in the St. Lawrence. Finally, three islands impact the physical dynamics of the
+St. Lawrence: the Anticosti Island to the north, the îles de la Madeleine in the
 middle of the Magdalen Shallows and Prince Edward Island to the south. See
@@
 The St. Lawrence drains over 25% of global freshwater reserves through its
-connexion to the Great Lakes Basin, which is home to over 45 million
-North-Americans, *i.e.* 15 and 30 million in Canada and the United States,
-respectively. The St. Lawrence itself boasts a much lower
+connection to the Great Lakes Basin, which is home to over 45 million
+North Americans, *i.e.* 15 and 30 million in Canada and the United States,
+respectively. The coasts of St. Lawrence System, as delimited by our study area (Figure \ref{egsl}), boast a much lower
 population of approximately 1 million Canadians living within 10 km of the
@@
 
-\textcolor{blue}{Drivers, as broadly defined in this study, are data-based indicators of environmental conditions and human activities that are often referred to as driving forces, stressors, pressures or states in the scientific and environmental assessment litterature} [*e.g.* @kristensen2004; @halpern2019].
+\textcolor{blue}{Drivers, as broadly defined in this study, are data-based indicators of environmental conditions and human activities that are often referred to as driving forces, stressors, pressures, or states in the scientific and environmental assessment litterature} [*e.g.* @kristensen2004; @halpern2019].
 \textcolor{blue}{Defining such categories, however, can be difficult and is often context- and ecosystem-specific} [@gari2015; @dempsey2018].
 \textcolor{blue}{As such, we refrain from articulating our work around a specific framework or imposing categories on data-based products that may change with a user's objective.
-We rather focus on available data-based indicators that contribute to evaluating an ecosystem's cumulative exposure to multiple threats.}
+Instead, we focus on available data-based indicators that contribute to evaluate the ecosystem's cumulative exposure to multiple threats.}
 
@@
 localized anthropogenic activities (*e.g.* fisheries). Drivers were divided
-into 4 groups: coastal, climate, fisheries and marine traffic (Table 1). All
+into 4 groups: coastal, climate, fisheries, and marine traffic (Table 1). All
 data layers and methodologies are described in the supplementary materials.
@@
 2019, drivers with non-normal frequency distributions were log-transformed to avoid underestimating intermediate driver values.
-All drivers were also scaled between 0 and 1 to allow driver comparisons.
-The 99th quantile of individual driver distribution was used as the upper bound for scaling to control for extreme values that may or may not be real observations.}
+All drivers were scaled between 0 and 1 to allow comparisons.
+The 99th quantile of individual driver distribution was used as the upper limit for scaling to control for extreme values that may or may not represent real observations.}
 ~~All drivers were embedded in a regular grid composed of 245604 $1 km^2$
@@
 \textcolor{blue}{The St. Lawrence System was divided into a regular grid of
-1 $km^2$ cells in which all drivers were integrated (Figure S2)}
+1 $km^2$ cells into which all drivers were integrated (Figure S2)}
 
@@
 ~~As an introduction to indicators of cumulative exposure
-(objective 1), we begin by presenting a simplifed 2-driver example focusing
+(objective 1), we begin by presenting a simplifed 2-driver example that focuses
 on hypoxia and demersal destructive fisheries, two drivers known to occur
@@
 grid cell (objective 1).~~
-\textcolor{blue}{We begin by providing a simplifed 2-driver example focusing on
-the co-occurrence of hypoxia and demersal destructive fisheries, two drivers
-occurring mainly in deeper waters of the St. Lawrence. Driver co-ccurrence was
+\textcolor{blue}{We begin by providing a simplified 2-driver example that focuses on
+the co-occurrence of hypoxia and demersal destructive fisheries, two drivers that mostly
+occur in deeper St. Lawrence waters. Driver co-occurrence was
 evaluated spatially by summing the scaled intensity of drivers in each grid cell.
@@
 co-occurrence was evaluated and serves as a stepping stone to the
-integrative indicators used after, \textit{i.e.} cumulative exposure and
+integrative indicators used hereafter, \textit{i.e.} cumulative exposure and
 cumulative hotspots (objective 1).}
@@
 drivers in each grid cell:~~
-\textcolor{blue}{We evaluated the cumulative exposure ($E_C$) of each grid
+\textcolor{blue}{We evaluated cumulative exposure ($E_C$) for each grid
 cell as the sum of scaled driver intensities:}
@@
 
-where $x$ is a grid cell, $i$ is a driver and $D$ is the scaled intensity of
+where $x$ is a grid cell, $i$ is a driver, and $D$ is the scaled intensity of
 driver $i$. ~~The cumulative footprint provides an estimate of the total relative
@@
 
-\textcolor{blue}{We identified areas with similar cumulative exposure profiles using a clustering approach} [@bowler2019].
+\textcolor{blue}{We identified areas with similar cumulative exposure profiles (objective 2) using a clustering approach} [@bowler2019].
 ~~Threat complexes were identified using~~ \textcolor{blue}{We used} a partional *k-medoids* clustering
@@
 
-The difference between clusters was explored by measuring the total inter-cluster
+Differences between clusters were explored by measuring the total inter-cluster
 dissimilarity and the contribution of each driver to the total inter-cluster
@@
 run the analysis and repeating the process over 50 iterations. We only used the
-bootstrapping procedure for clusters with less than 40000 observations since
+bootstrap procedure for clusters with less than 40000 observations since
 computation time was manageable.
@@
 destructive fisheries are located along the Laurentian Channel, the heads of the
-Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B).
+Anticosti and Esquiman Channels and around the îles de la Madeleine (Figure \ref{kernel}B).
 By combining both drivers, we can observe that hypoxia and demersal destructive
@@
 \textcolor{blue}{We first present the simplified hypoxia-fisheries example to demonstrate how driver co-occurrence was evaluated.
-Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels, with the head of the Channels most exposed (Figure \ref{kernel}A).
-Demersal destructive fisheries are located along the Laurentian Channel, the heads of the Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B).
-By combining both drivers, we observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C); these are the areas where we might expect interactions between these drivers to be more likely}.
+Hypoxic bottom waters area mainly found at the head of the Laurentian, Anticosti and Esquiman channels (Figure \ref{kernel}A).
+Demersal destructive fisheries are concentrated along the Laurentian Channel, the heads of the Anticosti and Esquiman channels and around the îles de la Madeleine (Figure \ref{kernel}B).
+By combining both drivers, we observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti channels (Figure \ref{kernel}C); these are the areas where we might expect interactions between these drivers to be more likely}.
 
@@
 ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
-generally highest along the coast (Figure \ref{footprint}), with hotspots
+generally highest along coasts (Figure \ref{footprint}), with hotspots
 located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
@@
 Anticosti Gyre being notable exceptions (Figures \ref{footprint} and
-\ref{hotspot}). This is not to say that offshore areas are free of exposure,
+\ref{hotspot}). This is not to say that offshore areas are free from exposure,
 as most of the St. Lawrence is exposed to multiple overlapping drivers
 (Figures \ref{footprint} and \ref{hotspot}).
-For example, \textcolor{blue}{the heads of the Anticosti and Esquiman Channels are highly exposed to cumulative drivers} ~~it is worthy to note high cumulative exposure observed at the heads of the Anticosti and Esquiman
+For example, \textcolor{blue}{the heads of the Anticosti and Esquiman channels are highly exposed to cumulative drivers} ~~it is worthy to note high cumulative exposure observed at the heads of the Anticosti and Esquiman
 Channels~~ (Figure \ref{footprint}).
@@
 areas [e.g. @halpern2015a; @feist2016; @mach2017; @stock2018]. These are areas
-where human activities (*e.g.* coastal development and shipping) and footprint
-(*e.g.* pollution runoff) are the most intense [@feist2016], and on which is
+where human activities (*e.g.* coastal development and shipping) and footprints
+(*e.g.* pollution runoff) are most intense [@feist2016], and on which is
 overlaid a background of natural disturbances [@micheli2016]. They are also the
 areas in which the most dramatic increases in exposure are expected, with
-populations increasing more rapidly along the coast than inland [@feist2016].
+populations increasing more rapidly along coasts than inland [@feist2016].
 In the St. Lawrence, large coastal cities are mostly located along the Estuary
-and the southwestern Gulf, while the northeastern Gulf is largely uninhabited
+and the southwestern Gulf, whereas the northeastern Gulf is largely uninhabited
 or home to small coastal communities. Certain smaller coastal communities with
@@
 As for offshore exposure, the Estuary, along with the St. Lawrence River,
-provide access to and serve as the primary drainage outflow of the Great Lakes
-Basin, which is home to over 45 million North-Americans and is the most densely
+provide access to and serves as the primary drainage outflow of the Great Lakes
+Basin, which is home to over 45 million North Americans and is the most densely
 populated region in Canada [@statistics-canada2017]. Most marine
-traffic thus converges to the Estuary.
+traffic thus converges into the Estuary.
 
-While we cannot ascertain that high exposure areas are the most impacted, we
+Whereas we cannot ascertain that high exposure areas are the most impacted, we
 can safely predict that these are the areas where studying ecosystem state will
@@
 exposure (Figure S3). Areas with high exposure such as the Estuary, the
-Anticosti Gyre and the southwestern Gulf coastline (Figure \ref{footprint} and
+Anticosti Gyre, and the southwestern Gulf coastline (Figure \ref{footprint} and
 \ref{hotspot}) are thus areas where driver interactions are
-most likely, and where they can arise between a host of different drivers.
+most likely and where they can arise between a host of different drivers.
 ~~The identification of threat complexes~~ \textcolor{blue}{Identifying areas with similar cumulative exposure profiles} provides a crucial tool to simplify the multi-dimensional complexity of overlapping drivers ~~to areas exposed to similar suites of drivers~~ [@bowler2019].
@@
 moving through ~~threat complexes and exposed to the combined effects of all
-drivers typical to those areas~~ \textcolor{blue}{areas exposed to similar suites of drivers}.
+drivers inherent to those areas~~ \textcolor{blue}{areas exposed to similar suites of drivers}.
 
-Six distinct ~~threat complexes~~\textcolor{blue}{clusters} were identified in the St. Lawrence using
-the *k-medoids* algorithm (Figures S4, S5). Based on their
+Six distinct ~~threat complexes~~\textcolor{blue}{clusters} were identified in the St. Lawrence (Figures S4, S5). Based on their
 distribution and representative drivers, ~~threat complexes~~\textcolor{blue}{clusters} can be divided into
@@
 Coastal ~~threat complexes~~\textcolor{blue}{clusters} (1 to 3; Figure \ref{cluster}) include all types
-of drivers besides hypoxia; they are also the most exposed ~~threat complexes~~\textcolor{blue}{clusters}, both in
+of drivers other than hypoxia; they are also the most exposed ~~threat complexes~~\textcolor{blue}{clusters}, both in
 terms of driver overlap and intensity. ~~Threat complex~~\textcolor{blue}{Cluster} 1 encompasses the
@@
 \ref{hotspot}). This ~~complex~~\textcolor{blue}{cluster} is characterized by high intensities of land-based
-drivers (*e.g.* nutrient input), demersal non-destructive high-bycath fisheries
+drivers (*e.g.* nutrient input), demersal non-destructive high bycatch fisheries
 (*e.g.* trap fishing), climate drivers and marine traffic drivers in
@@
 characterized by high intensity climate and marine traffic drivers.
-~~Threat complex~~\textcolor{blue}{Cluster} 4 is differentiated by demersal non-destructive high-bycath
-fisheries, higher marine traffic drivers compared to ~~complex~~\textcolor{blue}{cluster} 5 and generally
+~~Threat complex~~\textcolor{blue}{Cluster} 4 is differentiated by demersal non-destructive high bycatch
+fisheries, higher marine traffic drivers compared to ~~complex~~\textcolor{blue}{cluster} 5, and generally
 corresponds to the whole Southern Gulf. ~~Threat complex~~\textcolor{blue}{Cluster} 5 is characterized by
-more fisheries types (*i.e.* demersal destructive and pelagic high-bycatch),
-generally lower intensity marine traffic drivers and is located almost
+more fisheries types (*i.e.* demersal destructive and pelagic high bycatch),
+generally lower intensity marine traffic drivers, and is located almost
 exclusively in the Northern Gulf. Finally, ~~threat complex~~\textcolor{blue}{cluster} 6 is the most exposed
@@
 It is characterized by high intensity hypoxia, marine traffic and
-pollution, as well as demersal destructive and pelagic high-bycatch fisheries.
+pollution, as well as demersal destructive and pelagic high bycatch fisheries.
 This ~~threat complex~~\textcolor{blue}{cluster} corresponds primarily to the Laurentian Channel and
-incorporates parts of the Esquiman and Anticosti Channels.
+incorporates parts of the Esquiman and Anticosti channels.
 
@@
 spatial dynamics in areas intersecting multiple ~~threat complexes~~\textcolor{blue}{clusters}. For example,
-~~threat complexes~~\textcolor{blue}{clusters} 3 and 6 meet at the mouth of the River Saguenay. This area
+~~threat complexes~~\textcolor{blue}{clusters} 3 and 6 meet at the mouth of the Saguenay River. This area
 is particularly dynamic, with deep Atlantic waters advected through estuarine
-circulation mixing with surface waters from the St. Lawrence River and the
-Saguenay River [@dufour2007]. This results in the convergence of climate
+circulation mixing with surface waters from the St. Lawrence and
+Saguenay rivers [@dufour2007]. This results in the convergence of climate
 drivers from the bottom of the Laurentian Channel and marine traffic drivers
@@
 Sharing the knowledge acquired through the description of drivers in the
-St. Lawrence emerged as a priority to curtail the need to reach dozens
+St. Lawrence emerged as a priority to curtail the need to contact dozens
 of experts across multiple organizations and over extensive periods of time to
-assemble the data needed to apply integrated research and management. It is also
+assemble the data needed for integrated research and management. It is also
 a requirement to ensure that this manuscript will not quickly become an outdated
@@
 
-***Why***: Operating over such large scales in time, space and subject matter
+***Why***: Operating over such large scales in time, space, and subject matter
 requires a vast and diverse expertise that cannot possibly be possessed by any
@@
 networking of organizations and data holders to meet their needs and those of
-users, to improve knowledge and to assist decision-making in areas such as
+users, to improve knowledge, and to assist decision-making in areas such as
 public safety, climate change, transportation, resources and biodiversity
@@
 We are also developing collaborations with the Portal on water knowledge[^melcc],
-an initiative from the Québec provincial government. This portal aims at
-collecting and sharing accurate, complete and updated resources on water and
+an initiative from the Québec provincial government. The aim of this portal is to collect and share accurate, complete, and up-to-date resources on water and
 aquatic ecosystems to support the mandate of relevant actors and stakeholders
@@
 \textcolor{blue}{As the FAIR Data Principles already exist as a unified set of
-principles, we adopt them as a set of subprinciples guiding our initiative.}
+principles, we adopt them as a set of guiding subprinciples to our initiative.}
 
@@
 geoserver. The functions available in *eDrivers* to access the data have however
-been developed to ensure forward compatibility once the data migrate to SLGO's
+been developed to ensure forward compatibility once the data are migrated to SLGO's
 geoserver.
@@
 in place and will be increasingly accessible as the initiative evolves and other
-tools are developped to ease user experience. We provide an example of  the ease
+tools are developed to ease user experience. We provide an example of  the ease
 with which the data can be accessed and used with the \textit{eDrivers} R package
@@
 multi-driver studies [@dafforn2016; @stock2018; @bowler2019]. In the St.
-Lawrence, we found that few areas are free of cumulative exposure and that the
-whole Estuary, the Anticosti Gyre and coastal southwestern Gulf are particularly
+Lawrence, we found that few areas are free from cumulative exposure and that the
+whole Estuary, the Anticosti Gyre, and coastal southwestern Gulf are particularly
 exposed to cumulative drivers, especially close to urban areas. We also
@@
 cumulative exposure ~~regimes~~\textcolor{blue}{profiles}; these ~~complexes~~ reveal that coastal areas are
-particularly exposed to all types of drivers and that ~~multiple~~\textcolor{blue}{certain} driver \textcolor{blue}{combinations are typical of certain regions of the St.
+particularly exposed to all types of drivers and that ~~multiple~~\textcolor{blue}{certain} driver \textcolor{blue}{combinations are inherent to certain regions of the St.
 Lawrence}~~typically co-occur in space~~.
@@
 Our focus has been on a single element required for fully operational impact assessments.
-Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators [@dube2001; @dube2003; @jones2016].
-These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers and fail to diagnose impacts on valued components or non-additive effects [@dube2001; @dube2003].
+Data that provide knowledge on the exposure of ecosystems to drivers are called stressor-based indicators [@dube2001; @dube2003].
+These indicators efficently identify potential local impacts and can be proactively linked to decision-making, yet assume complete knowledge of drivers and fail to diagnose impacts on valued components or non-additive effects.
 In contrast, effect-based indicators are direct measurements of valued components (\textit{e.g.} species abundance and biodiversity) and inherently capture the effects of multiple drivers [@dube2001; @dube2003].
-Whereas effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers [@dube2001; @dube2003].
-Stressor-based and effect-based indicators are, therefore, both required to diagnose causes of ecosystem change [@dube2003; @jones2016].
+Whereas effect-based indicators are considered superior to stressor-based indicators, they fail to ascribe observed effects to specific drivers.
+Stressor-based and effect-based indicators are, therefore, both required to diagnose causes of ecosystem change [@jones2016].
 As a collection of knowledge on stressor-based indicators, \textit{eDrivers} should be weaved with other, comparable, collections of knowledge describing valued ecosystem components that can be linked to drivers and allow for a better understanding of cumulative impacts.
 Ultimately, interdisciplinary collections of knowledge could be weaved together through social-ecological metanetworks analyses [@dee2017].
-\textcolor{blue}{In turn, these could be used in conceptual frameworks that help in establishing causal relationships between drivers and valued ecosystem components such as the DPSIR (Driving forces -- Pressure -- State -- Impact -- Response) framework} [@kristensen2004; @gari2015].
-\textcolor{blue}{Through such frameworks, data-based indicators provided through \textit{eDrivers} could be categorized as driving forces, pressures or states, depending on the objective and context of a study.}
+\textcolor{blue}{In turn, these could be used in conceptual frameworks to help to establish causal relationships between drivers and valued ecosystem components such as the DPSIR (Driving forces -- Pressure -- State -- Impact -- Response) framework} [@kristensen2004; @gari2015].
+\textcolor{blue}{Within such frameworks, data-based indicators provided through \textit{eDrivers} could be categorized as driving forces, pressures or states, depending on the objective and context of a study.}
 
@@
 terrestrial and marine realms, considering coastlines as an impermeable
-barrier. While there is a sensible rationale for this division, we must strive
+barrier. Whereas there is a sensible rationale for this division, we must strive
 to eliminate it if we are to appropriately study and predict the
@@
 Lawrence in Eastern Canada.}\textcolor{blue}{Description of the St.
-Lawrence System in Eastern Canada, formed by the St.
+Lawrence System in Eastern Canada, composed of the St.
 Lawrence Estuary and the Gulf of St.
 Lawrence.
-The Estuary is defined by the seawater limit close to Îles d'Orléans to the west and by its connexion to the Gulf near Pointe-des-Monts.
-The Gulf is an interior sea connected to the Atlantic by the Cabot and Belle-Isle Straits to the south and north of Newfoundland, respectively.}}
+The Estuary is defined by the limit of seawater intrusion, close to Îles d'Orléans, to the west and by its connection to the Gulf near Pointe-des-Monts.
+The Gulf is an interior sea connected to the Atlantic by Cabot Strait and the Strait of Belle Isle to the south and north of Newfoundland, respectively.}}
 \label{egsl}

### 2020-05-01 Code/18-egsl.R

@@ -55,7 +55,7 @@ legendEGSL(range = c(0, 525),
            cexSub = .65,
            n = 6)
 
-colText <- '#717171'
+colText <- '#4b4b4b'
 
 city <- data.frame(latitude = numeric(0),
                    longitude = numeric(0),
@@ -116,8 +116,8 @@ text(y = 49.85, x = -62.25, labels = 'Anticosti Channel', srt = -15,col = colTex
 text(y = 49.85, x = -59, labels = 'Esquiman Channel', srt = 45,col = colText, cex = .8)
 text(y = 47.45, x = -63.25, labels = 'Magdalen\nShallows', srt = 0,col = colText, cex = .8, adj = c(.5,.5))
 text(y = 47.55, x = -60.1 , labels = 'Cabot\nStrait', srt = 0,col = colText, cex = .8, adj = c(.5,.5))
-text(y = 51.75, x = -56.25, labels = 'Belle-Isle\nStrait', srt = 0,col = colText, cex = .8, adj = c(.5,.5))
-text(y = 51.75, x = -56.25, labels = 'Belle-Isle\nStrait', srt = 0,col = colText, cex = .8, adj = c(.5,.5))
+text(y = 51.75, x = -56.25, labels = 'Strait of\nBelle Isle', srt = 0,col = colText, cex = .8, adj = c(.5,.5))
+text(y = 51.75, x = -56.25, labels = 'Strait of\nBelle Isle', srt = 0,col = colText, cex = .8, adj = c(.5,.5))
 text(y = 46.6, x = -59.8, 'Prince Edward\nIsland', srt = 0,col = colText, cex = 1, adj = c(0,.5), font = 2)
 text(y = 48.6, x = -58, 'Newfoundland', srt = 0,col = colText, cex = 1, adj = c(0,.5), font = 2)
 text(y = 46.4, x = -67.75, 'New-Brunswick', srt = 0,col = colText, cex = 1, adj = c(0,.5), font = 2)

### 2020-06-01 Code/0-Param.R
restructure whole repo and make it reproducible

@@ -1,0 +1,65 @@
+# Data
+load('./Data/drivers.RData')
+load('./Data/Spatial/egslGrid.RData')
+load('./Data/Spatial/egslSimple.RData')
+dr <- drivers[, -1]
+
+# Colors
+colsPalette <- c("#c7cbce", "#96a3a3", "#687677", "#222d3d", "#25364a", "#c77f20","#e69831", "#e3af16", "#e4be29", "#f2ea8b")
+pal <- colorRampPalette(colsPalette)
+rbPal <- colorRampPalette(c('#f4f4f4',colsPalette))
+focus  <- '#25364A'
+off    <- '#f2f2f2'
+uni <- colsPalette[7]
+gg_color_hue <- function(n) {
+  hues = seq(15, 375, length = n + 1)
+  hcl(h = hues, l = 65, c = 100, alpha = .5)[1:n]
+}
+
+# Extent
+library(sf)
+extFig <- as.numeric(st_bbox(egslSimple)) + c(-1.5,-.5,.5,.5)
+
+# Projections
+world <- '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs'
+
+# Parameters
+nDr <- ncol(dr) # Number of drivers
+
+# Driver names data.frame
+drNames <- data.frame(var = character(),
+                      full = character(),
+                      accr = character(),
+                      group = character(),
+                      stringsAsFactors = F)
+
+drNames[1, ] <- c("Acidification","Acidification","ACID","Cl")
+drNames[2, ] <- c("aquacultureInv","Aquaculture","AQUA","C")
+drNames[3, ] <- c("coastDev","Coastal development","CD","C")
+drNames[4, ] <- c("dirHumImpact","Direct human impact","DHI","C")
+drNames[5, ] <- c("fisheriesDD","Demersal destructive fisheries","DD","F")
+drNames[6, ] <- c("fisheriesDNH","Demersal, non-destructive, high-bycatch fisheries","DNH","F")
+drNames[7, ] <- c("fisheriesDNL","Demersal, non-destructive, low-bycatch fisheries","DNL","F")
+drNames[8, ] <- c("fisheriesPHB","Pelagic, high-bycatch fisheries","PHB","F")
+drNames[9, ] <- c("fisheriesPLB","Pelagic, low-bycatch fisheries","PLB","F")
+drNames[10, ] <- c("Hypoxia","Hypoxia","HYP","Cl")
[…]

### 2020-06-01 Code/1-data.R

@@ -1,18 +1,0 @@
-# This script is simply to load a version of the drivers and spatial objects data that will not
-# change through time, as will be the case with the data accessed through the
-# package. This will ensure that all the analyses and figures can be replicated.
-# Also, I transform the spatial objects to decimal degrees for the figures, as requested by co-authors
-library(slmeta)
-data(drivers, package = 'slmeta')
-save(drivers, file = './data/drivers.RData')
-
-data(driversRaw)
-save(driversRaw, file = './data/driversRaw.RData')
-
-data(egslGrid, package = 'slmeta')
-egslGrid <- st_transform(egslGrid, crs = 4326)
-save(egslGrid, file = './data/egslGrid.RData')
-
-data(egslSimple, package = 'slmeta')
-egslSimple <- st_transform(egslSimple, crs = 4326)
-save(egslSimple, file = './data/egslSimple.RData')

### 2020-06-01 Code/10-validateCluster.R

@@ -1,65 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-k <- 2:10
-
-# ~~~~~~~~~~~~~~~~~~~ Data ~~~~~~~~~~~~~~~~~~~ #
-load('./data/aswMedoid.RData')
-load('./data/wssMedoid.RData')
-load('./data/aswMean2.RData')
-load('./data/wssMean.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ Figure ~~~~~~~~~~~~~~~~~~~ #
-png('./figures/validation.png', width = 1280, height = 920, res = 200, pointsize = 7)
-layout(matrix(1:4, ncol = 2))
-par(mar = c(5,5,2,1))
-# k-medoid
-  # Silhouette
-  plot(aswMedoid ~ k,
-       pch = 20,
-       ylab = 'Average Silhouette Width',
-       type = 'b',
-       frame = F,
-       xlab = '',
-       xlim = c(0,10),
-       cex = 1.5,
-       cex.axis = .8)
-  mtext(side = 3, 'k-medoids clustering', font = 2, line = .75)
-
-  # WSS
-  plot(x = c(1,k),
-       y = wssMedoid,
-       type="b",
-       frame = F,
-       xlab="Number of Clusters (k)",
-       ylab="Total within-clusters sum of squares",
-       pch = 20,
-       xlim = c(0,10),
-       cex = 1.5,
-       cex.axis = .8)
-
-# k-means
-  # Silhouette
-  plot(aswMean2 ~ k,
-       pch = 20,
-       frame = F,
-       xlim = c(0,10),
-       xlab = '',
-       ylab = 'Average Silhouette Width',
-       type = 'b',
-       cex = 1.5,
-       cex.axis = .8)
-  mtext(side = 3, 'k-means clustering', font = 2, line = .75)
-
-  # WSS
-  plot(x = c(1,k),
-       y = wssMean,
-       type="b",
-       frame = F,
-       xlim = c(0,10),
-       xlab="Number of Clusters (k)",
-       ylab="Total within-clusters sum of squares",
-       pch = 20,
-       cex = 1.5,
-       cex.axis = .8)
-dev.off()

### 2020-06-01 Code/10-validatekmedoid.R

@@ -1,38 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-k <- 2:10
-
-# ~~~~~~~~~~~~~~~~~~~ Data ~~~~~~~~~~~~~~~~~~~ #
-load('./data/aswMedoid.RData')
-load('./data/wssMedoid.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ Figure ~~~~~~~~~~~~~~~~~~~ #
-png('./figures/validation.png', width = 920, height = 920, res = 200, pointsize = 7)
-layout(matrix(1:2, ncol = 1))
-par(mar = c(5,5,2,1))
-# k-medoid
-  # Silhouette
-  plot(aswMedoid ~ k,
-       pch = 20,
-       ylab = 'Average Silhouette Width',
-       type = 'b',
-       frame = F,
-       xlab = '',
-       xlim = c(0,10),
-       cex = 1.5,
-       cex.axis = .8)
-  mtext(side = 3, 'k-medoids clustering', font = 2, line = .75)
-
-  # WSS
-  plot(x = c(1,k),
-       y = wssMedoid,
-       type="b",
-       frame = F,
-       xlab="Number of Clusters (k)",
-       ylab="Total within-clusters sum of squares",
-       pch = 20,
-       xlim = c(0,10),
-       cex = 1.5,
-       cex.axis = .8)
-dev.off()

### 2020-06-01 Code/11-selectCluster.R

@@ -1,17 +1,0 @@
-# Simple script to select the number of clusters in light of the
-# clustering analyses and the validation, and export the simplified data
-# Number of clusters
-k <- 6
-
-# Load full datasets
-load('./data/kmedoid.RData')
-load('./data/kmeans.RData')
-
-# Object with only selected clustering results
-clMed <- kmedoid[[k-1]]$clustering
-clMean <- kMeans[[k-1]]$cluster
-
-# Export simplified dataset
-save(k, file = './data/k.RData')
-save(clMed, file = './data/clMed.RData')
-save(clMean, file = './data/clMean.RData')

### 2020-06-01 Code/12-compareCluster.R

@@ -1,66 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-load('./data/k.RData')
-load('./data/clMed.RData')
-load('./data/clMean.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ METHOD COMPARISON ~~~~~~~~~~~~~~~~~~~ #
-# Make clusters similar
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-# may need to be redone depending on the analyses results
-# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-id1 <- clMean == 1
-id2 <- clMean == 2
-id3 <- clMean == 3
-id4 <- clMean == 4
-id5 <- clMean == 5
-id6 <- clMean == 6
-clMean[id1] <- 1
-clMean[id2] <- 6
-clMean[id3] <- 5
-clMean[id4] <- 3
-clMean[id5] <- 2
-clMean[id6] <- 4
-
-# Differences
-cl12 <- as.numeric(clMed == clMean) + 1
-corr <- round((sum(clMed == clMean)/length(clMed)*100))
-
-png('./figures/clusterComp.png', width = 2560, height = 1840, res = 300, pointsize = 6)
-par(mfrow = c(2,2))
-
-# K-medoids clustering
-cols <- pal(k)[clMed]
-plotEGSL('egslSimple', prj = slmetaPrj('world'), extent = 'egslSimple')
-plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
-# plot(st_geometry(egslSimple), col = 'transparent', border = '#25364A', lwd = .75, add = T)
-text(x = mean(par('usr')[1:2]), y = par('usr')[4] - diff(par('usr')[3:4])*.1, 'k-medoids clustering', cex = 2, font = 2)
-
-# K-means clustering
-cols <- pal(k)[clMean]
-plotEGSL('egslSimple', prj = slmetaPrj('world'), extent = 'egslSimple')
-plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
[…]

### 2020-06-01 Code/13-cluster.R

@@ -1,77 +1,0 @@
-# source('Code/13-cluster.R')
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-load('./data/k.RData')
-load('./data/clMed.RData')
-nCl <- length(unique(clMed))
-load('./data/simperMean.RData')
-source('./code/barPlot.R')
-
-
-# Driver intensity in compared clusters
-drInt1 <- lapply(simperMean, function(x) x[,'ava']) %>%
-          as.data.frame()
-drInt2 <- lapply(simperMean, function(x) x[,'avb']) %>%
-          as.data.frame()
-
-dat <- vector('list', nCl)
-dat[[1]] = data.frame(x = drInt1$X1_2)
-dat[[2]] = data.frame(x = drInt1$X2_3)
-dat[[3]] = data.frame(x = drInt1$X3_4)
-dat[[4]] = data.frame(x = drInt1$X4_5)
-dat[[5]] = data.frame(x = drInt1$X5_6)
-dat[[6]] = data.frame(x = drInt2$X5_6)
-
-
-# ~~~~~~~~~~~~~~~~~~~ CLUSTER ~~~~~~~~~~~~~~~~~~~ #
-# png('./figures/cluster.png', width = 1280, height = 1280, res = 300, pointsize = 6)
-jpeg('./figures/cluster.jpg', width = 1280, height = 1280, res = 300, pointsize = 6)
-mat <- matrix(c(1,1,1,1,1,1,2,3,4,5,6,7), nrow = 2, byrow = T)
-layout(mat, heights = c(1, .4))
-cols <- pal(nCl)[clMed]
-plotEGSL(layers     = c('egslSimple', 'canada','usa'),
-         cols       = c('#00000000',off,off),
-         borders    = c('#00000000',focus,focus),
-         lwds       = c(.75,.75,.75),
-         mar        = c(2,2,2,2),
-         box        = T,
-         axes       = 1:4,
-         northArrow = F,
-         prj        = slmetaPrj('world'),
-         extent     = extFig)
[…]

### 2020-06-01 Code/14-interDissimilarity.R

@@ -1,73 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-load('./data/k.RData')
-load('./data/clMed.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
-source('./code/bootstrapSimper.R')
-source('./code/simperMan.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ SIMPER ~~~~~~~~~~~~~~~~~~~ #
-# Run iterative simper analysis
-iter <- 300
-prop <- .05
-system.time(bootSimper <- bootstrapSimper(dr, clMed, iter = iter, prop = prop))
-
-# # This is to run it multiple times simultaneously
-# bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '1')
-# bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '2')
-# bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '3')
-load('./data/bootSimper1.RData')
-x1 <- bootSimper
-load('./data/bootSimper2.RData')
-x2 <- bootSimper
-load('./data/bootSimper3.RData')
-x3 <- bootSimper
-bootSimper <- c(x1,x2,x3)
-save(bootSimper, file = './data/bootSimper.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ SIMPER SUMMARY ~~~~~~~~~~~~~~~~~~~ #
-# Summarize in array format
-iter <- length(bootSimper) # To delete
-nList <- length(bootSimper[[1]])
-
-# Name of variables to extract from simper analyses
-varNames <- c('contr','sd','ava','avb')
-
-# Empty list to store summaries
-simperSummary <- vector('list', nList)
-names(simperSummary) <- names(bootSimper[[1]])
-
-# Empty arrays to store inter-cluster dissimilarity results
-for(i in 1:nList) {
-  simperSummary[[i]] <- array(data = 0,
-                              dim = c(ncol(dr), length(varNames), iter),
-                              dimnames = list(drNames$accr, varNames))
-}
-
-# Extract summary for simper analyses interations
-for(i in 1:iter) {
[…]

### 2020-06-01 Code/15-intraSimilarity.R

@@ -1,57 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-load('./data/k.RData')
-load('./data/clMed.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
-source('./code/intraSim.R')
-
-# ~~~~~~~~~~~~~~~~~~~ INTRA-CLUSTER SIMILARITY ~~~~~~~~~~~~~~~~~~~ #
-# As with the simper analysis, the dataset is mostly too large to use
-# regular function in packages available in R. Also, while the software
-# PRIMER automatically measures the intra-cluster similarity using the
-# Bray-Curtis similarity, `vegan::simper` does not do it.
-# I therefore create a custom function that allows me to extract the
-# intra-cluster Bray-Curtis similarity and that does it using a bootstrap
-# procedure when the cluster has too many observations.
-#
-sim <- vector('list', k)
-for(i in 1:k) {
-  cat("   k: ", i, "\r")
-  id <- clMed == i
-  sim[[i]] <- bootIntraSim(dr[id, ], iter = 50, samp = .25)
-  save(sim, file = './data/intraSimilarity.RData')
-}
-
-# Extract species, contribution and sd
-# Summarize in array format
-iter <- length(sim[[4]])
-varNames <- c('contr')
-simSummary <- vector('list', k)
-for(i in 1:k) simSummary[[i]] <- array(data = 0, dim = c(ncol(dr), 1, iter), dimnames = list(drNames$accr, varNames))
-
-for(i in 1:k) {
-  # Check if there are multiple iterations or not.
-  # If length(sim[[i]]) == 2 it means that there were no iterations and the whole cluster was parsed through the simper2 function
-  # If not, then sim[[i]] should be equal to the number of iterations
-  if(i %in% c(1,2,3)) {
-    simSummary[[i]] <- matrix(ncol = 1,
-                              data = sim[[i]]$contr,
-                              dimnames = list(drNames$accr, 'contr'))
-  } else {
-    for(j in 1:iter) {
-      simSummary[[i]][,'contr',j] <- sim[[i]][[j]]$contr
-    }
-  }
-}
-
[…]

### 2020-06-01 Code/16-contributionDissimilarity.R

@@ -1,151 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# ~~~~~~~~~~~~~~~~~~~ EXTRA LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
-library(gridExtra)
-library(ggplot2)
-library(ggplotify)
-library(grid)
-
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-load('./data/k.RData')
-load('./data/clMed.RData')
-load('./data/simperMean.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
-source('./code/circularPlot.R')
-source('./code/butterfly.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ INDIVIDUAL CONTRIBUTION TO INTER-CLUSTER DISSIMILARITY ~~~~~~~~~~~~~~~~~~~ #
-# Driver percent contribution to dissimilarity
-contDiss <- lapply(simperMean, function(x) x[,'contr'] / sum(x[,'contr'])) %>%
-            as.data.frame()
-colnames(contDiss) <- names(simperMean)
-
-# Total dissimilarity
-totDiss <- unlist(lapply(simperMean, function(x) sum(x[, 'contr'])))
-
-# Driver intensity in compared clusters
-drInt1 <- lapply(simperMean, function(x) x[,'ava']) %>%
-          as.data.frame()
-drInt2 <- lapply(simperMean, function(x) x[,'avb']) %>%
-          as.data.frame()
-
-# Mean value in each cluster
-totMean <- numeric(k)
-for(i in 1:k) {
-  id <- clMed == i
-  totMean[i] <- mean(as.matrix(dr[id, ]))
-}
-
-# Fish
-fish <- matrix(ncol = 5, nrow = 7, dimnames = list(c(), drNames$var[5:9]))
-for(i in 1:k) {
-  id <- clMed == i
-  for(j in colnames(fish)) {
-    fish[i, j]<- sum((dr[id, j]))
-  }
-}
-
-fish <- apply(fish, 2, function(x) round(x / sum(x) * 100))
-
-# Set figure layout view
[…]

### 2020-06-01 Code/17-contributionSimilarity.R

@@ -1,112 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# ~~~~~~~~~~~~~~~~~~~ EXTRA LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
-library(gridExtra)
-library(ggplot2)
-library(ggplotify)
-library(grid)
-
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-load('./data/k.RData')
-load('./data/clMed.RData')
-load('./data/similarityMean.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
-source('./code/circularPlot.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ SIMILARITY CONTRIBUTION ~~~~~~~~~~~~~~~~~~~ #
-# Driver percent contribution to dissimilarity
-contSim <- lapply(similarityMean, function(x) x[,'contr'] / sum(x[,'contr'])) %>%
-            as.data.frame()
-colnames(contSim) <- names(similarityMean)
-
-# Total similarity
-totSim <- lapply(similarityMean, function(x) sum(x[,'contr'])) %>%
-          as.data.frame()
-
-# Circulat plots
-p1 <- vector('list', ncol(contSim))
-for(i in 1:length(contSim)) {
-  temp <- data.frame(individual = drNames$accr,
-                     group = drNames$group,
-                     value = round(contSim[, i]*100))
-  rownames(temp) <- NULL
-  temp2 <- round(totSim[i],2)
-  p1[[i]] <- circularPlot(temp, temp2)
-}
-
-# Individual clusters
-egsl <- st_centroid(egslGrid)
-p2 <- vector('list', k)
-for(i in 1:length(p2)) {
-  # Identify observations in cluster i
-  id <- clMed == i
-
-  # Select only grid cells in cluster i
-  temp <- egsl %>%
-          filter(clMed == i) %>%
-          st_coordinates() %>%
-          as.data.frame()
-          ext <- st_bbox(egsl)
-
-  p2[[i]] <- ggplot() +
-             geom_sf(data = egslSimple, fill = NA) +
-             geom_point(aes(x = X, y = Y), colour = '#187962', size = .01, data = temp) +
[…]

### 2020-06-01 Code/18-egsl.R

@@ -1,130 +1,0 @@
-# source('./Code/18-egsl.R')
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-load('./data/bathy.RData') # Bathymetry
-data(canada)
-data(usa)
-canada <- st_transform(canada, crs = 4326)
-usa <- st_transform(usa, crs = 4326)
-bathy <- st_transform(bathy, crs = 4326)
-deph <- bathy$DEPH
-
-
-library(raster)
-r <- raster(ncol=180, nrow=180)
-extent(r) <- extent(bathy)
-r <- fasterize::fasterize(bathy, r, field = 'DEPH')
-r <- mask(r, egslSimple)
-
-# ~~~~~~~~~~~~~~~~~~~ EGSL ~~~~~~~~~~~~~~~~~~~ #
-# Layout
-pal <- colorRampPalette(c('#E2FCF6', '#1B7CAA'))
-cols <- pal(101)[((deph/max(deph, na.rm = T))*100)+1]
-
-cexMain <- 1
-cexMain2 <- .75
-# png('./figures/egsl.png', width = 1280, height = 920, res = 300, pointsize = 6)
-jpeg('./figures/egsl.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
-par(bg = '#ffffff')
-plotEGSL(layers     = c('egslSimple', 'canada','usa'),
-         cols       = c('#00000000',off,off),
-         borders    = c('#00000000',focus,focus),
-         lwds       = c(.75,.75,.75),
-         mar        = c(2,2,2,2),
-         box        = T,
-         axes       = 1:4,
-         northArrow = F,
-         prj        = slmetaPrj('world'),
-         extent     = extFig)
-
-# plot(st_geometry(bathy), add = T, col = cols, border = cols, lwd = .1)
-# plot(st_geometry(bathy), add = T, col = cols, border = 'transparent', lwd = .1)
-image(r, col = pal(100), axes = F, ylab = '', xlab = '', add = T)
-plot(st_geometry(egslSimple), col = 'transparent', border = focus, lwd = .75, add = T)
-# plot(st_geometry(canada), col = off, border = focus, lwd = .75, add = T)
[…]

### 2020-06-01 Code/2-param.R

@@ -1,80 +1,0 @@
-# cd phdproj/platformms
-# Libraries
-library(graphicsutils)
-library(MASS)
-library(sf)
-library(magrittr)
-library(vegan)
-library(raster)
-library(cluster)
-library(slmeta)
-
-# Data
-load('./data/drivers.RData')
-load('./data/egslGrid.RData')
-load('./data/egslSimple.RData')
-dr <- drivers[, -1]
-
-# Parameters
-nDr <- ncol(dr) # Number of drivers
-
-# Driver names data.frame
-drNames <- data.frame(var = character(),
-                      full = character(),
-                      accr = character(),
-                      group = character(),
-                      stringsAsFactors = F)
-
-drNames[1, ] <- c("Acidification","Acidification","ACID","Cl")
-drNames[2, ] <- c("aquacultureInv","Aquaculture","AQUA","C")
-drNames[3, ] <- c("coastDev","Coastal development","CD","C")
-drNames[4, ] <- c("dirHumImpact","Direct human impact","DHI","C")
-drNames[5, ] <- c("fisheriesDD","Demersal destructive fisheries","DD","F")
-drNames[6, ] <- c("fisheriesDNH","Demersal, non-destructive, high-bycatch fisheries","DNH","F")
-drNames[7, ] <- c("fisheriesDNL","Demersal, non-destructive, low-bycatch fisheries","DNL","F")
-drNames[8, ] <- c("fisheriesPHB","Pelagic, high-bycatch fisheries","PHB","F")
-drNames[9, ] <- c("fisheriesPLB","Pelagic, low-bycatch fisheries","PLB","F")
-drNames[10, ] <- c("Hypoxia","Hypoxia","HYP","Cl")
-drNames[11, ] <- c("inorgPol","Inorganic pollution","IP","C")
-drNames[12, ] <- c("invasives", "Invasives species","INV","T")
-drNames[13, ] <- c("marinePol","Marine pollution","MP","T")
-drNames[14, ] <- c("nutrientInput","Nutrient input","NI","C")
-drNames[15, ] <- c("orgPol","Organic pollution","OP","C")
-drNames[16, ] <- c("negSBT","Negative sea bottom temperature anomalies","SBT-","Cl")
-drNames[17, ] <- c("posSBT","Positive sea bottom temperature anomalies","SBT+","Cl")
[…]

### 2020-06-01 Code/3-drivers.R

@@ -1,53 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-load('./data/driversRaw.RData')
-drRaw <- driversRaw[, -1]
-
-# ~~~~~~~~~~~~~~~~~~~ HISTOGRAMS ~~~~~~~~~~~~~~~~~~~ #
-# Value for `par` parameters for figures
-parVal <- c(ceiling(nDr / 4), 4)
-
-# Let's take a look with 0 removed from raw data
-# This is especially important for the coastal drivers, as including zeros
-# would include offshore cells, which vastly outnumber coastal cells and
-# to which coastal drivers to not extend
-png('./figures/drHist.png', width = 1280, height = 1500, res = 200)
-par(mfrow = parVal, mar = c(4,4,1,1))
-for(i in 1:nDr) {
-  id <- drRaw[, i] > 0
-  drID <- drNames$accr[drNames$var == colnames(drRaw)[i]]
-  hist(drRaw[id, i], prob = T, main = '', xlab = drID, col = focus)
-  lines(density(drRaw[id, i], from = 0, to = 1), col = uni, lwd=2)
-}
-dev.off()
-
-
-# ~~~~~~~~~~~~~~~~~~~ DRIVERS ~~~~~~~~~~~~~~~~~~~ #
-png('./figures/drivers.png', width = 1280, height = 1500, res = 200)
-par(mfrow = parVal, mar = c(4,4,1,1))
-# Drivers
-for(i in 1:nDr) {
-  drData <- dr[, i]
-  drID <- drNames$accr[drNames$var == colnames(dr)[i]]
-  id <- drData > 0
-  cols <- pal(101)[((drData[id]/max(drData[id]))*100)+1]
-  plotEGSL('egslSimple', prj = slmetaPrj('world'), extent = 'egslSimple')
-  plot(st_geometry(egslGrid[id, ]), add = T, col = cols, border = cols, lwd = .1)
-  text(x = mean(par('usr')[1:2]), y = par('usr')[4]-.5, labels = drID)
-}
-
-# Scale
-colRamp <- pal(100)
-scale = (length(colRamp) - 1)/(1 - 0)
-ticks <- seq(0, 1, len = 11)
-par(mar = c(.75,0,.75,11))
[…]

### 2020-06-01 Code/4-footprint.R

@@ -1,44 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-Fd <- rowSums(dr)
-
-
-# ~~~~~~~~~~~~~~~~~~~ FOOTPRINT ~~~~~~~~~~~~~~~~~~~ #
-# Layout
-# Figure 2
-cols <- pal(101)[((Fd/max(Fd, na.rm = T))*100)+1]
-cexMain <- 1
-cexMain2 <- .75
-# png('./figures/footprint.png', width = 1280, height = 920, res = 300, pointsize = 6)
-jpeg('./figures/footprint.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
-# par(bg = colBack)
-plotEGSL(layers     = c('egslSimple', 'canada','usa'),
-         cols       = c('#00000000',off,off),
-         borders    = c('#00000000',focus,focus),
-         lwds       = c(.75,.75,.75),
-         mar        = c(2,2,2,2),
-         box        = T,
-         axes       = 1:4,
-         northArrow = F,
-         prj        = slmetaPrj('world'),
-         extent     = extFig)
-plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
-plot(st_geometry(egslSimple),
-     col = 'transparent',
-     border = focus,
-     lwd = .75,
-     add = T)
-
-# Legend
-legendEGSL(range = c(min(Fd), max(Fd)),
-           mainTitle = 'Cumulative exposure',
-           subTitle = '',
-           cexMain = .9,
-           cexSub = .65,
-           n = 6)
-
-# City names
-cityEGSL(prj = 4326)
-dev.off()

### 2020-06-01 Code/5-marimekko.R

@@ -1,166 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-
-# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
-Fd <- rowSums(dr)
-
-# Create vector of bins
-nBins <- 512 # Just because
-binData <- seq(0,max(Fd), length.out = 513)
-bins <- data.frame(lower = binData[1:512],
-                   upper = binData[2:513])
-bins$mid <- (bins$upper + bins$lower) / 2
-
-# Sum of fd values within each bin
-bins$fdTot <- numeric(nBins)
-for(i in 1:nBins) {
-  id <- Fd > bins$lower[i] & Fd <= bins$upper[i]
-  bins$fdTot[i] <- sum(Fd[id])
-}
-
-# Fd values associated with each driver
-binDr <- matrix(nrow = nrow(bins),
-                ncol = ncol(dr),
-                data = 0,
-                dimnames = list(c(), colnames(dr)))
-
-for(j in 1:ncol(binDr)) {
-  for(i in 1:nBins) {
-    id <- Fd > bins$lower[i] & Fd <= bins$upper[i]
-    binDr[i, j] <- sum(dr[id, j])
-  }
-}
-
-# Transform into a marimekko graph
-# Begin by transforming the data into proportions of total Fd
-binDrProp <- binDr / bins$fdTot
-
-# Sort drivers by group
-sortNames <- as.factor(drNames$group) %>%
-             order() %>%
-             drNames[., ]
-sortNames$group <- as.factor(sortNames$group)
-
-# Add colors
-trans <- c('FF','DD','BB','99','77','55','33','11')
-sortNames$col <- gg_color_hue(nGroup)[as.numeric(sortNames$group)]
-for(i in levels(sortNames$group)) {
-  id <- which(sortNames$group == i)
-  nId <- length(id)
-  for(j in 1:nId) {
-    sortNames$col[id[j]] <- paste0(substr(sortNames$col[id[j]],1,7), trans[j])
-  }
-}
-
-# Sort by group
-binDrProp <- binDrProp[, sortNames$var]
-
-# # Sort drivers according to extent
[…]

### 2020-06-01 Code/6-hotspot.R

@@ -1,64 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-# Quantile
-# 80th quantile
-# These will be transformed as binary values
-# Then sum the drivers
-# Then look at distribution
-
-# Empty matrix to store results
-hot <- matrix(ncol = ncol(dr),
-              nrow = nrow(dr),
-              data = NA,
-              dimnames = list(c(), drNames$var))
-
-# Evaluate hotspots
-for(i in 1:nDr) {
-  # ID = 0
-  id0 <- dr[, i] > 0
-  # 80th quantile threshold
-  th <- quantile(dr[id0, i], .8)
-  # Values over hotspot threshold
-  hot[, i] <- dr[, i] > th
-}
-
-# Cumulative hotspots
-cumulHotspot <- rowSums(hot)
-cumulHotspot[cumulHotspot == 0] <- NA
-nOver <- max(cumulHotspot, na.rm = T)
-
-
-# Graph
-png('./figures/hotspot.png', width = 1280, height = 920, res = 300, pointsize = 6)
-# jpeg('./figures/hotspot.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
-cols <- pal(nOver)[cumulHotspot]
-plotEGSL(layers     = c('egslSimple', 'canada','usa'),
-         cols       = c('#00000000',off,off),
-         borders    = c('#00000000',focus,focus),
-         lwds       = c(.75,.75,.75),
-         mar        = c(2,2,2,2),
-         box        = T,
-         axes       = 1:4,
-         northArrow = F,
-         prj        = slmetaPrj('world'),
-         extent     = extFig)
-plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
-plot(st_geometry(egslSimple),
-     col = 'transparent',
-     border = focus,
-     lwd = .75,
-     add = T)
-
-# Legend
-legendEGSL(range = c(1, nOver),
-           mainTitle = 'Number of drivers',
-           subTitle = expression(80^th~quantile),
-           cexMain = .9,
-           cexSub = .65,
-           n = 6)
-
-# City names
-cityEGSL(prj = 4326)
-dev.off()

### 2020-06-01 Code/7-kernel.R

@@ -1,143 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# -----
-# Figure
-png('./figures/kernel.png', width = 1920, height = 1280, res = 300, pointsize = 6)
-# jpeg('./figures/footprint.jpg', width = 1920, height = 1280, res = 300, pointsize = 6)
-
-
-# Layout
-mat <- matrix(0, 28, 32)
-mat[1:12, 1:16] <- 1 # traffic
-mat[15:26, 1:16] <- 2 # demersal
-mat[1:12, 17:32] <- 3 # traffic + demersal
-mat[18:28, 17:29] <- 4 # kernel
-mat[15:17, 17:29] <- 5 # dens traffic (x)
-mat[18:28, 30:32] <- 6 # dens demersal (y)
-mat[13:14, 6:11] <- 7 # scale traffic
-mat[27:28, 6:11] <- 8 # scale demersal
-mat[13:14, 22:27] <- 9 # scale combine
-mat[1:2, 15:16] <- 10 # panel A
-mat[15:16, 15:16] <- 11 # panel B
-mat[1:2, 31:32] <- 12 # panel C
-mat[15:16, 31:32] <- 13 # panel D
-layout(mat)
-# layout.show(13)
-
-# Plot 1 - Hypoxia
-  cols <- rbPal(101)[(drivers$Hypoxia * 100)+1]
-  plotEGSL(layers = 'egslSimple', borders = '#C7CBCE66', prj = slmetaPrj('world'), extent = 'egslSimple')
-  plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
-  plot(st_geometry(egslSimple), border = '#00000077', lwd = 0.25, add = T)
-
-# Plot 2 - Demersal destructive
-  cols <- rbPal(101)[(drivers$fisheriesDD * 100)+1]
-  plotEGSL(layers = 'egslSimple', borders = '#C7CBCE66', prj = slmetaPrj('world'), extent = 'egslSimple')
-  plot(st_geometry(egslGrid), add = T, col = cols, border = cols, lwd = .1)
-  plot(st_geometry(egslSimple), border = '#00000077', lwd = 0.25, add = T)
-
-# Plot 3 - Added layers - hypoxia + demersal destructive
-   drivers$combine <- drivers$Hypoxia + drivers$fisheriesDD
[…]

### 2020-06-01 Code/8-kmedoids.R

@@ -1,49 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# ~~~~~~~~~~~~~~~~~~~ Data prep ~~~~~~~~~~~~~~~~~~~ #
-# Data needs to be scaled for clustering
-dr <- scale(dr)
-
-# ~~~~~~~~~~~~~~~~~~~ k-medoids clustering ~~~~~~~~~~~~~~~~~~~ #
-# Cluster analysis
-nSample <- 100
-sampleSize <- 10000
-k <- 2:10
-kmedoid <- vector('list', length(k)+1)
-for(i in k) {
-  cat("   k: ", i, "\r")
-  kmedoid[[i]] <- clara(dr,
-                        k = i,
-                        sampsize = sampleSize,
-                        samples = nSample,
-                        metric="manhattan",
-                        stand = TRUE,
-                        keep.data = FALSE)
-save(kmedoid, file = './data/kmedoid.RData')
-}
-kmedoid[[1]] <- NULL
-save(kmedoid, file = './data/kmedoid.RData')
-
-# ~~~~~~~~~~~~~~~~~~~ Average silhouette width ~~~~~~~~~~~~~~~~~~~ #
-aswMedoid <- unlist(lapply(kmedoid, function(x) x$silinfo$avg.width))
-save(aswMedoid, file = './data/aswMedoid.RData')
-
-# ~~~~~~~~~~~~~~~~~~~ Elbow method ~~~~~~~~~~~~~~~~~~~ #
-# Verify number of clusters by using the elbow method,
-# i.e. the inflexion point in the sum of squared error (SSE) scree plot
-# Extract within cluster sum of squared error (WSS) from each clustering run
-# Wss function
-wss <- function(x) (nrow(x)-1) * sum(apply(x,2,var))
-
-# Measure for each k and each cluster
-wssMedoid <- wss(dr)
-for(i in 1:length(kmedoid)) { # For each k value tested
-  wssTemp <- numeric()
-  for(j in 1:k[i]) { # For each cluster
-    id <- kmedoid[[i]]$clustering == j # Identify observations in cluster j
-    wssTemp[j] <- wss(dr[id, ])
-  }
-  wssMedoid[k[i]] <- sum(wssTemp)
-}
-save(wssMedoid, file = './data/wssMedoid.RData')

### 2020-06-01 Code/9-kmeans.R

@@ -1,111 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
-source('./code/2-param.R')
-
-# ~~~~~~~~~~~~~~~~~~~ Data prep ~~~~~~~~~~~~~~~~~~~ #
-# Data needs to be scaled for clustering
-dr <- scale(dr)
-
-# ~~~~~~~~~~~~~~~~~~~ k-means clustering ~~~~~~~~~~~~~~~~~~~ #
-# Cluster analysis
-k <- 2:10
-kMeans <- list('list', length(k))
-for(i in k) {
-  cat("   k: ", i, "\r")
-  kMeans[[i]] <- kmeans(x = dr,
-                      centers = i,
-                      iter.max = 1000,
-                      nstart = 25,
-                      algorithm = 'Lloyd')
-}
-kMeans[[1]] <- NULL
-save(kMeans, file = './data/kmeans.RData')
-
-
-# ~~~~~~~~~~~~~~~~~~~ Average silhouette width ~~~~~~~~~~~~~~~~~~~ #
-# Silhouette needs to be computed manually here, because the general function crashes
-# This will make a very inefficient function, but one that will work
-sil <- function(x, cl) {
-  # Euclidian distance function between a single point and another point or a string of points
-  euc <- function(x, y) {
-    if(is.null(dim(y))) {
-      sqrt(sum((x - y) ^ 2))
-    } else {
-      sqrt(colSums((x - t(y)) ^ 2))
-    }
-  }
-
-  # Number of rows in dataset x
-  nr <- length(cl)
-
-  # Clusters in dataset
-  k <- unique(cl)
-
-  # Empty data.frame to store silhouette information
-  df <- data.frame(cluster = numeric(nr),
-                   neighbor = numeric(nr),
-                   sil_width = numeric(nr),
-                   stringsAsFactors = F)
-  # Calculate silhouette for all data points in dataset x
-  for(i in 1:nr) {
-    # Cluster of data point i
-    df$cluster[i] <- cl[i]
-
-    # ID of points in cluster of point i
-    id <- cl == df$cluster[i]
-
-    # a = mean distance of point i to all points within its cluster
-      a <- euc(x[i, ], x[id, ]) %>%
-           mean()
-
-    # b =  bi is the smallest average distance of i to all points in any other
-    #      cluster of which i is not a member.
-      # Clusters of which i is not a member
[…]

### 2020-06-01 Code/Analyses/A1-Drivers.R

@@ -1,0 +1,126 @@
+# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! #
+# The initial version of these analyses and the paper were done with vectorized
+# data in `sf` format. The package has since transitioned to provide rasterized
+# data layers for their size and efficiency. Here, we provide the original code
+# using `sf` objects and the original vectorized data are provided in the `data`
+# folder as:
+# - './Data/Drivers.RData'
+# - './Data/DriversRaw.RData'
+#
+# We nonetheless provide the code, as comments, to load and transform the
+# rasterized data layers from the `eDrivers` package as a starting point.
+# We provide, also as comments, the code to reproduce the analyses with the
+# rasterized objects. However, the results and figures presented were obtained
+# with the vectorized objects.
+# !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! #
+
+# Load raw data
+load('./Data/DriversRaw.RData')
+drivers <- driversRaw
+
+# Log-transform drivers that should be transformed
+# Name of drivers to transform
+drNames <- colnames(drivers)
+drTrans <- c('coastDev','dirHumImpact','fisheriesDD',
+             'fisheriesDNH','fisheriesDNL','fisheriesPHB','fisheriesPLB',
+             'Hypoxia','inorgPol','invasives','nutrientInput',
+             'orgPol','negSBT','posSBT','shipping')
+
+# Log transform
+id <- which(colnames(drivers) %in% drTrans)
+for(i in id) drivers[,i] <- log(drivers[,i] + 1)
+
+# Scale drivers between 0 and 1 using the 99th quantile
+# Function
+quantNorm <- function(x) {
+  id <- x != 0
+  x <- x / quantile(x[id], probs = .99, na.rm = T)
[…]

### 2020-06-01 Code/Analyses/A2-Cumulative_Exposure.R

@@ -1,0 +1,30 @@
+# Load drivers data
+load('./Data/Drivers.RData')
+
+# Cumulative exposure
+CE <- data.frame(ID = drivers$ID,
+                 CumulativeExposure = rowSums(drivers[, -1]),
+                 stringsAsFactors = F)
+
+# Export
+save(CE, file = './Data/CumulativeExposure.RData')
+
+
+
+
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ RASTER CODE ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+
+# # Load drivers data
+# library(raster)
+# load('./Data/Drivers.RData')
+#
+# # Cumulative exposure
+# CE <- sum(drivers, na.rm = T)
+# names(CE) <- 'CumulativeExposure'
+#
+# # Export
+# save(CE, file = './Data/CumulativeExposure.RData')

### 2020-06-01 Code/Analyses/A3-Cumulative_Hotspots.R

@@ -1,0 +1,89 @@
+# Load drivers data
+load('./Data/Drivers.RData')
+
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# Quantile
+# 80th quantile
+# These will be transformed as binary values
+# Then sum the drivers
+# Then look at distribution
+dr <- drivers[,-1]
+
+# Empty matrix to store results
+hot <- matrix(ncol = ncol(dr),
+              nrow = nrow(dr),
+              data = NA,
+              dimnames = list(c(), colnames(dr)))
+
+# Evaluate hotspots
+for(i in 1:ncol(hot)) {
+  # ID = 0
+  id0 <- dr[, i] > 0
+  # 80th quantile threshold
+  th <- quantile(dr[id0, i], .8, na.rm = T)
+  # Values over hotspot threshold
+  hot[, i] <- dr[, i] > th
+}
+
+# Hotspots
+cumulHotspot <- rowSums(hot)
+cumulHotspot[cumulHotspot == 0] <- NA
+
+# Cumulative hotspots
+CH <- data.frame(ID = drivers$ID,
+                 CumulativeHotspots = cumulHotspot,
+                 stringsAsFactors = F)
+
+# Export
+save(CH, file = './Data/CumulativeHotspots.RData')
+
+
+
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ RASTER CODE ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+#
+# # Load drivers data
[…]

### 2020-06-01 Code/Analyses/A4-KMedoids.R

@@ -1,0 +1,76 @@
+# Load drivers data
+load('./Data/Drivers.RData')
+
+# As.matrix
+dr <- drivers[, -1]
+
+# Data needs to be scaled for clustering
+dr <- scale(dr)
+
+# ~~~~~~~~~~~~~~~~~~~ k-medoids clustering ~~~~~~~~~~~~~~~~~~~ #
+library(cluster)
+
+# Cluster analysis
+nSample <- 100
+sampleSize <- 10000
+k <- 2:10
+kmedoid <- vector('list', length(k)+1)
+for(i in k) {
+  cat("   k: ", i, "\r")
+  kmedoid[[i]] <- clara(dr,
+                        k = i,
+                        sampsize = sampleSize,
+                        samples = nSample,
+                        metric="manhattan",
+                        stand = TRUE,
+                        keep.data = FALSE)
+save(kmedoid, file = './Data/Clustering/kmedoid.RData')
+}
+kmedoid[[1]] <- NULL
+save(kmedoid, file = './Data/Clustering/kmedoid.RData')
+
+# ~~~~~~~~~~~~~~~~~~~ Average silhouette width ~~~~~~~~~~~~~~~~~~~ #
+aswMedoid <- unlist(lapply(kmedoid, function(x) x$silinfo$avg.width))
+save(aswMedoid, file = './Data/Clustering/aswMedoid.RData')
+
+# ~~~~~~~~~~~~~~~~~~~ Elbow method ~~~~~~~~~~~~~~~~~~~ #
+# Verify number of clusters by using the elbow method,
+# i.e. the inflexion point in the sum of squared error (SSE) scree plot
+# Extract within cluster sum of squared error (WSS) from each clustering run
+# Wss function
+wss <- function(x) (nrow(x)-1) * sum(apply(x,2,var))
+
+# Measure for each k and each cluster
+wssMedoid <- wss(dr)
+for(i in 1:length(kmedoid)) { # For each k value tested
+  wssTemp <- numeric()
+  for(j in 1:k[i]) { # For each cluster
+    id <- kmedoid[[i]]$clustering == j # Identify observations in cluster j
+    wssTemp[j] <- wss(dr[id, ])
+  }
+  wssMedoid[k[i]] <- sum(wssTemp)
+}
[…]

### 2020-06-01 Code/Analyses/A5-Inter_Cluster_Dissimilarity.R

@@ -1,0 +1,72 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAM ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Clustering/k.RData')
+load('./Data/Clustering/Clusters.RData')
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/bootstrapSimper.R')
+source('./Code/Functions/simperMan.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ SIMPER ~~~~~~~~~~~~~~~~~~~ #
+# Run iterative simper analysis
+iter <- 300
+prop <- .05
+system.time(bootSimper <- bootstrapSimper(dr, clMed, iter = iter, prop = prop))
+save(bootSimper, file = './Data/Clustering/bootSimper.RData')
+
+# # # This is to run it multiple times simultaneously
+# # bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '1')
+# # bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '2')
+# # bootSimper <- bootstrapSimper(dr, clMed, iter = 100, prop = prop, accr = '3')
+# load('./data/bootSimper1.RData')
+# x1 <- bootSimper
+# load('./data/bootSimper2.RData')
+# x2 <- bootSimper
+# load('./data/bootSimper3.RData')
+# x3 <- bootSimper
+# bootSimper <- c(x1,x2,x3)
+# save(bootSimper, file = './Data/Clustering/bootSimper.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ SIMPER SUMMARY ~~~~~~~~~~~~~~~~~~~ #
+# Summarize in array format
+iter <- length(bootSimper) # To delete
+nList <- length(bootSimper[[1]])
+
+# Name of variables to extract from simper analyses
+varNames <- c('contr','sd','ava','avb')
+
+# Empty list to store summaries
+simperSummary <- vector('list', nList)
+names(simperSummary) <- names(bootSimper[[1]])
+
+# Empty arrays to store inter-cluster dissimilarity results
+for(i in 1:nList) {
+  simperSummary[[i]] <- array(data = 0,
[…]

### 2020-06-01 Code/Analyses/A6-Intra_Cluster_Similarity.R

@@ -1,0 +1,58 @@
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Clustering/k.RData')
+load('./Data/Clustering/Clusters.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/intraSimManhattan.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ INTRA-CLUSTER SIMILARITY ~~~~~~~~~~~~~~~~~~~ #
+# As with the simper analysis, the dataset is mostly too large to use
+# regular function in packages available in R. Also, while the software
+# PRIMER automatically measures the intra-cluster similarity using the
+# Bray-Curtis similarity, `vegan::simper` does not do it.
+# I therefore create a custom function that allows me to extract the
+# intra-cluster Bray-Curtis similarity and that does it using a bootstrap
+# procedure when the cluster has too many observations.
+#
+sim <- vector('list', k)
+for(i in 1:k) {
+  cat("   k: ", i, "\r")
+  id <- clMed == i
+  sim[[i]] <- bootIntraSim(dr[id, ], iter = 50, samp = .25)
+  save(sim, file = './Data/Clustering/intraSimilarity.RData')
+}
+
+# Extract species, contribution and sd
+# Summarize in array format
+iter <- length(sim[[4]])
+varNames <- c('contr')
+simSummary <- vector('list', k)
+for(i in 1:k) simSummary[[i]] <- array(data = 0, dim = c(ncol(dr), 1, iter), dimnames = list(drNames$accr, varNames))
+
+for(i in 1:k) {
+  # Check if there are multiple iterations or not.
+  # If length(sim[[i]]) == 2 it means that there were no iterations and the whole cluster was parsed through the simper2 function
+  # If not, then sim[[i]] should be equal to the number of iterations
+  if(i %in% c(1,2,3)) {
+    simSummary[[i]] <- matrix(ncol = 1,
+                              data = sim[[i]]$contr,
+                              dimnames = list(drNames$accr, 'contr'))
+  } else {
+    for(j in 1:iter) {
[…]

### 2020-06-01 Code/Figures/F1-EGSL.R

@@ -1,0 +1,148 @@
+# source('./Code/Figures/F1-EGSL.R')
+# ~~~~~~~~~~~~~~~~~~~ LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+# Libraries
+library(raster)
+library(sf)
+
+# ~~~~~~~~~~~~~~~~~~~ Default parameters ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTIONS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/plotEGSL.R')
+source('./Code/Functions/legendEGSL.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Spatial/Bathymetry.RData') # Bathymetry
+load('./Data/Spatial/Canada.RData') # Canada
+load('./Data/Spatial/US.RData') # US
+load('./Data/Spatial/egslSimple.RData') # Simple EGSL geometry
+
+# ~~~~~~~~~~~~~~~~~~~ PROJECTIONS ~~~~~~~~~~~~~~~~~~~ #
+canada <- st_transform(canada, crs = 4326)
+usa <- st_transform(usa, crs = 4326)
+bathy <- st_transform(bathy, crs = 4326)
+deph <- bathy$DEPH
+
+# ~~~~~~~~~~~~~~~~~~~ FORMAT BATHYMETRY ~~~~~~~~~~~~~~~~~~~ #
+r <- raster(ncol=180, nrow=180)
+extent(r) <- extent(bathy)
+r <- fasterize::fasterize(bathy, r, field = 'DEPH')
+r <- mask(r, egslSimple)
+
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# EGSL FIGURE
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+jpeg('./figures/egsl.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
+
+# Colors
+pal <- colorRampPalette(c('#E2FCF6', '#1B7CAA'))
[…]

### 2020-06-01 Code/Figures/F2-Kernel.R

@@ -1,0 +1,156 @@
+# source('./Code/Figures/F2-Kernel.R')
+# ~~~~~~~~~~~~~~~~~~~ LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+# Libraries
+library(raster)
+library(sf)
+
+# ~~~~~~~~~~~~~~~~~~~ Default parameters ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTIONS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/plotEGSL.R')
+source('./Code/Functions/legendEGSL.R')
+source('./Code/Functions/colorBar.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Spatial/egslGrid.RData') # Vectorized grid
+load('./Data/Spatial/egslSimple.RData') # Simple EGSL geometry
+load('./Data/Drivers.RData')
+
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# EGSL FIGURE
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+jpeg('./figures/kernel.jpg', width = 1920, height = 1280, res = 300, pointsize = 6)
+
+# Layout
+mat <- matrix(0, 28, 32)
+mat[1:12, 1:16] <- 1 # traffic
+mat[15:26, 1:16] <- 2 # demersal
+mat[1:12, 17:32] <- 3 # traffic + demersal
+mat[18:28, 17:29] <- 4 # kernel
+mat[15:17, 17:29] <- 5 # dens traffic (x)
+mat[18:28, 30:32] <- 6 # dens demersal (y)
+mat[13:14, 6:11] <- 7 # scale traffic
+mat[27:28, 6:11] <- 8 # scale demersal
+mat[13:14, 22:27] <- 9 # scale combine
+mat[1:2, 15:16] <- 10 # panel A
+mat[15:16, 15:16] <- 11 # panel B
+mat[1:2, 31:32] <- 12 # panel C
[…]

### 2020-06-01 Code/Figures/F3-CumulativeExposure.R

@@ -1,0 +1,63 @@
+# source('./Code/Figures/F3-CumulativeExposure.R')
+# ~~~~~~~~~~~~~~~~~~~ LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+# Libraries
+library(raster)
+library(sf)
+
+# ~~~~~~~~~~~~~~~~~~~ Default parameters ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTIONS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/plotEGSL.R')
+source('./Code/Functions/legendEGSL.R')
+source('./Code/Functions/colorBar.R')
+source('./Code/Functions/cityEGSL.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Spatial/Canada.RData') # Canada
+load('./Data/Spatial/US.RData') # US
+load('./Data/Spatial/egslGrid.RData') # Vectorized grid
+load('./Data/Spatial/egslSimple.RData') # Simple EGSL geometry
+load('./Data/CumulativeExposure.RData')
+
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# Cumulative exposure figure
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+jpeg('./figures/CumulativeExposure.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
+
+# Main plot output
+par(bg = '#ffffff')
+plotEGSL(layers     = c('egslSimple', 'canada','usa'),
+         cols       = c('#00000000',off,off),
+         borders    = c('#00000000',focus,focus),
+         lwds       = c(.75,.75,.75),
+         mar        = c(2,2,2,2),
+         box        = T,
+         axes       = 1:4,
+         northArrow = F,
+         prj        = world,
+         extent     = extFig)
+
+# Cumulative exposure
+cols <- pal(101)[((CE$CumulativeExposure/max(CE$CumulativeExposure, na.rm = T))*100)+1]
[…]

### 2020-06-01 Code/Figures/F4-CumulativeHotspots.R

@@ -1,0 +1,64 @@
+# source('./Code/Figures/F4-CumulativeHotspots.R')
+# ~~~~~~~~~~~~~~~~~~~ LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+# Libraries
+library(raster)
+library(sf)
+
+# ~~~~~~~~~~~~~~~~~~~ Default parameters ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTIONS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/plotEGSL.R')
+source('./Code/Functions/legendEGSL.R')
+source('./Code/Functions/colorBar.R')
+source('./Code/Functions/cityEGSL.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Spatial/Canada.RData') # Canada
+load('./Data/Spatial/US.RData') # US
+load('./Data/Spatial/egslGrid.RData') # Vectorized grid
+load('./Data/Spatial/egslSimple.RData') # Simple EGSL geometry
+load('./Data/CumulativeHotspots.RData')
+
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+# Cumulative exposure figure
+# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= #
+jpeg('./figures/CumulativeHotspots.jpg', width = 1280, height = 920, res = 300, pointsize = 6)
+
+# Main plot output
+par(bg = '#ffffff')
+plotEGSL(layers     = c('egslSimple', 'canada','usa'),
+         cols       = c('#00000000',off,off),
+         borders    = c('#00000000',focus,focus),
+         lwds       = c(.75,.75,.75),
+         mar        = c(2,2,2,2),
+         box        = T,
+         axes       = 1:4,
+         northArrow = F,
+         prj        = world,
+         extent     = extFig)
+
+# Cumulative exposure
+nOver <- max(CH$CumulativeHotspots, na.rm = T)
+cols <- pal(nOver)[CH$CumulativeHotspots]
[…]

### 2020-06-01 Code/Figures/F5-Clusters.R

@@ -1,0 +1,97 @@
+# source('Code/Figures/F5-Clusters.R')
+# ~~~~~~~~~~~~~~~~~~~ LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+# Libraries
+library(raster)
+library(sf)
+
+# ~~~~~~~~~~~~~~~~~~~ Default parameters ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTIONS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/plotEGSL.R')
+source('./Code/Functions/legendEGSL.R')
+source('./Code/Functions/colorBar.R')
+source('./Code/Functions/cityEGSL.R')
+source('./Code/Functions/barPlot.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Spatial/Canada.RData') # Canada
+load('./Data/Spatial/US.RData') # US
+load('./Data/Spatial/egslGrid.RData') # Vectorized grid
+load('./Data/Spatial/egslSimple.RData') # Simple EGSL geometry
+load('./Data/Clustering/k.RData')
+load('./Data/Clustering/Clusters.RData')
+load('./Data/Clustering/simperMean.RData')
+nCl <- length(unique(clMed))
+
+# ~~~~~~~~~~~~~~~~~~~ INTER CLUSTER ~~~~~~~~~~~~~~~~~~~ #
+# Driver intensity in compared clusters
+drInt1 <- lapply(simperMean, function(x) x[,'ava']) %>%
+          as.data.frame()
+drInt2 <- lapply(simperMean, function(x) x[,'avb']) %>%
+          as.data.frame()
+
+dat <- vector('list', nCl)
+dat[[1]] = data.frame(x = drInt1$X1_2)
+dat[[2]] = data.frame(x = drInt1$X2_3)
+dat[[3]] = data.frame(x = drInt1$X3_4)
+dat[[4]] = data.frame(x = drInt1$X4_5)
+dat[[5]] = data.frame(x = drInt1$X5_6)
+dat[[6]] = data.frame(x = drInt2$X5_6)
+
[…]

### 2020-06-01 Code/Figures/S1-Histograms.R

@@ -1,0 +1,23 @@
+# source('./Code/Figures/S1-Histograms.R')
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+load('./Data/DriversRaw.RData')
+drRaw <- driversRaw[, -1]
+
+# ~~~~~~~~~~~~~~~~~~~ HISTOGRAMS ~~~~~~~~~~~~~~~~~~~ #
+# Value for `par` parameters for figures
+parVal <- c(ceiling(nDr / 4), 4)
+
+# Let's take a look with 0 removed from raw data
+# This is especially important for the coastal drivers, as including zeros
+# would include offshore cells, which vastly outnumber coastal cells and
+# to which coastal drivers to not extend
+jpeg('./figures/histograms.jpg', width = 1280, height = 1500, res = 200)
+par(mfrow = parVal, mar = c(4,4,1,1))
+for(i in 1:nDr) {
+  id <- drRaw[, i] > 0
+  drID <- drNames$accr[drNames$var == colnames(drRaw)[i]]
+  hist(drRaw[id, i], prob = T, main = '', xlab = drID, col = focus)
+  lines(density(drRaw[id, i], from = 0, to = 1), col = uni, lwd=2)
+}
+dev.off()

### 2020-06-01 Code/Figures/S2-Drivers.R

@@ -1,0 +1,41 @@
+# source('./Code/Figures/S2-Drivers.R')
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+load('./Data/DriversRaw.RData')
+drRaw <- driversRaw[, -1]
+
+# Value for `par` parameters for figures
+parVal <- c(ceiling(nDr / 4), 4)
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTIONS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/plotEGSL.R')
+
+# ~~~~~~~~~~~~~~~~~~~ DRIVERS ~~~~~~~~~~~~~~~~~~~ #
+jpeg('./figures/drivers.jpeg', width = 1280, height = 1500, res = 200)
+par(mfrow = parVal, mar = c(4,4,1,1))
+# Drivers
+for(i in 1:nDr) {
+  drData <- dr[, i]
+  drID <- drNames$accr[drNames$var == colnames(dr)[i]]
+  id <- drData > 0
+  cols <- pal(101)[((drData[id]/max(drData[id]))*100)+1]
+  plotEGSL('egslSimple', prj = world, extent = 'egslSimple')
+  plot(st_geometry(egslGrid[id, ]), add = T, col = cols, border = cols, lwd = .1)
+  text(x = mean(par('usr')[1:2]), y = par('usr')[4]-.5, labels = drID)
+}
+
+# Scale
+colRamp <- pal(100)
+scale = (length(colRamp) - 1)/(1 - 0)
+ticks <- seq(0, 1, len = 11)
+par(mar = c(.75,0,.75,11))
+plot(c(0, 10), c(0, 1), type = "n", bty = "n", xaxt = "n",
+    xlab = "", yaxt = "n", ylab = "")
+axis(4, ticks, las = 1)
+for (i in 1:(length(colRamp) - 1)) {
+    y = (i - 1)/scale + 0
+    rect(0, y, 10, y + 1/scale, col = colRamp[i], border = NA)
+}
+rect(xleft = 0, ybottom = 0, xright = 10, ytop = 1, lwd = 0.5)
+mtext('Relative\nintensity', 4, line = 4, cex = .8, adj = c(.5,.5))
+dev.off()

### 2020-06-01 Code/Figures/S3-Marimekko.R

@@ -1,0 +1,168 @@
+# source('./Code/Figures/S3-Marimekko.R')
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./code/0-Param.R')
+library(graphicsutils)
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+Fd <- rowSums(dr)
+
+# Create vector of bins
+nBins <- 512 # Just because
+binData <- seq(0,max(Fd), length.out = 513)
+bins <- data.frame(lower = binData[1:512],
+                   upper = binData[2:513])
+bins$mid <- (bins$upper + bins$lower) / 2
+
+# Sum of fd values within each bin
+bins$fdTot <- numeric(nBins)
+for(i in 1:nBins) {
+  id <- Fd > bins$lower[i] & Fd <= bins$upper[i]
+  bins$fdTot[i] <- sum(Fd[id])
+}
+
+# Fd values associated with each driver
+binDr <- matrix(nrow = nrow(bins),
+                ncol = ncol(dr),
+                data = 0,
+                dimnames = list(c(), colnames(dr)))
+
+for(j in 1:ncol(binDr)) {
+  for(i in 1:nBins) {
+    id <- Fd > bins$lower[i] & Fd <= bins$upper[i]
+    binDr[i, j] <- sum(dr[id, j])
+  }
+}
+
+# Transform into a marimekko graph
+# Begin by transforming the data into proportions of total Fd
+binDrProp <- binDr / bins$fdTot
+
+# Sort drivers by group
+sortNames <- as.factor(drNames$group) %>%
+             order() %>%
+             drNames[., ]
+sortNames$group <- as.factor(sortNames$group)
+
+# Add colors
+trans <- c('FF','DD','BB','99','77','55','33','11')
+sortNames$col <- gg_color_hue(nGroup)[as.numeric(sortNames$group)]
+for(i in levels(sortNames$group)) {
+  id <- which(sortNames$group == i)
+  nId <- length(id)
+  for(j in 1:nId) {
+    sortNames$col[id[j]] <- paste0(substr(sortNames$col[id[j]],1,7), trans[j])
+  }
+}
+
+# Sort by group
[…]

### 2020-06-01 Code/Figures/S4-ValidateClusters.R

@@ -1,0 +1,38 @@
+# source('./Code/Figures/S4-ValidateClusters.R')
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+k <- 2:10
+
+# ~~~~~~~~~~~~~~~~~~~ Data ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Clustering/aswMedoid.RData')
+load('./Data/Clustering/wssMedoid.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ Figure ~~~~~~~~~~~~~~~~~~~ #
+jpeg('./figures/validation.jpg', width = 920, height = 920, res = 200, pointsize = 7)
+layout(matrix(1:2, ncol = 1))
+par(mar = c(5,5,2,1))
+# k-medoid
+  # Silhouette
+  plot(aswMedoid ~ k,
+       pch = 20,
+       ylab = 'Average Silhouette Width',
+       type = 'b',
+       frame = F,
+       xlab = '',
+       xlim = c(0,10),
+       cex = 1.5,
+       cex.axis = .8)
+  mtext(side = 3, 'k-medoids clustering', font = 2, line = .75)
+
+  # WSS
+  plot(x = c(1,k),
+       y = wssMedoid,
+       type="b",
+       frame = F,
+       xlab="Number of Clusters (k)",
+       ylab="Total within-clusters sum of squares",
+       pch = 20,
+       xlim = c(0,10),
+       cex = 1.5,
+       cex.axis = .8)
+dev.off()

### 2020-06-01 Code/Figures/S5-Dissimilarity_Contribution.R

@@ -1,0 +1,152 @@
+# source('./Code/Figures/S5-Dissimilarity_Contribution.R')
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ EXTRA LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+library(gridExtra)
+library(ggplot2)
+library(ggplotify)
+library(grid)
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Clustering/k.RData')
+load('./Data/Clustering/Clusters.RData')
+load('./Data/Clustering/simperMean.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/circularPlot.R')
+source('./Code/Functions/butterfly.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ INDIVIDUAL CONTRIBUTION TO INTER-CLUSTER DISSIMILARITY ~~~~~~~~~~~~~~~~~~~ #
+# Driver percent contribution to dissimilarity
+contDiss <- lapply(simperMean, function(x) x[,'contr'] / sum(x[,'contr'])) %>%
+            as.data.frame()
+colnames(contDiss) <- names(simperMean)
+
+# Total dissimilarity
+totDiss <- unlist(lapply(simperMean, function(x) sum(x[, 'contr'])))
+
+# Driver intensity in compared clusters
+drInt1 <- lapply(simperMean, function(x) x[,'ava']) %>%
+          as.data.frame()
+drInt2 <- lapply(simperMean, function(x) x[,'avb']) %>%
+          as.data.frame()
+
+# Mean value in each cluster
+totMean <- numeric(k)
+for(i in 1:k) {
+  id <- clMed == i
+  totMean[i] <- mean(as.matrix(dr[id, ]))
+}
+
+# Fish
+fish <- matrix(ncol = 5, nrow = 7, dimnames = list(c(), drNames$var[5:9]))
+for(i in 1:k) {
+  id <- clMed == i
+  for(j in colnames(fish)) {
+    fish[i, j]<- sum((dr[id, j]))
+  }
+}
+
[…]

### 2020-06-01 Code/Figures/S6-Similarity_Contribution.R

@@ -1,0 +1,114 @@
+# source('./Code/Figures/S6-Similarity_Contribution.R')
+# ~~~~~~~~~~~~~~~~~~~ PARAMETERS ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/0-Param.R')
+
+# ~~~~~~~~~~~~~~~~~~~ EXTRA LIBRARIES ~~~~~~~~~~~~~~~~~~~ #
+library(gridExtra)
+library(ggplot2)
+library(ggplotify)
+library(grid)
+library(graphicsutils)
+
+
+# ~~~~~~~~~~~~~~~~~~~ DATA ~~~~~~~~~~~~~~~~~~~ #
+load('./Data/Clustering/k.RData')
+load('./Data/Clustering/Clusters.RData')
+load('./Data/Clustering/similarityMean.RData')
+
+
+# ~~~~~~~~~~~~~~~~~~~ FUNCTION ~~~~~~~~~~~~~~~~~~~ #
+source('./Code/Functions/circularPlot.R')
+
+
+# ~~~~~~~~~~~~~~~~~~~ SIMILARITY CONTRIBUTION ~~~~~~~~~~~~~~~~~~~ #
+# Driver percent contribution to dissimilarity
+contSim <- lapply(similarityMean, function(x) x[,'contr'] / sum(x[,'contr'])) %>%
+            as.data.frame()
+colnames(contSim) <- names(similarityMean)
+
+# Total similarity
+totSim <- lapply(similarityMean, function(x) sum(x[,'contr'])) %>%
+          as.data.frame()
+
+# Circulat plots
+p1 <- vector('list', ncol(contSim))
+for(i in 1:length(contSim)) {
+  temp <- data.frame(individual = drNames$accr,
+                     group = drNames$group,
+                     value = round(contSim[, i]*100))
+  rownames(temp) <- NULL
+  temp2 <- round(totSim[i],2)
+  p1[[i]] <- circularPlot(temp, temp2)
+}
+
+# Individual clusters
+egsl <- st_centroid(egslGrid)
+p2 <- vector('list', k)
+for(i in 1:length(p2)) {
+  # Identify observations in cluster i
+  id <- clMed == i
+
+  # Select only grid cells in cluster i
+  temp <- egsl %>%
+          filter(clMed == i) %>%
+          st_coordinates() %>%
+          as.data.frame()
+          ext <- st_bbox(egsl)
+
[…]

### 2020-06-01 Code/Functions/barPlot.R

@@ -1,0 +1,36 @@
+barPlot <- function(x2) {
+  # Gap
+  gap <- .25
+  yG <- .15
+
+  # Combine with drNames
+  # This makes the function unusable for another setting, but simply removing it would make it ok
+  x2$names = drNames$accr
+  x2$group = as.factor(drNames$group)
+
+  # Format data
+  x2$x <- x2$x + gap
+  x2 <- x2[order(x2$group), ]
+  x2$col <- as.numeric(x2$group)
+  x2 <- x2[nrow(x2):1, ]
+
+  # Colors
+  cols <- gg_color_hue(nGroup)
+  x2$col <- cols[x2$col]
+
+  # Plot
+  library(graphicsutils)
+  par(mar = c(4,0,2,0))
+  plot0(xlim = c(-.125, 1.25), ylim = c(0,length(x2$x)+1))
+  # axis(1, at = seq(-gap, -1-gap, by = -.25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
+  axis(1, at = seq(gap, 1+gap, by = .25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
+  # axis(1, at = seq(-gap, gap, by = 2*gap), labels = c('',''), lwd = 1, tck = -.005)
+  text(x = rep(0, length(x2$x)), y = 1:(length(x2$x)), labels = x2$names, cex = .7)
+  for(i in c(seq(-gap, -1-gap, by = -gap), seq(gap, 1+gap, by = gap))) {
+    lines(x = c(i,i), y = c(.5,length(x2$x)+1.5), lty = 'dotted', col = '#00000066')
+  }
+  # Add bar plot
+  for(i in 1:22) {
+    rect(xleft = x2$x[i], ybottom = i-yG, xright = gap, ytop = i+yG, col = x2$col[i], border = x2$col[i])
+  }
+}

### 2020-06-01 Code/Functions/bootstrapSimper.R

@@ -1,0 +1,28 @@
+# iter <- 10
+# prop <- .02 # .05: ~500s/iter | .01: ~20s/iter | .02: ~80s/iter
+# The analysis being very long, we will proceed using a
+# bootstrap procedure, i.e. iterative simper analysis
+bootstrapSimper <- function(x, cl, iter = 100, prop = .1, accr = '') {
+  # List to store simper results for post-processing
+  bootSimper <- vector('list', iter)
+
+  # Determine the number of data points to sample from each cluster
+    clSub <- ceiling(table(cl) * prop)
+
+  # Repeat the process for the desired number of iterations
+  for(i in 1:iter) {
+    cat("   i: ", i, ' of ', iter, "\r")
+
+    # Subsampled dataset
+    samp <- vector('list', length(clSub))
+    for(j in 1:length(clSub)) samp[[j]] <- sample(x = which(cl == j), size = clSub[j])
+    samp <- unlist(samp)
+
+    # Run simper analysis on subset of data
+    bootSimper[[i]] <- simperMan(x[samp, ], cl[samp])
+
+    # Save file
+    save(bootSimper, file = paste0('./data/bootSimper', accr, '.RData'))
+  }
+  return(bootSimper)
+}

### 2020-06-01 Code/Functions/butterfly.R

@@ -1,0 +1,44 @@
+butterfly <- function(x1, x2) {
+  # Gap
+  gap <- .25
+  yG <- .15
+
+  # Combine with drNames
+  # This makes the function unusable for another setting, but simply removing it would make it ok
+  x1 <- data.frame(x = x1, names = drNames$accr, group = drNames$group, stringsAsFactors = T)
+  x2 <- data.frame(x = x2, names = drNames$accr, group = drNames$group, stringsAsFactors = T)
+
+  # Format data
+  x1$x <- -x1$x - gap
+  x1 <- x1[order(x1$group), ]
+  x1$col <- as.numeric(x1$group)
+  x1 <- x1[nrow(x1):1, ]
+
+  x2$x <- x2$x + gap
+  x2 <- x2[order(x2$group), ]
+  x2$col <- as.numeric(x2$group)
+  x2 <- x2[nrow(x2):1, ]
+
+  # Colors
+  cols <- gg_color_hue(nGroup)
+  x1$col <- cols[x1$col]
+  x2$col <- cols[x2$col]
+
+  # Plot
+  library(graphicsutils)
+  par(mar = c(0,0,0,0))
+  plot0(xlim = c(-1.25, 1.25), ylim = c(0,length(x1$x)+1))
+  axis(1, at = seq(-gap, -1-gap, by = -.25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
+  axis(1, at = seq(gap, 1+gap, by = .25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
+  axis(1, at = seq(-gap, gap, by = 2*gap), labels = c('',''), lwd = 1, tck = -.005)
+  text(x = rep(0, length(x1$x)), y = 1:(length(x1$x)), labels = x1$names, cex = .7)
+  for(i in c(seq(-gap, -1-gap, by = -gap), seq(gap, 1+gap, by = gap))) {
+    lines(x = c(i,i), y = c(.5,length(x1$x)+1.5), lty = 'dotted', col = '#00000066')
+  }
+
+  # Add bar plot
+  for(i in 1:22) {
+    rect(xleft = x1$x[i], ybottom = i-yG, xright = -gap, ytop = i+yG, col = x1$col[i], border = x1$col[i])
[…]

### 2020-06-01 Code/Functions/circularPlot.R

@@ -1,0 +1,67 @@
+circularPlot <- function(data, center) {
+  # library
+  library(tidyverse)
+
+  # Set a number of 'empty bar' to add at the end of each group
+  empty_bar=3
+  to_add = data.frame( matrix(NA, empty_bar*nlevels(data$group), ncol(data)) )
+  colnames(to_add) = colnames(data)
+  to_add$group=rep(levels(data$group), each=empty_bar)
+  data=rbind(data, to_add)
+  data=data %>% arrange(group)
+  data$id=seq(1, nrow(data))
+
+  # Get the name and the y position of each label
+  label_data=data
+  number_of_bar=nrow(label_data)
+  angle= 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
+  label_data$hjust<-ifelse( angle < -90, 1, 0)
+  label_data$angle<-ifelse(angle < -90, angle+180, angle)
+
+  # prepare a data frame for base lines
+  base_data=data %>%
+    group_by(group) %>%
+    summarize(start=min(id), end=max(id) - empty_bar) %>%
+    rowwise() %>%
+    mutate(title=mean(c(start, end)))
+
+  # prepare a data frame for grid (scales)
+  grid_data = base_data
+  grid_data$end = grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
+  grid_data$start = grid_data$start - 1
+  grid_data=grid_data[-1,]
+
+  # Make the plot
+  p = ggplot(data, aes(x=as.factor(id), y=value, fill=group)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
+
+    geom_bar(aes(x=as.factor(id), y=value, fill=group), stat="identity", alpha=0.5) +
+
+    # Add a val=100/75/50/25 lines. I do it at the beginning to make sur barplots are OVER it.
+    geom_segment(data=grid_data, aes(x = end, y = 80, xend = start, yend = 80), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
+    geom_segment(data=grid_data, aes(x = end, y = 60, xend = start, yend = 60), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
[…]

### 2020-06-01 Code/Functions/cityEGSL.R

@@ -1,0 +1,73 @@
+cityEGSL <- function(prj = slmetaPrj('default')) {
+
+  city <- data.frame(latitude = numeric(0),
+                     longitude = numeric(0),
+                     city = character(0),
+                     yGap = numeric(0),
+                     xGap = numeric(0),
+                     adj1 = numeric(0),
+                     adj2 = numeric(0),
+                     stringsAsFactors = F)
+
+  city[1,] <- c(48.41, -68.42, 'Rimouski', .0056, .0076, 0, 1)
+  city[2,] <- c(47.79, -69.43, 'Rivière-du-Loup', 0, .0076, 0, 1)
+  city[3,] <- c(48.78, -67.45, 'Matane', 0, .0076, 0, 1)
+  city[4,] <- c(50.3, -66.6, 'Sept-Îles', .006, -.0076, 1, 0)
+  city[5,] <- c(49.26, -68.32, 'Baie-Comeau', 0, -.015, 1, .5)
+  city[6,] <- c(48.09, -69.9, 'Saguenay River', 0, -.0076, 1, 1)
+  city[7,] <- c(48.83164, -64.68, 'Gaspé', 0, -.0076, 1, 1)
+  city[8,] <- c(46.15, -64.53469, 'Shediac', 0, -.0076, 1, 1)
+  city[9,] <- c(47.05, -65.65, 'Miramichi', 0, -.015, 1, 1)
+  city[10,] <- c(45.53, -62.6, 'New Glasgow', 0, -.0076, 1, 1)
+  city[11,] <- c(48.61, -58.51, 'Stephenville', .006, .0076, 0, 0)
+  city[12,] <- c(48.94900, -57.86, 'Corner Brook', 0, .0076, 0, 0)
+  city[13,] <- c(49.7, -65.7, 'Anticosti Gyre', .1151, .011, 0, 0)
+  city[14,] <- c(49.41, -67.48, 'Pointe-des-Monts', .018, -.0036, 1, 0)
+  city[15,] <- c(46.28, -63.13107, 'Charlottetown', .061, .167, 0, .5)
+  city[16,] <- c(46.92953, -70.93119, 'Île d\'Orléans', 0, -.0076, 1, 0)
+
+  # city[8,] <- c(46.81388, -71.20798, 'Québec city', 0, -.0076, 1, 0)
+  # city[,] <- c(50.33, -63.59857, 'Havre-Saint-Pierre', 15000, -10000, 1, 0)
[…]

### 2020-06-01 Code/Functions/colorBar.r

@@ -1,0 +1,33 @@
+# Function to plot color bar
+# Modified from http://stackoverflow.com/questions/9314658/colorbar-from-custom-colorramppalette
+colorBar <- function(colRamp, min, max=-min, mar = NULL, nticks=11, ticks=seq(min, max, len=nticks), title='', align = 'horizontal', new = FALSE, ...) {
+    scale = (length(colRamp)-1)/(max-min)
+
+    if(align == 'vertical') {
+        if(new == TRUE) {
+            dev.new(width=1.75, height=5)
+        }
+        if(is.null(mar)) mar <- c(0,0,0,4)
+        par(mar = mar)
+        plot(c(0,10), c(min,max), type='n', bty='n', xaxt='n', xlab='', yaxt='n', ylab='', main=title, ...)
+        axis(4, ticks, las=1, ...)
+        for (i in 1:(length(colRamp)-1)) {
+            y = (i-1)/scale + min
+            rect(0,y,10,y+1/scale, col=colRamp[i], border=NA)
+        }
+    }
+
+    if(align == 'horizontal') {
+        if(new == TRUE) {
+            dev.new(width=5, height=1.75)
+        }
+        if(is.null(mar)) mar <- c(4,0,0,0)
+        par(mar = mar)
+        plot(c(min,max), c(0,1), type='n', bty='n', xaxt='n', xlab='', yaxt='n', ylab='', main=title, ...)
+        axis(1, ticks, las=1, ...)
+        for (i in 1:(length(colRamp)-1)) {
+            y = (i-1)/scale + min
+            rect(y,0,y+1/scale,10, col=colRamp[i], border=NA)
+        }
+    }
+}

### 2020-06-01 Code/Functions/intraSimManhattan.R

@@ -1,0 +1,62 @@
+# library(slmeta)
+# data(drivers)
+# dr <- drivers[,-1]
+# load('./data/clMed.RData')
+# samp <- which(clMed == 2)
+# comm <- dr[samp, ]
+# #
+# system.time(x <- intraSim(comm))
+
+intraSim <- function(comm) {
+  # Transform community data as matrix
+  comm <- as.matrix(comm)
+  # comm <- comm + 1
+
+  # Data frames to store results
+  nC <- ncol(comm)
+  # sp <- colnames(comm)
+  outlist <- data.frame(contr = numeric(nC),
+                        sd = numeric(nC),
+                        stringsAsFactors = F)
+
+  # Measure mean dissimilarity contribution in group per variable
+  for(i in seq_len(nC)) {
+    # Similarity for each column and data of group in comparison[j]
+    sim <- dist(comm[,i,drop=FALSE], method="manhattan")
+    sim <- as.vector(sim)
+
+    # Extract mean and sd contribution to inter-group dissimilarity
+    outlist$contr[i] <- mean(sim)
+    outlist$sd[i] <- sd(sim)
+  }
+
+  # Transform to similarity using .1 / .1+contribution
+  outlist$contr <- .1 / (.1 + outlist$contr)
+
+  return(outlist)
+}
+
+bootIntraSim <- function(x, iter = 250, samp = .25, boot = 40000) {
+  # If the dataset is too large, the analysis will be done through a bootstrap
+  # List to store simper results for post-processing
+  # Determine if the bootstrap procedure is necessary
+  # I evaluated the minimum number of observations necessary to initiate the bootstrap
+  # procedure to be 4000. That was chosen by running the analyses and opting for
+  # a number of observations that I deemed could run fast enough.
+  # It is however still a subjective choice and can definitely be changed.
+  if(nrow(x) > boot) {
+    sim <- vector('list', iter)
+
+    # Repeat the process for the desired number of iterations
+    for(i in 1:iter) {
+      cat("   i: ", i, ' of ', iter, "\r")
+      # Subsampled observation id
+      id <- sample(1:nrow(x), size = samp*nrow(x))
+      # Similarity analysis
+      sim[[i]] <- intraSim(x[id, ])
+    }
+  } else {
+    sim <- intraSim(x)
+  }
+  sim
+}

### 2020-06-01 Code/Functions/legendEGSL.R

@@ -1,0 +1,105 @@
+#' Function to create legend for the St. Lawrence base map `plotEGSL()`.
+#'
+#' @export
+#'
+#' @param range numeric, vector with minimal and maximal values
+#' @param pal character, vector of colors, or color palette
+#' @param cex.text numeric, cex for legend text
+#' @param mainTitle character, type of legend. Choices are 'continuous', 'binary' or 'categorical'
+#' @param subTitle character, type of legend. Choices are 'continuous', 'binary' or 'categorical'
+#' @param type character, type of legend. Choices are 'continuous', 'binary' or 'categorical'
+#' @param nTick numeric, number of ticks in the legend
+
+#'
+#' @return Opens a graphical interface with the plot
+#'
+#' @keywords plot, st. lawrence
+#'
+#' @examples
+#' plotEGSL()
+#' legendEGSL(mainTitle = 'Acidification', subTitle = expression(Aragonite~(m^2)))
+#'
+#' plotEGSL2('webdriver', driver = 'aquacultureInv')
+
+legendEGSL <- function (range = c(0,1),
+                        pal = NULL,
+                        cexMain = 1,
+                        cexSub = .75,
+                        mainTitle = NULL,
+                        subTitle = NULL,
+                        n = 5) {
+  # Legends
+  # Palette
+  if(is.null(pal)) {
+    pal <- colorRampPalette(slmetaPal('platform'))
+  } else {
+    if(class(pal) == 'character') {
+      pal <- colorRampPalette(pal)
+    }
+  }
+
+  # Determine plot boundaries, in units of the data
+  xmin <- par("usr")[1]
+  xmax <- par("usr")[2]
+  ymin <- par("usr")[3]
+  ymax <- par("usr")[4]
+  xR <- xmax - xmin
+  yR <- ymax - ymin
+
+  xinit <- xmin + .034*xR # minimum left side to write
+  yinit <- ymax - .056*yR # minimum upper side to write
+  ygap <- .056*yR
+  xgap <- .014*xR
+  ybarUp <- yinit - ygap/2 - .0041*yR
+  ybarDn <- yinit - ygap -ygap/2 + .0041*yR
+
+
+  # Plot
+   x <- seq(from = xinit, to = xinit + .17*xR, by = .0003*xR)
+   z <- data.frame(y1 = ybarUp,
[…]

### 2020-06-01 Code/Functions/plotEGSL.R

@@ -1,0 +1,174 @@
+#' Function to create base plots of the St. Lawrence and the data contained in the package
+#'
+#' @export
+#'
+#' @param layers character, vector of layers to plot on the graph, in plotting order.
+#' @param prj character, projection to use to plot the graph.
+#' @param extent numeric(c(xmn, ymn, xmx, ymx)) extent to plot in prj units OR  | character(data) to use extent of data (i.e. 'canada').
+#' @param cols character, vector of fill colors for all layers specified in layers param, in same order.
+#' @param borders character, vector of border colors for all layers specified in layers param, in same order.
+#' @param lwds numeric, vector of border widths for all layers specified in layers param, in same order.
+#' @param mar numeric(c(0,0,0,0)), plot margins.
+#' @param box logical, draw box around graph.
+#' @param axes numeric(c(1,2,3,4)), draw axes around graph, 1 = bottom; 2 = left; 3 = top; 4 = right.
+#' @param background character, graph background color.
+#' @param graticules character, graph graticules color.
+#' @param scale, logical, draw a scale bar (NOT YET IMPLEMENTED)
+#' @param northArrow, logical, draw a north arrow
+#' @param prjText, logical, add projection name to map
+#' @param prjCol, character, projection name font color
+#'
+#' @return Opens a graphical interface with the plot
+#'
+#' @keywords plot, st. lawrence
+#'
+#' @examples
+#' # Basic map
+#' plotEGSL()
+#'
+#' # Plot grid
+#' plotEGSL(grid = T)
+#'
+#' # Quebec
+#' plotEGSL(layers = c('egslOutline','quebec','canada','usa'),
+#'          cols = c('#2196a8', '#25364A','#25364A','#C7CBCE'),
+#'          borders = '#f4f4f4',
+#'          northArrow = T,
+#'          mar = c(2,2,2,2),
+#'          box = T,
+#'          axes = 1:4,
+#'          extent = 'quebec')
+#'
+#' # Canada
+#' plotEGSL(layers = c('egslOutline','canada','usa'),
+#'         cols = c('#2196a8','#25364A','#C7CBCE'),
+#'         borders = '#f4f4f4',
+#'         northArrow = T,
+#'         mar = c(2,2,2,2),
+#'         box = T,
+#'         axes = 1:4,
+#'         extent = 'canada',
[…]

### 2020-06-01 Code/Functions/simperMan.R

@@ -1,0 +1,63 @@
+# library(slmeta)
+# data(drivers)
+# dr <- drivers[,-1]
+# load('./data/clMed.RData')
+# # samp <- 100:200
+# prop = .05
+# clSub <- ceiling(table(clMed) * prop)
+# samp <- vector('list', length(clSub))
+# for(j in 1:length(clSub)) samp[[j]] <- sample(x = which(clMed == j), size = clSub[j])
+# samp <- unlist(samp)
+# comm <- dr[samp, ]
+# group <- clMed[samp]
+#
+# system.time(x <- simperMan(comm, group))
+
+simperMan <- function(comm, group) {
+  # Transform community data as matrix
+  comm <- as.matrix(comm)
+
+  ## Identify groups
+  comp <- t(combn(as.character(unique(group)), 2))
+  nG <- nrow(comp)
+
+  ## Data averages by group
+  spavg <- apply(comm, 2, function(x) tapply(x, group, mean))
+
+  # Data frames to store results
+  nC <- ncol(comm)
+  # sp <- colnames(comm)
+  outlist <- vector('list', nG)
+  names(outlist) <- apply(comp, 1, paste, collapse="_")
+  for(i in seq_len(nG)) {
+    outlist[[i]] <- data.frame(contr = numeric(nC),
+                               sd = numeric(nC),
+                               ava = spavg[as.numeric(comp[i, 1]), ],
+                               avb = spavg[as.numeric(comp[i, 2]), ],
+                               stringsAsFactors = F)
+  }
+
+  ## Function to match constrasts
+  contrmatch <- function(X, Y, patt) X != Y & X %in% patt & Y %in% patt
+
+  # Measure mean dissimilarity contribution per grouping
+  for(j in seq_len(nG)) {
+    # ID of group in comparison[j]
+    id <- which(group %in% as.numeric(comp[j, ]))
+
+    # Contrast of pairwise group comparison and extract ID of inter-group comparisons
+    contrast <- outer(group[id], group[id], FUN=contrmatch, patt=comp[j,])
+    contrast <- contrast[lower.tri(contrast)]
+
+    for(i in seq_len(nC)) {
+      # Similarity for each column and data of group in comparison[j]
+      sim <- dist(comm[id,i,drop=FALSE], method="manhattan")
+      sim <- as.vector(sim)
+
+      # Extract mean and sd contribution to inter-group dissimilarity
[…]

### 2020-06-01 Code/barPlot.R

@@ -1,36 +1,0 @@
-barPlot <- function(x2) {
-  # Gap
-  gap <- .25
-  yG <- .15
-
-  # Combine with drNames
-  # This makes the function unusable for another setting, but simply removing it would make it ok
-  x2$names = drNames$accr
-  x2$group = as.factor(drNames$group)
-
-  # Format data
-  x2$x <- x2$x + gap
-  x2 <- x2[order(x2$group), ]
-  x2$col <- as.numeric(x2$group)
-  x2 <- x2[nrow(x2):1, ]
-
-  # Colors
-  cols <- gg_color_hue(nGroup)
-  x2$col <- cols[x2$col]
-
-  # Plot
-  library(graphicsutils)
-  par(mar = c(4,0,2,0))
-  plot0(xlim = c(-.125, 1.25), ylim = c(0,length(x2$x)+1))
-  # axis(1, at = seq(-gap, -1-gap, by = -.25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
-  axis(1, at = seq(gap, 1+gap, by = .25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
-  # axis(1, at = seq(-gap, gap, by = 2*gap), labels = c('',''), lwd = 1, tck = -.005)
-  text(x = rep(0, length(x2$x)), y = 1:(length(x2$x)), labels = x2$names, cex = .7)
-  for(i in c(seq(-gap, -1-gap, by = -gap), seq(gap, 1+gap, by = gap))) {
-    lines(x = c(i,i), y = c(.5,length(x2$x)+1.5), lty = 'dotted', col = '#00000066')
-  }
-  # Add bar plot
-  for(i in 1:22) {
-    rect(xleft = x2$x[i], ybottom = i-yG, xright = gap, ytop = i+yG, col = x2$col[i], border = x2$col[i])
-  }
-}

### 2020-06-01 Code/bootstrapSimper.R

@@ -1,28 +1,0 @@
-# iter <- 10
-# prop <- .02 # .05: ~500s/iter | .01: ~20s/iter | .02: ~80s/iter
-# The analysis being very long, we will proceed using a
-# bootstrap procedure, i.e. iterative simper analysis
-bootstrapSimper <- function(x, cl, iter = 100, prop = .1, accr = '') {
-  # List to store simper results for post-processing
-  bootSimper <- vector('list', iter)
-
-  # Determine the number of data points to sample from each cluster
-    clSub <- ceiling(table(cl) * prop)
-
-  # Repeat the process for the desired number of iterations
-  for(i in 1:iter) {
-    cat("   i: ", i, ' of ', iter, "\r")
-
-    # Subsampled dataset
-    samp <- vector('list', length(clSub))
-    for(j in 1:length(clSub)) samp[[j]] <- sample(x = which(cl == j), size = clSub[j])
-    samp <- unlist(samp)
-
-    # Run simper analysis on subset of data
-    bootSimper[[i]] <- simperMan(x[samp, ], cl[samp])
-
-    # Save file
-    save(bootSimper, file = paste0('./data/bootSimper', accr, '.RData'))
-  }
-  return(bootSimper)
-}

### 2020-06-01 Code/butterfly.R

@@ -1,44 +1,0 @@
-butterfly <- function(x1, x2) {
-  # Gap
-  gap <- .25
-  yG <- .15
-
-  # Combine with drNames
-  # This makes the function unusable for another setting, but simply removing it would make it ok
-  x1 <- data.frame(x = x1, names = drNames$accr, group = drNames$group, stringsAsFactors = T)
-  x2 <- data.frame(x = x2, names = drNames$accr, group = drNames$group, stringsAsFactors = T)
-
-  # Format data
-  x1$x <- -x1$x - gap
-  x1 <- x1[order(x1$group), ]
-  x1$col <- as.numeric(x1$group)
-  x1 <- x1[nrow(x1):1, ]
-
-  x2$x <- x2$x + gap
-  x2 <- x2[order(x2$group), ]
-  x2$col <- as.numeric(x2$group)
-  x2 <- x2[nrow(x2):1, ]
-
-  # Colors
-  cols <- gg_color_hue(nGroup)
-  x1$col <- cols[x1$col]
-  x2$col <- cols[x2$col]
-
-  # Plot
-  library(graphicsutils)
-  par(mar = c(0,0,0,0))
-  plot0(xlim = c(-1.25, 1.25), ylim = c(0,length(x1$x)+1))
-  axis(1, at = seq(-gap, -1-gap, by = -.25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
-  axis(1, at = seq(gap, 1+gap, by = .25), labels = c('0','.25','.5','.75','1'), lwd = 1, font = 2, cex.axis = .7, tck = -.005, mgp = c(1, 0, 0))
-  axis(1, at = seq(-gap, gap, by = 2*gap), labels = c('',''), lwd = 1, tck = -.005)
-  text(x = rep(0, length(x1$x)), y = 1:(length(x1$x)), labels = x1$names, cex = .7)
-  for(i in c(seq(-gap, -1-gap, by = -gap), seq(gap, 1+gap, by = gap))) {
-    lines(x = c(i,i), y = c(.5,length(x1$x)+1.5), lty = 'dotted', col = '#00000066')
-  }
-
-  # Add bar plot
-  for(i in 1:22) {
-    rect(xleft = x1$x[i], ybottom = i-yG, xright = -gap, ytop = i+yG, col = x1$col[i], border = x1$col[i])
[…]

### 2020-06-01 Code/circularPlot.R

@@ -1,67 +1,0 @@
-circularPlot <- function(data, center) {
-  # library
-  library(tidyverse)
-
-  # Set a number of 'empty bar' to add at the end of each group
-  empty_bar=3
-  to_add = data.frame( matrix(NA, empty_bar*nlevels(data$group), ncol(data)) )
-  colnames(to_add) = colnames(data)
-  to_add$group=rep(levels(data$group), each=empty_bar)
-  data=rbind(data, to_add)
-  data=data %>% arrange(group)
-  data$id=seq(1, nrow(data))
-
-  # Get the name and the y position of each label
-  label_data=data
-  number_of_bar=nrow(label_data)
-  angle= 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
-  label_data$hjust<-ifelse( angle < -90, 1, 0)
-  label_data$angle<-ifelse(angle < -90, angle+180, angle)
-
-  # prepare a data frame for base lines
-  base_data=data %>%
-    group_by(group) %>%
-    summarize(start=min(id), end=max(id) - empty_bar) %>%
-    rowwise() %>%
-    mutate(title=mean(c(start, end)))
-
-  # prepare a data frame for grid (scales)
-  grid_data = base_data
-  grid_data$end = grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
-  grid_data$start = grid_data$start - 1
-  grid_data=grid_data[-1,]
-
-  # Make the plot
-  p = ggplot(data, aes(x=as.factor(id), y=value, fill=group)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
-
-    geom_bar(aes(x=as.factor(id), y=value, fill=group), stat="identity", alpha=0.5) +
-
-    # Add a val=100/75/50/25 lines. I do it at the beginning to make sur barplots are OVER it.
-    geom_segment(data=grid_data, aes(x = end, y = 80, xend = start, yend = 80), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
-    geom_segment(data=grid_data, aes(x = end, y = 60, xend = start, yend = 60), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
[…]

### 2020-06-01 Code/intraSimBCurtis.R

@@ -1,92 +1,0 @@
-# library(slmeta)
-# data(drivers, package = 'slmeta')
-# dr <- drivers[, !colnames(drivers) %in% c('UVpositive','UVnegative','ID')]
-#
-# x <- dr[1:500, ]
-# system.time(sim <- simper2(x))
-#
-
-intraSim <- function(x, samp = NULL, iter = NULL) {
-  # ~~~~~~~~~~~~~~~~~~ PARAMETERS & DATA SETUP ~~~~~~~~~~~~~~~~~~
-  # Transform data as matrix
-  x <- as.matrix(x)
-
-  # Create empty list to store results
-    sim <- vector('list', 2)
-    names(sim) <- c('totalSimilarity', 'similarityContribution')
-    class(sim) <- 'simper2'
-
-  # ~~~~~~~~~~~~~~~~~~ MEASUREMENTS FOR SPECIES CONTRIBUTION ~~~~~~~~~~~~~~~~~~
-  # Here is the function that we will be applying for each species i between pairs of observations j, k
-    # Numerator: 2 * min(A_{i,j}, A_{i,k})
-    # Denominator: sum(j, k)
-
-  # Denominator: The species contribution will be divided by the sum of all abundances in all pairs of observations
-    # Sum of rows of community matrix
-    xs <- rowSums(x)
-
-    # Sum of all pairs of observations
-    denom <- combn(xs, m = 2, FUN = sum)
-
-  # Numerator: 2 * min(id1, id2)
-    # Minimum
-    sp <- apply(x, 2, function(x) combn(x, m = 2, FUN = min))
-
-    # Minimum value * 2
-    sp <- 2 * sp
-
-    # Divide by denominator
-    sp <- apply(sp, 2, function(x) x/denom)
-
-  # ~~~~~~~~~~~~~~~~~~ TOTAL CONTRIBUTION ~~~~~~~~~~~~~~~~~~
-  # Now from the sp matrix, we can obtain the intra-cluster average similarity
-  # by taking the mean of row sums
-    # Sum of individual species contributions
-    simTot <- rowSums(sp)
-
-    # Average and standard deviation result for intra-cluster similarity
-    sim$totalSimilarity <- c(mean(simTot), sd(simTot))
-    names(sim$totalSimilarity) <- c('Mean','SD')
-
-  # ~~~~~~~~~~~~~~~~~~ SPECIES CONTRIBUTION ~~~~~~~~~~~~~~~~~~
-  # Data frame with summary results
[…]

### 2020-06-01 Code/intraSimManhattan.R

@@ -1,62 +1,0 @@
-# library(slmeta)
-# data(drivers)
-# dr <- drivers[,-1]
-# load('./data/clMed.RData')
-# samp <- which(clMed == 2)
-# comm <- dr[samp, ]
-# #
-# system.time(x <- intraSim(comm))
-
-intraSim <- function(comm) {
-  # Transform community data as matrix
-  comm <- as.matrix(comm)
-  # comm <- comm + 1
-
-  # Data frames to store results
-  nC <- ncol(comm)
-  # sp <- colnames(comm)
-  outlist <- data.frame(contr = numeric(nC),
-                        sd = numeric(nC),
-                        stringsAsFactors = F)
-
-  # Measure mean dissimilarity contribution in group per variable
-  for(i in seq_len(nC)) {
-    # Similarity for each column and data of group in comparison[j]
-    sim <- dist(comm[,i,drop=FALSE], method="manhattan")
-    sim <- as.vector(sim)
-
-    # Extract mean and sd contribution to inter-group dissimilarity
-    outlist$contr[i] <- mean(sim)
-    outlist$sd[i] <- sd(sim)
-  }
-
-  # Transform to similarity using .1 / .1+contribution
-  outlist$contr <- .1 / (.1 + outlist$contr)
-
-  return(outlist)
-}
-
-bootIntraSim <- function(x, iter = 250, samp = .25, boot = 40000) {
-  # If the dataset is too large, the analysis will be done through a bootstrap
-  # List to store simper results for post-processing
-  # Determine if the bootstrap procedure is necessary
-  # I evaluated the minimum number of observations necessary to initiate the bootstrap
-  # procedure to be 4000. That was chosen by running the analyses and opting for
-  # a number of observations that I deemed could run fast enough.
-  # It is however still a subjective choice and can definitely be changed.
-  if(nrow(x) > boot) {
-    sim <- vector('list', iter)
-
-    # Repeat the process for the desired number of iterations
-    for(i in 1:iter) {
-      cat("   i: ", i, ' of ', iter, "\r")
-      # Subsampled observation id
-      id <- sample(1:nrow(x), size = samp*nrow(x))
-      # Similarity analysis
-      sim[[i]] <- intraSim(x[id, ])
-    }
-  } else {
-    sim <- intraSim(x)
-  }
-  sim
-}

### 2020-06-01 Code/simperMan.R

@@ -1,63 +1,0 @@
-# library(slmeta)
-# data(drivers)
-# dr <- drivers[,-1]
-# load('./data/clMed.RData')
-# # samp <- 100:200
-# prop = .05
-# clSub <- ceiling(table(clMed) * prop)
-# samp <- vector('list', length(clSub))
-# for(j in 1:length(clSub)) samp[[j]] <- sample(x = which(clMed == j), size = clSub[j])
-# samp <- unlist(samp)
-# comm <- dr[samp, ]
-# group <- clMed[samp]
-#
-# system.time(x <- simperMan(comm, group))
-
-simperMan <- function(comm, group) {
-  # Transform community data as matrix
-  comm <- as.matrix(comm)
-
-  ## Identify groups
-  comp <- t(combn(as.character(unique(group)), 2))
-  nG <- nrow(comp)
-
-  ## Data averages by group
-  spavg <- apply(comm, 2, function(x) tapply(x, group, mean))
-
-  # Data frames to store results
-  nC <- ncol(comm)
-  # sp <- colnames(comm)
-  outlist <- vector('list', nG)
-  names(outlist) <- apply(comp, 1, paste, collapse="_")
-  for(i in seq_len(nG)) {
-    outlist[[i]] <- data.frame(contr = numeric(nC),
-                               sd = numeric(nC),
-                               ava = spavg[as.numeric(comp[i, 1]), ],
-                               avb = spavg[as.numeric(comp[i, 2]), ],
-                               stringsAsFactors = F)
-  }
-
-  ## Function to match constrasts
-  contrmatch <- function(X, Y, patt) X != Y & X %in% patt & Y %in% patt
-
-  # Measure mean dissimilarity contribution per grouping
-  for(j in seq_len(nG)) {
-    # ID of group in comparison[j]
-    id <- which(group %in% as.numeric(comp[j, ]))
-
-    # Contrast of pairwise group comparison and extract ID of inter-group comparisons
-    contrast <- outer(group[id], group[id], FUN=contrmatch, patt=comp[j,])
-    contrast <- contrast[lower.tri(contrast)]
-
-    for(i in seq_len(nC)) {
-      # Similarity for each column and data of group in comparison[j]
-      sim <- dist(comm[id,i,drop=FALSE], method="manhattan")
-      sim <- as.vector(sim)
-
-      # Extract mean and sd contribution to inter-group dissimilarity
[…]

### 2020-06-02 Beauchesne-eDriversMS-SI.Rmd
update figures and add latest versions of manuscript in main folder (to update again)

--- 2020-04-24 Revisions2/Beauchesne-eDriversMS-SI.rmd
@@
     - \onehalfspacing
-    - \usepackage[svgnames]{xcolor}
-    - \usepackage[normalem]{ulem}
-    - \definecolor{blue}{HTML}{20cad5}
 relativeurls: true
@@
 <!--
-rmarkdown::render('./Revisions2/Beauchesne-eDriversMS-SI.rmd')
+rmarkdown::render('./Beauchesne-eDriversMS-SI.rmd')
 -->
@@
 ```{r tableS1, echo=F, fontsize=4}
-load('../tables/tableS1.RData')
+load('./tables/tableS1.RData')
 tableS1
@@
 to cover all of Canada and are highly variable in shapes and sizes.
-\textcolor{blue}{For example, their sizes range from 0.0003 to over 44000 $km^2$ in our study area.}
+For example, their sizes range from 0.0003 to over 44000 $km^2$ in our study area.
 The census provides population count within the boundary
@@
 \centering
-\includegraphics{./FigureS1.png}
+\includegraphics{./Figures/histograms.jpg}
 \caption{Frequency distribution of the untransformed data for all driver layers.}
@@
 \centering
-\includegraphics{./FigureS2.png}
+\includegraphics{./Figures/drivers.jpg}
 \caption{Distribution and intensity of transformed and normalized drivers in the Estuary and Gulf of St. Lawrence available on *eDrivers*.}
@@
 \centering
-\includegraphics{./FigureS3.png}
+\includegraphics{./Figures/marimekko.jpg}
 \caption{Frequency distribution of cumulative exposure (\textit{i.e.} sum of normalized driver intensity in each grid cell) and percent contribution of each driver to the frequency distribution of cumulative exposure in the Estuary and Gulf of St.
@@
 
-# ~~Threat complexes~~ \textcolor{blue}{Cumulative exposure profiles}
+# Cumulative exposure profiles
 
@@
 \centering
-\includegraphics{./FigureS4.png}
+\includegraphics{./Figures/validation.jpg}
 \caption{Validation procedure for the \textit{k-medoids} and \textit{k-means} clustering algorithms based on the number of cluster that maximizes average silhouette width (upper panels; Kaufman and Rousseeuw, 1990) and minimizes the total within-cluster sum of squares (WSS; lower panels).}
@@
 \centering
-\includegraphics{./FigureS5.png}
+\includegraphics{./Figures/interDissimilarity.jpg}
 \caption{Evaluation of inter-cluster dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan distance (Clarke, 1993).
@@
 \centering
-\includegraphics{./FigureS6.png}
+\includegraphics{./Figures/intraSimilarity.jpg}
 \caption{Evaluation of intra-cluster similarity using the Manhattan distance transformed to a similarity index.

### 2020-06-02 Beauchesne-eDriversMS.Rmd
update figures and add latest versions of manuscript in main folder (to update again)

--- 2020-05-01 Revisions2/Beauchesne-eDriversMS.Rmd
@@
 <!--
-rmarkdown::render('./Revisions2/Beauchesne-eDriversMS.Rmd', 'pdf_document')
+rmarkdown::render('./Beauchesne-eDriversMS.Rmd', 'pdf_document')
 -->
@@
 \includegraphics{./Figure1.jpg}
-\caption{\sout{Description of the Estuary and Gulf of St.
-Lawrence in Eastern Canada.}\textcolor{blue}{Description of the St.
+\caption{Description of the St.
 Lawrence System in Eastern Canada, composed of the St.
@@
 Lawrence.
-The Estuary is defined by the limit of seawater intrusion, close to Îles d'Orléans, to the west and by its connection to the Gulf near Pointe-des-Monts.
-The Gulf is an interior sea connected to the Atlantic by Cabot Strait and the Strait of Belle Isle to the south and north of Newfoundland, respectively.}}
+The Estuary is defined by the limit of seawater intrusion, close to Île d'Orléans, to the west and by its connection to the Gulf near Pointe-des-Monts.
+The Gulf is an interior sea connected to the Atlantic by Cabot Strait and the Strait of Belle Isle to the south and north of Newfoundland, respectively.}
 \label{egsl}
@@
 An index of hypoxia ($A$) was created using bottom-water dissolved oxygen between 2013 and 2017 (Blais et al., 2018).
-Demersal destructive fisheries ($i.e.$ trawl and dredges) ($B$) intensity was evaluated from fisheries catch data collected between 2010 and 2015 used to measure annual area weighted total biomass ($kg$) in i $km^2$ grid cells (DFO, 2016b).
+Demersal destructive fisheries ($i.e.$ trawl and dredges) ($B$) intensity was evaluated from fisheries catch data collected between 2010 and 2015 used to measure annual area weighted total biomass ($kg$) in 1 $km^2$ grid cells (DFO, 2016b).
 See supplementary materials for more information on specific methodologies.
@@
 \includegraphics{./Figure5.jpg}
-\caption{Distribution of \sout{threat complexes}\textcolor{blue}{areas with similar cumlative exposure profiles} in the Estuary and Gulf of St.
-Lawrence\textcolor{blue}{, identified through a clustering approach} (upper panel).
-Mean intensity of all coastal (red), climate (green), fisheries (blue) and marine traffic (purple) drivers within each \sout{threat complex}\textcolor{blue}{cluster} (lower panel).
+\caption{Distribution of areas with similar cumulative exposure profiles in the Estuary and Gulf of St.
+Lawrence, identified through a clustering approach (Top).
+Mean intensity of all coastal (red), climate (green), fisheries (blue), and marine traffic (purple) drivers within each cluster (Bottom).
 Refer to Table 1 for acronym meaning and to the Supplementary Materials for more details.}
@@
 
-Table 1. List of drivers ~~currently~~ available on *eDrivers* and used for the analyses presented in this paper. Further details on methods and data are available in the Supplementary Material.
+Table 1. List of drivers available on *eDrivers* and used for the analyses presented in this paper. Further details on methods and data are available in the Supplementary Materials.
 

### 2020-06-02 Code/Figures/S2-Drivers.R

@@ -11,7 +11,7 @@ parVal <- c(ceiling(nDr / 4), 4)
 source('./Code/Functions/plotEGSL.R')
 
 # ~~~~~~~~~~~~~~~~~~~ DRIVERS ~~~~~~~~~~~~~~~~~~~ #
-jpeg('./figures/drivers.jpeg', width = 1280, height = 1500, res = 200)
+jpeg('./figures/drivers.jpg', width = 1280, height = 1500, res = 200)
 par(mfrow = parVal, mar = c(4,4,1,1))
 # Drivers
 for(i in 1:nDr) {

### 2020-06-02 Code/Figures/S3-Marimekko.R

@@ -83,7 +83,7 @@ x <- x[!id]
 bin <- rbind(numeric(length(x)), bin)
 
 # ~~~~~~~~~~~~~~~~~~~ LAYOUT ~~~~~~~~~~~~~~~~~~~ #
-jpeg('./figures/marimekko.jpeg', width = 1800, height = 800, res = 200, pointsize = 8)
+jpeg('./figures/marimekko.jpg', width = 1800, height = 800, res = 200, pointsize = 8)
 mat <- matrix(ncol = 20, nrow = 10, data = 1)
 mat[, 10:17] <- 2
 mat[, 18:20] <- 3
